### 2018-03-29 Datelife_paper1.Rmd
update

---
output: pdf_document
fontsize: 12pts
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'mg onna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
---

Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE <!--MAX 50 CHARACTERS, ALL CAPITALS-->

Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life

Authors: Luna L. Sánchez-Reyes^1^, Brian C. O’Meara^1^

Correspondence address:

1. *Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA*

Corresponding authors: sanchez.reyes.luna@gmail.com, bomeara@utk.edu

\newpage

**abstract.-** Here goes the abstract.

**Keywords:** Tree; Phylogeny; Scaling; Open; Ages; Congruify; Supertree;

\newpage

<!--INTRODUCTION-->
Divergence time of lineages<!--named biological entities--> <!--(taxon ages)--> constitutes in many ways the main knowledge necessary for evolutionary understanding.
<!--Along with phylogenetic relationships, it is the most important piece of information allowing to test alternative evolutionary hypothesis in many areas of research, from developmental to ecological questions.
Allong with migration rates and patterns, it is fundamental to test biogeographical hypothesis-->
Coupled to species number and distribution, it is the basic information for the study of diversification processes, such as the tempo and mode of speciation and extinction, crucial for the understanding of how biodiversity patterns are shaped across space and time [@Morlon2014].

When organisms are preserved in a fossil form, a time frame of taxon origin can be obtained directly from the age of rock strata.
However, not all organisms fossilize well or at all.
Fossilization success alone is highly circumstancial, and varies depending on a number of parameters including the nature of the habitat, population size, species range breadth and physical characteristics of the organism. <!--fossil recovery is also highly circumstancial-->
Thus, relying only on the fossil record to obtain a time frame of lineage divergence<!--taxon ages--> for all life is<!--, alas,--> not possible.
Relative rate of DNA or aminoacid substitution constitutes another important source of information on lineage divergence. <!--describe the two components of rel subst rate?-->
It can be obtained from hypothesis of character homology (alignments) when reconstructing phylogenetic relationships. Molecular dating techniques use external data such as absolute time calibrations (e.g., fossils, geologic events) or absolute substitution rates <!--birds--> to generate dated phylogenies (chronograms) which contain information on absolute times of node divergence and taxon ages. <!--With molecular dating methods, an external calibration point is still necessary to estimate the absolute time of origin. Hence, the accumulation of data on fossils for calibration is also continual-->
<!--This method has been particularly useful to obtain time frame of origin of organisms with poor or no fossil record, but is also largely used for organisms with rich fossil records-->
<!--Models allow to take into account the nature of the fossil record: Given the characteristics of the fossilization process, the quality of dates inferred directly from it is not always considered the best.-->

In the past decades, the possibility to obtain DNA sequences in large quantities from a wide variety of organisms became a reality, which, coupled to methodological development in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the amount of<!--organisms with--> data on taxon ages across the tree of life.
To date, there is a large amount of both fossil and molecular-based data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE and Open Tree of Life (OToL).
OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
devtools::load_all(pkg = "~/Desktop/datelife")
length(opentree_chronograms$trees)
names(opentree_chronograms)
-->
<!--How many trees? How many dated trees?-->
Methods to include living and fossil lineages are in continued development and increased usage by the community<!--total evidence and revbayes (fossilized birth-death)-->, which coupled to better sharing data practices, are greatly contributing to the accumulation in number and type of available data on taxon ages.

<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->
Despite its importance, analytical tools to summarize available information on taxon ages are still lacking.
Data might not have been exploited because:
Data is in different repositories and formats;
Lineage names are different among studies and difficult to reconcile;
Taxonomy is also different among studies and difficult to reconcile.
<!--Data curation is necessary at some point (is this always true?)-->
Also, data curation is an important part of any biological study. The research community considers it as an important or even crucial step before data analysis. Hence, automated processes for large data analysis are frequently received with skepticism.
Recent work on this area (e.g.,, supersmart<!--and, which others?-->) aims to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of data curation is still necessary.
<!--General issues with dating techniques:-->
Choosing correct fossils for calibration requires a lot of expertise and knowledge on the group.
Incorrect use of fossils can generate severe bias in dating results [@Sauquet2012c].
DateLife palliates this by using only information available from already published studies, which are ideally constructed using robust information, such as sequence data and curated fossil calibrations.
<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
<!--Importance of datelife:-->
DateLife can summarize this information in several formats that can be easily inspected by users.
This  allows rapidly obtaining a time frame of taxon divergence for a wide number of taxa.
DateLife can also generate chronograms for taxa with little available information, by using the available data as calibration points.
DateLife is the main service for scaling phylogenetic trees in Phylotastic! system [@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
It can be used through an R package <link to documentation>, a web interface <http://www.datelife.org/query/> and an API <!--(still not up, right?)-->

<!--DESCRIPTION of Datelife-->
<span style="font-variant:small-caps;">
Description
</span>

<!--(BOLD and OToL species names are homogeneous?-->

DateLife is a service for searching and processing information on ages of any taxon <!--or group of taxa-->of interest, available in chronograms from public data repositories of published peer reviewed studies. With this data, it can also generate new taxon age information<!--by using expert available phylogenies and scaling them to absolute time using information on ages obtained from published chronograms,--> by calling a variety of external services and proved methods.
It only requires as input a set of taxon names, in the form of a comma separated listing or vector, or of a phylogeny with taxon names on the tips.
Taxon names can correspond to binomial species names or clades<!--common names (will we implement it?)-->.
When taxon names are clades, DateLife pulls all accepted species names within the clade from OToL's reference taxonomy using a service of rphylotastic R package<!--citation-->.
Names belonging to subspecies or any other infraspecific category are treated as species.
DateLife can process input names with the taxon name resolution service (TNRS), <!--which "corrects misspelled names and authorities, standardizes variant spellings, and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->which corrects misspelled names or typos, and standardizes variation in spelling and synonyms [@Boyle2013], increasing the probability to correctly find the queried taxa in the chronogram database. DateLife uses TNRS to compare names against OToL's reference taxonomy using a service from the R package rotl [@Michonneau2016]. <!--we'll implement the option to use other reference taxonomies in the future -->

DateLife's main function searches taxon names across the chronogram database specified by the user. At the moment, it queries chronograms from OToL [@Hinchliff2015] and TreeBASE [@Piel2002]<!-- there is an earlier reference from 1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive tool for browsing the phylogeny of life. Am J Bot. 1994, 81 (6): 183--> repositories.
DateLife identifies chronograms having at least two taxon names, and subset them to contain only the taxa of interest within each chronogram. It then stores taxon age information from each chronogram individually as a patristic matrix, named with the citation of the original study. This format allows a rapid summary in a number of different ways, including: 1) citations of the original studies containing the subset chronograms, 2) a list of subset chronograms' mrca ages, 3) a list of subset chronograms in newick or phylo format, 4) a table containing all information in html or R's data frame format, or 5) as a single chronogram summarized from subset chronograms using the supertree Super Distance Matrix (SDM) approach [@Criscuolo2006] or using the median of branch lengths. <!--(how to better explain the latter?? Also, I just discovered the median gives non ultrametric trees...).-->

DateLife also stores information on input taxon presence/absence across subset chronograms. Users can choose to add ages of missing taxa to subset chronograms in different ways, depending on the amount of knowledge or how much they want to be involved in the steps of the addition process.
If users have no access to biological information (i.e., a character, DNA or protein matrix), missing taxa can be added to any chronogram simply at random, or by following taxonomic or phylogenetic knowledge from expert sources. There are a wide number of open reference taxonomies available, such as the Catalogue of Life [@Roskob2017] or the NCBI taxonomy database [@Federhen2012]. Expert phylogenies to be used as topological constraint or backbone (a tree with no branch lengths) can also be obtained from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002] and Dryad <https://www.datadryad.org//>.
At the moment, DateLife only uses OToL's synthetic tree and reference taxonomy as expert knowledge to automatically add missing taxa to chronograms. Alternatively, users can input a reference taxonomy or topological constraint of their choosing or making. <!-- at some point we will implement the possibility for users to choose different reference taxonomies automatically-->
If OToL's synthetic tree is not satisfactorily resolved for the taxa of interest, DateLife can construct a sequence data matrix from DNA markers available from the Barcode of Life Database (BOLD; @Ratnasingham2007), to attempt to further resolve polytomies. It will follow OToL's synthetic tree as backbone. <!--This molecular tree can then be dated with chronosMPL from the ape package, but is it really needed?? To estimate node ages chronosMPL uses mean path length method from Britton et al. 2002-->
<!--It can rapidly construct trees with branch lengths from
a set of lineages with sequence data from the Barcode Of Life Database (BOLD, <http://www.boldsystems.org/>), using the synthetic OToL (<https://tree.opentreeoflife.org>) as backbone, and to obtain divergence dates.
It allows direct comparison of dates obtained with different markers
available in BOLD (in plants and fungi in particular).-->
To use information from a topological constraint, DateLife calls the congruification method described in [@Eastman2013] to find shared nodes between trees (congruent nodes). It then fixes their ages, and add ages to remaining nodes with a dating method that can be specified by the user. If users have access to biological data, thay can input a tree with branch lengths proportional to relative substitution rates as topological constraint. In this case, age data from congruent nodes will be used as calibration points.
Age data from several chronograms can be combined and congruified to be used as calibration points in a single analysis.

Several dating methods are implemented in DateLife.
Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
DateLife implements BLADJ from the development R version of phylocom's R package [@Webb2008], phylocomr <https://github.com/ropensci/phylocomr>. It can only be used when there is a topological constraint with no branch lengths.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. It is also called through R. <!--It could be used with a reference taxonomy, if we provide a tree with politomies-->
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method [@Smith2012]. It is called through R.
MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at random, following a reference taxonomy or a topological constraint. It draws ages from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
DateLife calls MrBayes trough an R function.

DateLife can also correct negative branch lengths in several ways.

<!--**Benchmark: Testing DateLife computing performance**-->
<span style="font-variant:small-caps;">
Benchmark
</span>

DateLife's performance, speed and scalability were tested on a 6 months old computer
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time relative to 1) number of input names, 2) number of chronograms in database and 3) service used.
<!--
Speed with different number of input lineages and types of analysis:
-->
Results show that searching time increases linearly with number of input names and number of chronograms in database (Fig. X1).

Summarizing DateLife results processing times

Adding dates processing time

get_bold_otol_tree running time
<!-- test improvement in synthetic tree after using bold sequences. Is it really better resolved?-->

<!--
Speed of web interface and of r package (in computers with different capacities?)
-->

<!--**Biological example: Testing DateLife accuracy**-->
<span style="font-variant:small-caps;">
Biological example
</span>

Find a clade with at least one chronogram containing all clade's species.
(Penguis look good, but they are giving weird results in SDM)

Remove this chronogram from datelife Results.

Make sdm and median trees and Compare

add taxa with different methods and Compare

Use ltts to compare for now.
Fig. X2 shows comparison of available chronograms for Felidae species and chronograms generated through DateLife

think of a test to compare trees, topology- and date-wise

<!--Bird (or reptile) chronograms, too long time...
finches is good
nothofagus
Look for all chronograms containing any birds
Or, look for chronograms containing small and old lineages
Determine which clade of birds has the more chronograms (have been dated
more times) and use that as biological example-->

<!--**Discussion/Conclusions** there is not a discussion section in this type of sys Bio contributions-->
<span style="font-variant:small-caps;">
Conclusions
</span>

Taxon ages are key to many areas of evolutionary studies: trait evolution, species diversification, biogeography, macroecology and more. Obtaining these ages is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to develop the necessary knowledge or data curation skills to produce new chronograms. Knowledge on taxon ages is also important for non-biological studies and the non-academic community.
The combination of new analytical techniques, availability of more fossil and molecular data, and better practices in data sharing has resulted in a steady accumulation of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree of Life, for a large quantity and diversity of organisms. However, this information remains difficult to synthesize for many biologists and the non-academic community<!-- or just non biologists-->.

<!--Potential applications shown here-->Here, we have shown that DateLife allows an easy and fast obtention of all publicly available information on taxon ages, which can be used to generate new data. <!--Potential applications not shown here-->This information can be used to account for the effect of phylogenetic signal in studies of trait evolution; to explore potential speciation and extinction dynamics of interest within a clade; to obtain a time frame of biogeographical events; for science communication and outreach, amongst others.
<!--Why you should use it-->Compared to similar platforms such as time tree of life and supermart, it offers several advantages.
It is fast;
source data is completely open;
it requires no expert biological knowledge for any of their functionalities;
it allows exploration of alternative taxonomic and phylogenetic schemes;
it allows rapid exploration of the effect of alternative divergence time hypothesis;
it allows rapid synthesis in a number of different formats; <!-- we should add graphical options too-->
it facilitates reproducibility of analyses;

Improvements, short and long-term:
* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations [@Schenk2016].
* bayesian congruification
* topological congruification

<!--**Availability**-->
<span style="font-variant:small-caps;">
Availability
</span>

DateLife is free and open source and it can be used through its current website <http://www.datelife.org/query/>, or through Phylotastic!'s web portal <http://phylo.cs.nmsu.edu:3000/>.
DateLife can also be used locally through its R package. The stable version is available for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>) using the command `install.packages(pkgs = "datelife")` from R. Development versions are available from GitHub repository (<https://github.com/phylotastic/datelife>) and can be installed using the devtools R package command `install_github("phylotastic/datelife")`.

<!--**Supplementary Material**-->
<span style="font-variant:small-caps;">
Supplementary Material
</span>

Supplementary material, including code files and
online-only appendices, can be found in the GitHub
repository <!-- Dryad also? since it is all in github already-->

<!--**Funding**-->
<span style="font-variant:small-caps;">
Funding
</span>

Funding was provided by NSF grant 1458603

NESCent

Open Tree of Life

University of Tennessee, Knoxville

<!--**Acknowledgements**-->
<span style="font-variant:small-caps;">
Acknowledgements
</span>

<!--**References**-->
<span style="font-variant:small-caps;">
References
</span>

### 2018-03-29 Datelife_paper1.Rmd
added word

@@
 ---
-output: pdf_document
-fontsize: 12pts
-textalign: left  # this does not work...
-geometry: margin = 1in
 bibliography: library.bib
 csl: systematic-biology.csl
-link-citations: yes
-# indent: true # this works
-# header-includes:
-#    - \setlength\parindent{24pt}  # this works too, but I'mg onna indent manually with 4 spaces
-# nope, 4 spaces indents are not recognized, they do something else
-linestretch: 2
+fontsize: 12pts
+geometry: margin = 1in
 header-includes:
@@
 - \linenumbers
+linestretch: 2
+link-citations: yes
+output:
+  word_document: default
+  pdf_document: default
+textalign: left
 ---

### 2018-03-29 Datelife_paper1.Rmd
paper improvement

--- 2018-03-29 Datelife_paper1.Rmd
@@
 <!--INTRODUCTION-->
-Divergence time of lineages<!--named biological entities--> <!--(taxon ages)--> constitutes in many ways the main knowledge necessary for evolutionary understanding.
+Divergence time of lineages<!--named biological entities, taxon ages?--> constitutes in many ways the main knowledge necessary for evolutionary understanding.
 <!--Along with phylogenetic relationships, it is the most important piece of information allowing to test alternative evolutionary hypothesis in many areas of research, from developmental to ecological questions.
-Allong with migration rates and patterns, it is fundamental to test biogeographical hypothesis-->
+Along with migration rates and patterns, it is fundamental to test biogeographical hypothesis-->
 Coupled to species number and distribution, it is the basic information for the study of diversification processes, such as the tempo and mode of speciation and extinction, crucial for the understanding of how biodiversity patterns are shaped across space and time [@Morlon2014].
@@
 Thus, relying only on the fossil record to obtain a time frame of lineage divergence<!--taxon ages--> for all life is<!--, alas,--> not possible.
-Relative rate of DNA or aminoacid substitution constitutes another important source of information on lineage divergence. <!--describe the two components of rel subst rate?-->
-It can be obtained from hypothesis of character homology (alignments) when reconstructing phylogenetic relationships.
+Relative rate of DNA or aminoacid substitution constitutes another important source of information on lineage divergence.
+<!--describe the two components of rel subst rate: which, contain tne information on absolute rate of substitution and absolute time.-->
+It is usually obtained from hypothesis of character homology (alignments) when reconstructing phylogenetic relationships.
 Molecular dating techniques use external data such as absolute time calibrations (e.g., fossils, geologic events) or absolute substitution rates <!--birds--> to generate dated phylogenies (chronograms) which contain information on absolute times of node divergence and taxon ages.
@@
 Hence, automated processes for large data analysis are frequently received with skepticism.
-Recent work on this area (e.g.,, supersmart<!--and, which others?-->) aims to:
+Recent work on this area (e.g., supersmart<!--and, which others?-->) aims to:
 Generate new dates using all available DNA sequence information;
@@
 DateLife is the main service for scaling phylogenetic trees in Phylotastic! system [@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
-It can be used through an R package <link to documentation>, a web interface <http://www.datelife.org/query/> and an API <!--(still not up, right?)-->
+It can be used through an R package<link to documentation>, a web interface (<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
 
@@
 There are a wide number of open reference taxonomies available, such as the Catalogue of Life [@Roskob2017] or the NCBI taxonomy database [@Federhen2012].
-Expert phylogenies to be used as topological constraint or backbone (a tree with no branch lengths) can also be obtained from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002] and Dryad <https://www.datadryad.org//>.
+Expert phylogenies (with or without branch lengths) to be used as topological constraint (backbone) can also be obtained from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002] and Dryad (<https://www.datadryad.org//>).
 At the moment, DateLife only uses OToL's synthetic tree and reference taxonomy as expert knowledge to automatically add missing taxa to chronograms.
@@
 Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
-DateLife implements BLADJ from the development R version of phylocom's R package [@Webb2008], phylocomr <https://github.com/ropensci/phylocomr>.
+DateLife implements BLADJ from the development R version of phylocom's R package [@Webb2008], phylocomr (<https://github.com/ropensci/phylocomr>).
 It can only be used when there is a topological constraint with no branch lengths.
@@
 source data is completely open;
-it requires no expert biological knowledge for any of their functionalities;
+it requires no expert biological knowledge from users for any of its functionalities;
 it allows exploration of alternative taxonomic and phylogenetic schemes;

### 2018-04-03 Datelife_paper1.Rmd
missing references

@@
 If users have no access to biological information (i.e., a character, DNA or protein matrix), missing taxa can be added to any chronogram simply at random, or by following taxonomic or phylogenetic knowledge from expert sources.
-There are a wide number of open reference taxonomies available, such as the Catalogue of Life [@Roskob2017] or the NCBI taxonomy database [@Federhen2012].
+There are a wide number of open reference taxonomies available, such as the Catalogue of Life [@Roskov2017] or the NCBI taxonomy database [@Federhen2012].
 Expert phylogenies (with or without branch lengths) to be used as topological constraint (backbone) can also be obtained from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002] and Dryad (<https://www.datadryad.org//>).

### 2018-05-23 Datelife_paper1.Rmd
timetree description

@@
 ---
+```{r eval = TRUE, echo = FALSE}
+figure_runtime_main <- "Fig. 1"  # make_datelife_query_plus_get_datelife_result
+```
 
@@
 
-In the past decades, the possibility to obtain DNA sequences in large quantities from a wide variety of organisms became a reality, which, coupled to methodological development in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the amount of<!--organisms with--> data on taxon ages across the tree of life.
+In the past decades, the possibility to obtain DNA sequences in large quantities from a wide variety of organisms became a reality, which, coupled to methodological development in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon ages across the tree of life.
 To date, there is a large amount of both fossil and molecular-based data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE and Open Tree of Life (OToL).
@@
 -->
-Despite its importance, analytical tools to summarize available information on taxon ages are still lacking.
+Data from chronograms and data tables of divergence time estimates reported in scientific studies published since 1987 have been synthesised in the TimeTree of Life resource [@Hedges2006; @Hedges2015; @Kumar2017].
+TimeTree of Life platform allows searching ages of taxa from a chronogram synthesizing age data from more than three thousand (3 163) studies, encompassing almost a hundred thousand (97 085) species [@Kumar2017].
+<!--
+-->
+This chronogram samples only between 0.097 and 3.236% of total earth species diversity (following taxonomic expert opinion on the global, extant species numbers, which ranges from 3 to 100 millions species [Mayr2010; Moran2011]),
+<!--
+[@Mora2011]
+eukaryotes and prokaryotes:
+on earth: 1.24 million catalogued, 8.74 million predicted
+on the oceans: 0.19 million catalogued, 2.21 predicted
+
+Total ssp number estimated to be between 3 and 100 million species, as mentiones in [@Mora2011]:
+ttol_spp <- 97085
+(ttol_spp * 100) / 3e+6  # 3.236167
+(ttol_spp * 100) / 100e+6  # 0.097085
+-->
+but is currently one of the most important sources on divergence times, since it includes all main lineages of life (plants, animals, fungi, bacteria and archaea) and is currenty the largest chronogram in existence.
+As a science communication tool it is very powerful, since it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them.
+However, the primary data source and the synthetic chornogram are not publicly available, and large-scale, automated, data-mining is not allowed, minimizing its potential usefulness for the scientific community.
+<!--
+While its node-time search is very powerful, using available data to infer divergence time from lineages that have never been sequenced before, its timetree option is less powerful, their main flaw being that it does not provide source data.
+However, their primary database is not fully open, giving information on a per node basis.
+So, it is not possible to easily access information from original studies for a group of taxa (One would need to perform a pair of taxa search for all pairs from the group of interest).
+Divergence time estimates for all taxa in the TToL are precomputed and stored in the TimeTree database
+All searches allow species names and higher group names. Common names are allowed in all searches but a list of taxa.
+one-taxon search was nor working for dog, though.
+1) ttol one lineage: it gives the evolutionary timeline of a lineage, showing the age of main divergence events occuring along the lineage evolutionary history, i.e., when it shared a comon ancestor with all main lineages from which it diverged a long time ago.
+It shows results only graphically on the website.
+2) ttol parwise search: allows common names, species names and higher group names
+    how it handles names missing from tree: it looks in the NCBI taxonomy for the closest relative of the species or higher taxa. Then uses the closest relative to get the lineage pair age from ttol.
+3) ttol group of taxa
+4) ttol list of taxa: Cannot get it to work; does not work with cat, dog, turtle search, felidae, hominidae, phrynops search, some_ants species name search.
+What kind of files does it accepts: txt, rtf, doc, docx?
+-->
+
+<!--
+retrieves information takes available data on lineage ages and phylogenetic relationships to construct a tree of all life scaled to absolute time: a chronogram of all life, while allowing the general public to search for age data within the tree for one taxon of interest (timeline option), a pair of taxa (node time option), a group of taxa (it can only be a single name of a taxonomic group), or a list of taxa of interest (timetree option for the two latter).
+Good things:
+Allows to get ages of species with no data, by extrapolating from closely related species with data.
+Allows comparison of timelines to geologic events, such as asteroid impacts, oxygen and carbon dioxide levels, solar luminosity.
+Not so good things:
+Only allows to download some data from original sources if performing a pair-of-taxa search.
+Only allows to see age ranges or confidence intervals in graphical form, when hovering over nodes.
+These CIs come from ttol.
+It does not have CIs from original studies.
+Node hovering is activated on one-taxon evolutionary history, and group-of-taxa searches (we could not get to work list-of-taxa search).
+Pair-of-taxa search shows the confidence interval of the only ttol node that constitutes the taxon-pair mrca.
+You can only download data from pair-of-taxa and group-of-taxa search.
+It generates a table with ages from original studies and a tree pruned from ttol, respectively.
+In both cases it gives single point age estimates.
+We cannot download CI information from ttol.
+-->
+
+Despite its importance, analytical tools to summarize available information on taxon ages for the scientific community are still lacking.
 Data might not have been exploited because:
@@
 DateLife can summarize this information in several formats that can be easily inspected by users.
-This  allows rapidly obtaining a time frame of taxon divergence for a wide number of taxa.
+This allows rapidly obtaining a time frame of lineage divergence for a wide number of taxa.
 DateLife can also generate chronograms for taxa with little available information, by using the available data as calibration points.
 DateLife is the main service for scaling phylogenetic trees in Phylotastic! system [@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
-It can be used through an R package<link to documentation>, a web interface (<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
+It can be used through an R package <link to documentation>, a web interface (<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
 
@@
 1994, 81 (6): 183--> repositories.
-DateLife identifies chronograms having at least two taxon names, and subset them to contain only the taxa of interest within each chronogram.
+DateLife identifies chronograms having at least two taxon names, and subset them to contain only the taxa of interest.
 It then stores taxon age information from each chronogram individually as a patristic matrix, named with the citation of the original study.
@@
 DateLife also stores information on input taxon presence/absence across subset chronograms.
-Users can choose to add ages of missing taxa to subset chronograms in different ways, depending on the amount of knowledge or how much they want to be involved in the steps of the addition process.
+Users can choose to add ages of missing taxa to subset chronograms in different ways, depending on the amount of knowledge they want to input or how much they want to be involved in the steps of the addition process.
 If users have no access to biological information (i.e., a character, DNA or protein matrix), missing taxa can be added to any chronogram simply at random, or by following taxonomic or phylogenetic knowledge from expert sources.
@@
 </span>
+<!-- Good to test 3 features on a package: performance, speed and scalability-->
 
-DateLife's performance, speed and scalability were tested on a 6 months old computer
+DateLife's code speed was tested on a 6 months old computer
 with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
-We registered variation in computing time relative to 1) number of input names, 2) number of chronograms in database and 3) service used.
+We registered variation in computing time relative to number of input names <!--number of chronograms in database--> and DateLife service.
+Input processing increases roughly linearly with number of input taxon names, and increases considerably if tnrs service is activated (`r figure_runtime_main`).
 <!--
@@
 -->
-Results show that searching time increases linearly with number of input names and number of chronograms in database (Fig. X1).
+Results show that searching time increases linearly with number of input names and number of chronograms in database.
 
@@
 -->
+DateLife's code performance was evaluated with a set of unit tests designed and implemented with the R package testthat [@RCoreTeam2018].
+These tests were run both locally –using the devtools package [@RCoreTeam2018]– and on a public server –via GitHub– using the continuous integration tool Travis CI (<//travis-ci.org>).
+At present, unit tests cover around 30% (for now) of DateLife's code (<https://codecov.io/gh/phylotastic/datelife>).
+<!-- unit test make code robust, verifiable, debuggable, -->
 
@@
 Find a clade with at least one chronogram containing all clade's species.
-(Penguis look good, but they are giving weird results in SDM)
+(Penguins look good, but they are giving weird results in SDM)
 
@@
 Taxon ages are key to many areas of evolutionary studies: trait evolution, species diversification, biogeography, macroecology and more.
-Obtaining these ages is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to develop the necessary knowledge or data curation skills to produce new chronograms.
+Obtaining these ages is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to develop the necessary knowledge and data curation skills to produce new chronograms.
 Knowledge on taxon ages is also important for non-biological studies and the non-academic community.

### 2018-05-23 Datelife_paper1.Rmd
modify timetree section

@@
 -->
-Data from chronograms and data tables of divergence time estimates reported in scientific studies published since 1987 have been synthesised in the TimeTree of Life resource [@Hedges2006; @Hedges2015; @Kumar2017].
-TimeTree of Life platform allows searching ages of taxa from a chronogram synthesizing age data from more than three thousand (3 163) studies, encompassing almost a hundred thousand (97 085) species [@Kumar2017].
-<!--
--->
-This chronogram samples only between 0.097 and 3.236% of total earth species diversity (following taxonomic expert opinion on the global, extant species numbers, which ranges from 3 to 100 millions species [Mayr2010; Moran2011]),
+One inspiration for the DateLife project was the TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017].
+This project has aggregated chronograms from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
+However, even in this gold standard resource, the included taxa only include between 0.097 and 3.236% of total species diversity (following taxonomic expert opinion on the global, extant species numbers, which ranges from 3 to 100 million species [Mayr2010; Moran2011]).
+One advantage of TimeTree is that it includes taxa from across the tree of life, versus more specialized chronograms focusing on plants [PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups.
+Users can use a web interface or a mobile app to receive information on divergence times for pairs of species, ______.
+It is of limited utility for scientific studies, however.
+The thousands of trees that have been entered are unavailable for examination or reuse; according to their FAQ, methods for allowing download of the data have been under discussion since 2015.
+There is no API allowing programmatic access to any data.
+The nearly hundred thousand taxon chronogram is not available with [@Kumar2017] or the TimeTree website, though the still substantial chronogram from @Hedges2015 is available.
+Large-scale, automated, data-mining is not allowed under the site's terms of use.
+As a science communication tool it is very powerful, since it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them.
+
+With this inspiration, a prototype DateLife service was developed over a series of phylotastic hackathons [__CITE__] at the National Evolutionary Synthesis Center.
+NSF funding allowed for the further development leading to this paper.
+A core goal has been openness of both the data sources and the code underlying the analyses.
+
 <!--
@@
 -->
-but is currently one of the most important sources on divergence times, since it includes all main lineages of life (plants, animals, fungi, bacteria and archaea) and is currenty the largest chronogram in existence.
-As a science communication tool it is very powerful, since it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them.
-However, the primary data source and the synthetic chornogram are not publicly available, and large-scale, automated, data-mining is not allowed, minimizing its potential usefulness for the scientific community.
 <!--
@@
 Taxon names can correspond to binomial species names or clades<!--common names (will we implement it?)-->.
-When taxon names are clades, DateLife pulls all accepted species names within the clade from OToL's reference taxonomy using a service of rphylotastic R package<!--citation-->.
+When taxon names are clades, DateLife pulls all accepted species names within the clade (up to OToL's limit of _____ species) from OToL's reference taxonomy using a service of rphylotastic R package<!--citation-->.
 Names belonging to subspecies or any other infraspecific category are treated as species.
@@
 DateLife's main function searches taxon names across the chronogram database specified by the user.
-At the moment, it queries chronograms from OToL [@Hinchliff2015] and TreeBASE [@Piel2002]<!-- there is an earlier reference from 1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive tool for browsing the phylogeny of life.
+At the moment, it queries chronograms from the OToL [@Hinchliff2015]<!-- there is an earlier reference from 1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive tool for browsing the phylogeny of life.
 Am J Bot.
-1994, 81 (6): 183--> repositories.
+1994, 81 (6): 183--> repository.
 DateLife identifies chronograms having at least two taxon names, and subset them to contain only the taxa of interest.
@@
 It then fixes their ages, and add ages to remaining nodes with a dating method that can be specified by the user.
-If users have access to biological data, thay can input a tree with branch lengths proportional to relative substitution rates as topological constraint.
+If users have access to biological data, they can input a tree with branch lengths proportional to relative substitution rates as topological constraint.
 In this case, age data from congruent nodes will be used as calibration points.
@@
 
-DateLife's code speed was tested on a 6 months old computer
+DateLife's code speed was tested on an Apple iMac
 with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
@@
 
+Problems and caveats:
+
 <!--**Availability**-->

### 2018-05-23 Datelife_paper1.Rmd
something

--- 2018-05-23 Datelife_paper1.Rmd
@@
 In the past decades, the possibility to obtain DNA sequences in large quantities from a wide variety of organisms became a reality, which, coupled to methodological development in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon ages across the tree of life.
+<!-- figure showing growth on studies with age data from references table from ttol-->
 To date, there is a large amount of both fossil and molecular-based data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE and Open Tree of Life (OToL).
@@
 As a science communication tool it is very powerful, since it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them.
-However, the primary data source and the synthetic chornogram are not publicly available, and large-scale, automated, data-mining is not allowed, minimizing its potential usefulness for the scientific community.
+However, the primary source of data and the synthetic chronogram are not publicly available, and large-scale, automated, data-mining is not allowed, minimizing its potential usefulness for the scientific community.
 <!--

### 2018-05-24 Datelife_paper1.Rmd

--- 2018-05-23 Datelife_paper1.Rmd
@@
 One advantage of TimeTree is that it includes taxa from across the tree of life, versus more specialized chronograms focusing on plants [PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups.
-Users can use a web interface or a mobile app to receive information on divergence times for pairs of species, ______.
+Users can choose between a web interface or a mobile app to receive information on divergence times for the evolutionary history of a lineage, pairs of taxa, all lineages within a taxon, or a list of taxa.
+As a science communication tool, TimeTree project is a very powerful tool: it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them.
 It is of limited utility for scientific studies, however.
 The thousands of trees that have been entered are unavailable for examination or reuse; according to their FAQ, methods for allowing download of the data have been under discussion since 2015.
-There is no API allowing programmatic access to any data.
-The nearly hundred thousand taxon chronogram is not available with [@Kumar2017] or the TimeTree website, though the still substantial chronogram from @Hedges2015 is available.
-Large-scale, automated, data-mining is not allowed under the site's terms of use.
-As a science communication tool it is very powerful, since it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them.
+Also, there is no Application Programming Interface (API) allowing programmatic access to any data.
+Moreover, large-scale, automated, data-mining is not allowed under the site's terms of use.
+The nearly hundred thousand taxon chronogram is not available with [@Kumar2017] or the TimeTree website, though the still substantial chronogram from [@Hedges2015] is available at OToL.
 

### 2019-01-16 Datelife_paper1.Rmd
last paragraph of intro

@@
 <!--INTRODUCTION-->
-Divergence time of lineages<!--named biological entities, taxon ages?--> constitutes in many ways the main knowledge necessary for evolutionary understanding.
+Time of lineage divergence<!--named biological entities, taxon ages?--> constitutes in many ways the fundamental/main knowledge necessary for evolutionary understanding.
 <!--Along with phylogenetic relationships, it is the most important piece of information allowing to test alternative evolutionary hypothesis in many areas of research, from developmental to ecological questions.
 Along with migration rates and patterns, it is fundamental to test biogeographical hypothesis-->
-Coupled to species number and distribution, it is the basic information for the study of diversification processes, such as the tempo and mode of speciation and extinction, crucial for the understanding of how biodiversity patterns are shaped across space and time [@Morlon2014].
-
+Coupled to species number and distribution, it is the main information necessary for the study of diversification processes (i.e., the tempo and mode of speciation and extinction),
+<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity patterns are shaped across space, time and clades [@Morlon2014].
+Evolutionary understanding also relies on comparative studies, for which knowing the time context for all life is crucial.
+Efforts to have a whole tree of life have been great and here are some examples.
+<!--INTRODUCTION: 2nd paragraph-->
 When organisms are preserved in a fossil form, a time frame of taxon origin can be obtained directly from the age of rock strata.
 However, not all organisms fossilize well or at all.
-Fossilization success alone is highly circumstancial, and varies depending on a number of parameters including the nature of the habitat, population size, species range breadth and physical characteristics of the organism.
+Fossilization success alone is highly circumstancial, and varies depending on a number of parameters including the nature of the habitat, population size, species range breadth and physical characteristics of the organism, all of which greatly vary across different organisms and through time.
 <!--fossil recovery is also highly circumstancial-->
 Thus, relying only on the fossil record to obtain a time frame of lineage divergence<!--taxon ages--> for all life is<!--, alas,--> not possible.
-Relative rate of DNA or aminoacid substitution constitutes another important source of information on lineage divergence.
-<!--describe the two components of rel subst rate: which, contain tne information on absolute rate of substitution and absolute time.-->
-It is usually obtained from hypothesis of character homology (alignments) when reconstructing phylogenetic relationships.
-Molecular dating techniques use external data such as absolute time calibrations (e.g., fossils, geologic events) or absolute substitution rates <!--birds--> to generate dated phylogenies (chronograms) which contain information on absolute times of node divergence and taxon ages.
+
+Another caveat is that we usually cannot have a point age estimates of lineage origin because of the nature of the fossilization process: when a fossil first appears on the fossil record it is not necessarily because the lineage just originated, it might just be bc the conditions for fossilization started at that point but the organism might have been around for a long time before that.
+Because of this, we use ranges from the age of strata where fossils are found.
+But this does not solve the problem.
+The organism could even have been around previously, but not fossilize under the conditions prevailing around the time of formation of rock strata below its first appearance in the fossil record.
+In this sense, fossil ages can only be considered as minimum time of origin of lineages.
+<!--D Bapst work?-->
+
+Another source that has been widely used to inform about the timing of lineage origin is the relative rate of DNA or aminoacid substitution.
+It is estimated from hypothesis of character homology (alignments) for reconstructing phylogenetic relationships.
+<!--describe the two components of rel subst rate? bc they contain the information on absolute rate of substitution and absolute time.-->
+Molecular dating techniques use external data such as absolute time calibrations (e.g., fossils, geologic events) or absolute substitution rates <!--obtained in the lab?? birds-->
+to <!--dissociate rel subs rates components and--> generate dated phylogenies (chronograms) which contain information on absolute times of node divergence and taxon ages.
 <!--With molecular dating methods, an external calibration point is still necessary to estimate the absolute time of origin.
@@
 <!--Models allow to take into account the nature of the fossil record: Given the characteristics of the fossilization process, the quality of dates inferred directly from it is not always considered the best.-->
-
-In the past decades, the possibility to obtain DNA sequences in large quantities from a wide variety of organisms became a reality, which, coupled to methodological development in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon ages across the tree of life.
+<!--INTRODUCTION: 3rd paragraph-->
+In the past two decades, the possibility to obtain good quality DNA sequences
+<!-- from a wide variety of organisms became a reality, which,  -->
+coupled to methodological developments in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon ages across the tree of life.
 <!-- figure showing growth on studies with age data from references table from ttol-->
@@
 Methods to include living and fossil lineages are in continued development and increased usage by the community<!--total evidence and revbayes (fossilized birth-death)-->, which coupled to better sharing data practices, are greatly contributing to the accumulation in number and type of available data on taxon ages.
-
 <!--What led to datelife development?
@@
 -->
-One inspiration for the DateLife project was the TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017].
-This project has aggregated chronograms from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
-However, even in this gold standard resource, the included taxa only include between 0.097 and 3.236% of total species diversity (following taxonomic expert opinion on the global, extant species numbers, which ranges from 3 to 100 million species [Mayr2010; Moran2011]).
+
+<!--INTRODUCTION: 4th paragraph-->
+The TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017] has aggregated chronograms from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
+However, even in this gold standard resource, the included taxa only encompass between 0.097 and 3.236% of total species diversity (following taxonomic expert opinion on the global, extant species numbers, which ranges from 3 to 100 million species [Mayr2010; Moran2011]).
 One advantage of TimeTree is that it includes taxa from across the tree of life, versus more specialized chronograms focusing on plants [PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups.
 Users can choose between a web interface or a mobile app to receive information on divergence times for the evolutionary history of a lineage, pairs of taxa, all lineages within a taxon, or a list of taxa.
-As a science communication tool, TimeTree project is a very powerful tool: it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them.
+As a science communication tool, TimeTree project is very powerful: it has a friendly graphical interface, with informative and colorful outputs, that allows the general public to satisfy curiosity regarding a particular organism of interest or group of them.
 It is of limited utility for scientific studies, however.
-The thousands of trees that have been entered are unavailable for examination or reuse; according to their FAQ, methods for allowing download of the data have been under discussion since 2015.
-Also, there is no Application Programming Interface (API) allowing programmatic access to any data.
-Moreover, large-scale, automated, data-mining is not allowed under the site's terms of use.
-The nearly hundred thousand taxon chronogram is not available with [@Kumar2017] or the TimeTree website, though the still substantial chronogram from [@Hedges2015] is available at OToL.
+The thousands of trees that have been entered are unavailable for examination or reuse; according to the creators (see TimeTree web FAQ), methods for allowing data downloading have been under discussion for the past several years yet the primary data remain closed.
+Moreover, there is no Application Programming Interface (API) allowing programmatic access to any data, greatly impairing the possibility of large-scale, automated data-mining, which is not allowed under TimeTree website's terms of use.
+The nearly hundred thousand taxon summary chronogram generated from TimeTree resources is not available with its publication [@Kumar2017] or the TimeTree website, though the still substantial chronogram from a previous publication [@Hedges2015] was made available at OToL.
 
 With this inspiration, a prototype DateLife service was developed over a series of phylotastic hackathons [__CITE__] at the National Evolutionary Synthesis Center.
-NSF funding allowed for the further development leading to this paper.
-A core goal has been openness of both the data sources and the code underlying the analyses.
+NSF funding allowed for further development leading to this paper.
+A core goal of DateLife is openness of both the data sources and the code underlying the analyses, so scientific community can take advantage of this tool to leverage available information to the advancement of discoveries in biology.
 
@@
 All searches allow species names and higher group names. Common names are allowed in all searches but a list of taxa.
-one-taxon search was nor working for dog, though.
+one-taxon search was not working for dog, though.
 1) ttol one lineage: it gives the evolutionary timeline of a lineage, showing the age of main divergence events occuring along the lineage evolutionary history, i.e., when it shared a comon ancestor with all main lineages from which it diverged a long time ago.
@@
 
-Despite its importance, analytical tools to summarize available information on taxon ages for the scientific community are still lacking.
-Data might not have been exploited because:
-Data is in different repositories and formats;
-Lineage names are different among studies and difficult to reconcile;
-Taxonomy is also different among studies and difficult to reconcile.
+Despite its great importance, analytical tools to summarize available information on taxon ages for the scientific community are still lacking.
+We identified several aspects that might have so far delayed the exploitation of existing data.
+First, original chronograms available publicly are scattered across various repositories (otol tree store, dryad, treebase, journals supplementary data) usually with different formats too.
+Second, lineage names due to taxonomic idiosincracy can be different among studies and manual curation of that is usually necessary.
+Third, data curation
+Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
+Generate new dates using all available DNA sequence information;
+Perform one global analysis using all available information;
+Problems or downsides: This might be time consuming for large groups and a lot of data curation and knowledge on the group of interest is still necessary.
+For example, <!--General issues with dating techniques:--> choosing correct fossils for calibration requires a lot of expertise and knowledge on the group.
+An incorrect use of fossils can generate severe bias in dating results [@Sauquet2012c].
 <!--Data curation is necessary at some point (is this always true?)-->
-Also, data curation is an important part of any biological study.
+Hence, data curation is still an important part of any biological study.
 The research community considers it as an important or even crucial step before data analysis.
 Hence, automated processes for large data analysis are frequently received with skepticism.
-Recent work on this area (e.g., supersmart<!--and, which others?-->) aims to:
-Generate new dates using all available DNA sequence information;
-Perform one global analysis using all available information;
-Problems or downsides: This might be time consuming for large groups and a lot of data curation is still necessary.
-<!--General issues with dating techniques:-->
-Choosing correct fossils for calibration requires a lot of expertise and knowledge on the group.
-Incorrect use of fossils can generate severe bias in dating results [@Sauquet2012c].
-DateLife palliates this by using only information available from already published studies, which are ideally constructed using robust information, such as sequence data and curated fossil calibrations.
+<!--INTRODUCTION: 5th paragraph-->
+
+DateLife palliates this by only using information available from already published studies, which are ideally constructed using robust information, such as sequence data and thoughtfully curated fossil calibrations.
 <!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
@@
 It can be used through an R package <link to documentation>, a web interface (<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
+In here we present the first release of DateLife.
+It contains an improved databse of chronograms, more methods to summarize trees, and new functions to visualize data.
+It also allows comparison of summary methods.
 
@@
 
-<!--(BOLD and OToL species names are homogeneous?-->
+<!--(BOLD and OToL species names, are they homogeneous/standardized?-->
 
 DateLife is a service for searching and processing information on ages of any taxon <!--or group of taxa-->of interest, available in chronograms from public data repositories of published peer reviewed studies.
-With this data, it can also generate new taxon age information<!--by using expert available phylogenies and scaling them to absolute time using information on ages obtained from published chronograms,--> by calling a variety of external services and proved methods.
-It only requires as input a set of taxon names, in the form of a comma separated listing or vector, or of a phylogeny with taxon names on the tips.
+From these data, it can also generate new taxon age information<!--by using expert available phylogenies and scaling them to absolute time using information on ages obtained from published chronograms,--> by calling a variety of external services and tested algorithms.
+It takes advatage of the rotl, ape, and geiger packages to gather, process, and present information.
+
+It only requires a set of taxon names as input, in the form of a comma separated listing or vector, or of a phylogeny with taxon names on the tips.
 Taxon names can correspond to binomial species names or clades<!--common names (will we implement it?)-->.
@@
 DateLife's main function searches taxon names across the chronogram database specified by the user.
-At the moment, it queries chronograms from the OToL [@Hinchliff2015]<!-- there is an earlier reference from 1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive tool for browsing the phylogeny of life.
+At the moment, it queries chronograms from the OToL [@Hinchliff2015]<!-- it cannot query chronograms from TreeBASE -there is an earlier reference from 1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive tool for browsing the phylogeny of life.
 Am J Bot.
@@
 It then stores taxon age information from each chronogram individually as a patristic matrix, named with the citation of the original study.
-This format allows a rapid summary in a number of different ways, including: 1) citations of the original studies containing the subset chronograms, 2) a list of subset chronograms' mrca ages, 3) a list of subset chronograms in newick or phylo format, 4) a table containing all information in html or R's data frame format, or 5) as a single chronogram summarized from subset chronograms using the supertree Super Distance Matrix (SDM) approach [@Criscuolo2006] or using the median of branch lengths.
+This format allows a rapid summary in a number of different ways, including: 1) citations of the original studies containing the subset chronograms, 2) a list of mrca ages of subset chronograms, 3) a list of complete subset chronograms in newick or phylo format, 4) a table containing all information retrieved in html or R's data frame format, or 5) a single chronogram summarized from subset chronograms using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths.
 <!--(how to better explain the latter??
@@
 Problems and caveats:
+Not many databases, only OToL
+Why TreeBase is not very useful for us? Be precise.
 
@@
 DateLife is free and open source and it can be used through its current website <http://www.datelife.org/query/>, or through Phylotastic!'s web portal <http://phylo.cs.nmsu.edu:3000/>. 
+RStudio's Shiny Server and the shiny package open infrastructure are used to maintain the former. Also Docker.
 DateLife can also be used locally through its R package.
@@
 </span>
+We thank colleagues (students and postdocs) at the O'Meara Lab at the University of Tennesse Knoxville for suggestions, discussions and software testing.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons that led to initial work on this project.
+The Open Tree of Life project that provides the open, metadata rich repository of trees used for DateLife.
+The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in OpenTree.
+The US National Science Foundation (NSF) for funding nearly all the above, in addition to the ABI grant that funded this project itself.
 

### 2019-01-20 Datelife_paper1.Rmd
some figures

@@
 ```{r eval = TRUE, echo = FALSE}
-figure_runtime_main <- "Fig. 1"  # make_datelife_query_plus_get_datelife_result
+fig_workflow_main <- "Fig. 1"
+fig_runtime_main <- "Fig. 2"  
+# make_datelife_query_plus_get_datelife_result
 ```
@@
 <!--INTRODUCTION-->
-Time of lineage divergence<!--named biological entities, taxon ages?--> constitutes in many ways the fundamental/main knowledge necessary for evolutionary understanding.
-<!--Along with phylogenetic relationships, it is the most important piece of information allowing to test alternative evolutionary hypothesis in many areas of research, from developmental to ecological questions.
-Along with migration rates and patterns, it is fundamental to test biogeographical hypothesis-->
-Coupled to species number and distribution, it is the main information necessary for the study of diversification processes (i.e., the tempo and mode of speciation and extinction),
-<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity patterns are shaped across space, time and clades [@Morlon2014].
-Evolutionary understanding also relies on comparative studies, for which knowing the time context for all life is crucial.
-Efforts to have a whole tree of life have been great and here are some examples.
+Time of lineage divergence<!--alt lineages: named biological entities, taxon ages?-->
+constitutes in many ways the fundamental/main knowledge necessary for evolutionary
+understanding.
+<!--Along with phylogenetic relationships, it is the most important piece of information
+allowing to test alternative evolutionary hypothesis in many areas of research,
+from developmental to ecological questions.
+Along with migration rates and patterns, it is fundamental to test biogeographical
+hypothesis-->
+Coupled to species number and distribution, it is the main information necessary
+for the study of diversification processes (i.e., the tempo and mode of speciation
+and extinction),
+<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity
+patterns are shaped across space, time and clades [@Morlon2014].
+Evolutionary understanding also relies on comparative studies, for which knowing
+the time context for all life is crucial. Efforts to have a whole tree of life have
+been great and here are some examples.
 <!--INTRODUCTION: 2nd paragraph-->
-When organisms are preserved in a fossil form, a time frame of taxon origin can be obtained directly from the age of rock strata.
+When organisms are preserved in a fossil form, a time frame of taxon origin can
+be obtained directly from the age of rock strata.
 However, not all organisms fossilize well or at all.
-Fossilization success alone is highly circumstancial, and varies depending on a number of parameters including the nature of the habitat, population size, species range breadth and physical characteristics of the organism, all of which greatly vary across different organisms and through time.
+Fossilization success alone is highly circumstancial, and varies depending on a
+number of parameters including the nature of the habitat, population size, species
+range breadth and physical characteristics of the organism, all of which greatly
+vary across different organisms and through time.
 <!--fossil recovery is also highly circumstancial-->
-Thus, relying only on the fossil record to obtain a time frame of lineage divergence<!--taxon ages--> for all life is<!--, alas,--> not possible.
+Thus, relying only on the fossil record to obtain a time frame of lineage divergence
+<!--taxon ages--> for all life is<!--, alas,--> not possible.
 
-Another caveat is that we usually cannot have a point age estimates of lineage origin because of the nature of the fossilization process: when a fossil first appears on the fossil record it is not necessarily because the lineage just originated, it might just be bc the conditions for fossilization started at that point but the organism might have been around for a long time before that.
-Because of this, we use ranges from the age of strata where fossils are found.
-But this does not solve the problem.
-The organism could even have been around previously, but not fossilize under the conditions prevailing around the time of formation of rock strata below its first appearance in the fossil record.
-In this sense, fossil ages can only be considered as minimum time of origin of lineages.
+Another caveat is that we usually cannot have a point age estimates of lineage origin
+because of the nature of the fossilization process: when a fossil first appears
+on the fossil record it is not necessarily because the lineage just originated,
+it might just be bc the conditions for fossilization started at that point but the
+organism might have been around for a long time before that. Because of this, we
+use ranges from the age of strata where fossils are found. But this does not solve
+the problem. The organism could even have been around previously, but not fossilize
+under the conditions prevailing around the time of formation of rock strata below
+its first appearance in the fossil record. In this sense, fossil ages can only be
+considered as minimum time of origin of lineages.
 <!--D Bapst work?-->
 
-Another source that has been widely used to inform about the timing of lineage origin is the relative rate of DNA or aminoacid substitution.
-It is estimated from hypothesis of character homology (alignments) for reconstructing phylogenetic relationships.
-<!--describe the two components of rel subst rate? bc they contain the information on absolute rate of substitution and absolute time.-->
-Molecular dating techniques use external data such as absolute time calibrations (e.g., fossils, geologic events) or absolute substitution rates <!--obtained in the lab?? birds-->
-to <!--dissociate rel subs rates components and--> generate dated phylogenies (chronograms) which contain information on absolute times of node divergence and taxon ages.
-<!--With molecular dating methods, an external calibration point is still necessary to estimate the absolute time of origin.
-Hence, the accumulation of data on fossils for calibration is also continual-->
-<!--This method has been particularly useful to obtain time frame of origin of organisms with poor or no fossil record, but is also largely used for organisms with rich fossil records-->
-<!--Models allow to take into account the nature of the fossil record: Given the characteristics of the fossilization process, the quality of dates inferred directly from it is not always considered the best.-->
+Another source that has been widely used to inform about the timing of lineage origin
+is the relative rate of DNA or aminoacid substitution.
+It is estimated from hypothesis of character homology (alignments) for reconstructing
+phylogenetic relationships.
+<!--describe the two components of rel subst rate? bc they contain the information
+on absolute rate of substitution and absolute time.-->
+Molecular dating techniques use external data such as absolute time calibrations
+(e.g., fossils, geologic events) or absolute substitution rates <!--obtained in
+the lab?? birds-->to <!--dissociate rel subs rates components and--> generate dated
+phylogenies (chronograms) which contain information on absolute times of node divergence
+and taxon ages. <!--With molecular dating methods, an external calibration point
+is still necessary to estimate the absolute time of origin. Hence, the accumulation
+of data on fossils for calibration is also continual-->
+<!--This method has been particularly useful to obtain time frame of origin of organisms
+with poor or no fossil record, but is also largely used for organisms with rich
+fossil records-->
+<!--Models allow to take into account the nature of the fossil record: Given the
+characteristics of the fossilization process, the quality of dates inferred directly
+from it is not always considered the best.-->
 <!--INTRODUCTION: 3rd paragraph-->
@@
 <!-- from a wide variety of organisms became a reality, which,  -->
-coupled to methodological developments in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon ages across the tree of life.
+coupled to methodological developments in phylogenetic and dating inference, allowed
+the application of molecular dating methods on a very large amount and diversity
+of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon
+ages across the tree of life.
 <!-- figure showing growth on studies with age data from references table from ttol-->
-To date, there is a large amount of both fossil and molecular-based data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE and Open Tree of Life (OToL).
+To date, there is a large amount of both fossil and molecular-based data on taxon
+ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
+and Open Tree of Life (OToL).
 OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
@@
 <!--How many trees? How many dated trees?-->
-Methods to include living and fossil lineages are in continued development and increased usage by the community<!--total evidence and revbayes (fossilized birth-death)-->, which coupled to better sharing data practices, are greatly contributing to the accumulation in number and type of available data on taxon ages.
+Methods to include living and fossil lineages are in continued development and increased
+usage by the community<!--total evidence and revbayes (fossilized birth-death)-->,
+which coupled to better sharing data practices, are greatly contributing to the
+accumulation in number and type of available data on taxon ages.
 <!--What led to datelife development?
@@
 <!--INTRODUCTION: 4th paragraph-->
-The TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017] has aggregated chronograms from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
-However, even in this gold standard resource, the included taxa only encompass between 0.097 and 3.236% of total species diversity (following taxonomic expert opinion on the global, extant species numbers, which ranges from 3 to 100 million species [Mayr2010; Moran2011]).
-One advantage of TimeTree is that it includes taxa from across the tree of life, versus more specialized chronograms focusing on plants [PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups.
-Users can choose between a web interface or a mobile app to receive information on divergence times for the evolutionary history of a lineage, pairs of taxa, all lineages within a taxon, or a list of taxa.
-As a science communication tool, TimeTree project is very powerful: it has a friendly graphical interface, with informative and colorful outputs, that allows the general public to satisfy curiosity regarding a particular organism of interest or group of them.
-It is of limited utility for scientific studies, however.
-The thousands of trees that have been entered are unavailable for examination or reuse; according to the creators (see TimeTree web FAQ), methods for allowing data downloading have been under discussion for the past several years yet the primary data remain closed.
-Moreover, there is no Application Programming Interface (API) allowing programmatic access to any data, greatly impairing the possibility of large-scale, automated data-mining, which is not allowed under TimeTree website's terms of use.
-The nearly hundred thousand taxon summary chronogram generated from TimeTree resources is not available with its publication [@Kumar2017] or the TimeTree website, though the still substantial chronogram from a previous publication [@Hedges2015] was made available at OToL.
+The TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017] has aggregated chronograms
+from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
+However, even in this gold standard resource, the included taxa only encompass between
+0.097 and 3.236% of total species diversity (following taxonomic expert opinion
+on the global, extant species numbers, which ranges from 3 to 100 million species
+[Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from
+across the tree of life, versus more specialized chronograms focusing on plants
+[PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups. Users can choose
+between a web interface or a mobile app to receive information on divergence times
+for the evolutionary history of a lineage, pairs of taxa, all lineages within a
+taxon, or a list of taxa. As a science communication tool, TimeTree project is
+very powerful: it has a friendly graphical interface, with informative and colorful
+outputs, that allows the general public to satisfy curiosity regarding a particular
+organism of interest or group of them. It is of limited utility for scientific
+studies, however. The thousands of trees that have been entered are unavailable
+for examination or reuse; according to the creators (see TimeTree web FAQ), methods
+for allowing data downloading have been under discussion for the past several years
+yet the primary data remain closed. Moreover, there is no Application Programming
+Interface (API) allowing programmatic access to any data, greatly impairing the
+possibility of large-scale, automated data-mining, which is not allowed under TimeTree
+website's terms of use. The nearly hundred thousand taxon summary chronogram generated
+from TimeTree resources is not available with its publication [@Kumar2017] or the
+TimeTree website, though the still substantial chronogram from a previous publication
+[@Hedges2015] was made available at OToL.
 
-With this inspiration, a prototype DateLife service was developed over a series of phylotastic hackathons [__CITE__] at the National Evolutionary Synthesis Center.
-NSF funding allowed for further development leading to this paper.
-A core goal of DateLife is openness of both the data sources and the code underlying the analyses, so scientific community can take advantage of this tool to leverage available information to the advancement of discoveries in biology.
+With this inspiration, a prototype DateLife service was developed over a series
+of phylotastic hackathons [__CITE__] at the National Evolutionary Synthesis Center.
+NSF funding allowed for further development leading to this paper. A core goal of
+DateLife is openness of both the data sources and the code underlying the analyses,
+so scientific community can take advantage of this tool to leverage available information
+to the advancement of discoveries in biology.
 
@@
 <!--
-While its node-time search is very powerful, using available data to infer divergence time from lineages that have never been sequenced before, its timetree option is less powerful, their main flaw being that it does not provide source data.
-However, their primary database is not fully open, giving information on a per node basis.
-So, it is not possible to easily access information from original studies for a group of taxa (One would need to perform a pair of taxa search for all pairs from the group of interest).
-Divergence time estimates for all taxa in the TToL are precomputed and stored in the TimeTree database
-All searches allow species names and higher group names. Common names are allowed in all searches but a list of taxa.
+While its node-time search is very powerful, using available data to infer divergence
+time from lineages that have never been sequenced before, its timetree option is
+less powerful, their main flaw being that it does not provide source data.
+However, their primary database is not fully open, giving information on a per node
+basis. So, it is not possible to easily access information from original studies
+for a group of taxa (One would need to perform a pair of taxa search for all pairs
+from the group of interest).
+Divergence time estimates for all taxa in the TToL are precomputed and stored in
+the TimeTree database
+All searches allow species names and higher group names. Common names are allowed
+in all searches but a list of taxa.
 one-taxon search was not working for dog, though.
-1) ttol one lineage: it gives the evolutionary timeline of a lineage, showing the age of main divergence events occuring along the lineage evolutionary history, i.e., when it shared a comon ancestor with all main lineages from which it diverged a long time ago.
-It shows results only graphically on the website.
+1) ttol one lineage: it gives the evolutionary timeline of a lineage, showing the
+age of main divergence events occuring along the lineage evolutionary history, i.e.,
+when it shared a comon ancestor with all main lineages from which it diverged a
+long time ago.
+It only shows results graphically on the website.
 2) ttol parwise search: allows common names, species names and higher group names
-    how it handles names missing from tree: it looks in the NCBI taxonomy for the closest relative of the species or higher taxa. Then uses the closest relative to get the lineage pair age from ttol.
+    how it handles names missing from tree: it looks in the NCBI taxonomy for the
+    closest relative of the species or higher taxa. Then uses the closest relative
+    to get the lineage pair age from ttol.
 3) ttol group of taxa
-4) ttol list of taxa: Cannot get it to work; does not work with cat, dog, turtle search, felidae, hominidae, phrynops search, some_ants species name search.
-What kind of files does it accepts: txt, rtf, doc, docx?
+4) ttol list of taxa: Cannot get it to work; does not work with cat, dog, turtle
+search, felidae, hominidae, phrynops search, some_ants species name search. What
+kind of files does it accepts: txt, rtf, doc, docx?
 -->
@@
 <!--
-retrieves information takes available data on lineage ages and phylogenetic relationships to construct a tree of all life scaled to absolute time: a chronogram of all life, while allowing the general public to search for age data within the tree for one taxon of interest (timeline option), a pair of taxa (node time option), a group of taxa (it can only be a single name of a taxonomic group), or a list of taxa of interest (timetree option for the two latter).
+retrieves information takes available data on lineage ages and phylogenetic relationships
+to construct a tree of all life scaled to absolute time: a chronogram of all life,
+while allowing the general public to search for age data within the tree for one
+taxon of interest (timeline option), a pair of taxa (node time option), a group
+of taxa (it can only be a single name of a taxonomic group), or a list of taxa of
+interest (timetree option for the two latter).
 Good things:
-Allows to get ages of species with no data, by extrapolating from closely related species with data.
-Allows comparison of timelines to geologic events, such as asteroid impacts, oxygen and carbon dioxide levels, solar luminosity.
+Allows to get ages of species with no data, by extrapolating from closely related
+species with data.
+Allows comparison of timelines to geologic events, such as asteroid impacts, oxygen
+and carbon dioxide levels, solar luminosity.
 Not so good things:
-Only allows to download some data from original sources if performing a pair-of-taxa search.
-Only allows to see age ranges or confidence intervals in graphical form, when hovering over nodes.
-These CIs come from ttol.
-It does not have CIs from original studies.
-Node hovering is activated on one-taxon evolutionary history, and group-of-taxa searches (we could not get to work list-of-taxa search).
-Pair-of-taxa search shows the confidence interval of the only ttol node that constitutes the taxon-pair mrca.
-You can only download data from pair-of-taxa and group-of-taxa search.
-It generates a table with ages from original studies and a tree pruned from ttol, respectively.
-In both cases it gives single point age estimates.
-We cannot download CI information from ttol.
+Only allows to download some data from original sources if performing a pair-of-taxa
+search.
+Only allows to see age ranges or confidence intervals in graphical form, when hovering
+over nodes. These CIs come from ttol. It does not have CIs from original studies.
+Node hovering is activated on one-taxon evolutionary history, and group-of-taxa
+searches (we could not get to work list-of-taxa search). Pair-of-taxa search shows
+the confidence interval of the only ttol node that constitutes the taxon-pair mrca.
+You can only download data from pair-of-taxa and group-of-taxa search. It generates
+a table with ages from original studies and a tree pruned from ttol, respectively.
+In both cases it gives single point age estimates. We cannot download CI information
+from ttol.
 -->
 
-Despite its great importance, analytical tools to summarize available information on taxon ages for the scientific community are still lacking.
-We identified several aspects that might have so far delayed the exploitation of existing data.
-First, original chronograms available publicly are scattered across various repositories (otol tree store, dryad, treebase, journals supplementary data) usually with different formats too.
-Second, lineage names due to taxonomic idiosincracy can be different among studies and manual curation of that is usually necessary.
+Despite its great importance, analytical tools to summarize available information
+on taxon ages for the scientific community are still lacking.
+We identified several aspects that might have so far delayed the exploitation of
+existing data. First, original chronograms available publicly are scattered across
+various repositories (otol tree store, dryad, treebase, journals supplementary data)
+usually with different formats too.
+Second, lineage names due to taxonomic idiosincracy can be different among studies
+and manual curation of that is usually necessary.
 Third, data curation
@@
 Perform one global analysis using all available information;
-Problems or downsides: This might be time consuming for large groups and a lot of data curation and knowledge on the group of interest is still necessary.
-For example, <!--General issues with dating techniques:--> choosing correct fossils for calibration requires a lot of expertise and knowledge on the group.
-An incorrect use of fossils can generate severe bias in dating results [@Sauquet2012c].
+Problems or downsides: This might be time consuming for large groups and a lot of
+data curation and knowledge on the group of interest is still necessary. For example,
+<!--General issues with dating techniques:--> choosing correct fossils for calibration r
+equires a lot of expertise and knowledge on the group. An incorrect use of fossils
+can generate severe bias in dating results [@Sauquet2012c].
 <!--Data curation is necessary at some point (is this always true?)-->
-Hence, data curation is still an important part of any biological study.
-The research community considers it as an important or even crucial step before data analysis.
+Hence, data curation is still an important part of any biological study. The research
+community considers it as an important or even crucial step before data analysis.
 Hence, automated processes for large data analysis are frequently received with skepticism.
@@
 
-DateLife palliates this by only using information available from already published studies, which are ideally constructed using robust information, such as sequence data and thoughtfully curated fossil calibrations.
+DateLife palliates this by only using information available from already published
+studies, which are ideally constructed using robust information, such as sequence
+data and thoughtfully curated fossil calibrations.
 <!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
 <!--Importance of datelife:-->
-DateLife can summarize this information in several formats that can be easily inspected by users.
-This allows rapidly obtaining a time frame of lineage divergence for a wide number of taxa.
-DateLife can also generate chronograms for taxa with little available information, by using the available data as calibration points.
-DateLife is the main service for scaling phylogenetic trees in Phylotastic! system [@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
-It can be used through an R package <link to documentation>, a web interface (<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
+DateLife can summarize this information in several formats that can be easily inspected
+by users.
+This allows rapidly obtaining a time frame of lineage divergence for a wide number
+of taxa.
+DateLife can also generate chronograms for taxa with little available information,
+by using the available data as calibration points.
+DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
+[@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
+It can be used through an R package <link to documentation>, a web interface
+(<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
 In here we present the first release of DateLife.
-It contains an improved databse of chronograms, more methods to summarize trees, and new functions to visualize data.
+It contains an improved databse of chronograms, more methods to summarize trees,
+and new functions to visualize data.
 It also allows comparison of summary methods.
@@
 </span>
-
+```{r, echo = FALSE}
+fig_workflow_main_foot <- "Figure 1. Summarized DateLife workflow. Only main actions are shown."
+```
 <!--(BOLD and OToL species names, are they homogeneous/standardized?-->
 
-DateLife is a service for searching and processing information on ages of any taxon <!--or group of taxa-->of interest, available in chronograms from public data repositories of published peer reviewed studies.
-From these data, it can also generate new taxon age information<!--by using expert available phylogenies and scaling them to absolute time using information on ages obtained from published chronograms,--> by calling a variety of external services and tested algorithms.
-It takes advatage of the rotl, ape, and geiger packages to gather, process, and present information.
+DateLife is a service for searching and processing information on ages of any number
+of taxa <!--or group of taxa-->of interest, across chronograms available in public
+data repositories coming from published peer reviewed studies. It can also generate
+new taxon age information<!--by using expert available phylogenies and scaling them
+to absolute time using information on ages obtained from published chronograms,--> by l
+inking several external services and tested algorithms.
+It takes advantage of the rotl, ape, and geiger packages to gather, process, and
+present information.
 
-It only requires a set of taxon names as input, in the form of a comma separated listing or vector, or of a phylogeny with taxon names on the tips.
-Taxon names can correspond to binomial species names or clades<!--common names (will we implement it?)-->.
-When taxon names are clades, DateLife pulls all accepted species names within the clade (up to OToL's limit of _____ species) from OToL's reference taxonomy using a service of rphylotastic R package<!--citation-->.
-Names belonging to subspecies or any other infraspecific category are treated as species.
-DateLife can process input names with the taxon name resolution service (TNRS), <!--which "corrects misspelled names and authorities, standardizes variant spellings, and converts nomenclatural synonyms to accepted names" (Boyle et al.
-2013)-->which corrects misspelled names or typos, and standardizes variation in spelling and synonyms [@Boyle2013], increasing the probability to correctly find the queried taxa in the chronogram database.
-DateLife uses TNRS to compare names against OToL's reference taxonomy using a service from the R package rotl [@Michonneau2016].
+It only requires a set of taxon names as input, in the form of a comma separated
+listing or vector, or of a phylogeny with taxon names on the tips.
+Taxon names can correspond to binomial species names or clades
+<!--common names (will we implement it?)-->.
+When taxon names are clades, DateLife pulls all accepted species names within the
+clade (up to OToL's limit of _____ species) from OToL's reference taxonomy using
+a service of rphylotastic R package<!--citation-->.
+Names belonging to subspecies or any other infraspecific category are treated as
+species.
+DateLife can process input names with the taxon name resolution service (TNRS),
+<!--which "corrects misspelled names and authorities, standardizes variant spellings,
+and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
+which corrects misspelled names or typos, and standardizes variation in spelling
+and synonyms [@Boyle2013], increasing the probability to correctly find the queried
+taxa in the chronogram database. DateLife uses TNRS to compare names against OToL's
+reference taxonomy using a service from the R package rotl [@Michonneau2016].
 <!--we'll implement the option to use other reference taxonomies in the future -->
 
-DateLife's main function searches taxon names across the chronogram database specified by the user.
-At the moment, it queries chronograms from the OToL [@Hinchliff2015]<!-- it cannot query chronograms from TreeBASE -there is an earlier reference from 1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive tool for browsing the phylogeny of life.
-Am J Bot.
-1994, 81 (6): 183--> repository.
-DateLife identifies chronograms having at least two taxon names, and subset them to contain only the taxa of interest.
-It then stores taxon age information from each chronogram individually as a patristic matrix, named with the citation of the original study.
-This format allows a rapid summary in a number of different ways, including: 1) citations of the original studies containing the subset chronograms, 2) a list of mrca ages of subset chronograms, 3) a list of complete subset chronograms in newick or phylo format, 4) a table containing all information retrieved in html or R's data frame format, or 5) a single chronogram summarized from subset chronograms using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths.
+DateLife's main function searches taxon names across the chronogram database specified
+by the user. At the moment, it queries chronograms from OToL [@Hinchliff2015]
+<!-- it cannot query chronograms from TreeBASE -there is an earlier reference from
+1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel
+WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive
+tool for browsing the phylogeny of life. Am J Bot. 1994, 81 (6): 183--> repository.
+DateLife identifies chronograms having at least two taxon names, and subset them
+to contain only the taxa of interest. It then stores taxon age information from
+each chronogram individually as a patristic matrix, named with the citation of the
+original study. This format allows a rapid summary in a number of different ways,
+including:
+1) citations of the original studies containing the subset chronograms,
+2) a list of mrca ages of subset chronograms,
+3) a list of complete subset chronograms in newick or phylo format,
+4) a table containing all information retrieved in html or R's data frame format, or
+5) a single chronogram summarized from subset chronograms using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths.
 <!--(how to better explain the latter??
@@
 DateLife also stores information on input taxon presence/absence across subset chronograms.
-Users can choose to add ages of missing taxa to subset chronograms in different ways, depending on the amount of knowledge they want to input or how much they want to be involved in the steps of the addition process.
-If users have no access to biological information (i.e., a character, DNA or protein matrix), missing taxa can be added to any chronogram simply at random, or by following taxonomic or phylogenetic knowledge from expert sources.
-There are a wide number of open reference taxonomies available, such as the Catalogue of Life [@Roskov2017] or the NCBI taxonomy database [@Federhen2012].
-Expert phylogenies (with or without branch lengths) to be used as topological constraint (backbone) can also be obtained from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002] and Dryad (<https://www.datadryad.org//>).
-At the moment, DateLife only uses OToL's synthetic tree and reference taxonomy as expert knowledge to automatically add missing taxa to chronograms.
-Alternatively, users can input a reference taxonomy or topological constraint of their choosing or making.
-<!-- at some point we will implement the possibility for users to choose different reference taxonomies automatically-->
-If OToL's synthetic tree is not satisfactorily resolved for the taxa of interest, DateLife can construct a sequence data matrix from DNA markers available from the Barcode of Life Database (BOLD; @Ratnasingham2007), to attempt to further resolve polytomies.
-It will follow OToL's synthetic tree as backbone.
-<!--This molecular tree can then be dated with chronosMPL from the ape package, but is it really needed??
-To estimate node ages chronosMPL uses mean path length method from Britton et al.
-2002-->
+Users can choose to add ages of missing taxa to subset chronograms in different
+ways, depending on the amount of knowledge they want to input or how much they want
+to be involved in the steps of the addition process.
+If users have no access to biological information (i.e., a character, DNA or protein
+matrix), missing taxa can be added to any chronogram simply at random, or by following
+taxonomic or phylogenetic knowledge from expert sources. There are a wide number
+of open reference taxonomies available, such as the Catalogue of Life [@Roskov2017]
+or the NCBI taxonomy database [@Federhen2012]. Expert phylogenies (with or without
+branch lengths) to be used as topological constraint (backbone) can also be obtained
+from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002]
+and Dryad (<https://www.datadryad.org//>).
+At the moment, DateLife only uses OToL's synthetic tree and reference taxonomy as
+expert knowledge to automatically add missing taxa to chronograms. Alternatively,
+users can input a reference taxonomy or topological constraint of their choosing
+or making. <!-- at some point we will implement the possibility for users to choose
+different reference taxonomies automatically-->
+If OToL's synthetic tree is not satisfactorily resolved for the taxa of interest,
+DateLife can construct a sequence data matrix from DNA markers available from the
+Barcode of Life Database (BOLD; @Ratnasingham2007), to attempt to further resolve
+polytomies. It will follow OToL's synthetic tree as backbone.
+<!--This molecular tree can then be dated with chronosMPL from the ape package, but is it really needed?? To estimate node ages chronosMPL uses mean path length method from Britton et al. 2002-->
 <!--It can rapidly construct trees with branch lengths from
@@
 available in BOLD (in plants and fungi in particular).-->
-To use information from a topological constraint, DateLife calls the congruification method described in [@Eastman2013] to find shared nodes between trees (congruent nodes).
-It then fixes their ages, and add ages to remaining nodes with a dating method that can be specified by the user.
-If users have access to biological data, they can input a tree with branch lengths proportional to relative substitution rates as topological constraint.
-In this case, age data from congruent nodes will be used as calibration points.
-Age data from several chronograms can be combined and congruified to be used as calibration points in a single analysis.
+To use information from a topological constraint, DateLife calls the congruification
+method described in [@Eastman2013] to find shared nodes between trees (congruent
+nodes). It then fixes their ages, and add ages to remaining nodes with a dating
+method that can be specified by the user. If users have access to biological data,
+they can input a tree with branch lengths proportional to relative substitution
+rates as topological constraint. In this case, age data from congruent nodes will
+be used as calibration points.
+Age data from several chronograms can be combined and congruified to be used as
+calibration points in a single analysis.
 
 Several dating methods are implemented in DateLife.
-Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
-DateLife implements BLADJ from the development R version of phylocom's R package [@Webb2008], phylocomr (<https://github.com/ropensci/phylocomr>).
-It can only be used when there is a topological constraint with no branch lengths.
-PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. It is also called through R. <!--It could be used with a reference taxonomy, if we provide a tree with politomies-->
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method [@Smith2012]. It is called through R.
-MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at random, following a reference taxonomy or a topological constraint.
-It draws ages from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
+Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
+nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
+DateLife implements BLADJ from the development R version of phylocom's R package
+[@Webb2008], phylocomr (<https://github.com/ropensci/phylocomr>). It can only be
+used when there is a topological constraint with no branch lengths.
+PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. It is
+also called through R. <!--It could be used with a reference taxonomy, if we provide a tree with politomies-->
+treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
+[@Smith2012]. It is called through R.
+MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
+random, following a reference taxonomy or a topological constraint. It draws ages
+from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
 DateLife calls MrBayes trough an R function.
@@
 with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
-We registered variation in computing time relative to number of input names <!--number of chronograms in database--> and DateLife service.
-Input processing increases roughly linearly with number of input taxon names, and increases considerably if tnrs service is activated (`r figure_runtime_main`).
+We registered variation in computing time relative to number of input names
+<!--number of chronograms in database--> and DateLife service.
+Input processing increases roughly linearly with number of input taxon names, and
+increases considerably if tnrs service is activated (`r fig_runtime_main`).
 <!--
@@
 -->
-Results show that searching time increases linearly with number of input names and number of chronograms in database.
+Results show that searching time increases linearly with number of input names and
+number of chronograms in database.
 
@@
 -->
-DateLife's code performance was evaluated with a set of unit tests designed and implemented with the R package testthat [@RCoreTeam2018].
-These tests were run both locally –using the devtools package [@RCoreTeam2018]– and on a public server –via GitHub– using the continuous integration tool Travis CI (<//travis-ci.org>).
-At present, unit tests cover around 30% (for now) of DateLife's code (<https://codecov.io/gh/phylotastic/datelife>).
-<!-- unit test make code robust, verifiable, debuggable, -->
+DateLife's code performance was evaluated with a set of unit tests designed and
+implemented with the R package testthat [@RCoreTeam2018]. These tests were run both
+locally –using the devtools package [@RCoreTeam2018]– and on a public server –via
+GitHub– using the continuous integration tool Travis CI (<//travis-ci.org>). At
+present, unit tests cover around 30% (for now) of DateLife's code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
 
@@
 Use ltts to compare for now.
-Fig. X2 shows comparison of available chronograms for Felidae species and chronograms generated through DateLife
+Fig. X2 shows comparison of available chronograms for Felidae species and chronograms
+generated through DateLife
 
@@
 
-Taxon ages are key to many areas of evolutionary studies: trait evolution, species diversification, biogeography, macroecology and more.
-Obtaining these ages is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to develop the necessary knowledge and data curation skills to produce new chronograms.
-Knowledge on taxon ages is also important for non-biological studies and the non-academic community.
-The combination of new analytical techniques, availability of more fossil and molecular data, and better practices in data sharing has resulted in a steady accumulation of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree of Life, for a large quantity and diversity of organisms.
-However, this information remains difficult to synthesize for many biologists and the non-academic community<!-- or just non biologists-->.
+Taxon ages are key to many areas of evolutionary studies: trait evolution, species
+diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
+especially for those who want to use phylogenies but who are not systematists, or
+do not have the time to develop the necessary knowledge and data curation skills
+to produce new chronograms. Knowledge on taxon ages is also important for non-biological
+studies and the non-academic community.
+The combination of new analytical techniques, availability of more fossil and molecular
+data, and better practices in data sharing has resulted in a steady accumulation
+of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree
+of Life, for a large quantity and diversity of organisms. However, this information
+remains difficult to synthesize for many biologists and the non-academic community<!-- or just non biologists-->.
 
-<!--Potential applications shown here-->Here, we have shown that DateLife allows an easy and fast obtention of all publicly available information on taxon ages, which can be used to generate new data.
-<!--Potential applications not shown here-->This information can be used to account for the effect of phylogenetic signal in studies of trait evolution; to explore potential speciation and extinction dynamics of interest within a clade; to obtain a time frame of biogeographical events; for science communication and outreach, amongst others.
-<!--Why you should use it-->Compared to similar platforms such as time tree of life and supermart, it offers several advantages.
+<!--Potential applications shown here-->Here, we have shown that DateLife allows
+an easy and fast obtention of all publicly available information on taxon ages,
+which can be used to generate new data. <!--Potential applications not shown here-->
+This information can be used to account for the effect of phylogenetic signal in
+studies of trait evolution; to explore potential speciation and extinction dynamics
+of interest within a clade; to obtain a time frame of biogeographical events; for
+science communication and outreach, amongst others.
+<!--Why you should use it-->Compared to similar platforms such as time tree of life
+and supermart, it offers several advantages.
 It is fast;
@@
 
-DateLife is free and open source and it can be used through its current website <http://www.datelife.org/query/>, or through Phylotastic!'s web portal <http://phylo.cs.nmsu.edu:3000/>. 
-RStudio's Shiny Server and the shiny package open infrastructure are used to maintain the former. Also Docker.
-DateLife can also be used locally through its R package.
-The stable version is available for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>) using the command `install.packages(pkgs = "datelife")` from R.
-Development versions are available from GitHub repository (<https://github.com/phylotastic/datelife>) and can be installed using the devtools R package command `install_github("phylotastic/datelife")`.
+DateLife is free and open source and it can be used through its current website
+<http://www.datelife.org/query/>, or through Phylotastic!'s web portal <http://phylo.cs.nmsu.edu:3000/>.
+RStudio's Shiny Server and the shiny package open infrastructure are used to maintain
+the former. Also Docker.
+DateLife can also be used locally through its R package. The stable version is available
+for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
+using the command `install.packages(pkgs = "datelife")` from R. Development versions
+are available from GitHub repository (<https://github.com/phylotastic/datelife>)
+and can be installed using the devtools R package command `install_github("phylotastic/datelife")`.
 
@@
 </span>
-We thank colleagues (students and postdocs) at the O'Meara Lab at the University of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons that led to initial work on this project.
-The Open Tree of Life project that provides the open, metadata rich repository of trees used for DateLife.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in OpenTree.
-The US National Science Foundation (NSF) for funding nearly all the above, in addition to the ABI grant that funded this project itself.
+We thank colleagues (students and postdocs) at the O'Meara Lab at the University
+of Tennesse Knoxville for suggestions, discussions and software testing.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
+that led to initial work on this project.
+The Open Tree of Life project that provides the open, metadata rich repository of
+trees used for DateLife.
+The many scientists who publish their chronograms in an open, reusable form, and
+the scientists who curate them for deposition in OpenTree.
+The US National Science Foundation (NSF) for funding nearly all the above, in addition
+to the ABI grant that funded this project itself.
 
@@
 </span>
+
+```{r pressure, out.width="100%", include=FALSE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
+knitr::include_graphics("fig1.pdf", dpi = 360)
+```

### 2019-01-20 Datelife_paper1.R

@@ -1,0 +1,9 @@
+# Domingo 14 de enero 2018
+library(citer)
+library(rmarkdown)
+setwd("~/Library/Mobile Documents/com~apple~CloudDocs/Desktop/datelife_paper1")
+pandoc_convert("Datelife_paper1.md", to="pdf", citeproc = TRUE, output = "Datelife_paper1.pdf")
+# pandoc: Unknown writer: pdf
+# To create a pdf with pandoc, use the latex or beamer writer and specify
+# an output file with .pdf extension (pandoc -t latex -o filename.pdf).
+# Error: pandoc document conversion failed with error 9

### 2019-01-21 Datelife_paper1.Rmd
finished description and updating refs

@@
 geometry: margin = 1in
-bibliography: library.bib
+bibliography: library_red.bib
 csl: systematic-biology.csl
@@
 fig_workflow_main <- "Fig. 1"
+fig_workflow_all_functions <- "Vignette Fig. 1"
 fig_runtime_main <- "Fig. 2"  
@@
 data curation and knowledge on the group of interest is still necessary. For example,
-<!--General issues with dating techniques:--> choosing correct fossils for calibration r
-equires a lot of expertise and knowledge on the group. An incorrect use of fossils
+<!--General issues with dating techniques:--> choosing correct fossils for calibration 
+requires a lot of expertise and knowledge on the group. An incorrect use of fossils
 can generate severe bias in dating results [@Sauquet2012c].
@@
 In here we present the first release of DateLife.
-It contains an improved databse of chronograms, more methods to summarize trees,
+It contains an improved database of chronograms, more methods to summarize trees,
 and new functions to visualize data.
@@
 ```{r, echo = FALSE}
-fig_workflow_main_foot <- "Figure 1. Summarized DateLife workflow. Only main actions are shown."
+fig_workflow_main_foot <- paste("Stylized DateLife workflow. Detailed steps are shown in",  fig_workflow_all_functions)
 ```
 <!--(BOLD and OToL species names, are they homogeneous/standardized?-->
+The basic `DateLife` workflow is shown in `r fig_workflow_main` and consists of:
 
-DateLife is a service for searching and processing information on ages of any number
-of taxa <!--or group of taxa-->of interest, across chronograms available in public
-data repositories coming from published peer reviewed studies. It can also generate
-new taxon age information<!--by using expert available phylogenies and scaling them
-to absolute time using information on ages obtained from published chronograms,--> by l
-inking several external services and tested algorithms.
-It takes advantage of the rotl, ape, and geiger packages to gather, process, and
-present information.
+1) A user providing as input at least two taxon names, either as tip labels on a tree, or as a simple comma separated character string.
+The tree can be in newick or phylo format, and can be with or without branch lengths.
+2) `DateLife` then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
+This format facilitates and greatly speeds up all further analyses and summary algorithms.
+3) The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
+4) At this point, users can choose to use all or some source data as calibration points to date a tree of their own making or choosing. <!--, a taxonomic tree-->
+5) Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
+6) Finally, users can easily view results graphically as well as construct their own graphs using inbuilt `DateLife` graphic generators.
 
-It only requires a set of taxon names as input, in the form of a comma separated
-listing or vector, or of a phylogeny with taxon names on the tips.
-Taxon names can correspond to binomial species names or clades
-<!--common names (will we implement it?)-->.
-When taxon names are clades, DateLife pulls all accepted species names within the
-clade (up to OToL's limit of _____ species) from OToL's reference taxonomy using
-a service of rphylotastic R package<!--citation-->.
-Names belonging to subspecies or any other infraspecific category are treated as
+<!--To perform the former tasks:-->
+`DateLife`'s chronogram database is currently built from Open Tree of Life (OToL)'s [@Hinchliff2015] tree repository.
+Among currently existing repositories (e.g., TreeBAse, Dryad), OToL's metadata rich tree store  is the only one meeting the requirements for proper/accurate automatized handling of trees.
+Input taxon names accepted by `DateLife` are binomial species names or clades.
+<!--common names (implemented yet?)--> Taxon searches are performed at the species level, so when input names correspond to higher clades, `DateLife` pulls all accepted species names within the
+clade from OToL's reference taxonomy to perform the search.
+Input names belonging to subspecies or any other infraspecific category are treated as
 species.
-DateLife can process input names with the taxon name resolution service (TNRS),
+DateLife also processes input names with the taxon name resolution service (TNRS),
 <!--which "corrects misspelled names and authorities, standardizes variant spellings,
 and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
-which corrects misspelled names or typos, and standardizes variation in spelling
-and synonyms [@Boyle2013], increasing the probability to correctly find the queried
-taxa in the chronogram database. DateLife uses TNRS to compare names against OToL's
-reference taxonomy using a service from the R package rotl [@Michonneau2016].
-<!--we'll implement the option to use other reference taxonomies in the future -->
+which corrects potentially misspelled names and typos, and standardizes variation in spelling
+and synonyms <!--to a taxonomic reference of users choosing, degfault to otol taxonomy -->[@Boyle2013], increasing the probability to correctly find the queried
+taxa in `DateLife`'s chronogram database.
 
-DateLife's main function searches taxon names across the chronogram database specified
-by the user. At the moment, it queries chronograms from OToL [@Hinchliff2015]
-<!-- it cannot query chronograms from TreeBASE -there is an earlier reference from
-1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel
-WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive
-tool for browsing the phylogeny of life. Am J Bot. 1994, 81 (6): 183--> repository.
-DateLife identifies chronograms having at least two taxon names, and subset them
-to contain only the taxa of interest. It then stores taxon age information from
-each chronogram individually as a patristic matrix, named with the citation of the
-original study. This format allows a rapid summary in a number of different ways,
-including:
-1) citations of the original studies containing the subset chronograms,
-2) a list of mrca ages of subset chronograms,
-3) a list of complete subset chronograms in newick or phylo format,
-4) a table containing all information retrieved in html or R's data frame format, or
-5) a single chronogram summarized from subset chronograms using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths.
-<!--(how to better explain the latter??
-Also, I just discovered the median gives non ultrametric trees...).-->
-
-DateLife also stores information on input taxon presence/absence across subset chronograms.
-Users can choose to add ages of missing taxa to subset chronograms in different
-ways, depending on the amount of knowledge they want to input or how much they want
-to be involved in the steps of the addition process.
-If users have no access to biological information (i.e., a character, DNA or protein
-matrix), missing taxa can be added to any chronogram simply at random, or by following
-taxonomic or phylogenetic knowledge from expert sources. There are a wide number
-of open reference taxonomies available, such as the Catalogue of Life [@Roskov2017]
-or the NCBI taxonomy database [@Federhen2012]. Expert phylogenies (with or without
-branch lengths) to be used as topological constraint (backbone) can also be obtained
-from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002]
-and Dryad (<https://www.datadryad.org//>).
-At the moment, DateLife only uses OToL's synthetic tree and reference taxonomy as
-expert knowledge to automatically add missing taxa to chronograms. Alternatively,
-users can input a reference taxonomy or topological constraint of their choosing
-or making. <!-- at some point we will implement the possibility for users to choose
-different reference taxonomies automatically-->
-If OToL's synthetic tree is not satisfactorily resolved for the taxa of interest,
-DateLife can construct a sequence data matrix from DNA markers available from the
-Barcode of Life Database (BOLD; @Ratnasingham2007), to attempt to further resolve
-polytomies. It will follow OToL's synthetic tree as backbone.
-<!--This molecular tree can then be dated with chronosMPL from the ape package, but is it really needed?? To estimate node ages chronosMPL uses mean path length method from Britton et al. 2002-->
-<!--It can rapidly construct trees with branch lengths from
-a set of lineages with sequence data from the Barcode Of Life Database (BOLD, <http://www.boldsystems.org/>), using the synthetic OToL (<https://tree.opentreeoflife.org>) as backbone, and to obtain divergence dates.
-It allows direct comparison of dates obtained with different markers
-available in BOLD (in plants and fungi in particular).-->
-To use information from a topological constraint, DateLife calls the congruification
-method described in [@Eastman2013] to find shared nodes between trees (congruent
-nodes). It then fixes their ages, and add ages to remaining nodes with a dating
-method that can be specified by the user. If users have access to biological data,
-they can input a tree with branch lengths proportional to relative substitution
-rates as topological constraint. In this case, age data from congruent nodes will
-be used as calibration points.
-Age data from several chronograms can be combined and congruified to be used as
-calibration points in a single analysis.
-
-Several dating methods are implemented in DateLife.
+Source chronogram summary tree can be assembled using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths and the hierarchical clustering method <!-- of ttol-->.
+<!--Also, I just discovered the median gives non ultrametric trees...
+but I thonk it is fixed now, check it!).-->
+Tree dating and simulation options are performed with various algorithms:
 Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
 nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
-DateLife implements BLADJ from the development R version of phylocom's R package
-[@Webb2008], phylocomr (<https://github.com/ropensci/phylocomr>). It can only be
-used when there is a topological constraint with no branch lengths.
-PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. It is
-also called through R. <!--It could be used with a reference taxonomy, if we provide a tree with politomies-->
+PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. 
 treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-[@Smith2012]. It is called through R.
-MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
+[@Smith2012].
+MrBayes [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
 random, following a reference taxonomy or a topological constraint. It draws ages
 from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
-DateLife calls MrBayes trough an R function.
+To apply calibrations to a tree, congruification method described in [@Eastman2013] to find shared nodes between trees (congruent nodes).
 
-DateLife can also correct negative branch lengths in several ways.
+To gather, process, and present information, `DateLife` builds up from functions available in several R packages including rotl [@Michonneau2016], ape[@Paradis2004], geiger[@Harmon2008], paleotree[@Bapst2012a], bold[@Chamberlain2018], phytools[@Revell2012], taxize[@Chamberlain2013; @Chamberlain2018], phyloch[@Heibl2008], phylocomr[@Ooms2018] and rphylotastic[@Omeara2019].
+
+Details on each step are further developed in `DateLife`'s R package documentation `datelife workflow` vignette at (<https://LINK>).
 
@@
 Improvements, short and long-term:
-* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations [@Schenk2016].
+* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
+[@Schenk2016].
 * bayesian congruification
@@
 Supplementary material, including code files and
-online-only appendices, can be found in the GitHub
-repository <!-- Dryad also? since it is all in github already-->
+online-only appendices and vignettes, can be found in the GitHub
+repository of the paper at (<https://github.com/LunaSare/datelife_paper1>) as well as in the package vignettes <!-- Dryad also? since it is all in github already-->
+and are also available from the Dryad Digital Repository at LINK.
 
@@
 
+```{r pressure, out.width="100%", include=TRUE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
+knitr::include_graphics("fig1.pdf", dpi = 360)
+```
+
 <!--**References**-->
@@
 </span>
-
-```{r pressure, out.width="100%", include=FALSE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
-knitr::include_graphics("fig1.pdf", dpi = 360)
-```

### 2019-02-21 Datelife_paper1.Rmd
some figures

@@
 # header-includes:
-#    - \setlength\parindent{24pt}  # this works too, but I'mg onna indent manually with 4 spaces
+#    - \setlength\parindent{24pt}  # this works too, but I'm gonna indent manually with 4 spaces
 # nope, 4 spaces indents are not recognized, they do something else
@@
 ```{r eval = TRUE, echo = FALSE}
-fig_workflow_main <- "Fig. 1"
-fig_workflow_all_functions <- "Vignette Fig. 1"
-fig_runtime_main <- "Fig. 2"  
-# make_datelife_query_plus_get_datelife_result
+source("Datelife_paper1_global.R")
 ```
@@
 <!--INTRODUCTION-->
+Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
+
+<!--INTRODUCTION: 1st paragraph: general statement that the field knows to be true and important-->
 Time of lineage divergence<!--alt lineages: named biological entities, taxon ages?-->
@@
 been great and here are some examples.
-<!--INTRODUCTION: 2nd paragraph-->
-When organisms are preserved in a fossil form, a time frame of taxon origin can
-be obtained directly from the age of rock strata.
-However, not all organisms fossilize well or at all.
-Fossilization success alone is highly circumstancial, and varies depending on a
-number of parameters including the nature of the habitat, population size, species
-range breadth and physical characteristics of the organism, all of which greatly
-vary across different organisms and through time.
-<!--fossil recovery is also highly circumstancial-->
-Thus, relying only on the fossil record to obtain a time frame of lineage divergence
-<!--taxon ages--> for all life is<!--, alas,--> not possible.
-
-Another caveat is that we usually cannot have a point age estimates of lineage origin
-because of the nature of the fossilization process: when a fossil first appears
-on the fossil record it is not necessarily because the lineage just originated,
-it might just be bc the conditions for fossilization started at that point but the
-organism might have been around for a long time before that. Because of this, we
-use ranges from the age of strata where fossils are found. But this does not solve
-the problem. The organism could even have been around previously, but not fossilize
-under the conditions prevailing around the time of formation of rock strata below
-its first appearance in the fossil record. In this sense, fossil ages can only be
-considered as minimum time of origin of lineages.
-<!--D Bapst work?-->
-
-Another source that has been widely used to inform about the timing of lineage origin
-is the relative rate of DNA or aminoacid substitution.
-It is estimated from hypothesis of character homology (alignments) for reconstructing
-phylogenetic relationships.
-<!--describe the two components of rel subst rate? bc they contain the information
-on absolute rate of substitution and absolute time.-->
-Molecular dating techniques use external data such as absolute time calibrations
-(e.g., fossils, geologic events) or absolute substitution rates <!--obtained in
-the lab?? birds-->to <!--dissociate rel subs rates components and--> generate dated
-phylogenies (chronograms) which contain information on absolute times of node divergence
-and taxon ages. <!--With molecular dating methods, an external calibration point
-is still necessary to estimate the absolute time of origin. Hence, the accumulation
-of data on fossils for calibration is also continual-->
-<!--This method has been particularly useful to obtain time frame of origin of organisms
-with poor or no fossil record, but is also largely used for organisms with rich
-fossil records-->
-<!--Models allow to take into account the nature of the fossil record: Given the
-characteristics of the fossilization process, the quality of dates inferred directly
-from it is not always considered the best.-->
-<!--INTRODUCTION: 3rd paragraph-->
+<!--INTRODUCTION: 2nd paragraph: important advances in the field-->
 In the past two decades, the possibility to obtain good quality DNA sequences
@@
 devtools::load_all(pkg = "~/Desktop/datelife")
-length(opentree_chronograms$trees)
-names(opentree_chronograms)
+length(opentree_chronograms$trees) names(opentree_chronograms)
 -->
@@
 accumulation in number and type of available data on taxon ages.
-<!--What led to datelife development?
-Describe public and research necessities covered by datelife
-It is important to use available data on time frame of lineage origin:
-To know the state of dating for a group of interest:
-What range of estimated ages exist already? Are fossil and molecular
-time frames coherent? (e.g., [@Magallon2015c]).
-To construct a time tree of life.
-For science communication, improve scientific discussions, time-framing
-other events of importance in other research areas.
-Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
--->
 
-<!--INTRODUCTION: 4th paragraph-->
+<!--INTRODUCTION: 3rd paragraph: problems associated-->
 The TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017] has aggregated chronograms
@@
 [@Hedges2015] was made available at OToL.
-
-With this inspiration, a prototype DateLife service was developed over a series
-of phylotastic hackathons [__CITE__] at the National Evolutionary Synthesis Center.
-NSF funding allowed for further development leading to this paper. A core goal of
-DateLife is openness of both the data sources and the code underlying the analyses,
-so scientific community can take advantage of this tool to leverage available information
-to the advancement of discoveries in biology.
-
-<!--
-[@Mora2011]
-eukaryotes and prokaryotes:
-on earth: 1.24 million catalogued, 8.74 million predicted
-on the oceans: 0.19 million catalogued, 2.21 predicted
-
-Total ssp number estimated to be between 3 and 100 million species, as mentiones in [@Mora2011]:
-ttol_spp <- 97085
-(ttol_spp * 100) / 3e+6  # 3.236167
-(ttol_spp * 100) / 100e+6  # 0.097085
--->
-<!--
-While its node-time search is very powerful, using available data to infer divergence
-time from lineages that have never been sequenced before, its timetree option is
-less powerful, their main flaw being that it does not provide source data.
-However, their primary database is not fully open, giving information on a per node
-basis. So, it is not possible to easily access information from original studies
-for a group of taxa (One would need to perform a pair of taxa search for all pairs
-from the group of interest).
-Divergence time estimates for all taxa in the TToL are precomputed and stored in
-the TimeTree database
-All searches allow species names and higher group names. Common names are allowed
-in all searches but a list of taxa.
-one-taxon search was not working for dog, though.
-1) ttol one lineage: it gives the evolutionary timeline of a lineage, showing the
-age of main divergence events occuring along the lineage evolutionary history, i.e.,
-when it shared a comon ancestor with all main lineages from which it diverged a
-long time ago.
-It only shows results graphically on the website.
-2) ttol parwise search: allows common names, species names and higher group names
-    how it handles names missing from tree: it looks in the NCBI taxonomy for the
-    closest relative of the species or higher taxa. Then uses the closest relative
-    to get the lineage pair age from ttol.
-3) ttol group of taxa
-4) ttol list of taxa: Cannot get it to work; does not work with cat, dog, turtle
-search, felidae, hominidae, phrynops search, some_ants species name search. What
-kind of files does it accepts: txt, rtf, doc, docx?
+<!--What led to datelife development?
+Describe public and research necessities covered by datelife
+It is important to use available data on time frame of lineage origin:
+To know the state of dating for a group of interest:
+What range of estimated ages exist already? Are fossil and molecular
+time frames coherent? (e.g., [@Magallon2015c]).
+To construct a time tree of life.
+For science communication, improve scientific discussions, time-framing
+other events of importance in other research areas.
+Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
 -->
 
-<!--
-retrieves information takes available data on lineage ages and phylogenetic relationships
-to construct a tree of all life scaled to absolute time: a chronogram of all life,
-while allowing the general public to search for age data within the tree for one
-taxon of interest (timeline option), a pair of taxa (node time option), a group
-of taxa (it can only be a single name of a taxonomic group), or a list of taxa of
-interest (timetree option for the two latter).
-Good things:
-Allows to get ages of species with no data, by extrapolating from closely related
-species with data.
-Allows comparison of timelines to geologic events, such as asteroid impacts, oxygen
-and carbon dioxide levels, solar luminosity.
-Not so good things:
-Only allows to download some data from original sources if performing a pair-of-taxa
-search.
-Only allows to see age ranges or confidence intervals in graphical form, when hovering
-over nodes. These CIs come from ttol. It does not have CIs from original studies.
-Node hovering is activated on one-taxon evolutionary history, and group-of-taxa
-searches (we could not get to work list-of-taxa search). Pair-of-taxa search shows
-the confidence interval of the only ttol node that constitutes the taxon-pair mrca.
-You can only download data from pair-of-taxa and group-of-taxa search. It generates
-a table with ages from original studies and a tree pruned from ttol, respectively.
-In both cases it gives single point age estimates. We cannot download CI information
-from ttol.
--->
-
+<!--INTRODUCTION: 4th paragraph-->
 Despite its great importance, analytical tools to summarize available information
@@
 Hence, automated processes for large data analysis are frequently received with skepticism.
-<!--INTRODUCTION: 5th paragraph-->
 
@@
 <!--Importance of datelife:-->
-DateLife can summarize this information in several formats that can be easily inspected
-by users.
-This allows rapidly obtaining a time frame of lineage divergence for a wide number
-of taxa.
-DateLife can also generate chronograms for taxa with little available information,
-by using the available data as calibration points.
+
+<!--INTRODUCTION: 5th paragraph: Goal, objectives and predictions-->
+Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
+inspired the development of a prototype `DateLife` service over a series
+of phylotastic hackathons [@Stoltzfus2013] at the National Evolutionary Synthesis Center.
+In this paper we present the first formal description of `DateLife`, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
 DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
@@
 (<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
-In here we present the first release of DateLife.
-It contains an improved database of chronograms, more methods to summarize trees,
-and new functions to visualize data.
-It also allows comparison of summary methods.
 
@@
 </span>
-```{r, echo = FALSE}
-fig_workflow_main_foot <- paste("Stylized DateLife workflow. Detailed steps are shown in",  fig_workflow_all_functions)
-```
+
 <!--(BOLD and OToL species names, are they homogeneous/standardized?-->
@@
 <!--To perform the former tasks:-->
-`DateLife`'s chronogram database is currently built from Open Tree of Life (OToL)'s [@Hinchliff2015] tree repository.
-Among currently existing repositories (e.g., TreeBAse, Dryad), OToL's metadata rich tree store  is the only one meeting the requirements for proper/accurate automatized handling of trees.
+`DateLife`'s chronogram database is currently built from Open Tree of Life (OToL)'s
+[@Hinchliff2015] tree repository. Among currently existing repositories (e.g., TreeBAse,
+Dryad), OToL's metadata rich tree store is the only one meeting the requirements
+for proper/accurate automatized handling of trees.
 Input taxon names accepted by `DateLife` are binomial species names or clades.
-<!--common names (implemented yet?)--> Taxon searches are performed at the species level, so when input names correspond to higher clades, `DateLife` pulls all accepted species names within the
+<!--common names (implemented yet?)-->
+Taxon searches are performed at the species level, so when input names correspond
+to higher clades,
+`DateLife` pulls all accepted species names within the
 clade from OToL's reference taxonomy to perform the search.
-Input names belonging to subspecies or any other infraspecific category are treated as
-species.
-DateLife also processes input names with the taxon name resolution service (TNRS),
+Currently, searches at the infrapsecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are treated
+as species.
+`DateLife` also processes input names with the taxon name resolution service (TNRS),
 <!--which "corrects misspelled names and authorities, standardizes variant spellings,
 and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
-which corrects potentially misspelled names and typos, and standardizes variation in spelling
-and synonyms <!--to a taxonomic reference of users choosing, degfault to otol taxonomy -->[@Boyle2013], increasing the probability to correctly find the queried
+which corrects potentially misspelled names and typos, and standardizes variation
+in spelling and synonyms <!--to a taxonomic reference of users choosing, default
+to otol taxonomy -->[@Boyle2013], increasing the probability to correctly find the
+queried
 taxa in `DateLife`'s chronogram database.
 
-Source chronogram summary tree can be assembled using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths and the hierarchical clustering method <!-- of ttol-->.
-<!--Also, I just discovered the median gives non ultrametric trees...
-but I thonk it is fixed now, check it!).-->
+Source chronogram summary tree can be assembled using the Super Distance Matrix
+(SDM) supertree construction approach [@Criscuolo2006] or using the median of branch
+lengths and the hierarchical clustering method<!-- of ttol-->. <!--Also, I discovered the median gives non ultrametric trees... but I thonk it is fixed now, check it!).-->
 Tree dating and simulation options are performed with various algorithms:
@@
 from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
-To apply calibrations to a tree, congruification method described in [@Eastman2013] to find shared nodes between trees (congruent nodes).
+To apply calibrations to a tree, th econgruification algorithm described in [@Eastman2013]
+is used to find shared nodes between trees (congruent nodes).
 
-To gather, process, and present information, `DateLife` builds up from functions available in several R packages including rotl [@Michonneau2016], ape[@Paradis2004], geiger[@Harmon2008], paleotree[@Bapst2012a], bold[@Chamberlain2018], phytools[@Revell2012], taxize[@Chamberlain2013; @Chamberlain2018], phyloch[@Heibl2008], phylocomr[@Ooms2018] and rphylotastic[@Omeara2019].
+To gather, process, and present information, `DateLife` builds up from functions
+available in several R packages including rotl [@Michonneau2016], ape [@Paradis2004],
+geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012],
+taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], phylocomr [@Ooms2018]
+and rphylotastic [@Omeara2019].
 
-Details on each step are further developed in `DateLife`'s R package documentation `datelife workflow` vignette at (<https://LINK>).
+Details on each step are further developed in `DateLife`'s R package documentation
+`datelife workflow` vignette at (<https://LINK>).
 
@@
 
-DateLife's code speed was tested on an Apple iMac
+`DateLife`'s code speed was tested on an Apple iMac
 with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
 We registered variation in computing time relative to number of input names
-<!--number of chronograms in database--> and DateLife service.
+<!--number of chronograms in database--> and `DateLife` service.
 Input processing increases roughly linearly with number of input taxon names, and
@@
 -->
-DateLife's code performance was evaluated with a set of unit tests designed and
+`DateLife`'s code performance was evaluated with a set of unit tests designed and
 implemented with the R package testthat [@RCoreTeam2018]. These tests were run both
@@
 GitHub– using the continuous integration tool Travis CI (<//travis-ci.org>). At
-present, unit tests cover around 30% (for now) of DateLife's code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
+present, unit tests cover around 30% (for now) of `DateLife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
 
@@
 online-only appendices and vignettes, can be found in the GitHub
-repository of the paper at (<https://github.com/LunaSare/datelife_paper1>) as well as in the package vignettes <!-- Dryad also? since it is all in github already-->
+repository of the paper at (<https://github.com/LunaSare/datelife_paper1>) as well
+as in the package vignettes <!-- Dryad also? since it is all in github already-->
 and are also available from the Dryad Digital Repository at LINK.
@@
 
-```{r pressure, out.width="100%", include=TRUE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
-knitr::include_graphics("fig1.pdf", dpi = 360)
-```
-
 <!--**References**-->

### 2019-02-21 Datelife_paper1_figures.Rmd
some figures

---
output: pdf_document
fontsize: 12pts
textalign: left  
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
linestretch: 2
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

```{r setup, echo = FALSE}
source("Datelife_paper1_global.R")
```

```{r pressure, out.width="100%", include=TRUE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
knitr::include_graphics("fig1.pdf", dpi = 560)
```

```{r eval = TRUE, results = 'asis', out.width="100%", include=TRUE, fig.align="center", fig.cap=fig_runtime_main_foot, echo=FALSE}
download.file(url = "https://github.com/phylotastic/datelife/raw/master/data-raw/benchmark/runtime_tests/2_tests/2_random_spp_names/mdq_plus_gdr_2018.05.29_fill.pdf", destfile = "fig_runtime_main.pdf", mode="wb")
knitr::include_graphics("fig_runtime_main.pdf", dpi = 560)

```

### 2019-02-21 Datelife_paper1_global.R

@@ -1,0 +1,9 @@
+fig_workflow_main <- "Fig. 1"
+fig_workflow_all_functions <- "Vignette Fig. 1"
+# make_datelife_query_plus_get_datelife_result
+fig_workflow_main_foot <- paste("Stylized DateLife workflow. Detailed steps are shown in",  fig_workflow_all_functions)
+
+fig_runtime_main <- "Fig. 2"
+fig_runtime_main_foot <- "Computation time of input processing and search across `DateLife`s chronogram database"
+fig_runtime_datagen <- "Fig. 3"
+fig_runtime_datasim <- "Fig. 4"

### 2019-05-01 Datelife_paper1.Rmd
discussion-a

@@
 
-1) A user providing as input at least two taxon names, either as tip labels on a tree, or as a simple comma separated character string.
+1) A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
 The tree can be in newick or phylo format, and can be with or without branch lengths.
 2) `DateLife` then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summary algorithms.
+This format facilitates and greatly speeds up all further analyses and summarization algorithms.
 3) The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
@@
 `DateLife`'s chronogram database is currently built from Open Tree of Life (OToL)'s
-[@Hinchliff2015] tree repository. Among currently existing repositories (e.g., TreeBAse,
+[@Hinchliff2015] tree repository. Among currently existing repositories (e.g., TreeBase,
 Dryad), OToL's metadata rich tree store is the only one meeting the requirements
@@
 Why TreeBase is not very useful for us? Be precise.
+Are these chronograms reliable to study evolutionary patterns, such as species diversification?
+DateLife can be seen as an open resource to know the current state of knowledge on lineage divergence times.
+Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
+If all, la facilidad para obtener hipotesis de tiempo de divergencia nos ayudará a evaluar la capacidad de los cronogramas para estudiar otros fenomenos evolutivos.
+Por ahora, no podemos aseverar que estos cronogramas puedan usarse para todo tipo de analisis.
 

### 2019-05-20 Datelife_paper1.Rmd
figure cap format

@@
 ---
-output: pdf_document
-fontsize: 12pts
+output: 
+  bookdown::pdf_document2:
+    keep_tex: yes
+    toc: false
+    includes:
+      # in_header: preamble.tex
+      after_body: [Datelife_paper1_figures_noyaml.tex]
+    # fig_caption: no
 textalign: left  # this does not work...
@@
 header-includes:
+    # - \usepackage{caption}
     - \usepackage[left]{lineno}
     - \linenumbers  
+  # - \documentclass[oupdraft]{sysbio} # this is duplicated in tex generated by knitr
+  # - \usepackage{xr}
+  # - \externaldocument{Datelife_paper1_figures}
+    # - \usepackage[utf8]{inputenc}
+    # - \usepackage[english]{babel}
+    # - \usepackage[document]{ragged2e}
 ---
-```{r eval = TRUE, echo = FALSE}
+
+```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE}
 source("Datelife_paper1_global.R")
@@
 <!--(BOLD and OToL species names, are they homogeneous/standardized?-->
-The basic `DateLife` workflow is shown in `r fig_workflow_main` and consists of:
+The basic `DateLife` workflow is shown in figure \ref{fig:workflow} and consists of:
 
@@
 (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch
-lengths and the hierarchical clustering method<!-- of ttol-->. <!--Also, I discovered the median gives non ultrametric trees... but I thonk it is fixed now, check it!).-->
+lengths and the hierarchical clustering method<!-- of ttol-->. <!--Also, I discovered the median gives non ultrametric trees... but I think it is fixed now, check it!).-->
 Tree dating and simulation options are performed with various algorithms:
@@
 Input processing increases roughly linearly with number of input taxon names, and
-increases considerably if tnrs service is activated (`r fig_runtime_main`).
+increases considerably if tnrs service is activated (Fig. \ref{fig:runtime1}).
 <!--
@@
 GitHub– using the continuous integration tool Travis CI (<//travis-ci.org>). At
-present, unit tests cover around 30% (for now) of `DateLife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
+present, unit tests cover more than 50% (for now) of `DateLife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
 
@@
 
+<!--Bird (or reptile) chronograms, too long time...
+finches is good
+nothofagus
+Look for all chronograms containing any birds
+Or, look for chronograms containing small and old lineages
+Determine which clade of birds has the more chronograms (have been dated
+more times) and use that as biological example. Fringillidae is a good candidate
 Find a clade with at least one chronogram containing all clade's species.
@@
 Use ltts to compare for now.
-Fig. X2 shows comparison of available chronograms for Felidae species and chronograms
-generated through DateLife
 
 think of a test to compare trees, topology- and date-wise
+-->
+In this section we demonstrate the types of outputs that can be obtained with `datelife`, using the bird family Fringillidae of true finches as example.
+We performed a higher-taxon search to obtain all data on lineage divergence available from `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp.
+according to the Open Tree of Life taxonomy).
+We found `r length(dr)` trees across `r length(unique(names(dr)))` studies (Fig.
+\ref{fig:schronograms}.
 
-<!--Bird (or reptile) chronograms, too long time...
-finches is good
-nothofagus
-Look for all chronograms containing any birds
-Or, look for chronograms containing small and old lineages
-Determine which clade of birds has the more chronograms (have been dated
-more times) and use that as biological example-->
+<!--Show code here, in the paper? Or show it only in the examples? Or only in the vignette? Write down what about datelife will be shown in each of these.
+Paper should show outputs from web site and package. 
+Vignette will show the R code used to generate the outputs displayed here and in the website.
+datelife examples will explain more on what is done in every step.
+-->
 
@@
 </span>
+
 We thank colleagues (students and postdocs) at the O'Meara Lab at the University
@@
 
-<!--**References**-->
-<span style="font-variant:small-caps;">
-References
-</span>
+\newpage
+
+\begin{center}
+\textsc{References}
+\end{center}

### 2019-05-20 Datelife_paper1_figures.Rmd
figure cap format

@@
 ---
-output: pdf_document
-fontsize: 12pts
+output: 
+  bookdown::pdf_document2:
+    keep_tex: yes
 textalign: left
@@
   - \linenumbers
+  - \usepackage{caption}
+  # - \usepackage{xr}
+  #- \externaldocument{name of external document to reference fugures from there in here}
 ---
 
-```{r setup, echo = FALSE}
+```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
 source("Datelife_paper1_global.R")
@@
 
-```{r pressure, out.width="100%", include=TRUE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
-knitr::include_graphics("fig1.pdf", dpi = 560)
-```
+<!-- ```{r myfigure, fig.cap = "My figure"} -->
+<!-- # fig.show = "hide" AND include = FALSE # does not generate the label for the figure -->
+<!-- # fig.cap = " " inactivates the label reference -->
+<!-- plot(1:10) -->
+<!-- ``` -->
 
-```{r eval = TRUE, results = 'asis', out.width="100%", include=TRUE, fig.align="center", fig.cap=fig_runtime_main_foot, echo=FALSE}
-download.file(url = "https://github.com/phylotastic/datelife/raw/master/data-raw/benchmark/runtime_tests/2_tests/2_random_spp_names/mdq_plus_gdr_2018.05.29_fill.pdf", destfile = "fig_runtime_main.pdf", mode="wb")
-knitr::include_graphics("fig_runtime_main.pdf", dpi = 560)
+<!-- My plot is called \@ref(fig:myfigure). -->
 
-```
+<!-- \begin{figure}[!ht] -->
+<!-- \includegraphics{blank.png} -->
+<!-- \caption{} -->
+<!-- \label{fig:workflow} -->
+<!-- \end{figure} -->
+<!-- Stylized DateLife workflow.
+This shows the general worflows and analyses that can be performed with DateLife, via the R package or through the website.
+Details on the functions involved on each workflow are shown in `datelife`'s R package vignette.
+-->
+
+<!-- \begin{figure}[!ht] -->
+<!-- \includegraphics{blank.png} -->
+<!-- \caption{} -->
+<!-- \label{fig:runtime1} -->
+<!-- \end{figure} -->
+<!-- Computation time of input processing and search across `datelife`s  -->
+<!--   chronogram database. -->
+
+<!-- \begin{figure}[!ht] -->
+<!-- \includegraphics{blank.png} -->
+<!-- \caption{} -->
+<!-- \label{fig:schronograms} -->
+<!-- \end{figure} -->
+<!-- Lineage through time (LTT) plots of Fringillidae (bird family of true finches) source chronograms.
+The latter are obtained by searching a chronogram database and subsetting chronograms conaning at least two species.
+Arrows indicate maximum age of each source chronogram and numbers indicate the reference of original studies 1: @barker2012going, 2: @barker2015new, 3: @burns2014phylogenetics, 4: @claramunt2015new, 5: @gibb2015new, 6: @Hedges2015, 7: @hooper2017chromosomal, 8: @Jetz2012, 9: @price2014niche.
+-->
+
+<!-- \begin{figure}[!ht] -->
+<!-- \includegraphics{blank.png} -->
+<!-- \caption{} -->
+<!-- \label{fig:cvbladj} -->
+<!-- \end{figure} -->
+<!-- LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarizing information from source chronograms found for the Fringillidae. Arrows indicate maximum age. -->
+
+<!-- \newpage -->
+
+Figure 1. 
+Stylized DateLife workflow.
+This shows the general worflows and analyses that can be performed with DateLife, via the R package or through the website.
+Details on the functions involved on each workflow are shown in `datelife`'s R package vignette.
+
+Figure 2. 
+Computation time of input processing and search across `datelife`s chronogram database.
+
+Figure 3. 
+Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
+Arrows indicate maximum age of each chronogram.
+Numbers reference to chronograms' original publications 1: @barker2012going, 2: @barker2015new, 3: @burns2014phylogenetics, 4: @claramunt2015new, 5: @gibb2015new, 6: @Hedges2015, 7: @hooper2017chromosomal, 8: @Jetz2012, 9: @price2014niche.
+
+Figure 4.
+LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarizing information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
+
+\newpage
+
+\begin{figure}[!h]
+\includegraphics{Fig1.pdf}
+\caption{}
+\label{fig:workflow}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\includegraphics{fig_runtime1.pdf}
+\caption{}
+\label{fig:runtime1}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\includegraphics{fig_schronograms1.pdf}
+\caption{}
+\label{fig:schronograms}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!ht]
+\includegraphics{fig_crossval_bladj.pdf}
+\caption{}
+\label{fig:cvbladj}
+\end{figure}

### 2019-05-20 Datelife_paper1.R

@@ -7,3 +7,5 @@ pandoc_convert("Datelife_paper1.md", to="pdf", citeproc = TRUE, output = "Dateli
 # To create a pdf with pandoc, use the latex or beamer writer and specify
 # an output file with .pdf extension (pandoc -t latex -o filename.pdf).
 # Error: pandoc document conversion failed with error 9
+
+rmarkdown::render(input = "~/Desktop/datelife_paper1/Datelife_paper1.Rmd", output_format = "pdf_document")

### 2019-05-20 Datelife_paper1_global.R

@@ -1,9 +1,21 @@
-fig_workflow_main <- "Fig. 1"
-fig_workflow_all_functions <- "Vignette Fig. 1"
-# make_datelife_query_plus_get_datelife_result
-fig_workflow_main_foot <- paste("Stylized DateLife workflow. Detailed steps are shown in",  fig_workflow_all_functions)
-
-fig_runtime_main <- "Fig. 2"
-fig_runtime_main_foot <- "Computation time of input processing and search across `DateLife`s chronogram database"
-fig_runtime_datagen <- "Fig. 3"
-fig_runtime_datasim <- "Fig. 4"
+# fig_workflow_cap <- paste("Stylized DateLife workflow. This shows the general worflows 
+#                           and analyses that can be performed with DateLife, via the R 
+#                           package or through the website. Details on the functions involved 
+#                           on each workflow are shown in datelife's R package vignette.")
+download.file(url = "https://github.com/phylotastic/datelife/raw/b3532e70b679272093658731d249534e84e15e28/data-raw/benchmark/runtime_tests/2_tests/2_random_spp_names/mdq_plus_gdr_2018.05.29_fill.pdf", 
+              destfile = "fig_runtime1.pdf", mode="wb")
+# fig_runtime1_cap <- "Computation time of input processing and search across `DateLife`s 
+#   chronogram database"
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringilidae_LTTplot_phyloall.pdf",
+              destfile = "fig_schronograms1.pdf", mode="wb")
+# fig_schronograms_cap <- "Lineage through time (LTT) plots of source chronograms containing 
+#   all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers 
+#   indicate chrinograms' original publication: 1) [@barker2012going], 2) [@barker2015new], 3) [@burns2014phylogenetics],
+#   4) [@claramunt2015new], 5) [@gibb2015new], 6) [@Hedges2015], 7) [@hooper2017chromosomal],
+#   8) [@Jetz2012], 9) [@price2014niche]"
+  
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringilidae_LTTplot_crossval_bladj.pdf",
+              destfile = "fig_crossval_bladj.pdf", mode="wb")
+# fig_crossval_bladj
+source("frin_plan.R")
+loadd() # loads all the targets listed in the workflow plan of the previous `make()`

### 2019-05-21 Datelife_paper1.Rmd
figure finally well formatted

@@
 
-<!--DESCRIPTION of Datelife-->
-<span style="font-variant:small-caps;">
-Description
-</span>
-
+\begin{center}
+\textsc{Description}
+\end{center}
 <!--(BOLD and OToL species names, are they homogeneous/standardized?-->
@@
 
+\begin{center}
+\textsc{Benchmark}
+\end{center}
 <!--**Benchmark: Testing DateLife computing performance**-->
-<span style="font-variant:small-caps;">
-Benchmark
-</span>
 <!-- Good to test 3 features on a package: performance, speed and scalability-->
@@
 
-<!--
-Speed of web interface and of r package (in computers with different capacities?)
--->
+<!--Speed of web interface and of r package (in computers with different capacities?)-->
 `DateLife`'s code performance was evaluated with a set of unit tests designed and
@@
 
+\begin{center}
+\textsc{Example}
+\end{center}
 <!--**Biological example: Testing DateLife accuracy**-->
-<span style="font-variant:small-caps;">
-Biological example
-</span>
-
 <!--Bird (or reptile) chronograms, too long time...
@@
 
-<!--**Discussion/Conclusions** there is not a discussion section in this type of sys Bio contributions-->
-<span style="font-variant:small-caps;">
-Conclusions
-</span>
+\begin{center}
+\textsc{Conclusions}
+\end{center}
 
@@
 
-<!--**Availability**-->
-<span style="font-variant:small-caps;">
-Availability
-</span>
+\begin{center}
+\textsc{Availability}
+\end{center}
 
-DateLife is free and open source and it can be used through its current website
-<http://www.datelife.org/query/>, or through Phylotastic!'s web portal <http://phylo.cs.nmsu.edu:3000/>.
-RStudio's Shiny Server and the shiny package open infrastructure are used to maintain
-the former. Also Docker.
-DateLife can also be used locally through its R package. The stable version is available
+`Datelife` is free and open source and it can be used through its current website
+<http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
+`Datelife`'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+`DateLife`'s R package stable version is available
 for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
-using the command `install.packages(pkgs = "datelife")` from R. Development versions
+using the command `install.packages(pkgs = "datelife")` from within R. Development versions
 are available from GitHub repository (<https://github.com/phylotastic/datelife>)
-and can be installed using the devtools R package command `install_github("phylotastic/datelife")`.
+and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
 
-<!--**Supplementary Material**-->
-<span style="font-variant:small-caps;">
-Supplementary Material
-</span>
+\begin{center}
+\textsc{Supplementary Material}
+\end{center}
 
-Supplementary material, including code files and
-online-only appendices and vignettes, can be found in the GitHub
-repository of the paper at (<https://github.com/LunaSare/datelife_paper1>) as well
-as in the package vignettes <!-- Dryad also? since it is all in github already-->
-and are also available from the Dryad Digital Repository at LINK.
+Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in the GitHub repositories <https://github.com/LunaSare/datelife_paper1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>, respectively.
+<!-- Vignettes
+Dryad also?? since it is all in github already, maybe not necessary-->
 
-<!--**Funding**-->
-<span style="font-variant:small-caps;">
-Funding
-</span>
+\begin{center}
+\textsc{Funding}
+\end{center}
 
@@
 
-<!--**Acknowledgements**-->
-<span style="font-variant:small-caps;">
-Acknowledgements
-</span>
+\begin{center}
+\textsc{Acknowledgements}
+\end{center}
 

### 2019-05-21 Datelife_paper1.tex
figure finally well formatted

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{2}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin = 1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage[left]{lineno}
\linenumbers

\begin{document}

Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE

Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life

Authors: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}

Correspondence address:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA}
\end{enumerate}

Corresponding authors: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}

\newpage

\textbf{abstract.-} Here goes the abstract.

\textbf{Keywords:} Tree; Phylogeny; Scaling; Open; Ages; Congruify; Supertree;

\newpage

Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.

Time of lineage divergence
constitutes in many ways the fundamental/main knowledge necessary for evolutionary
understanding.
Coupled to species number and distribution, it is the main information necessary
for the study of diversification processes (i.e., the tempo and mode of speciation
and extinction),
central for the understanding of how biodiversity
patterns are shaped across space, time and clades (Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
Evolutionary understanding also relies on comparative studies, for which knowing
the time context for all life is crucial. Efforts to have a whole tree of life have
been great and here are some examples.
In the past two decades, the possibility to obtain good quality DNA sequences
coupled to methodological developments in phylogenetic and dating inference, allowed
the application of molecular dating methods on a very large amount and diversity
of organisms, greatly increasing the quantity of data on taxon
ages across the tree of life.
To date, there is a large amount of both fossil and molecular-based data on taxon
ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
and Open Tree of Life (OToL).
OToL alone holds more than 200 chronograms.
Methods to include living and fossil lineages are in continued development and increased
usage by the community,
which coupled to better sharing data practices, are greatly contributing to the
accumulation in number and type of available data on taxon ages.

The TimeTree project (Hedges et al. \protect\hyperlink{ref-Hedges2006}{2006}, \protect\hyperlink{ref-Hedges2015}{2015}; Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) has aggregated chronograms
from 3,163 studies, encompassing 97,085 species (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}), and continues to grow.
However, even in this gold standard resource, the included taxa only encompass between
0.097 and 3.236\% of total species diversity (following taxonomic expert opinion
on the global, extant species numbers, which ranges from 3 to 100 million species
{[}Mayr2010; Moran2011{]}). One advantage of TimeTree is that it includes taxa from
across the tree of life, versus more specialized chronograms focusing on plants
{[}PHYLOMATIC{]}, birds {[}JETZ ET AL BIRDTREE.ORG{]}, and other groups. Users can choose
between a web interface or a mobile app to receive information on divergence times
for the evolutionary history of a lineage, pairs of taxa, all lineages within a
taxon, or a list of taxa. As a science communication tool, TimeTree project is
very powerful: it has a friendly graphical interface, with informative and colorful
outputs, that allows the general public to satisfy curiosity regarding a particular
organism of interest or group of them. It is of limited utility for scientific
studies, however. The thousands of trees that have been entered are unavailable
for examination or reuse; according to the creators (see TimeTree web FAQ), methods
for allowing data downloading have been under discussion for the past several years
yet the primary data remain closed. Moreover, there is no Application Programming
Interface (API) allowing programmatic access to any data, greatly impairing the
possibility of large-scale, automated data-mining, which is not allowed under TimeTree
website's terms of use. The nearly hundred thousand taxon summary chronogram generated
from TimeTree resources is not available with its publication (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) or the
TimeTree website, though the still substantial chronogram from a previous publication
(Hedges et al. \protect\hyperlink{ref-Hedges2015}{2015}) was made available at OToL.

Despite its great importance, analytical tools to summarize available information
on taxon ages for the scientific community are still lacking.
We identified several aspects that might have so far delayed the exploitation of
existing data. First, original chronograms available publicly are scattered across
various repositories (otol tree store, dryad, treebase, journals supplementary data)
usually with different formats too.
Second, lineage names due to taxonomic idiosincracy can be different among studies
and manual curation of that is usually necessary.
Third, data curation
Recent advances on this area (e.g., supersmart) aim to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of
data curation and knowledge on the group of interest is still necessary. For example,
choosing correct fossils for calibration
requires a lot of expertise and knowledge on the group. An incorrect use of fossils
can generate severe bias in dating results (Sauquet et al. \protect\hyperlink{ref-Sauquet2012c}{2012}).
Hence, data curation is still an important part of any biological study. The research
community considers it as an important or even crucial step before data analysis.
Hence, automated processes for large data analysis are frequently received with skepticism.

DateLife palliates this by only using information available from already published
studies, which are ideally constructed using robust information, such as sequence
data and thoughtfully curated fossil calibrations.

Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
inspired the development of a prototype \texttt{DateLife} service over a series
of phylotastic hackathons (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}) at the National Evolutionary Synthesis Center.
In this paper we present the first formal description of \texttt{DateLife}, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
(Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013})
It can be used through an R package , a web interface
(\url{http://www.datelife.org/query/}) and an API.

\begin{center}
\textsc{Description}
\end{center}

The basic \texttt{DateLife} workflow is shown in figure \ref{fig:workflow} and consists of:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string. The tree can be in newick or phylo format, and can be with or without branch lengths.
\item
  \texttt{DateLife} then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study. This format facilitates and greatly speeds up all further analyses and summarization algorithms.
\item
  The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
\item
  At this point, users can choose to use all or some source data as calibration points to date a tree of their own making or choosing. 
\item
  Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
\item
  Finally, users can easily view results graphically as well as construct their own graphs using inbuilt \texttt{DateLife} graphic generators.
\end{enumerate}

\texttt{DateLife}'s chronogram database is currently built from Open Tree of Life (OToL)'s
(Hinchliff et al. \protect\hyperlink{ref-Hinchliff2015}{2015}) tree repository. Among currently existing repositories (e.g., TreeBase,
Dryad), OToL's metadata rich tree store is the only one meeting the requirements
for proper/accurate automatized handling of trees.
Input taxon names accepted by \texttt{DateLife} are binomial species names or clades.
Taxon searches are performed at the species level, so when input names correspond
to higher clades,
\texttt{DateLife} pulls all accepted species names within the
clade from OToL's reference taxonomy to perform the search.
Currently, searches at the infrapsecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are treated
as species.
\texttt{DateLife} also processes input names with the taxon name resolution service (TNRS),
which corrects potentially misspelled names and typos, and standardizes variation
in spelling and synonyms (Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}), increasing the probability to correctly find the
queried
taxa in \texttt{DateLife}'s chronogram database.

Source chronogram summary tree can be assembled using the Super Distance Matrix
(SDM) supertree construction approach (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}) or using the median of branch
lengths and the hierarchical clustering method.
Tree dating and simulation options are performed with various algorithms:
Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
nodes evenly, which minimizes age variance in the chronogram (Webb et al. \protect\hyperlink{ref-Webb2008}{2008}).
PATHd8 is a non-clock, rate-smoothing method (Britton et al. \protect\hyperlink{ref-Britton2007}{2007}) to date trees.
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
(Smith and O'Meara \protect\hyperlink{ref-Smith2012}{2012}).
MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) can be used when adding taxa at
random, following a reference taxonomy or a topological constraint. It draws ages
from a pure birth model, as implemented by Jetz and collaborators (\protect\hyperlink{ref-Jetz2012}{2012}).
To apply calibrations to a tree, th econgruification algorithm described in (Eastman et al. \protect\hyperlink{ref-Eastman2013}{2013})
is used to find shared nodes between trees (congruent nodes).

To gather, process, and present information, \texttt{DateLife} builds up from functions
available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
geiger (Harmon et al. \protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}),
taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018})
and rphylotastic (O'Meara et al. \protect\hyperlink{ref-Omeara2019}{2019}).

Details on each step are further developed in \texttt{DateLife}'s R package documentation
\texttt{datelife\ workflow} vignette at (\url{https://LINK}).

\begin{center}
\textsc{Benchmark}
\end{center}

\texttt{DateLife}'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor.
We registered variation in computing time relative to number of input names
and \texttt{DateLife} service.
Input processing increases roughly linearly with number of input taxon names, and
increases considerably if tnrs service is activated (Fig. \ref{fig:runtime1}).
Results show that searching time increases linearly with number of input names and
number of chronograms in database.

Summarizing DateLife results processing times

Adding dates processing time

get\_bold\_otol\_tree running time

\texttt{DateLife}'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}). These tests were run both
locally --using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018})-- and on a public server --via
GitHub-- using the continuous integration tool Travis CI (\textless{}//travis-ci.org\textgreater{}). At
present, unit tests cover more than 50\% (for now) of \texttt{DateLife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).

\begin{center}
\textsc{Example}
\end{center}

In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using the bird family Fringillidae of true finches as example. We performed a higher-taxon search to obtain all data on lineage divergence available from \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp. according to the Open Tree of Life taxonomy). We found 13 trees across 9 studies (Fig. \ref{fig:schronograms}.

\begin{center}
\textsc{Conclusions}
\end{center}

Taxon ages are key to many areas of evolutionary studies: trait evolution, species
diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
especially for those who want to use phylogenies but who are not systematists, or
do not have the time to develop the necessary knowledge and data curation skills
to produce new chronograms. Knowledge on taxon ages is also important for non-biological
studies and the non-academic community.
The combination of new analytical techniques, availability of more fossil and molecular
data, and better practices in data sharing has resulted in a steady accumulation
of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree
of Life, for a large quantity and diversity of organisms. However, this information
remains difficult to synthesize for many biologists and the non-academic community.

Here, we have shown that DateLife allows
an easy and fast obtention of all publicly available information on taxon ages,
which can be used to generate new data.
This information can be used to account for the effect of phylogenetic signal in
studies of trait evolution; to explore potential speciation and extinction dynamics
of interest within a clade; to obtain a time frame of biogeographical events; for
science communication and outreach, amongst others.
Compared to similar platforms such as time tree of life
and supermart, it offers several advantages.
It is fast;
source data is completely open;
it requires no expert biological knowledge from users for any of its functionalities;
it allows exploration of alternative taxonomic and phylogenetic schemes;
it allows rapid exploration of the effect of alternative divergence time hypothesis;
it allows rapid synthesis in a number of different formats;
it facilitates reproducibility of analyses;

Improvements, short and long-term:
* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
(Schenk \protect\hyperlink{ref-Schenk2016}{2016}).
* bayesian congruification
* topological congruification

Problems and caveats:
Not many databases, only OToL
Why TreeBase is not very useful for us? Be precise.
Are these chronograms reliable to study evolutionary patterns, such as species diversification?
DateLife can be seen as an open resource to know the current state of knowledge on lineage divergence times.
Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
If all, la facilidad para obtener hipotesis de tiempo de divergencia nos ayudará a evaluar la capacidad de los cronogramas para estudiar otros fenomenos evolutivos.
Por ahora, no podemos aseverar que estos cronogramas puedan usarse para todo tipo de analisis.

\begin{center}
\textsc{Availability}
\end{center}

\texttt{Datelife} is free and open source and it can be used through its current website
\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
\texttt{Datelife}'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
\texttt{DateLife}'s R package stable version is available
for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
are available from GitHub repository (\url{https://github.com/phylotastic/datelife})
and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.

\begin{center}
\textsc{Supplementary Material}
\end{center}

Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in the GitHub repositories \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.

\begin{center}
\textsc{Funding}
\end{center}

Funding was provided by NSF grant 1458603

NESCent

Open Tree of Life

University of Tennessee, Knoxville

\begin{center}
\textsc{Acknowledgements}
\end{center}

We thank colleagues (students and postdocs) at the O'Meara Lab at the University
of Tennesse Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used for DateLife.
The many scientists who publish their chronograms in an open, reusable form, and
the scientists who curate them for deposition in OpenTree.
The US National Science Foundation (NSF) for funding nearly all the above, in addition
to the ABI grant that funded this project itself.

\newpage

\begin{center}
\textsc{References}
\end{center}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Bapst2012a}{}%
Bapst D.W. 2012. Paleotree: An R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution. 3:803--807.

\leavevmode\hypertarget{ref-Boyle2013}{}%
Boyle B., Hopkins N., Lu Z., Raygoza Garay J.A., Mozzherin D., Rees T., Matasci N., Narro M.L., Piel W.H., Mckay S.J., Lowry S., Freeland C., Peet R.K., Enquist B.J. 2013. The taxonomic name resolution service: An online tool for automated standardization of plant names. BMC Bioinformatics. 14.

\leavevmode\hypertarget{ref-Britton2007}{}%
Britton T., Anderson C.L., Jacquet D., Lundqvist S., Bremer K. 2007. Estimating Divergence Times in Large Phylogenetic Trees. Systematic Biology. 56:741--752.

\leavevmode\hypertarget{ref-Chamberlain2018}{}%
Chamberlain S. 2018. bold: Interface to Bold Systems API..

\leavevmode\hypertarget{ref-Chamberlain2013}{}%
Chamberlain S.A., Szöcs E. 2013. taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}. F1000Research. 2:1--29.

\leavevmode\hypertarget{ref-Criscuolo2006}{}%
Criscuolo A., Berry V., Douzery E.J., Gascuel O. 2006. SDM: A fast distance-based approach for (super)tree building in phylogenomics. Systematic Biology. 55:740--755.

\leavevmode\hypertarget{ref-Eastman2013}{}%
Eastman J.M., Harmon L.J., Tank D.C. 2013. Congruification: Support for time scaling large phylogenetic trees. Methods in Ecology and Evolution. 4:688--691.

\leavevmode\hypertarget{ref-Harmon2008}{}%
Harmon L., Weir J., Brock C., Glor R., Challenger W. 2008. GEIGER: investigating evolutionary radiations. Bioinformatics. 24:129--131.

\leavevmode\hypertarget{ref-Hedges2006}{}%
Hedges S.B., Dudley J., Kumar S. 2006. TimeTree: A public knowledge-base of divergence times among organisms. Bioinformatics. 22:2971--2972.

\leavevmode\hypertarget{ref-Hedges2015}{}%
Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.

\leavevmode\hypertarget{ref-Heibl2008}{}%
Heibl C. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..

\leavevmode\hypertarget{ref-Hinchliff2015}{}%
Hinchliff C.E., Smith S.A., Allman J.F., Burleigh J.G., Chaudhary R., Coghill L.M., Crandall K.A., Deng J., Drew B.T., Gazis R., Gude K., Hibbett D.S., Katz L.A., Laughinghouse H.D., McTavish E.J., Midford P.E., Owen C.L., Ree R.H., Rees J.A., Soltis D.E., Williams T., Cranston K.A. 2015. Synthesis of phylogeny and taxonomy into a comprehensive tree of life. Proceedings of the National Academy of Sciences. 112:12764--12769.

\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
Huelsenbeck J.P., Ronquist F. 2001. MRBAYES: Bayesian inference of phylogenetic trees. Bioinformatics. 17:754--755.

\leavevmode\hypertarget{ref-Jetz2012}{}%
Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.

\leavevmode\hypertarget{ref-Kumar2017}{}%
Kumar S., Stecher G., Suleski M., Hedges S.B. 2017. TimeTree: A Resource for Timelines, Timetrees, and Divergence Times. Molecular biology and evolution. 34:1812--1819.

\leavevmode\hypertarget{ref-Michonneau2016}{}%
Michonneau F., Brown J.W., Winter D.J. 2016. rotl: an R package to interact with the Open Tree of Life data. Methods in Ecology and Evolution. 7:1476--1481.

\leavevmode\hypertarget{ref-Morlon2014}{}%
Morlon H. 2014. Phylogenetic approaches for studying diversification. Ecology Letters. 17:508--525.

\leavevmode\hypertarget{ref-Omeara2019}{}%
O'Meara B., Md Tayeen A.S., Sanchez Reyes L.L. 2019. Rphylotastic: An r interface to 'phylotastic' web services..

\leavevmode\hypertarget{ref-Ooms2018}{}%
Ooms J., Chamberlain S. 2018. Phylocomr: Interface to 'phylocom'..

\leavevmode\hypertarget{ref-Paradis2004}{}%
Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.

\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
R Core Team. 2018. R: a language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing.

\leavevmode\hypertarget{ref-Revell2012}{}%
Revell L.J. 2012. Phytools: An r package for phylogenetic comparative biology (and other things). Methods in Ecology and Evolution. 3:217--223.

\leavevmode\hypertarget{ref-Ronquist2003}{}%
Ronquist F., Huelsenbeck J.P. 2003. MrBayes 3: Bayesian phylogenetic inference under mixed models. Bioinformatics. 19:1572--1574.

\leavevmode\hypertarget{ref-Sauquet2012c}{}%
Sauquet H., Ho S.Y.W., Gandolfo M. a, Jordan G.J., Wilf P., Cantrill D.J., Bayly M.J., Bromham L., Brown G.K., Carpenter R.J., Lee D.M., Murphy D.J., Sniderman J.M.K., Udovicic F. 2012. Testing the impact of calibration on molecular divergence times using a fossil-rich group: the case of Nothofagus (Fagales). Systematic Biology. 61:289--313.

\leavevmode\hypertarget{ref-Schenk2016}{}%
Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.

\leavevmode\hypertarget{ref-Smith2012}{}%
Smith S.A., O'Meara B.C. 2012. TreePL: Divergence time estimation using penalized likelihood for large phylogenies. Bioinformatics. 28:2689--2690.

\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
Stoltzfus A., Lapp H., Matasci N., Deus H., Sidlauskas B., Zmasek C.M., Vaidya G., Pontelli E., Cranston K., Vos R., Webb C.O., Harmon L.J., Pirrung M., O'Meara B., Pennell M.W., Mirarab S., Rosenberg M.S., Balhoff J.P., Bik H.M., Heath T.A., Midford P.E., Brown J.W., McTavish E.J., Sukumaran J., Westneat M., Alfaro M.E., Steele A., Jordan G. 2013. Phylotastic! Making tree-of-life knowledge accessible, reusable and convenient. BMC Bioinformatics. 14.

\leavevmode\hypertarget{ref-Webb2008}{}%
Webb C.O., Ackerly D.D., Kembel S.W. 2008. Phylocom: Software for the analysis of phylogenetic community structure and trait evolution. Bioinformatics. 24:2098--2100.

\leavevmode\hypertarget{ref-barker2012going}{}%
Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J. 2012. Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds. Systematic biology. 62:298--320.

\leavevmode\hypertarget{ref-barker2015new}{}%
Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J. 2015. New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk: Ornithological Advances. 132:333--348.

\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
Burns K.J., Shultz A.J., Title P.O., Mason N.A., Barker F.K., Klicka J., Lanyon S.M., Lovette I.J. 2014. Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds. Molecular Phylogenetics and Evolution. 75:41--77.

\leavevmode\hypertarget{ref-claramunt2015new}{}%
Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.

\leavevmode\hypertarget{ref-gibb2015new}{}%
Gibb G.C., England R., Hartig G., McLenachan P.A., Taylor Smith B.L., McComish B.J., Cooper A., Penny D. 2015. New zealand passerines help clarify the diversification of major songbird lineages during the oligocene. Genome biology and evolution. 7:2983--2995.

\leavevmode\hypertarget{ref-Hedges2015}{}%
Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.

\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.

\leavevmode\hypertarget{ref-Jetz2012}{}%
Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.

\leavevmode\hypertarget{ref-price2014niche}{}%
Price T.D., Hooper D.M., Buchanan C.D., Johansson U.S., Tietze D.T., Alström P., Olsson U., Ghosh-Harihar M., Ishtiaq F., Gupta S.K., others. 2014. Niche filling slows the diversification of himalayan songbirds. Nature. 509:222.

\newpage

\begin{center}
\textsc{Figure 1}
\end{center}
Stylized DateLife workflow. This shows the general worflows and analyses that can be performed with DateLife, via the R package or through the website. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.

\begin{center}
\textsc{Figure 2}
\end{center}
Computation time of input processing and search across \texttt{datelife}'s chronogram database.

\begin{center}
\textsc{Figure 3}
\end{center}
Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers reference to chronograms' original publications 1: Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al. (\protect\hyperlink{ref-price2014niche}{2014}).

\begin{center}
\textsc{Figure 4}
\end{center}
LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarizing information from source chronograms found for the Fringillidae. Arrows indicate maximum age.

\newpage

\begin{figure}[!h]
\includegraphics{Fig1.pdf}
\caption{}
\label{fig:workflow}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=1\linewidth]{fig_runtime1.pdf}
\caption{}
\label{fig:runtime1}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
\caption{}
\label{fig:schronograms}
\end{figure}

\newpage

\begin{figure}[!ht]
\includegraphics{fig_crossval_bladj.pdf}
\caption{}
\label{fig:cvbladj}
\end{figure}

\end{document}

### 2019-05-21 Datelife_paper1_figures.tex
figure finally well formatted

--- 2019-05-21 Datelife_paper1.tex
@@
 \linenumbers
+\usepackage{caption}
 
@@
 
-Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE
+{
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
+Figure 1.
+Stylized DateLife workflow.
+This shows the general worflows and analyses that can be performed with DateLife, via the R package or through the website.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
 
-Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
+Figure 2.
+Computation time of input processing and search across \texttt{datelife}s chronogram database.
 
-Authors: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}
+Figure 3.
+Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
+Arrows indicate maximum age of each chronogram.
+Numbers reference to chronograms' original publications 1: Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
+(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
+(\protect\hyperlink{ref-price2014niche}{2014}).
 
-Correspondence address:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA}
-\end{enumerate}
-
-Corresponding authors: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
+Figure 4.
+LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarizing information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
 
@@
 
-\textbf{abstract.-} Here goes the abstract.
-
-\textbf{Keywords:} Tree; Phylogeny; Scaling; Open; Ages; Congruify; Supertree;
+\begin{figure}[!h]
+\includegraphics{Fig1.pdf}
+\caption{}
+\label{fig:workflow}
+\end{figure}
 
@@
 
-Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
-
-Time of lineage divergence
-constitutes in many ways the fundamental/main knowledge necessary for evolutionary
-understanding.
-Coupled to species number and distribution, it is the main information necessary
-for the study of diversification processes (i.e., the tempo and mode of speciation
-and extinction),
-central for the understanding of how biodiversity
-patterns are shaped across space, time and clades (Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-Evolutionary understanding also relies on comparative studies, for which knowing
-the time context for all life is crucial. Efforts to have a whole tree of life have
-been great and here are some examples.
-In the past two decades, the possibility to obtain good quality DNA sequences
-coupled to methodological developments in phylogenetic and dating inference, allowed
-the application of molecular dating methods on a very large amount and diversity
-of organisms, greatly increasing the quantity of data on taxon
-ages across the tree of life.
-To date, there is a large amount of both fossil and molecular-based data on taxon
-ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
-and Open Tree of Life (OToL).
-OToL alone holds more than 200 chronograms.
-Methods to include living and fossil lineages are in continued development and increased
-usage by the community,
-which coupled to better sharing data practices, are greatly contributing to the
-accumulation in number and type of available data on taxon ages.
-
-The TimeTree project (Hedges et al. \protect\hyperlink{ref-Hedges2006}{2006}, \protect\hyperlink{ref-Hedges2015}{2015}; Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) has aggregated chronograms
-from 3,163 studies, encompassing 97,085 species (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}), and continues to grow.
-However, even in this gold standard resource, the included taxa only encompass between
-0.097 and 3.236\% of total species diversity (following taxonomic expert opinion
-on the global, extant species numbers, which ranges from 3 to 100 million species
-{[}Mayr2010; Moran2011{]}). One advantage of TimeTree is that it includes taxa from
-across the tree of life, versus more specialized chronograms focusing on plants
-{[}PHYLOMATIC{]}, birds {[}JETZ ET AL BIRDTREE.ORG{]}, and other groups. Users can choose
-between a web interface or a mobile app to receive information on divergence times
-for the evolutionary history of a lineage, pairs of taxa, all lineages within a
-taxon, or a list of taxa. As a science communication tool, TimeTree project is
-very powerful: it has a friendly graphical interface, with informative and colorful
-outputs, that allows the general public to satisfy curiosity regarding a particular
-organism of interest or group of them. It is of limited utility for scientific
-studies, however. The thousands of trees that have been entered are unavailable
-for examination or reuse; according to the creators (see TimeTree web FAQ), methods
-for allowing data downloading have been under discussion for the past several years
-yet the primary data remain closed. Moreover, there is no Application Programming
-Interface (API) allowing programmatic access to any data, greatly impairing the
-possibility of large-scale, automated data-mining, which is not allowed under TimeTree
-website's terms of use. The nearly hundred thousand taxon summary chronogram generated
-from TimeTree resources is not available with its publication (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) or the
-TimeTree website, though the still substantial chronogram from a previous publication
-(Hedges et al. \protect\hyperlink{ref-Hedges2015}{2015}) was made available at OToL.
-
-Despite its great importance, analytical tools to summarize available information
-on taxon ages for the scientific community are still lacking.
-We identified several aspects that might have so far delayed the exploitation of
-existing data. First, original chronograms available publicly are scattered across
-various repositories (otol tree store, dryad, treebase, journals supplementary data)
-usually with different formats too.
-Second, lineage names due to taxonomic idiosincracy can be different among studies
-and manual curation of that is usually necessary.
-Third, data curation
-Recent advances on this area (e.g., supersmart) aim to:
-Generate new dates using all available DNA sequence information;
-Perform one global analysis using all available information;
-Problems or downsides: This might be time consuming for large groups and a lot of
-data curation and knowledge on the group of interest is still necessary. For example,
-choosing correct fossils for calibration
-requires a lot of expertise and knowledge on the group. An incorrect use of fossils
-can generate severe bias in dating results (Sauquet et al. \protect\hyperlink{ref-Sauquet2012c}{2012}).
-Hence, data curation is still an important part of any biological study. The research
-community considers it as an important or even crucial step before data analysis.
-Hence, automated processes for large data analysis are frequently received with skepticism.
-
-DateLife palliates this by only using information available from already published
-studies, which are ideally constructed using robust information, such as sequence
-data and thoughtfully curated fossil calibrations.
-
-Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
-inspired the development of a prototype \texttt{DateLife} service over a series
-of phylotastic hackathons (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}) at the National Evolutionary Synthesis Center.
-In this paper we present the first formal description of \texttt{DateLife}, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
-DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
-(Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013})
-It can be used through an R package , a web interface
-(\url{http://www.datelife.org/query/}) and an API.
-
-\begin{center}
-\textsc{Description}
-\end{center}
-
-The basic \texttt{DateLife} workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
-\tightlist
-\item
-  A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item
-  \texttt{DateLife} then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarization algorithms.
-\item
-  The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
-\item
-  At this point, users can choose to use all or some source data as calibration points to date a tree of their own making or choosing. 
-\item
-  Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
-\item
-  Finally, users can easily view results graphically as well as construct their own graphs using inbuilt \texttt{DateLife} graphic generators.
-\end{enumerate}
-
-\texttt{DateLife}'s chronogram database is currently built from Open Tree of Life (OToL)'s
-(Hinchliff et al. \protect\hyperlink{ref-Hinchliff2015}{2015}) tree repository. Among currently existing repositories (e.g., TreeBase,
-Dryad), OToL's metadata rich tree store is the only one meeting the requirements
-for proper/accurate automatized handling of trees.
-Input taxon names accepted by \texttt{DateLife} are binomial species names or clades.
-Taxon searches are performed at the species level, so when input names correspond
-to higher clades,
-\texttt{DateLife} pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
-Currently, searches at the infrapsecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are treated
-as species.
-\texttt{DateLife} also processes input names with the taxon name resolution service (TNRS),
-which corrects potentially misspelled names and typos, and standardizes variation
-in spelling and synonyms (Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}), increasing the probability to correctly find the
-queried
-taxa in \texttt{DateLife}'s chronogram database.
-
-Source chronogram summary tree can be assembled using the Super Distance Matrix
-(SDM) supertree construction approach (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}) or using the median of branch
-lengths and the hierarchical clustering method.
-Tree dating and simulation options are performed with various algorithms:
-Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
-nodes evenly, which minimizes age variance in the chronogram (Webb et al. \protect\hyperlink{ref-Webb2008}{2008}).
-PATHd8 is a non-clock, rate-smoothing method (Britton et al. \protect\hyperlink{ref-Britton2007}{2007}) to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Smith and O'Meara \protect\hyperlink{ref-Smith2012}{2012}).
-MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) can be used when adding taxa at
-random, following a reference taxonomy or a topological constraint. It draws ages
-from a pure birth model, as implemented by Jetz and collaborators (\protect\hyperlink{ref-Jetz2012}{2012}).
-To apply calibrations to a tree, th econgruification algorithm described in (Eastman et al. \protect\hyperlink{ref-Eastman2013}{2013})
-is used to find shared nodes between trees (congruent nodes).
-
-To gather, process, and present information, \texttt{DateLife} builds up from functions
-available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
-geiger (Harmon et al.
-\protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}),
-taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018})
-and rphylotastic (O'Meara et al. \protect\hyperlink{ref-Omeara2019}{2019}).
-
-Details on each step are further developed in \texttt{DateLife}'s R package documentation
-\texttt{datelife\ workflow} vignette at (\url{https://LINK}).
-
-\begin{center}
-\textsc{Benchmark}
-\end{center}
-
-\texttt{DateLife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time relative to number of input names
-and \texttt{DateLife} service.
-Input processing increases roughly linearly with number of input taxon names, and
-increases considerably if tnrs service is activated (Fig. \ref{fig:runtime1}).
-Results show that searching time increases linearly with number of input names and
-number of chronograms in database.
-
-Summarizing DateLife results processing times
-
-Adding dates processing time
-
-get\_bold\_otol\_tree running time
-
-\texttt{DateLife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}). These tests were run both
-locally --using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018})-- and on a public server --via
-GitHub-- using the continuous integration tool Travis CI (\textless{}//travis-ci.org\textgreater{}). At
-present, unit tests cover more than 50\% (for now) of \texttt{DateLife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\begin{center}
-\textsc{Example}
-\end{center}
-
-In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using the bird family Fringillidae of true finches as example.
-We performed a higher-taxon search to obtain all data on lineage divergence available from \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
-according to the Open Tree of Life taxonomy).
-We found 13 trees across 9 studies (Fig.
-\ref{fig:schronograms}.
-
-\begin{center}
-\textsc{Conclusions}
-\end{center}
-
-Taxon ages are key to many areas of evolutionary studies: trait evolution, species
-diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to develop the necessary knowledge and data curation skills
-to produce new chronograms. Knowledge on taxon ages is also important for non-biological
-studies and the non-academic community.
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree
-of Life, for a large quantity and diversity of organisms. However, this information
-remains difficult to synthesize for many biologists and the non-academic community.
-
-Here, we have shown that DateLife allows
-an easy and fast obtention of all publicly available information on taxon ages,
-which can be used to generate new data.
-This information can be used to account for the effect of phylogenetic signal in
-studies of trait evolution; to explore potential speciation and extinction dynamics
-of interest within a clade; to obtain a time frame of biogeographical events; for
-science communication and outreach, amongst others.
-Compared to similar platforms such as time tree of life
-and supermart, it offers several advantages.
-It is fast;
-source data is completely open;
-it requires no expert biological knowledge from users for any of its functionalities;
-it allows exploration of alternative taxonomic and phylogenetic schemes;
-it allows rapid exploration of the effect of alternative divergence time hypothesis;
-it allows rapid synthesis in a number of different formats;
-it facilitates reproducibility of analyses;
-
-Improvements, short and long-term:
-* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
-(Schenk \protect\hyperlink{ref-Schenk2016}{2016}).
-* bayesian congruification
-* topological congruification
-
-Problems and caveats:
-Not many databases, only OToL
-Why TreeBase is not very useful for us? Be precise.
-Are these chronograms reliable to study evolutionary patterns, such as species diversification?
-DateLife can be seen as an open resource to know the current state of knowledge on lineage divergence times.
-Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
-If all, la facilidad para obtener hipotesis de tiempo de divergencia nos ayudará a evaluar la capacidad de los cronogramas para estudiar otros fenomenos evolutivos.
-Por ahora, no podemos aseverar que estos cronogramas puedan usarse para todo tipo de analisis.
-
-\begin{center}
-\textsc{Availability}
-\end{center}
-
-\texttt{Datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{Datelife}'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{DateLife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\begin{center}
-\textsc{Supplementary Material}
-\end{center}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in the GitHub repositories \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
-
-\begin{center}
-\textsc{Funding}
-\end{center}
-
-Funding was provided by NSF grant 1458603
-
-NESCent
-
-Open Tree of Life
-
-University of Tennessee, Knoxville
-
-\begin{center}
-\textsc{Acknowledgements}
-\end{center}
-
-We thank colleagues (students and postdocs) at the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for DateLife.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in OpenTree.
-The US National Science Foundation (NSF) for funding nearly all the above, in addition
-to the ABI grant that funded this project itself.
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{fig_runtime1.pdf}
+\caption{}
+\label{fig:runtime1}
+\end{figure}
 
@@
 
-\begin{center}
-\textsc{References}
-\end{center}
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
+\caption{}
+\label{fig:schronograms}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!ht]
+\includegraphics{fig_crossval_bladj.pdf}
+\caption{}
+\label{fig:cvbladj}
+\end{figure}
 
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst D.W. 2012. Paleotree: An R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution. 3:803--807.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle B., Hopkins N., Lu Z., Raygoza Garay J.A., Mozzherin D., Rees T., Matasci N., Narro M.L., Piel W.H., Mckay S.J., Lowry S., Freeland C., Peet R.K., Enquist B.J.
-2013.
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton T., Anderson C.L., Jacquet D., Lundqvist S., Bremer K. 2007. Estimating Divergence Times in Large Phylogenetic Trees. Systematic Biology. 56:741--752.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain S. 2018. bold: Interface to Bold Systems API..
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain S.A., Szöcs E. 2013. taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}. F1000Research. 2:1--29.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo A., Berry V., Douzery E.J., Gascuel O. 2006. SDM: A fast distance-based approach for (super)tree building in phylogenomics. Systematic Biology. 55:740--755.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman J.M., Harmon L.J., Tank D.C. 2013. Congruification: Support for time scaling large phylogenetic trees. Methods in Ecology and Evolution. 4:688--691.
-
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon L., Weir J., Brock C., Glor R., Challenger W. 2008. GEIGER: investigating evolutionary radiations. Bioinformatics. 24:129--131.
-
-\leavevmode\hypertarget{ref-Hedges2006}{}%
-Hedges S.B., Dudley J., Kumar S. 2006. TimeTree: A public knowledge-base of divergence times among organisms. Bioinformatics. 22:2971--2972.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl C. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..
-
-\leavevmode\hypertarget{ref-Hinchliff2015}{}%
-Hinchliff C.E., Smith S.A., Allman J.F., Burleigh J.G., Chaudhary R., Coghill L.M., Crandall K.A., Deng J., Drew B.T., Gazis R., Gude K., Hibbett D.S., Katz L.A., Laughinghouse H.D., McTavish E.J., Midford P.E., Owen C.L., Ree R.H., Rees J.A., Soltis D.E., Williams T., Cranston K.A.
-2015.
-Synthesis of phylogeny and taxonomy into a comprehensive tree of life.
-Proceedings of the National Academy of Sciences.
-112:12764--12769.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck J.P., Ronquist F. 2001. MRBAYES: Bayesian inference of phylogenetic trees. Bioinformatics. 17:754--755.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.
-
-\leavevmode\hypertarget{ref-Kumar2017}{}%
-Kumar S., Stecher G., Suleski M., Hedges S.B. 2017. TimeTree: A Resource for Timelines, Timetrees, and Divergence Times. Molecular biology and evolution. 34:1812--1819.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau F., Brown J.W., Winter D.J. 2016. rotl: an R package to interact with the Open Tree of Life data. Methods in Ecology and Evolution. 7:1476--1481.
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon H. 2014. Phylogenetic approaches for studying diversification. Ecology Letters. 17:508--525.
-
-\leavevmode\hypertarget{ref-Omeara2019}{}%
-O'Meara B., Md Tayeen A.S., Sanchez Reyes L.L. 2019. Rphylotastic: An r interface to 'phylotastic' web services..
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms J., Chamberlain S. 2018. Phylocomr: Interface to 'phylocom'..
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. 2018. R: a language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell L.J. 2012. Phytools: An r package for phylogenetic comparative biology (and other things). Methods in Ecology and Evolution. 3:217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist F., Huelsenbeck J.P. 2003. MrBayes 3: Bayesian phylogenetic inference under mixed models. Bioinformatics. 19:1572--1574.
-
-\leavevmode\hypertarget{ref-Sauquet2012c}{}%
-Sauquet H., Ho S.Y.W., Gandolfo M.
-a, Jordan G.J., Wilf P., Cantrill D.J., Bayly M.J., Bromham L., Brown G.K., Carpenter R.J., Lee D.M., Murphy D.J., Sniderman J.M.K., Udovicic F.
-2012.
-Testing the impact of calibration on molecular divergence times using a fossil-rich group: the case of Nothofagus (Fagales).
-Systematic Biology.
-61:289--313.
-
-\leavevmode\hypertarget{ref-Schenk2016}{}%
-Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
-
-\leavevmode\hypertarget{ref-Smith2012}{}%
-Smith S.A., O'Meara B.C. 2012. TreePL: Divergence time estimation using penalized likelihood for large phylogenies. Bioinformatics. 28:2689--2690.
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus A., Lapp H., Matasci N., Deus H., Sidlauskas B., Zmasek C.M., Vaidya G., Pontelli E., Cranston K., Vos R., Webb C.O., Harmon L.J., Pirrung M., O'Meara B., Pennell M.W., Mirarab S., Rosenberg M.S., Balhoff J.P., Bik H.M., Heath T.A., Midford P.E., Brown J.W., McTavish E.J., Sukumaran J., Westneat M., Alfaro M.E., Steele A., Jordan G.
-2013.
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb C.O., Ackerly D.D., Kembel S.W. 2008. Phylocom: Software for the analysis of phylogenetic community structure and trait evolution. Bioinformatics. 24:2098--2100.
-
 \leavevmode\hypertarget{ref-barker2012going}{}%
@@
 
-\newpage
-
-\begin{center}
-\textsc{Figure 1}
-\end{center}
-Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with DateLife, via the R package or through the website.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-
-\begin{center}
-\textsc{Figure 2}
-\end{center}
-Computation time of input processing and search across \texttt{datelife}'s chronogram database.
-
-\begin{center}
-\textsc{Figure 3}
-\end{center}
-Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
-
-\begin{center}
-\textsc{Figure 4}
-\end{center}
-LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarizing information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics{Fig1.pdf}
-\caption{}
-\label{fig:workflow}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_runtime1.pdf}
-\caption{}
-\label{fig:runtime1}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
-\caption{}
-\label{fig:schronograms}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!ht]
-\includegraphics{fig_crossval_bladj.pdf}
-\caption{}
-\label{fig:cvbladj}
-\end{figure}
-
 \end{document}

### 2019-05-21 Datelife_paper1_figures_noyaml.tex
figure finally well formatted

--- 2019-05-21 Datelife_paper1.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{setspace}
-\setstretch{2}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin = 1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{longtable,booktabs}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\providecommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{}
-    \pretitle{\vspace{\droptitle}}
-  \posttitle{}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
-\usepackage[left]{lineno}
-\linenumbers
-
-\begin{document}
-
-Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE
-
-Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
-
-Authors: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}
-
-Correspondence address:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA}
-\end{enumerate}
-
-Corresponding authors: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
-
-\newpage
-
-\textbf{abstract.-} Here goes the abstract.
-
-\textbf{Keywords:} Tree; Phylogeny; Scaling; Open; Ages; Congruify; Supertree;
-
-\newpage
-
-Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
-
-Time of lineage divergence
-constitutes in many ways the fundamental/main knowledge necessary for evolutionary
-understanding.
-Coupled to species number and distribution, it is the main information necessary
-for the study of diversification processes (i.e., the tempo and mode of speciation
-and extinction),
-central for the understanding of how biodiversity
-patterns are shaped across space, time and clades (Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-Evolutionary understanding also relies on comparative studies, for which knowing
-the time context for all life is crucial. Efforts to have a whole tree of life have
-been great and here are some examples.
-In the past two decades, the possibility to obtain good quality DNA sequences
-coupled to methodological developments in phylogenetic and dating inference, allowed
-the application of molecular dating methods on a very large amount and diversity
-of organisms, greatly increasing the quantity of data on taxon
-ages across the tree of life.
-To date, there is a large amount of both fossil and molecular-based data on taxon
-ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
-and Open Tree of Life (OToL).
-OToL alone holds more than 200 chronograms.
-Methods to include living and fossil lineages are in continued development and increased
-usage by the community,
-which coupled to better sharing data practices, are greatly contributing to the
-accumulation in number and type of available data on taxon ages.
-
-The TimeTree project (Hedges et al. \protect\hyperlink{ref-Hedges2006}{2006}, \protect\hyperlink{ref-Hedges2015}{2015}; Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) has aggregated chronograms
-from 3,163 studies, encompassing 97,085 species (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}), and continues to grow.
-However, even in this gold standard resource, the included taxa only encompass between
-0.097 and 3.236\% of total species diversity (following taxonomic expert opinion
-on the global, extant species numbers, which ranges from 3 to 100 million species
-{[}Mayr2010; Moran2011{]}). One advantage of TimeTree is that it includes taxa from
-across the tree of life, versus more specialized chronograms focusing on plants
-{[}PHYLOMATIC{]}, birds {[}JETZ ET AL BIRDTREE.ORG{]}, and other groups. Users can choose
-between a web interface or a mobile app to receive information on divergence times
-for the evolutionary history of a lineage, pairs of taxa, all lineages within a
-taxon, or a list of taxa. As a science communication tool, TimeTree project is
-very powerful: it has a friendly graphical interface, with informative and colorful
-outputs, that allows the general public to satisfy curiosity regarding a particular
-organism of interest or group of them. It is of limited utility for scientific
-studies, however. The thousands of trees that have been entered are unavailable
-for examination or reuse; according to the creators (see TimeTree web FAQ), methods
-for allowing data downloading have been under discussion for the past several years
-yet the primary data remain closed. Moreover, there is no Application Programming
-Interface (API) allowing programmatic access to any data, greatly impairing the
-possibility of large-scale, automated data-mining, which is not allowed under TimeTree
-website's terms of use. The nearly hundred thousand taxon summary chronogram generated
-from TimeTree resources is not available with its publication (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) or the
-TimeTree website, though the still substantial chronogram from a previous publication
-(Hedges et al. \protect\hyperlink{ref-Hedges2015}{2015}) was made available at OToL.
-
-Despite its great importance, analytical tools to summarize available information
-on taxon ages for the scientific community are still lacking.
-We identified several aspects that might have so far delayed the exploitation of
-existing data. First, original chronograms available publicly are scattered across
-various repositories (otol tree store, dryad, treebase, journals supplementary data)
-usually with different formats too.
-Second, lineage names due to taxonomic idiosincracy can be different among studies
-and manual curation of that is usually necessary.
-Third, data curation
-Recent advances on this area (e.g., supersmart) aim to:
-Generate new dates using all available DNA sequence information;
-Perform one global analysis using all available information;
-Problems or downsides: This might be time consuming for large groups and a lot of
-data curation and knowledge on the group of interest is still necessary. For example,
-choosing correct fossils for calibration
-requires a lot of expertise and knowledge on the group. An incorrect use of fossils
-can generate severe bias in dating results (Sauquet et al. \protect\hyperlink{ref-Sauquet2012c}{2012}).
-Hence, data curation is still an important part of any biological study. The research
-community considers it as an important or even crucial step before data analysis.
-Hence, automated processes for large data analysis are frequently received with skepticism.
-
-DateLife palliates this by only using information available from already published
-studies, which are ideally constructed using robust information, such as sequence
-data and thoughtfully curated fossil calibrations.
-
-Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
-inspired the development of a prototype \texttt{DateLife} service over a series
-of phylotastic hackathons (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}) at the National Evolutionary Synthesis Center.
-In this paper we present the first formal description of \texttt{DateLife}, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
-DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
-(Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013})
-It can be used through an R package , a web interface
-(\url{http://www.datelife.org/query/}) and an API.
-
-\begin{center}
-\textsc{Description}
-\end{center}
-
-The basic \texttt{DateLife} workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
-\tightlist
-\item
-  A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item
-  \texttt{DateLife} then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarization algorithms.
-\item
-  The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
-\item
-  At this point, users can choose to use all or some source data as calibration points to date a tree of their own making or choosing. 
-\item
-  Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
-\item
-  Finally, users can easily view results graphically as well as construct their own graphs using inbuilt \texttt{DateLife} graphic generators.
-\end{enumerate}
-
-\texttt{DateLife}'s chronogram database is currently built from Open Tree of Life (OToL)'s
-(Hinchliff et al. \protect\hyperlink{ref-Hinchliff2015}{2015}) tree repository. Among currently existing repositories (e.g., TreeBase,
-Dryad), OToL's metadata rich tree store is the only one meeting the requirements
-for proper/accurate automatized handling of trees.
-Input taxon names accepted by \texttt{DateLife} are binomial species names or clades.
-Taxon searches are performed at the species level, so when input names correspond
-to higher clades,
-\texttt{DateLife} pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
-Currently, searches at the infrapsecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are treated
-as species.
-\texttt{DateLife} also processes input names with the taxon name resolution service (TNRS),
-which corrects potentially misspelled names and typos, and standardizes variation
-in spelling and synonyms (Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}), increasing the probability to correctly find the
-queried
-taxa in \texttt{DateLife}'s chronogram database.
-
-Source chronogram summary tree can be assembled using the Super Distance Matrix
-(SDM) supertree construction approach (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}) or using the median of branch
-lengths and the hierarchical clustering method.
-Tree dating and simulation options are performed with various algorithms:
-Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
-nodes evenly, which minimizes age variance in the chronogram (Webb et al. \protect\hyperlink{ref-Webb2008}{2008}).
-PATHd8 is a non-clock, rate-smoothing method (Britton et al. \protect\hyperlink{ref-Britton2007}{2007}) to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Smith and O'Meara \protect\hyperlink{ref-Smith2012}{2012}).
-MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) can be used when adding taxa at
-random, following a reference taxonomy or a topological constraint. It draws ages
-from a pure birth model, as implemented by Jetz and collaborators (\protect\hyperlink{ref-Jetz2012}{2012}).
-To apply calibrations to a tree, th econgruification algorithm described in (Eastman et al. \protect\hyperlink{ref-Eastman2013}{2013})
-is used to find shared nodes between trees (congruent nodes).
-
-To gather, process, and present information, \texttt{DateLife} builds up from functions
-available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
-geiger (Harmon et al.
-\protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}),
-taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018})
-and rphylotastic (O'Meara et al. \protect\hyperlink{ref-Omeara2019}{2019}).
-
-Details on each step are further developed in \texttt{DateLife}'s R package documentation
-\texttt{datelife\ workflow} vignette at (\url{https://LINK}).
-
-\begin{center}
-\textsc{Benchmark}
-\end{center}
-
-\texttt{DateLife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time relative to number of input names
-and \texttt{DateLife} service.
-Input processing increases roughly linearly with number of input taxon names, and
-increases considerably if tnrs service is activated (Fig. \ref{fig:runtime1}).
-Results show that searching time increases linearly with number of input names and
-number of chronograms in database.
-
-Summarizing DateLife results processing times
-
-Adding dates processing time
-
-get\_bold\_otol\_tree running time
-
-\texttt{DateLife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}). These tests were run both
-locally --using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018})-- and on a public server --via
-GitHub-- using the continuous integration tool Travis CI (\textless{}//travis-ci.org\textgreater{}). At
-present, unit tests cover more than 50\% (for now) of \texttt{DateLife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\begin{center}
-\textsc{Example}
-\end{center}
-
-In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using the bird family Fringillidae of true finches as example.
-We performed a higher-taxon search to obtain all data on lineage divergence available from \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
-according to the Open Tree of Life taxonomy).
-We found 13 trees across 9 studies (Fig.
-\ref{fig:schronograms}.
-
-\begin{center}
-\textsc{Conclusions}
-\end{center}
-
-Taxon ages are key to many areas of evolutionary studies: trait evolution, species
-diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to develop the necessary knowledge and data curation skills
-to produce new chronograms. Knowledge on taxon ages is also important for non-biological
-studies and the non-academic community.
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree
-of Life, for a large quantity and diversity of organisms. However, this information
-remains difficult to synthesize for many biologists and the non-academic community.
-
-Here, we have shown that DateLife allows
-an easy and fast obtention of all publicly available information on taxon ages,
-which can be used to generate new data.
-This information can be used to account for the effect of phylogenetic signal in
-studies of trait evolution; to explore potential speciation and extinction dynamics
-of interest within a clade; to obtain a time frame of biogeographical events; for
-science communication and outreach, amongst others.
-Compared to similar platforms such as time tree of life
-and supermart, it offers several advantages.
-It is fast;
-source data is completely open;
-it requires no expert biological knowledge from users for any of its functionalities;
-it allows exploration of alternative taxonomic and phylogenetic schemes;
-it allows rapid exploration of the effect of alternative divergence time hypothesis;
-it allows rapid synthesis in a number of different formats;
-it facilitates reproducibility of analyses;
-
-Improvements, short and long-term:
-* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
-(Schenk \protect\hyperlink{ref-Schenk2016}{2016}).
-* bayesian congruification
-* topological congruification
-
-Problems and caveats:
-Not many databases, only OToL
-Why TreeBase is not very useful for us? Be precise.
-Are these chronograms reliable to study evolutionary patterns, such as species diversification?
-DateLife can be seen as an open resource to know the current state of knowledge on lineage divergence times.
-Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
-If all, la facilidad para obtener hipotesis de tiempo de divergencia nos ayudará a evaluar la capacidad de los cronogramas para estudiar otros fenomenos evolutivos.
-Por ahora, no podemos aseverar que estos cronogramas puedan usarse para todo tipo de analisis.
-
-\begin{center}
-\textsc{Availability}
-\end{center}
-
-\texttt{Datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{Datelife}'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{DateLife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\begin{center}
-\textsc{Supplementary Material}
-\end{center}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in the GitHub repositories \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
-
-\begin{center}
-\textsc{Funding}
-\end{center}
-
-Funding was provided by NSF grant 1458603
-
-NESCent
-
-Open Tree of Life
-
-University of Tennessee, Knoxville
-
-\begin{center}
-\textsc{Acknowledgements}
-\end{center}
-
-We thank colleagues (students and postdocs) at the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for DateLife.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in OpenTree.
-The US National Science Foundation (NSF) for funding nearly all the above, in addition
-to the ABI grant that funded this project itself.
-
-\newpage
-
-\begin{center}
-\textsc{References}
-\end{center}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst D.W. 2012. Paleotree: An R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution. 3:803--807.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle B., Hopkins N., Lu Z., Raygoza Garay J.A., Mozzherin D., Rees T., Matasci N., Narro M.L., Piel W.H., Mckay S.J., Lowry S., Freeland C., Peet R.K., Enquist B.J.
-2013.
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton T., Anderson C.L., Jacquet D., Lundqvist S., Bremer K. 2007. Estimating Divergence Times in Large Phylogenetic Trees. Systematic Biology. 56:741--752.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain S. 2018. bold: Interface to Bold Systems API..
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain S.A., Szöcs E. 2013. taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}. F1000Research. 2:1--29.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo A., Berry V., Douzery E.J., Gascuel O. 2006. SDM: A fast distance-based approach for (super)tree building in phylogenomics. Systematic Biology. 55:740--755.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman J.M., Harmon L.J., Tank D.C. 2013. Congruification: Support for time scaling large phylogenetic trees. Methods in Ecology and Evolution. 4:688--691.
-
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon L., Weir J., Brock C., Glor R., Challenger W. 2008. GEIGER: investigating evolutionary radiations. Bioinformatics. 24:129--131.
-
-\leavevmode\hypertarget{ref-Hedges2006}{}%
-Hedges S.B., Dudley J., Kumar S. 2006. TimeTree: A public knowledge-base of divergence times among organisms. Bioinformatics. 22:2971--2972.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl C. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..
-
-\leavevmode\hypertarget{ref-Hinchliff2015}{}%
-Hinchliff C.E., Smith S.A., Allman J.F., Burleigh J.G., Chaudhary R., Coghill L.M., Crandall K.A., Deng J., Drew B.T., Gazis R., Gude K., Hibbett D.S., Katz L.A., Laughinghouse H.D., McTavish E.J., Midford P.E., Owen C.L., Ree R.H., Rees J.A., Soltis D.E., Williams T., Cranston K.A.
-2015.
-Synthesis of phylogeny and taxonomy into a comprehensive tree of life.
-Proceedings of the National Academy of Sciences.
-112:12764--12769.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck J.P., Ronquist F. 2001. MRBAYES: Bayesian inference of phylogenetic trees. Bioinformatics. 17:754--755.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.
-
-\leavevmode\hypertarget{ref-Kumar2017}{}%
-Kumar S., Stecher G., Suleski M., Hedges S.B. 2017. TimeTree: A Resource for Timelines, Timetrees, and Divergence Times. Molecular biology and evolution. 34:1812--1819.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau F., Brown J.W., Winter D.J. 2016. rotl: an R package to interact with the Open Tree of Life data. Methods in Ecology and Evolution. 7:1476--1481.
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon H. 2014. Phylogenetic approaches for studying diversification. Ecology Letters. 17:508--525.
-
-\leavevmode\hypertarget{ref-Omeara2019}{}%
-O'Meara B., Md Tayeen A.S., Sanchez Reyes L.L. 2019. Rphylotastic: An r interface to 'phylotastic' web services..
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms J., Chamberlain S. 2018. Phylocomr: Interface to 'phylocom'..
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. 2018. R: a language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell L.J. 2012. Phytools: An r package for phylogenetic comparative biology (and other things). Methods in Ecology and Evolution. 3:217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist F., Huelsenbeck J.P. 2003. MrBayes 3: Bayesian phylogenetic inference under mixed models. Bioinformatics. 19:1572--1574.
-
-\leavevmode\hypertarget{ref-Sauquet2012c}{}%
-Sauquet H., Ho S.Y.W., Gandolfo M.
-a, Jordan G.J., Wilf P., Cantrill D.J., Bayly M.J., Bromham L., Brown G.K., Carpenter R.J., Lee D.M., Murphy D.J., Sniderman J.M.K., Udovicic F.
-2012.
-Testing the impact of calibration on molecular divergence times using a fossil-rich group: the case of Nothofagus (Fagales).
-Systematic Biology.
-61:289--313.
-
-\leavevmode\hypertarget{ref-Schenk2016}{}%
-Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
-
-\leavevmode\hypertarget{ref-Smith2012}{}%
-Smith S.A., O'Meara B.C. 2012. TreePL: Divergence time estimation using penalized likelihood for large phylogenies. Bioinformatics. 28:2689--2690.
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus A., Lapp H., Matasci N., Deus H., Sidlauskas B., Zmasek C.M., Vaidya G., Pontelli E., Cranston K., Vos R., Webb C.O., Harmon L.J., Pirrung M., O'Meara B., Pennell M.W., Mirarab S., Rosenberg M.S., Balhoff J.P., Bik H.M., Heath T.A., Midford P.E., Brown J.W., McTavish E.J., Sukumaran J., Westneat M., Alfaro M.E., Steele A., Jordan G.
-2013.
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb C.O., Ackerly D.D., Kembel S.W. 2008. Phylocom: Software for the analysis of phylogenetic community structure and trait evolution. Bioinformatics. 24:2098--2100.
-
 \leavevmode\hypertarget{ref-barker2012going}{}%
@@
 \end{figure}
-
-\end{document}

### 2019-05-21 frin_plan.R

@@ -1,0 +1,16 @@
+library(datelife)
+library(drake)
+frin_plan <- drake_plan(
+  strings_in_dots = "literals",
+  dq = make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE),
+  dr = get_datelife_result(input = dq),
+  summ = get_taxon_summary(datelife_query = dq, datelife_result = dr),
+  phyloall = summarize_datelife_result(datelife_query = dq,
+    datelife_result = dr, summary_format = "phylo_all",
+    taxon_summary = "none"),
+  datedotol = get_dated_otol_induced_subtree(input = dq,
+    ott_ids = dq$ott_ids),
+  otol = get_otol_synthetic_tree(input = dq),
+  bestgrove = get_best_grove(datelife_result = dr)
+)
+make(frin_plan)

### 2019-05-23 Datelife_paper1.Rmd
all figures, yay

@@
     # - \usepackage[document]{ragged2e}
+nocite: | 
+  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
 ---
 
-```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE}
+```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
 source("Datelife_paper1_global.R")
@@
 
-DateLife palliates this by only using information available from already published
+`datelife` palliates this by only using information available from already published
 studies, which are ideally constructed using robust information, such as sequence
@@
 Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
-inspired the development of a prototype `DateLife` service over a series
+inspired the development of a prototype `datelife` service over a series
 of phylotastic hackathons [@Stoltzfus2013] at the National Evolutionary Synthesis Center.
-In this paper we present the first formal description of `DateLife`, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
-DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
+In this paper we present the first formal description of `datelife`, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
+`datelife` is the main service for scaling phylogenetic trees in Phylotastic! system
 [@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
@@
 <!--(BOLD and OToL species names, are they homogeneous/standardized?-->
-The basic `DateLife` workflow is shown in figure \ref{fig:workflow} and consists of:
+The basic `datelife` workflow is shown in figure \ref{fig:workflow} and consists of:
 
@@
 The tree can be in newick or phylo format, and can be with or without branch lengths.
-2) `DateLife` then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
+2) `datelife` then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
 This format facilitates and greatly speeds up all further analyses and summarization algorithms.
@@
 5) Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
-6) Finally, users can easily view results graphically as well as construct their own graphs using inbuilt `DateLife` graphic generators.
+6) Finally, users can easily view results graphically as well as construct their own graphs using inbuilt `datelife` graphic generators.
 
 <!--To perform the former tasks:-->
-`DateLife`'s chronogram database is currently built from Open Tree of Life (OToL)'s
+`datelife`'s chronogram database is currently built from Open Tree of Life (OToL)'s
 [@Hinchliff2015] tree repository. Among currently existing repositories (e.g., TreeBase,
@@
 for proper/accurate automatized handling of trees.
-Input taxon names accepted by `DateLife` are binomial species names or clades.
+Input taxon names accepted by `datelife` are binomial species names or clades.
 <!--common names (implemented yet?)-->
@@
 to higher clades,
-`DateLife` pulls all accepted species names within the
+`datelife` pulls all accepted species names within the
 clade from OToL's reference taxonomy to perform the search.
@@
 as species.
-`DateLife` also processes input names with the taxon name resolution service (TNRS),
+`datelife` also processes input names with the taxon name resolution service (TNRS),
 <!--which "corrects misspelled names and authorities, standardizes variant spellings,
@@
 to otol taxonomy -->[@Boyle2013], increasing the probability to correctly find the
-queried
-taxa in `DateLife`'s chronogram database.
+queried taxa in `datelife`'s chronogram database.
 
@@
 
-To gather, process, and present information, `DateLife` builds up from functions
+To gather, process, and present information, `datelife` builds up from functions
 available in several R packages including rotl [@Michonneau2016], ape [@Paradis2004],
@@
 
-Details on each step are further developed in `DateLife`'s R package documentation
-`datelife workflow` vignette at (<https://LINK>).
+<!-- Details on each step are further developed in `datelife`'s R package documentation -->
+<!-- `datelife workflow` vignette at (<https://LINK>). -->
 
@@
 
-`DateLife`'s code speed was tested on an Apple iMac
+`datelife`'s code speed was tested on an Apple iMac
 with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
-We registered variation in computing time relative to number of input names
-<!--number of chronograms in database--> and `DateLife` service.
-Input processing increases roughly linearly with number of input taxon names, and
-increases considerably if tnrs service is activated (Fig. \ref{fig:runtime1}).
+We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
+Query processing increases roughly linearly with number of input taxon names, and
+increases considerably if TNRS service is activated.
+Up to ten thousand names can be processed and searched in less than 30 minutes.
+A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig.
+\ref{fig:runtime1}).
 <!--
 Speed with different number of input lineages and types of analysis:
--->
 Results show that searching time increases linearly with number of input names and
 number of chronograms in database.
+Summarizing `datelife` results processing times: sdm speed, clustering methods speed
 
-Summarizing DateLife results processing times
+generating new trees: bladj speed
+Adding dates processing time: reflects more mrbayes speed than anything
 
-Adding dates processing time
-
-get_bold_otol_tree running time
+get_bold_otol_tree running time: not really useful since that is an external service
+-->
 <!-- test improvement in synthetic tree after using bold sequences. Is it really better resolved?-->
@@
 <!--Speed of web interface and of r package (in computers with different capacities?)-->
-`DateLife`'s code performance was evaluated with a set of unit tests designed and
+`datelife`'s code performance was evaluated with a set of unit tests designed and
 implemented with the R package testthat [@RCoreTeam2018]. These tests were run both
 locally –using the devtools package [@RCoreTeam2018]– and on a public server –via
-GitHub– using the continuous integration tool Travis CI (<//travis-ci.org>). At
-present, unit tests cover more than 50% (for now) of `DateLife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
+GitHub– using the continuous integration tool Travis CI (<https://travis-ci.org>). At
+present, unit tests cover more than 50% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
 
@@
 -->
-In this section we demonstrate the types of outputs that can be obtained with `datelife`, using the bird family Fringillidae of true finches as example.
+In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches.
 We performed a higher-taxon search to obtain all data on lineage divergence available from `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp.
 according to the Open Tree of Life taxonomy).
-We found `r length(dr)` trees across `r length(unique(names(dr)))` studies (Fig.
-\ref{fig:schronograms}.
+There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig.
+\ref{fig:schronograms1}).
+<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig.
+\ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
+-->
+Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
+As explained in the `Description`, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
+\ref{fig:summaries}).
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+In both cases, ages are coherent with source ages.
+<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
+<!-- and hierarchical clustering method which appears to push ages back -->
+It is not certain if these chronograms can be used to perform downstream evolutionary analyses.
+There is currently wide interest in determining this.
+However, it is certain that these chronograms are useful for...
+
+Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs.
+\ref{fig:cvbladj} and \ref{fig:cvbold}).
+As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms.
+In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig.
+\ref{fig:cvbladj}).
+Maximum tree ages were only approximately recovered in one case (study 2; Fig.
+\ref{fig:cvbladj}).
+Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in  
+<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
+three cases (studies 3, 5, and 9; Fig.
+\ref{fig:cvbold}).
+From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD.
+Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+More examples and details can be consulted in <https://github.com/LunaSare/datelife_examples>.
 
@@
 
-<!--Potential applications shown here-->Here, we have shown that DateLife allows
+<!--Potential applications shown here-->Here, we have shown that `datelife` allows
 an easy and fast obtention of all publicly available information on taxon ages,
@@
 Are these chronograms reliable to study evolutionary patterns, such as species diversification?
-DateLife can be seen as an open resource to know the current state of knowledge on lineage divergence times.
+`datelife` can be seen as an open resource to know the current state of knowledge on lineage divergence times.
 Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
@@
 
-`Datelife` is free and open source and it can be used through its current website
+`datelife` is free and open source and it can be used through its current website
 <http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
-`Datelife`'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-`DateLife`'s R package stable version is available
+`datelife`'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+`datelife`'s R package stable version is available
 for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
 using the command `install.packages(pkgs = "datelife")` from within R. Development versions
-are available from GitHub repository (<https://github.com/phylotastic/datelife>)
+are available from the GitHub repository (<https://github.com/phylotastic/datelife>)
 and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in the GitHub repositories <https://github.com/LunaSare/datelife_paper1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>, respectively.
+Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in GitHub repositories at <https://github.com/LunaSare/datelife_paper1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>, respectively.
 <!-- Vignettes
@@
 The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for DateLife.
+trees used for `datelife`.
 The many scientists who publish their chronograms in an open, reusable form, and

### 2019-05-23 Datelife_paper1.tex
all figures, yay

@@
 
-DateLife palliates this by only using information available from already published
+\texttt{datelife} palliates this by only using information available from already published
 studies, which are ideally constructed using robust information, such as sequence
@@
 Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
-inspired the development of a prototype \texttt{DateLife} service over a series
+inspired the development of a prototype \texttt{datelife} service over a series
 of phylotastic hackathons (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}) at the National Evolutionary Synthesis Center.
-In this paper we present the first formal description of \texttt{DateLife}, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
-DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
+In this paper we present the first formal description of \texttt{datelife}, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
+\texttt{datelife} is the main service for scaling phylogenetic trees in Phylotastic! system
 (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013})
@@
 
-The basic \texttt{DateLife} workflow is shown in figure \ref{fig:workflow} and consists of:
+The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
 
@@
 \item
-  \texttt{DateLife} then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
+  \texttt{datelife} then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
 This format facilitates and greatly speeds up all further analyses and summarization algorithms.
@@
 \item
-  Finally, users can easily view results graphically as well as construct their own graphs using inbuilt \texttt{DateLife} graphic generators.
+  Finally, users can easily view results graphically as well as construct their own graphs using inbuilt \texttt{datelife} graphic generators.
 \end{enumerate}
 
-\texttt{DateLife}'s chronogram database is currently built from Open Tree of Life (OToL)'s
+\texttt{datelife}'s chronogram database is currently built from Open Tree of Life (OToL)'s
 (Hinchliff et al. \protect\hyperlink{ref-Hinchliff2015}{2015}) tree repository. Among currently existing repositories (e.g., TreeBase,
@@
 for proper/accurate automatized handling of trees.
-Input taxon names accepted by \texttt{DateLife} are binomial species names or clades.
+Input taxon names accepted by \texttt{datelife} are binomial species names or clades.
 Taxon searches are performed at the species level, so when input names correspond
 to higher clades,
-\texttt{DateLife} pulls all accepted species names within the
+\texttt{datelife} pulls all accepted species names within the
 clade from OToL's reference taxonomy to perform the search.
@@
 as species.
-\texttt{DateLife} also processes input names with the taxon name resolution service (TNRS),
+\texttt{datelife} also processes input names with the taxon name resolution service (TNRS),
 which corrects potentially misspelled names and typos, and standardizes variation
 in spelling and synonyms (Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}), increasing the probability to correctly find the
-queried
-taxa in \texttt{DateLife}'s chronogram database.
+queried taxa in \texttt{datelife}'s chronogram database.
 
@@
 
-To gather, process, and present information, \texttt{DateLife} builds up from functions
+To gather, process, and present information, \texttt{datelife} builds up from functions
 available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
@@
 \protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}),
-taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018})
+taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl \protect\hyperlink{ref-Heibl2008}{2008}), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018})
 and rphylotastic (O'Meara et al. \protect\hyperlink{ref-Omeara2019}{2019}).
-
-Details on each step are further developed in \texttt{DateLife}'s R package documentation
-\texttt{datelife\ workflow} vignette at (\url{https://LINK}).
 
@@
 
-\texttt{DateLife}'s code speed was tested on an Apple iMac
+\texttt{datelife}'s code speed was tested on an Apple iMac
 with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time relative to number of input names
-and \texttt{DateLife} service.
-Input processing increases roughly linearly with number of input taxon names, and
-increases considerably if tnrs service is activated (Fig. \ref{fig:runtime1}).
-Results show that searching time increases linearly with number of input names and
-number of chronograms in database.
+We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
+Query processing increases roughly linearly with number of input taxon names, and
+increases considerably if TNRS service is activated.
+Up to ten thousand names can be processed and searched in less than 30 minutes.
+A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig.
+\ref{fig:runtime1}).
 
-Summarizing DateLife results processing times
-
-Adding dates processing time
-
-get\_bold\_otol\_tree running time
-
-\texttt{DateLife}'s code performance was evaluated with a set of unit tests designed and
+\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
 implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}). These tests were run both
 locally --using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018})-- and on a public server --via
-GitHub-- using the continuous integration tool Travis CI (\textless{}//travis-ci.org\textgreater{}). At
-present, unit tests cover more than 50\% (for now) of \texttt{DateLife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
+GitHub-- using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
+present, unit tests cover more than 50\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
 
@@
 
-In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using the bird family Fringillidae of true finches as example.
+In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using as an example the bird family Fringillidae of true finches.
 We performed a higher-taxon search to obtain all data on lineage divergence available from \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
 according to the Open Tree of Life taxonomy).
-We found 13 trees across 9 studies (Fig.
-\ref{fig:schronograms}.
+There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig.
+\ref{fig:schronograms1}).
+Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
+As explained in the \texttt{Description}, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
+\ref{fig:summaries}).
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+In both cases, ages are coherent with source ages.
+It is not certain if these chronograms can be used to perform downstream evolutionary analyses.
+There is currently wide interest in determining this.
+However, it is certain that these chronograms are useful for\ldots{}
+
+Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs.
+\ref{fig:cvbladj} and \ref{fig:cvbold}).
+As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms.
+In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig.
+\ref{fig:cvbladj}).
+Maximum tree ages were only approximately recovered in one case (study 2; Fig.
+\ref{fig:cvbladj}).
+Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in\\
+three cases (studies 3, 5, and 9; Fig.
+\ref{fig:cvbold}).
+From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD.
+Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+More examples and details can be consulted in \url{https://github.com/LunaSare/datelife_examples}.
 
@@
 
-Here, we have shown that DateLife allows
+Here, we have shown that \texttt{datelife} allows
 an easy and fast obtention of all publicly available information on taxon ages,
@@
 Are these chronograms reliable to study evolutionary patterns, such as species diversification?
-DateLife can be seen as an open resource to know the current state of knowledge on lineage divergence times.
+\texttt{datelife} can be seen as an open resource to know the current state of knowledge on lineage divergence times.
 Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
@@
 
-\texttt{Datelife} is free and open source and it can be used through its current website
+\texttt{datelife} is free and open source and it can be used through its current website
 \url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{Datelife}'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{DateLife}'s R package stable version is available
+\texttt{datelife}'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+\texttt{datelife}'s R package stable version is available
 for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
 using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from GitHub repository (\url{https://github.com/phylotastic/datelife})
+are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
 and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in the GitHub repositories \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
+Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in GitHub repositories at \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
 
@@
 The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for DateLife.
+trees used for \texttt{datelife}.
 The many scientists who publish their chronograms in an open, reusable form, and
@@
 
+\leavevmode\hypertarget{ref-barker2012going}{}%
+Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
+2012.
+Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
+Systematic biology.
+62:298--320.
+
+\leavevmode\hypertarget{ref-barker2015new}{}%
+Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
+2015.
+New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
+The Auk: Ornithological Advances.
+132:333--348.
+
 \leavevmode\hypertarget{ref-Boyle2013}{}%
@@
 
+\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
+Burns K.J., Shultz A.J., Title P.O., Mason N.A., Barker F.K., Klicka J., Lanyon S.M., Lovette I.J.
+2014.
+Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
+Molecular Phylogenetics and Evolution.
+75:41--77.
+
 \leavevmode\hypertarget{ref-Chamberlain2018}{}%
@@
 
+\leavevmode\hypertarget{ref-claramunt2015new}{}%
+Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.
+
 \leavevmode\hypertarget{ref-Criscuolo2006}{}%
@@
 Eastman J.M., Harmon L.J., Tank D.C. 2013. Congruification: Support for time scaling large phylogenetic trees. Methods in Ecology and Evolution. 4:688--691.
+
+\leavevmode\hypertarget{ref-gibb2015new}{}%
+Gibb G.C., England R., Hartig G., McLenachan P.A., Taylor Smith B.L., McComish B.J., Cooper A., Penny D.
+2015.
+New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
+Genome biology and evolution.
+7:2983--2995.
 
@@
 \leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl C. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..
+Heibl C. 2008. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..
 
@@
 112:12764--12769.
+
+\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
+Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.
 
@@
 Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.
+
+\leavevmode\hypertarget{ref-price2014niche}{}%
+Price T.D., Hooper D.M., Buchanan C.D., Johansson U.S., Tietze D.T., Alström P., Olsson U., Ghosh-Harihar M., Ishtiaq F., Gupta S.K., others.
+2014.
+Niche filling slows the diversification of himalayan songbirds.
+Nature.
+509:222.
 
@@
 
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2012.
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-Systematic biology.
-62:298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2015.
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk: Ornithological Advances.
-132:333--348.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns K.J., Shultz A.J., Title P.O., Mason N.A., Barker F.K., Klicka J., Lanyon S.M., Lovette I.J.
-2014.
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-Molecular Phylogenetics and Evolution.
-75:41--77.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb G.C., England R., Hartig G., McLenachan P.A., Taylor Smith B.L., McComish B.J., Cooper A., Penny D.
-2015.
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-Genome biology and evolution.
-7:2983--2995.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price T.D., Hooper D.M., Buchanan C.D., Johansson U.S., Tietze D.T., Alström P., Olsson U., Ghosh-Harihar M., Ishtiaq F., Gupta S.K., others.
-2014.
-Niche filling slows the diversification of himalayan songbirds.
-Nature.
-509:222.
-
 \newpage
@@
 \begin{center}
-\textsc{Figure 1}
+\textsc{Figure \ref{fig:workflow}}
 \end{center}
 Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with DateLife, via the R package or through the website.
+This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website.
 Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
@@
 \begin{center}
-\textsc{Figure 2}
+\textsc{Figure \ref{fig:runtime1}}
 \end{center}
-Computation time of input processing and search across \texttt{datelife}'s chronogram database.
+Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
+We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
+We also performed a search using the already processed query for comparison (light blue).
 
 \begin{center}
-\textsc{Figure 3}
+\textsc{Figure \ref{fig:schronograms1}}
 \end{center}
@@
 \begin{center}
-\textsc{Figure 4}
+\textsc{Figure \ref{fig:summaries}}
 \end{center}
-LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarizing information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
+LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
+
+\begin{center}
+\textsc{Figure \ref{fig:cvbladj}}
+\end{center}
+LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
+
+\begin{center}
+\textsc{Figure \ref{fig:cvbold}}
+\end{center}
+LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successfull in three source chronograms shown here.
 
@@
 \caption{}
-\label{fig:schronograms}
+\label{fig:schronograms1}
 \end{figure}
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!h]
+\includegraphics{fig_summaries.pdf}
+%\includegraphics{fig_summaries.pdf}
+\caption{}
+\label{fig:summaries}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
 \includegraphics{fig_crossval_bladj.pdf}
@@
 
+\newpage
+
+\begin{figure}[!h]
+\includegraphics{fig_crossval_boldsumm.pdf}
+\caption{}
+\label{fig:cvbold}
+\end{figure}
+
 \end{document}

### 2019-05-23 Datelife_paper1_figures_noyaml.tex
all figures, yay

@@
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2012.
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-Systematic biology.
-62:298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2015.
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk: Ornithological Advances.
-132:333--348.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns K.J., Shultz A.J., Title P.O., Mason N.A., Barker F.K., Klicka J., Lanyon S.M., Lovette I.J.
-2014.
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-Molecular Phylogenetics and Evolution.
-75:41--77.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb G.C., England R., Hartig G., McLenachan P.A., Taylor Smith B.L., McComish B.J., Cooper A., Penny D.
-2015.
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-Genome biology and evolution.
-7:2983--2995.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price T.D., Hooper D.M., Buchanan C.D., Johansson U.S., Tietze D.T., Alström P., Olsson U., Ghosh-Harihar M., Ishtiaq F., Gupta S.K., others.
-2014.
-Niche filling slows the diversification of himalayan songbirds.
-Nature.
-509:222.
-
 \newpage
@@
 \begin{center}
-\textsc{Figure 4}
+\textsc{Figure \ref{fig:summaries}}
 \end{center}
-LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarizing information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
+LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
+
+\begin{center}
+\textsc{Figure \ref{fig:cvbladj}}
+\end{center}
+LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
+
+\begin{center}
+\textsc{Figure \ref{fig:cvbold}}
+\end{center}
+LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successful in three source chronograms shown here.
 
@@
 \end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\includegraphics{fig_crossval_boldsumm.pdf}
+\caption{}
+\label{fig:cvbold}
+\end{figure}

### 2019-05-23 Datelife_paper1_global.R

@@ -6,7 +6,7 @@ download.file(url = "https://github.com/phylotastic/datelife/raw/b3532e70b679272
               destfile = "fig_runtime1.pdf", mode="wb")
 # fig_runtime1_cap <- "Computation time of input processing and search across `DateLife`s 
 #   chronogram database"
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringilidae_LTTplot_phyloall.pdf",
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_phyloall.pdf",
               destfile = "fig_schronograms1.pdf", mode="wb")
 # fig_schronograms_cap <- "Lineage through time (LTT) plots of source chronograms containing 
 #   all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers 
@@ -14,8 +14,13 @@ download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/do
 #   4) [@claramunt2015new], 5) [@gibb2015new], 6) [@Hedges2015], 7) [@hooper2017chromosomal],
 #   8) [@Jetz2012], 9) [@price2014niche]"
   
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringilidae_LTTplot_crossval_bladj.pdf",
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_summary_chronograms2.pdf",
+              destfile = "fig_summaries.pdf", mode="wb")
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_crossval_bladj.pdf",
               destfile = "fig_crossval_bladj.pdf", mode="wb")
-# fig_crossval_bladj
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_lttplot_crossval_pathd8_summ1.pdf",
+              destfile = "fig_crossval_boldsumm.pdf", mode="wb")
+
+
 source("frin_plan.R")
 loadd() # loads all the targets listed in the workflow plan of the previous `make()`

### 2019-05-23 Datelife_paper1.Rmd
all figures, yay

@@
 \newpage
-
-<!--INTRODUCTION-->
-Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
-
 <!--INTRODUCTION: 1st paragraph: general statement that the field knows to be true and important-->
 Time of lineage divergence<!--alt lineages: named biological entities, taxon ages?-->
-constitutes in many ways the fundamental/main knowledge necessary for evolutionary
+constitutes in many ways a fundamental piece of information for evolutionary
 understanding.
-<!--Along with phylogenetic relationships, it is the most important piece of information
-allowing to test alternative evolutionary hypothesis in many areas of research,
-from developmental to ecological questions.
+<!--Along with phylogenetic relationships, it is crucial for testing alternative
+evolutionary hypothesis in many areas of research,
+from developmental to ecological questions @Felsenstein1985a @Webb2008.
+It has been shown that even a little bit of information on a lineages time of origin is good to study ecological stuff @Webb2008
+comparative studies, in general, for which knowing
+the time context for all life is crucial.
 Along with migration rates and patterns, it is fundamental to test biogeographical
 hypothesis-->
-Coupled to species number and distribution, it is the main information necessary
-for the study of diversification processes (i.e., the tempo and mode of speciation
+Coupled to species number and distribution, it allows studying the tempo and mode of speciation
 and extinction),
-<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity
+<!-- crucial/vital/critical/essential --> central for understanding of how biodiversity
 patterns are shaped across space, time and clades [@Morlon2014].
-Evolutionary understanding also relies on comparative studies, for which knowing
-the time context for all life is crucial. Efforts to have a whole tree of life have
-been great and here are some examples.
+<!-- Efforts to have a whole tree of life have
+been great and here are some examples. -->
 <!--INTRODUCTION: 2nd paragraph: important advances in the field-->
@@
 the application of molecular dating methods on a very large amount and diversity
-of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon
+of organisms, greatly increasing the quantity and quality of<!--organisms with--> data on taxon
 ages across the tree of life.
 <!-- figure showing growth on studies with age data from references table from ttol-->
-To date, there is a large amount of both fossil and molecular-based data on taxon
-ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
+To date, there is a large amount of
+<!-- both fossil and molecular-based  -->
+data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
 and Open Tree of Life (OToL).
-OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
+<!-- pbdb ??-->
+OToL alone hosts now more than 200 chronograms. <!--cache updated jan 15 2018
 devtools::load_all(pkg = "~/Desktop/datelife")
@@
 [Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from
-across the tree of life, versus more specialized chronograms focusing on plants
-[PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups. Users can choose
+across the tree of life, versus more specialized chronogram repositories focusing on plants
+[@webb2005phylomatic], birds [@Jetz2012], and fish . Users can choose
 between a web interface or a mobile app to receive information on divergence times
@@
 We identified several aspects that might have so far delayed the exploitation of
-existing data. First, original chronograms available publicly are scattered across
-various repositories (otol tree store, dryad, treebase, journals supplementary data)
-usually with different formats too.
+existing data. First, original chronograms available publicly are scattered across repositories
+<!-- (otol tree store, dryad, treebase,  -->
+and journals supplementary data, usually with different formats too.
 Second, lineage names due to taxonomic idiosincracy can be different among studies
@@
 Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
+Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
 Generate new dates using all available DNA sequence information;
@@
 [@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
-It can be used through an R package <link to documentation>, a web interface
-(<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
+It can be used through its R package <link to documentation>, and a web interface
+(<http://www.datelife.org/query/>)<!-- and an API (still not up, tho)-->.
 

### 2019-05-23 Datelife_paper1.tex
all figures, yay

@@
 
-Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
-
 Time of lineage divergence
-constitutes in many ways the fundamental/main knowledge necessary for evolutionary
+constitutes in many ways a fundamental piece of information for evolutionary
 understanding.
-Coupled to species number and distribution, it is the main information necessary
-for the study of diversification processes (i.e., the tempo and mode of speciation
+Coupled to species number and distribution, it allows studying the tempo and mode of speciation
 and extinction),
-central for the understanding of how biodiversity
+central for understanding of how biodiversity
 patterns are shaped across space, time and clades (Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-Evolutionary understanding also relies on comparative studies, for which knowing
-the time context for all life is crucial. Efforts to have a whole tree of life have
-been great and here are some examples.
 In the past two decades, the possibility to obtain good quality DNA sequences
@@
 the application of molecular dating methods on a very large amount and diversity
-of organisms, greatly increasing the quantity of data on taxon
+of organisms, greatly increasing the quantity and quality of data on taxon
 ages across the tree of life.
-To date, there is a large amount of both fossil and molecular-based data on taxon
-ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
+To date, there is a large amount of
+data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
 and Open Tree of Life (OToL).
-OToL alone holds more than 200 chronograms.
+OToL alone hosts now more than 200 chronograms.
 Methods to include living and fossil lineages are in continued development and increased
@@
 {[}Mayr2010; Moran2011{]}). One advantage of TimeTree is that it includes taxa from
-across the tree of life, versus more specialized chronograms focusing on plants
-{[}PHYLOMATIC{]}, birds {[}JETZ ET AL BIRDTREE.ORG{]}, and other groups. Users can choose
+across the tree of life, versus more specialized chronogram repositories focusing on plants
+(Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}), birds (Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012}), and fish . Users can choose
 between a web interface or a mobile app to receive information on divergence times
@@
 We identified several aspects that might have so far delayed the exploitation of
-existing data. First, original chronograms available publicly are scattered across
-various repositories (otol tree store, dryad, treebase, journals supplementary data)
-usually with different formats too.
+existing data. First, original chronograms available publicly are scattered across repositories
+and journals supplementary data, usually with different formats too.
 Second, lineage names due to taxonomic idiosincracy can be different among studies
@@
 Recent advances on this area (e.g., supersmart) aim to:
+Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
 Generate new dates using all available DNA sequence information;
@@
 (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013})
-It can be used through an R package , a web interface
-(\url{http://www.datelife.org/query/}) and an API.
+It can be used through its R package , and a web interface
+(\url{http://www.datelife.org/query/}).
 
@@
 \ref{fig:cvbladj}).
-Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in\\
+Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
 three cases (studies 3, 5, and 9; Fig.
@@
 
+\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
+Webb C.O., Donoghue M.J. 2005. Phylomatic: Tree assembly for applied phylogenetics. Molecular Ecology Notes. 5:181--183.
+
 \newpage
@@
 We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successfull in three source chronograms shown here.
+However, dating with PATHd8 was only successful in three source chronograms shown here.
 

### 2019-05-29 Datelife_paper1.Rmd
intro

--- 2019-05-23 Datelife_paper1.Rmd
@@
 
-**Keywords:** Tree; Phylogeny; Scaling; Open; Ages; Congruify; Supertree;
+**Keywords:** Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
 
 \newpage
+Time of lineage divergence constitutes a fundamental piece of information for evolutionary
+understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014].
+The primary information needed for time of lineage divergence estimation comes from the fossil record.
+Coupled to molecular phylogenies, molecular dating methods are used to reconstruct the time of divergence of extant lineages.
 
-<!--INTRODUCTION-->
-Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
+Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
+Still, generating a chronogram is not easy unless you have specialized training.
+That's why there has been an urge for reuse of the vast amount and still accuulating already available information for the advantage of researchers not specialized in this area.
 
-<!--INTRODUCTION: 1st paragraph: general statement that the field knows to be true and important-->
-Time of lineage divergence<!--alt lineages: named biological entities, taxon ages?-->
-constitutes in many ways the fundamental/main knowledge necessary for evolutionary
-understanding.
-<!--Along with phylogenetic relationships, it is the most important piece of information
-allowing to test alternative evolutionary hypothesis in many areas of research,
-from developmental to ecological questions.
-Along with migration rates and patterns, it is fundamental to test biogeographical
-hypothesis-->
-Coupled to species number and distribution, it is the main information necessary
-for the study of diversification processes (i.e., the tempo and mode of speciation
-and extinction),
-<!-- crucial/vital/critical/essential  -->central for the understanding of how biodiversity
-patterns are shaped across space, time and clades [@Morlon2014].
-Evolutionary understanding also relies on comparative studies, for which knowing
-the time context for all life is crucial. Efforts to have a whole tree of life have
-been great and here are some examples.
-<!--INTRODUCTION: 2nd paragraph: important advances in the field-->
-In the past two decades, the possibility to obtain good quality DNA sequences
-<!-- from a wide variety of organisms became a reality, which,  -->
-coupled to methodological developments in phylogenetic and dating inference, allowed
-the application of molecular dating methods on a very large amount and diversity
-of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon
-ages across the tree of life.
-<!-- figure showing growth on studies with age data from references table from ttol-->
-To date, there is a large amount of both fossil and molecular-based data on taxon
-ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
-and Open Tree of Life (OToL).
-OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
-devtools::load_all(pkg = "~/Desktop/datelife")
-length(opentree_chronograms$trees) names(opentree_chronograms)
+Wide interest from the scientific community to make chronogram information available for consultation and reuse has spurred the creation of public platforms with various goals.
+TreeBASE [@morell1996roots; @Piel2002], Phylomatic [@webb2005phylomatic], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] platforms are dedicated to storing and making available published trees and chronograms for easy scientific reuse.
+For chronogram reuse for a wide range of living organisms, only OToL is generally suitable right now.
+Treebase does not store branch length information and dryad does not distinguish between data sets, so the time of lineage div information cannot be accessed.
+Phylomatic does contain chronograms but is restricted to plants and mammals, and if it tries to add more data it gives many polytomies back.
+
+OToL also has the primary goal of synthesizing stored trees into a single tree of life, to show phylogenetic relationships among known extant species and subspecies.
+All or parts of OToL's synthetic tree can be reused for any purpose; however, it currently does not include information on time of divergence, so the platform does not summarize source chronogram data.
+The Timetree of Life platform main goal is the synthesis of a single chronogram of life [@Hedges2006].
+However, both the latest version of their synthetic chronogram [@Kumar2017] and the thousands of chronograms compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
+
+Other platforms such as SuperSmart [@antonelli2017supersmart] and phylogenerator [@pearse2013phylogenerator] are focused in _de novo_ chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
+However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
+Moreover, they do not provide information from available chronograms.
+
+<!-- discussion:
+The Angiosperm Phylogeny Website [APW; @stevens2001apw] also specializes in constructing a single synthetic tree, but it focuses in showing phylogenetic relationships between seed plant families only, and it does not store or make available the source trees.
+
+Platforms dedicated to store data on fossils and fossil calibrations: PBDB and The fossil calibration database
+can provide data on time of lineage divergence, but it is often not in a phylogenetic context.
+
+To our knowledge, there is no system allowing the comparison of available source information (apw does this a little bit, but it is a verbal comparison more than a graphical or visual comparison).
+Other types of technical comparisons: which trees are better (which trees represent reality better), but this is of course a very difficult issue.
  -->
-<!--How many trees? How many dated trees?-->
-Methods to include living and fossil lineages are in continued development and increased
-usage by the community<!--total evidence and revbayes (fossilized birth-death)-->,
-which coupled to better sharing data practices, are greatly contributing to the
-accumulation in number and type of available data on taxon ages.
-
-<!--INTRODUCTION: 3rd paragraph: problems associated-->
-The TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017] has aggregated chronograms
-from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
-However, even in this gold standard resource, the included taxa only encompass between
-0.097 and 3.236% of total species diversity (following taxonomic expert opinion
-on the global, extant species numbers, which ranges from 3 to 100 million species
-[Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from
-across the tree of life, versus more specialized chronograms focusing on plants
-[PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups. Users can choose
-between a web interface or a mobile app to receive information on divergence times
-for the evolutionary history of a lineage, pairs of taxa, all lineages within a
-taxon, or a list of taxa. As a science communication tool, TimeTree project is
-very powerful: it has a friendly graphical interface, with informative and colorful
-outputs, that allows the general public to satisfy curiosity regarding a particular
-organism of interest or group of them. It is of limited utility for scientific
-studies, however. The thousands of trees that have been entered are unavailable
-for examination or reuse; according to the creators (see TimeTree web FAQ), methods
-for allowing data downloading have been under discussion for the past several years
-yet the primary data remain closed. Moreover, there is no Application Programming
-Interface (API) allowing programmatic access to any data, greatly impairing the
-possibility of large-scale, automated data-mining, which is not allowed under TimeTree
-website's terms of use. The nearly hundred thousand taxon summary chronogram generated
-from TimeTree resources is not available with its publication [@Kumar2017] or the
-TimeTree website, though the still substantial chronogram from a previous publication
-[@Hedges2015] was made available at OToL.
-<!--What led to datelife development?
-Describe public and research necessities covered by datelife
-It is important to use available data on time frame of lineage origin:
-To know the state of dating for a group of interest:
-What range of estimated ages exist already? Are fossil and molecular
-time frames coherent? (e.g., [@Magallon2015c]).
-To construct a time tree of life.
-For science communication, improve scientific discussions, time-framing
-other events of importance in other research areas.
-Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
--->
-
-<!--INTRODUCTION: 4th paragraph-->
-Despite its great importance, analytical tools to summarize available information
-on taxon ages for the scientific community are still lacking.
-We identified several aspects that might have so far delayed the exploitation of
-existing data. First, original chronograms available publicly are scattered across
-various repositories (otol tree store, dryad, treebase, journals supplementary data)
-usually with different formats too.
-Second, lineage names due to taxonomic idiosincracy can be different among studies
-and manual curation of that is usually necessary.
-Third, data curation
-Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
-Generate new dates using all available DNA sequence information;
-Perform one global analysis using all available information;
-Problems or downsides: This might be time consuming for large groups and a lot of
-data curation and knowledge on the group of interest is still necessary. For example,
-<!--General issues with dating techniques:--> choosing correct fossils for calibration
-requires a lot of expertise and knowledge on the group. An incorrect use of fossils
-can generate severe bias in dating results [@Sauquet2012c].
-<!--Data curation is necessary at some point (is this always true?)-->
-Hence, data curation is still an important part of any biological study. The research
-community considers it as an important or even crucial step before data analysis.
-Hence, automated processes for large data analysis are frequently received with skepticism.
-
-`datelife` palliates this by only using information available from already published
-studies, which are ideally constructed using robust information, such as sequence
-data and thoughtfully curated fossil calibrations.
-<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
-<!--Importance of datelife:-->
-
-<!--INTRODUCTION: 5th paragraph: Goal, objectives and predictions-->
-Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
-inspired the development of a prototype `datelife` service over a series
-of phylotastic hackathons [@Stoltzfus2013] at the National Evolutionary Synthesis Center.
-In this paper we present the first formal description of `datelife`, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
-`datelife` is the main service for scaling phylogenetic trees in Phylotastic! system
-[@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
-It can be used through an R package <link to documentation>, a web interface
-(<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
+A platform for efficient reuse of expert, published data on time of lineage divergence should have an open source database in a format suitable for scientific reuse, and straightforward means of accessing, comparing and summarizing the source data as needed by the user.
+A prototype service striving for this characteristics was initially developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
+In here we present the first formal description of the `datelife` service.
+Constituted by and R package and web site, it features new methods to summarize chronograms, an improved system for chronogram database maintenance, new functions to visualize source data in the web interface (<http://www.datelife.org/query/>), as well as tools for comparison of source and summary trees.
+R packages for benchmarking of functionalities and exemplifying services were also developped.
 
@@
 <!--To perform the former tasks:-->
-`datelife`'s chronogram database is currently built from Open Tree of Life (OToL)'s
-[@Hinchliff2015] tree repository. Among currently existing repositories (e.g., TreeBase,
+`datelife`'s chronogram database is currently built from OToL's
+tree repository. Among currently existing repositories (i.e., TreeBASE,
 Dryad), OToL's metadata rich tree store is the only one meeting the requirements
-for proper/accurate automatized handling of trees.
-Input taxon names accepted by `datelife` are binomial species names or clades.
+for automatized handling of source trees.
+TreeBASE trees are stored without branch lengths.
+Dryad metadata does not allow differentiating between different types of data sets, and trees would need to be manually curated anyways.
+`datelife` currently accepts scientific names, from named clades to binomial specifics.
 <!--common names (implemented yet?)-->
 Taxon searches are performed at the species level, so when input names correspond
-to higher clades,
-`datelife` pulls all accepted species names within the
+to named clades, `datelife` pulls all accepted species names within the
 clade from OToL's reference taxonomy to perform the search.
-Currently, searches at the infrapsecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are treated
-as species.
-`datelife` also processes input names with the taxon name resolution service (TNRS),
+Currently, searches at the infraspecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
+`datelife` also processes input names with the taxon name resolution service [TNRS; @Boyle2013],
 <!--which "corrects misspelled names and authorities, standardizes variant spellings,
 and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
-which corrects potentially misspelled names and typos, and standardizes variation
-in spelling and synonyms <!--to a taxonomic reference of users choosing, default
-to otol taxonomy -->[@Boyle2013], increasing the probability to correctly find the
+which corrects potentially misspelled names and typos, and standardizes spelling
+variations and synonyms <!--to a taxonomic reference of users choosing, default
+to otol taxonomy -->, increasing the probability to correctly find the
 queried taxa in `datelife`'s chronogram database.
@@
 (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch
-lengths and the hierarchical clustering method<!-- of ttol-->. <!--Also, I discovered the median gives non ultrametric trees... but I think it is fixed now, check it!).-->
+lengths <!-- and the hierarchical clustering method of ttol-->.
 Tree dating and simulation options are performed with various algorithms:
@@
 from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
-To apply calibrations to a tree, th econgruification algorithm described in [@Eastman2013]
-is used to find shared nodes between trees (congruent nodes).
+To apply calibrations to a tree, the congruification algorithm described by @Eastman2013
+is implemented to find shared nodes between trees (congruent nodes).
 
@@
 \ref{fig:runtime1}).
-<!--
-Speed with different number of input lineages and types of analysis:
-Results show that searching time increases linearly with number of input names and
-number of chronograms in database.
-Summarizing `datelife` results processing times: sdm speed, clustering methods speed
-
-generating new trees: bladj speed
-Adding dates processing time: reflects more mrbayes speed than anything
-
-get_bold_otol_tree running time: not really useful since that is an external service
--->
-<!-- test improvement in synthetic tree after using bold sequences. Is it really better resolved?-->
-
-<!--Speed of web interface and of r package (in computers with different capacities?)-->
+<!--Speed of other datelife workflows? NO, bc it refkects speed from other software mainly. But I will put a sentence about it...
+Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
+generating new trees: shows bladj speed;
+Adding dates processing time: reflects more mrbayes speed than anything;
+get_bold_otol_tree running time: reflects bold downloading data speed.-->
+<!--Speed with different number of chronograms in database? NO-->
+<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
 `datelife`'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat [@RCoreTeam2018]. These tests were run both
-locally –using the devtools package [@RCoreTeam2018]– and on a public server –via
-GitHub– using the continuous integration tool Travis CI (<https://travis-ci.org>). At
+implemented with the R package testthat [@RCoreTeam2018] that were run locally
+--using the devtools package [@RCoreTeam2018], and on a public server --via
+GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
 present, unit tests cover more than 50% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
@@
 think of a test to compare trees, topology- and date-wise
+
+test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
+Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
 -->
 In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain all data on lineage divergence available from `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp.
+We performed a higher-taxon search to obtain all data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp.
 according to the Open Tree of Life taxonomy).
@@
 There is currently wide interest in determining this.
-However, it is certain that these chronograms are useful for...
+However, we know that these chronograms are useful for...
 
@@
 -->
+
+\begin{center}
+\textsc{Deficiencies, Limitations and Prospects}
+\end{center}
+
+The main goal of `datelife` is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It depends
 
@@
 
-Funding was provided by NSF grant 1458603
+Funding was provided by US National Science Foundation (NSF) grant 1458603
 
@@
 
-We thank colleagues (students and postdocs) at the O'Meara Lab at the University
+We thank colleagues from the O'Meara Lab at the University
 of Tennesse Knoxville for suggestions, discussions and software testing.
@@
 the scientists who curate them for deposition in OpenTree.
-The US National Science Foundation (NSF) for funding nearly all the above, in addition
+The NSF for funding nearly all the above, in addition
 to the ABI grant that funded this project itself.

### 2019-05-29 Datelife_paper1.tex
intro

@@
 
-\textbf{Keywords:} Tree; Phylogeny; Scaling; Open; Ages; Congruify; Supertree;
+\textbf{Keywords:} Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
 
@@
 
-Time of lineage divergence
-constitutes in many ways a fundamental piece of information for evolutionary
-understanding.
-Coupled to species number and distribution, it allows studying the tempo and mode of speciation
-and extinction),
-central for understanding of how biodiversity
-patterns are shaped across space, time and clades (Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-In the past two decades, the possibility to obtain good quality DNA sequences
-coupled to methodological developments in phylogenetic and dating inference, allowed
-the application of molecular dating methods on a very large amount and diversity
-of organisms, greatly increasing the quantity and quality of data on taxon
-ages across the tree of life.
-To date, there is a large amount of
-data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
-and Open Tree of Life (OToL).
-OToL alone hosts now more than 200 chronograms.
-Methods to include living and fossil lineages are in continued development and increased
-usage by the community,
-which coupled to better sharing data practices, are greatly contributing to the
-accumulation in number and type of available data on taxon ages.
+Time of lineage divergence constitutes a fundamental piece of information for evolutionary
+understanding in many areas of research, from developmental to conservation biology (Felsenstein \protect\hyperlink{ref-Felsenstein1985a}{1985}; Webb \protect\hyperlink{ref-Webb2000}{2000}), from historical biogeography to species diversification studies (Posadas et al.
+\protect\hyperlink{ref-posadas2006historical}{2006}; Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
+The primary information needed for time of lineage divergence estimation comes from the fossil record.
+Coupled to molecular phylogenies, molecular dating methods are used to reconstruct the time of divergence of extant lineages.
 
-The TimeTree project (Hedges et al. \protect\hyperlink{ref-Hedges2006}{2006}, \protect\hyperlink{ref-Hedges2015}{2015}; Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) has aggregated chronograms
-from 3,163 studies, encompassing 97,085 species (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}), and continues to grow.
-However, even in this gold standard resource, the included taxa only encompass between
-0.097 and 3.236\% of total species diversity (following taxonomic expert opinion
-on the global, extant species numbers, which ranges from 3 to 100 million species
-{[}Mayr2010; Moran2011{]}). One advantage of TimeTree is that it includes taxa from
-across the tree of life, versus more specialized chronogram repositories focusing on plants
-(Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}), birds (Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012}), and fish . Users can choose
-between a web interface or a mobile app to receive information on divergence times
-for the evolutionary history of a lineage, pairs of taxa, all lineages within a
-taxon, or a list of taxa. As a science communication tool, TimeTree project is
-very powerful: it has a friendly graphical interface, with informative and colorful
-outputs, that allows the general public to satisfy curiosity regarding a particular
-organism of interest or group of them. It is of limited utility for scientific
-studies, however. The thousands of trees that have been entered are unavailable
-for examination or reuse; according to the creators (see TimeTree web FAQ), methods
-for allowing data downloading have been under discussion for the past several years
-yet the primary data remain closed. Moreover, there is no Application Programming
-Interface (API) allowing programmatic access to any data, greatly impairing the
-possibility of large-scale, automated data-mining, which is not allowed under TimeTree
-website's terms of use. The nearly hundred thousand taxon summary chronogram generated
-from TimeTree resources is not available with its publication (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) or the
-TimeTree website, though the still substantial chronogram from a previous publication
-(Hedges et al. \protect\hyperlink{ref-Hedges2015}{2015}) was made available at OToL.
+Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al.
+\protect\hyperlink{ref-Kumar2017}{2017}).
+Still, generating a chronogram is not easy unless you have specialized training.
+That's why there has been an urge for reuse of the vast amount and still accuulating already available information for the advantage of researchers not specialized in this area.
 
-Despite its great importance, analytical tools to summarize available information
-on taxon ages for the scientific community are still lacking.
-We identified several aspects that might have so far delayed the exploitation of
-existing data. First, original chronograms available publicly are scattered across repositories
-and journals supplementary data, usually with different formats too.
-Second, lineage names due to taxonomic idiosincracy can be different among studies
-and manual curation of that is usually necessary.
-Third, data curation
-Recent advances on this area (e.g., supersmart) aim to:
-Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
-Generate new dates using all available DNA sequence information;
-Perform one global analysis using all available information;
-Problems or downsides: This might be time consuming for large groups and a lot of
-data curation and knowledge on the group of interest is still necessary. For example,
-choosing correct fossils for calibration
-requires a lot of expertise and knowledge on the group. An incorrect use of fossils
-can generate severe bias in dating results (Sauquet et al. \protect\hyperlink{ref-Sauquet2012c}{2012}).
-Hence, data curation is still an important part of any biological study. The research
-community considers it as an important or even crucial step before data analysis.
-Hence, automated processes for large data analysis are frequently received with skepticism.
+Wide interest from the scientific community to make chronogram information available for consultation and reuse has spurred the creation of public platforms with various goals.
+TreeBASE (Morell \protect\hyperlink{ref-morell1996roots}{1996}; Piel et al.
+\protect\hyperlink{ref-Piel2002}{2002}), Phylomatic (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al.
+\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms are dedicated to storing and making available published trees and chronograms for easy scientific reuse.
+For chronogram reuse for a wide range of living organisms, only OToL is generally suitable right now.
+Treebase does not store branch length information and dryad does not distinguish between data sets, so the time of lineage div information cannot be accessed.
+Phylomatic does contain chronograms but is restricted to plants and mammals, and if it tries to add more data it gives many polytomies back.
 
-\texttt{datelife} palliates this by only using information available from already published
-studies, which are ideally constructed using robust information, such as sequence
-data and thoughtfully curated fossil calibrations.
+OToL also has the primary goal of synthesizing stored trees into a single tree of life, to show phylogenetic relationships among known extant species and subspecies.
+All or parts of OToL's synthetic tree can be reused for any purpose; however, it currently does not include information on time of divergence, so the platform does not summarize source chronogram data.
+The Timetree of Life platform main goal is the synthesis of a single chronogram of life (Hedges et al.
+\protect\hyperlink{ref-Hedges2006}{2006}).
+However, both the latest version of their synthetic chronogram (Kumar et al.
+\protect\hyperlink{ref-Kumar2017}{2017}) and the thousands of chronograms compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
 
-Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
-inspired the development of a prototype \texttt{datelife} service over a series
-of phylotastic hackathons (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}) at the National Evolutionary Synthesis Center.
-In this paper we present the first formal description of \texttt{datelife}, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
-\texttt{datelife} is the main service for scaling phylogenetic trees in Phylotastic! system
-(Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013})
-It can be used through its R package , and a web interface
-(\url{http://www.datelife.org/query/}).
+Other platforms such as SuperSmart (Antonelli et al.
+\protect\hyperlink{ref-antonelli2017supersmart}{2017}) and phylogenerator (Pearse and Purvis \protect\hyperlink{ref-pearse2013phylogenerator}{2013}) are focused in \emph{de novo} chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
+However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
+Moreover, they do not provide information from available chronograms.
+
+A platform for efficient reuse of expert, published data on time of lineage divergence should have an open source database in a format suitable for scientific reuse, and straightforward means of accessing, comparing and summarizing the source data as needed by the user.
+A prototype service striving for this characteristics was initially developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
+\protect\hyperlink{ref-Stoltzfus2013}{2013}).
+In here we present the first formal description of the \texttt{datelife} service.
+Constituted by and R package and web site, it features new methods to summarize chronograms, an improved system for chronogram database maintenance, new functions to visualize source data in the web interface (\url{http://www.datelife.org/query/}), as well as tools for comparison of source and summary trees.
+R packages for benchmarking of functionalities and exemplifying services were also developped.
 
@@
 
-\texttt{datelife}'s chronogram database is currently built from Open Tree of Life (OToL)'s
-(Hinchliff et al. \protect\hyperlink{ref-Hinchliff2015}{2015}) tree repository. Among currently existing repositories (e.g., TreeBase,
+\texttt{datelife}'s chronogram database is currently built from OToL's
+tree repository. Among currently existing repositories (i.e., TreeBASE,
 Dryad), OToL's metadata rich tree store is the only one meeting the requirements
-for proper/accurate automatized handling of trees.
-Input taxon names accepted by \texttt{datelife} are binomial species names or clades.
+for automatized handling of source trees.
+TreeBASE trees are stored without branch lengths.
+Dryad metadata does not allow differentiating between different types of data sets, and trees would need to be manually curated anyways.
+\texttt{datelife} currently accepts scientific names, from named clades to binomial specifics.
 Taxon searches are performed at the species level, so when input names correspond
-to higher clades,
-\texttt{datelife} pulls all accepted species names within the
+to named clades, \texttt{datelife} pulls all accepted species names within the
 clade from OToL's reference taxonomy to perform the search.
-Currently, searches at the infrapsecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are treated
-as species.
-\texttt{datelife} also processes input names with the taxon name resolution service (TNRS),
-which corrects potentially misspelled names and typos, and standardizes variation
-in spelling and synonyms (Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}), increasing the probability to correctly find the
+Currently, searches at the infraspecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
+\texttt{datelife} also processes input names with the taxon name resolution service (TNRS; Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}),
+which corrects potentially misspelled names and typos, and standardizes spelling
+variations and synonyms , increasing the probability to correctly find the
 queried taxa in \texttt{datelife}'s chronogram database.
@@
 (SDM) supertree construction approach (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}) or using the median of branch
-lengths and the hierarchical clustering method.
+lengths .
 Tree dating and simulation options are performed with various algorithms:
@@
 from a pure birth model, as implemented by Jetz and collaborators (\protect\hyperlink{ref-Jetz2012}{2012}).
-To apply calibrations to a tree, th econgruification algorithm described in (Eastman et al. \protect\hyperlink{ref-Eastman2013}{2013})
-is used to find shared nodes between trees (congruent nodes).
+To apply calibrations to a tree, the congruification algorithm described by Eastman et al. (\protect\hyperlink{ref-Eastman2013}{2013})
+is implemented to find shared nodes between trees (congruent nodes).
 
@@
 \ref{fig:runtime1}).
-
 \texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}). These tests were run both
-locally --using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018})-- and on a public server --via
-GitHub-- using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
+implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}) that were run locally
+--using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}), and on a public server --via
+GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
 present, unit tests cover more than 50\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
@@
 In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain all data on lineage divergence available from \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
+We performed a higher-taxon search to obtain all data on lineage divergence available in \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
 according to the Open Tree of Life taxonomy).
@@
 There is currently wide interest in determining this.
-However, it is certain that these chronograms are useful for\ldots{}
+However, we know that these chronograms are useful for\ldots{}
 
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
+Antonelli A., Hettling H., Condamine F.L., Vos K., Nilsson R.H., Sanderson M.J., Sauquet H., Scharn R., Silvestro D., Töpel M., Bacon C.D., Oxelman B., Vos R.A.
+2017.
+Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
+Systematic Biology.
+66:153--166.
+
 \leavevmode\hypertarget{ref-Bapst2012a}{}%
@@
 
+\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
+Felsenstein J. 1985. Phylogenies and the Comparative Method. The American Naturalist. 125:1--15.
+
 \leavevmode\hypertarget{ref-gibb2015new}{}%
@@
 
+\leavevmode\hypertarget{ref-morell1996roots}{}%
+Morell V. 1996. The roots of phylogeny. Science. 273:569.
+
 \leavevmode\hypertarget{ref-Morlon2014}{}%
@@
 Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.
+
+\leavevmode\hypertarget{ref-pearse2013phylogenerator}{}%
+Pearse W.D., Purvis A. 2013. PhyloGenerator: An automated phylogeny generation tool for ecologists. Methods in Ecology and Evolution. 4:692--698.
+
+\leavevmode\hypertarget{ref-Piel2002}{}%
+Piel W.H., Donoghue M., Sanderson M.
+2002.
+TreeBASE : A database of phylogenetic information.
+In: Shimura J., Wilson K., Gordon D., editors.
+To the interoperable ``catalog of life'' with partners.
+Tsukuba, Japan: National Institute for Environmental Studies.
+p.
+41--47.
+
+\leavevmode\hypertarget{ref-posadas2006historical}{}%
+Posadas P., Crisci J.V., Katinas L. 2006. Historical biogeography: A review of its basic concepts and critical issues. Journal of Arid Environments. 66:389--403.
 
@@
 
-\leavevmode\hypertarget{ref-Sauquet2012c}{}%
-Sauquet H., Ho S.Y.W., Gandolfo M.
-a, Jordan G.J., Wilf P., Cantrill D.J., Bayly M.J., Bromham L., Brown G.K., Carpenter R.J., Lee D.M., Murphy D.J., Sniderman J.M.K., Udovicic F.
-2012.
-Testing the impact of calibration on molecular divergence times using a fossil-rich group: the case of Nothofagus (Fagales).
-Systematic Biology.
-61:289--313.
-
 \leavevmode\hypertarget{ref-Schenk2016}{}%
@@
 14.
+
+\leavevmode\hypertarget{ref-Webb2000}{}%
+Webb C.O. 2000. Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. The American Naturalist. 156:145--155.
 

### 2019-05-29 Intro.Rmd
intro

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---

<!--INTRODUCTION: 1st paragraph: general statement that the field knows to be true and important-->
Time of lineage divergence <!--alt lineages: named biological entities, taxon ages?-->
constitutes a fundamental piece of information for evolutionary
understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000]. This information can be obtained from fossils and from molecular dating studies. The developments in DNA sequencing and phylogenetic dating inferences have encouraged/spurred the application of molecular dating methods to a very large diversity of organisms, greatly increasing the quantity 
<!-- and improving the quality  (do we know this? we would expect thatm but there is no analysis on that really...)-->
of data on taxon ages across the tree of life.
Along with phylogenetic relationships, it allows testing alternative
evolutionary hypothesis in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000].
It has been shown that only a bit of information on time of divergence can improve inferences of phylogenetic distance @Webb2008 bladj
<!--comparative studies, in general, for which knowing
the time context for all life is crucial.-->
Together with geographical distribution patterns, it is a crucial component of historical biogeography studies [@posadas2006historical].
Coupled to species number, it allows studying the tempo and mode of speciation and extinction,
<!-- crucial/vital/critical/essential -->
central for understanding how biodiversity
patterns are shaped across space, time, and clades [@Morlon2014].
<!-- Efforts to have a whole tree of life have been great and here are some examples. -->
<!--INTRODUCTION: 2nd paragraph: important advances in the field-->
In the past two decades, the possibility to obtain good quality DNA sequences
<!-- from a wide variety of organisms became a reality, which,  -->
coupled to methodological developments in phylogenetic and dating inference, allowed
the application of molecular dating methods to a very large diversity
of organisms, greatly increasing the quantity and quality of<!--organisms with--> data on taxon ages across the tree of life.
<!-- figure showing growth on studies with age data from references table from ttol-->
To date, there is a large amount of
<!-- both fossil and molecular-based  -->
data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
and Open Tree of Life (OToL).
<!-- pbdb ??-->
<!--OToL alone hosts now more than 200 chronograms. Cache updated jan 15 2018
devtools::load_all(pkg = "~/Desktop/datelife")
length(opentree_chronograms$trees) names(opentree_chronograms)
-->
<!--How many trees? How many dated trees?-->
<!-- Methods to include living and fossil lineages are in continued development and increased -->
<!-- usage by the community<!--total evidence and revbayes (fossilized birth-death)-->, -->
<!-- which coupled to better sharing data practices, are greatly contributing to the -->
<!-- accumulation in number and type of available data on taxon ages. -->

<!--INTRODUCTION: 3rd paragraph: problems associated-->
The TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017] has aggregated chronograms
from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow.
However, even in this gold standard resource, the included taxa only encompass between
0.097 and 3.236% of total species diversity (following taxonomic expert opinion
on the global, extant species numbers, which ranges from 3 to 100 million species
[Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from
across the tree of life, versus more specialized chronogram repositories focusing on land plants
[@stevens2001apw] and mammals [@webb2005phylomatic], birds [@Jetz2012], and fish [@chang2019r]. Users can choose
between a web interface or a mobile app to receive information on divergence times
for the evolutionary history of a lineage, pairs of taxa, all lineages within a
taxon, or a list of taxa. As a science communication tool, TimeTree project is
very powerful: it has a friendly graphical interface, with informative and colorful
outputs, that allows the general public to satisfy curiosity regarding a particular
organism of interest or group of them. It is of limited utility for scientific
studies, however. The thousands of trees that have been entered are unavailable
for examination or reuse; according to the creators (see TimeTree web FAQ), methods
for allowing data downloading have been under discussion for the past several years
yet the primary data remain closed. Moreover, there is no Application Programming
Interface (API) allowing programmatic access to any data, greatly impairing the
possibility of large-scale, automated data-mining, which is not allowed under TimeTree
website's terms of use. The nearly hundred thousand taxon summary chronogram generated
from TimeTree resources is not available with its publication [@Kumar2017] or the
TimeTree website, though the still substantial chronogram from a previous publication
[@Hedges2015] was made available at OToL.
<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->

<!--INTRODUCTION: 4th paragraph-->
Despite its great importance, analytical tools to summarize available information
on taxon ages for the scientific community are still lacking.
We identified several aspects that might have so far delayed the exploitation of
existing data. First, original chronograms available publicly are scattered across repositories
<!-- (otol tree store, dryad, treebase,  -->
and journals supplementary data, usually with different formats too.
Second, lineage names due to taxonomic idiosincracy can be different among studies
and manual curation of that is usually necessary.
Third, data curation

Platforms that contribute to knowledge in this areas:

Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.

Phylogenerator [@pearse2013]

Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;

Problems or downsides: This might be time consuming for large groups and a lot of
data curation and knowledge on the group of interest is still necessary. For example,
<!--General issues with dating techniques:--> choosing correct fossils for calibration
requires a lot of expertise and knowledge on the group. An incorrect use of fossils
can generate severe bias in dating results [@Sauquet2012c].
<!--Data curation is necessary at some point (is this always true?)-->
Hence, data curation is still an important part of any biological study. The research
community considers it as an important or even crucial step before data analysis.
Hence, automated processes for large data analysis are frequently received with skepticism.

`datelife` palliates this by only using information available from already published
studies, which are ideally constructed using robust information, such as sequence
data and thoughtfully curated fossil calibrations.
<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
<!--Importance of datelife:-->

<!--INTRODUCTION: 5th paragraph: Goal, objectives and predictions-->
Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
inspired the development of a prototype `datelife` service over a series
of phylotastic hackathons [@Stoltzfus2013] at the National Evolutionary Synthesis Center.
In this paper we present the first formal description of `datelife`, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
`datelife` is the main service for scaling phylogenetic trees in Phylotastic! system
[@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
It can be used through its R package <link to documentation>, and a web interface
(<http://www.datelife.org/query/>)<!-- and an API (still not up, tho)-->.

### 2019-05-29 Intro.tex
intro

--- 2019-05-23 Datelife_paper1.tex
@@
 
-Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE
-
-Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
-
-Authors: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}
-
-Correspondence address:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA}
-\end{enumerate}
-
-Corresponding authors: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
-
-\newpage
-
-\textbf{abstract.-} Here goes the abstract.
-
-\textbf{Keywords:} Tree; Phylogeny; Scaling; Open; Ages; Congruify; Supertree;
-
-\newpage
-
 Time of lineage divergence
-constitutes in many ways a fundamental piece of information for evolutionary
-understanding.
-Coupled to species number and distribution, it allows studying the tempo and mode of speciation
-and extinction),
-central for understanding of how biodiversity
-patterns are shaped across space, time and clades (Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
+constitutes a fundamental piece of information for evolutionary
+understanding in many areas of research, from developmental to conservation biology ({\textbf{???}}; Webb \protect\hyperlink{ref-Webb2000}{2000}).
+This information can be obtained from fossils and from molecular dating studies.
+The developments in DNA sequencing and phylogenetic dating inferences have encouraged/spurred the application of molecular dating methods to a very large diversity of organisms, greatly increasing the quantity
+of data on taxon ages across the tree of life.
+Along with phylogenetic relationships, it allows testing alternative
+evolutionary hypothesis in many areas of research, from developmental to conservation biology ({\textbf{???}}; Webb \protect\hyperlink{ref-Webb2000}{2000}).
+It has been shown that only a bit of information on time of divergence can improve inferences of phylogenetic distance Webb et al. (\protect\hyperlink{ref-Webb2008}{2008}) bladj
+Together with geographical distribution patterns, it is a crucial component of historical biogeography studies (Posadas et al. \protect\hyperlink{ref-posadas2006historical}{2006}).
+Coupled to species number, it allows studying the tempo and mode of speciation and extinction,
+central for understanding how biodiversity
+patterns are shaped across space, time, and clades (Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
 In the past two decades, the possibility to obtain good quality DNA sequences
 coupled to methodological developments in phylogenetic and dating inference, allowed
-the application of molecular dating methods on a very large amount and diversity
-of organisms, greatly increasing the quantity and quality of data on taxon
-ages across the tree of life.
+the application of molecular dating methods to a very large diversity
+of organisms, greatly increasing the quantity and quality of data on taxon ages across the tree of life.
 To date, there is a large amount of
@@
 and Open Tree of Life (OToL).
-OToL alone hosts now more than 200 chronograms.
-Methods to include living and fossil lineages are in continued development and increased
-usage by the community,
-which coupled to better sharing data practices, are greatly contributing to the
-accumulation in number and type of available data on taxon ages.
+, --\textgreater{}
 
@@
 {[}Mayr2010; Moran2011{]}). One advantage of TimeTree is that it includes taxa from
-across the tree of life, versus more specialized chronogram repositories focusing on plants
-(Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}), birds (Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012}), and fish . Users can choose
+across the tree of life, versus more specialized chronogram repositories focusing on land plants
+(Stevens and Davis \protect\hyperlink{ref-stevens2001apw}{2001}) and mammals (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}), birds (Jetz et al.
+\protect\hyperlink{ref-Jetz2012}{2012}), and fish (Chang et al.
+\protect\hyperlink{ref-chang2019r}{2019}).
+Users can choose
 between a web interface or a mobile app to receive information on divergence times
@@
 Third, data curation
-Recent advances on this area (e.g., supersmart) aim to:
+
+Platforms that contribute to knowledge in this areas:
+
 Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
+
+Phylogenerator ({\textbf{???}})
+
 Generate new dates using all available DNA sequence information;
 Perform one global analysis using all available information;
+
 Problems or downsides: This might be time consuming for large groups and a lot of
@@
 
-\begin{center}
-\textsc{Description}
-\end{center}
-
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
-\tightlist
-\item
-  A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item
-  \texttt{datelife} then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarization algorithms.
-\item
-  The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
-\item
-  At this point, users can choose to use all or some source data as calibration points to date a tree of their own making or choosing. 
-\item
-  Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
-\item
-  Finally, users can easily view results graphically as well as construct their own graphs using inbuilt \texttt{datelife} graphic generators.
-\end{enumerate}
-
-\texttt{datelife}'s chronogram database is currently built from Open Tree of Life (OToL)'s
-(Hinchliff et al. \protect\hyperlink{ref-Hinchliff2015}{2015}) tree repository. Among currently existing repositories (e.g., TreeBase,
-Dryad), OToL's metadata rich tree store is the only one meeting the requirements
-for proper/accurate automatized handling of trees.
-Input taxon names accepted by \texttt{datelife} are binomial species names or clades.
-Taxon searches are performed at the species level, so when input names correspond
-to higher clades,
-\texttt{datelife} pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
-Currently, searches at the infrapsecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are treated
-as species.
-\texttt{datelife} also processes input names with the taxon name resolution service (TNRS),
-which corrects potentially misspelled names and typos, and standardizes variation
-in spelling and synonyms (Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}), increasing the probability to correctly find the
-queried taxa in \texttt{datelife}'s chronogram database.
-
-Source chronogram summary tree can be assembled using the Super Distance Matrix
-(SDM) supertree construction approach (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}) or using the median of branch
-lengths and the hierarchical clustering method.
-Tree dating and simulation options are performed with various algorithms:
-Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
-nodes evenly, which minimizes age variance in the chronogram (Webb et al. \protect\hyperlink{ref-Webb2008}{2008}).
-PATHd8 is a non-clock, rate-smoothing method (Britton et al. \protect\hyperlink{ref-Britton2007}{2007}) to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Smith and O'Meara \protect\hyperlink{ref-Smith2012}{2012}).
-MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) can be used when adding taxa at
-random, following a reference taxonomy or a topological constraint. It draws ages
-from a pure birth model, as implemented by Jetz and collaborators (\protect\hyperlink{ref-Jetz2012}{2012}).
-To apply calibrations to a tree, th econgruification algorithm described in (Eastman et al. \protect\hyperlink{ref-Eastman2013}{2013})
-is used to find shared nodes between trees (congruent nodes).
-
-To gather, process, and present information, \texttt{datelife} builds up from functions
-available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
-geiger (Harmon et al.
-\protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}),
-taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl \protect\hyperlink{ref-Heibl2008}{2008}), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018})
-and rphylotastic (O'Meara et al. \protect\hyperlink{ref-Omeara2019}{2019}).
-
-\begin{center}
-\textsc{Benchmark}
-\end{center}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS service is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes.
-A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}). These tests were run both
-locally --using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018})-- and on a public server --via
-GitHub-- using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 50\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\begin{center}
-\textsc{Example}
-\end{center}
-
-In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain all data on lineage divergence available from \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
-according to the Open Tree of Life taxonomy).
-There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig.
-\ref{fig:schronograms1}).
-Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
-As explained in the \texttt{Description}, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are coherent with source ages.
-It is not certain if these chronograms can be used to perform downstream evolutionary analyses.
-There is currently wide interest in determining this.
-However, it is certain that these chronograms are useful for\ldots{}
-
-Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs.
-\ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms.
-In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
-Maximum tree ages were only approximately recovered in one case (study 2; Fig.
-\ref{fig:cvbladj}).
-Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
-three cases (studies 3, 5, and 9; Fig.
-\ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and details can be consulted in \url{https://github.com/LunaSare/datelife_examples}.
-
-\begin{center}
-\textsc{Conclusions}
-\end{center}
-
-Taxon ages are key to many areas of evolutionary studies: trait evolution, species
-diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to develop the necessary knowledge and data curation skills
-to produce new chronograms. Knowledge on taxon ages is also important for non-biological
-studies and the non-academic community.
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree
-of Life, for a large quantity and diversity of organisms. However, this information
-remains difficult to synthesize for many biologists and the non-academic community.
-
-Here, we have shown that \texttt{datelife} allows
-an easy and fast obtention of all publicly available information on taxon ages,
-which can be used to generate new data.
-This information can be used to account for the effect of phylogenetic signal in
-studies of trait evolution; to explore potential speciation and extinction dynamics
-of interest within a clade; to obtain a time frame of biogeographical events; for
-science communication and outreach, amongst others.
-Compared to similar platforms such as time tree of life
-and supermart, it offers several advantages.
-It is fast;
-source data is completely open;
-it requires no expert biological knowledge from users for any of its functionalities;
-it allows exploration of alternative taxonomic and phylogenetic schemes;
-it allows rapid exploration of the effect of alternative divergence time hypothesis;
-it allows rapid synthesis in a number of different formats;
-it facilitates reproducibility of analyses;
-
-Improvements, short and long-term:
-* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
-(Schenk \protect\hyperlink{ref-Schenk2016}{2016}).
-* bayesian congruification
-* topological congruification
-
-Problems and caveats:
-Not many databases, only OToL
-Why TreeBase is not very useful for us? Be precise.
-Are these chronograms reliable to study evolutionary patterns, such as species diversification?
-\texttt{datelife} can be seen as an open resource to know the current state of knowledge on lineage divergence times.
-Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
-If all, la facilidad para obtener hipotesis de tiempo de divergencia nos ayudará a evaluar la capacidad de los cronogramas para estudiar otros fenomenos evolutivos.
-Por ahora, no podemos aseverar que estos cronogramas puedan usarse para todo tipo de analisis.
-
-\begin{center}
-\textsc{Availability}
-\end{center}
-
-\texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\begin{center}
-\textsc{Supplementary Material}
-\end{center}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in GitHub repositories at \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
-
-\begin{center}
-\textsc{Funding}
-\end{center}
-
-Funding was provided by NSF grant 1458603
-
-NESCent
-
-Open Tree of Life
-
-University of Tennessee, Knoxville
-
-\begin{center}
-\textsc{Acknowledgements}
-\end{center}
-
-We thank colleagues (students and postdocs) at the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in OpenTree.
-The US National Science Foundation (NSF) for funding nearly all the above, in addition
-to the ABI grant that funded this project itself.
-
-\newpage
-
-\begin{center}
-\textsc{References}
-\end{center}
-
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst D.W. 2012. Paleotree: An R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution. 3:803--807.
-
 \leavevmode\hypertarget{ref-barker2012going}{}%
@@
 
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle B., Hopkins N., Lu Z., Raygoza Garay J.A., Mozzherin D., Rees T., Matasci N., Narro M.L., Piel W.H., Mckay S.J., Lowry S., Freeland C., Peet R.K., Enquist B.J.
-2013.
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton T., Anderson C.L., Jacquet D., Lundqvist S., Bremer K. 2007. Estimating Divergence Times in Large Phylogenetic Trees. Systematic Biology. 56:741--752.
-
 \leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
@@
 
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain S. 2018. bold: Interface to Bold Systems API..
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain S.A., Szöcs E. 2013. taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}. F1000Research. 2:1--29.
+\leavevmode\hypertarget{ref-chang2019r}{}%
+Chang J., Rabosky D.L., Smith S.A., Alfaro M.E. 2019. An r package and online resource for macroevolutionary studies using the ray-finned fish tree of life. Methods in Ecology and Evolution. 0:1--7.
 
@@
 Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo A., Berry V., Douzery E.J., Gascuel O. 2006. SDM: A fast distance-based approach for (super)tree building in phylogenomics. Systematic Biology. 55:740--755.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman J.M., Harmon L.J., Tank D.C. 2013. Congruification: Support for time scaling large phylogenetic trees. Methods in Ecology and Evolution. 4:688--691.
 
@@
 
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon L., Weir J., Brock C., Glor R., Challenger W. 2008. GEIGER: investigating evolutionary radiations. Bioinformatics. 24:129--131.
-
 \leavevmode\hypertarget{ref-Hedges2006}{}%
@@
 
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl C. 2008. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..
-
-\leavevmode\hypertarget{ref-Hinchliff2015}{}%
-Hinchliff C.E., Smith S.A., Allman J.F., Burleigh J.G., Chaudhary R., Coghill L.M., Crandall K.A., Deng J., Drew B.T., Gazis R., Gude K., Hibbett D.S., Katz L.A., Laughinghouse H.D., McTavish E.J., Midford P.E., Owen C.L., Ree R.H., Rees J.A., Soltis D.E., Williams T., Cranston K.A.
-2015.
-Synthesis of phylogeny and taxonomy into a comprehensive tree of life.
-Proceedings of the National Academy of Sciences.
-112:12764--12769.
-
 \leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
 Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck J.P., Ronquist F. 2001. MRBAYES: Bayesian inference of phylogenetic trees. Bioinformatics. 17:754--755.
 
@@
 
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau F., Brown J.W., Winter D.J. 2016. rotl: an R package to interact with the Open Tree of Life data. Methods in Ecology and Evolution. 7:1476--1481.
-
 \leavevmode\hypertarget{ref-Morlon2014}{}%
@@
 
-\leavevmode\hypertarget{ref-Omeara2019}{}%
-O'Meara B., Md Tayeen A.S., Sanchez Reyes L.L. 2019. Rphylotastic: An r interface to 'phylotastic' web services..
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms J., Chamberlain S. 2018. Phylocomr: Interface to 'phylocom'..
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.
+\leavevmode\hypertarget{ref-posadas2006historical}{}%
+Posadas P., Crisci J.V., Katinas L. 2006. Historical biogeography: A review of its basic concepts and critical issues. Journal of Arid Environments. 66:389--403.
 
@@
 509:222.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. 2018. R: a language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell L.J. 2012. Phytools: An r package for phylogenetic comparative biology (and other things). Methods in Ecology and Evolution. 3:217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist F., Huelsenbeck J.P. 2003. MrBayes 3: Bayesian phylogenetic inference under mixed models. Bioinformatics. 19:1572--1574.
 
@@
 
-\leavevmode\hypertarget{ref-Schenk2016}{}%
-Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
-
-\leavevmode\hypertarget{ref-Smith2012}{}%
-Smith S.A., O'Meara B.C. 2012. TreePL: Divergence time estimation using penalized likelihood for large phylogenies. Bioinformatics. 28:2689--2690.
+\leavevmode\hypertarget{ref-stevens2001apw}{}%
+Stevens P.F., Davis H. 2001. Angiosperm phylogeny website..
 

### 2019-06-01 Datelife_paper1.Rmd
discussion

@@
 
-**abstract.-** Here goes the abstract.
+**abstract.-**
+The combination of new analytical techniques, availability of more fossil and molecular
+data, and better practices in data sharing has resulted in a steady accumulation
+of chronograms in public and open databases such as Dryad, TreeBASE and Open Tree
+of Life, for a large quantity and diversity of organisms. However, this information
+remains difficult to synthesize for many biologists and the non-academic community.
 
@@
 The primary information needed for time of lineage divergence estimation comes from the fossil record.
-Coupled to molecular phylogenies, molecular dating methods are used to reconstruct the time of divergence of extant lineages.
+Coupled to molecular phylogenies (and more recenyly to morphological data), the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
+Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
+Still, generating a chronogram is not an easy task unless you have specialized training.
+That's why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
 
-Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
-Still, generating a chronogram is not easy unless you have specialized training.
-That's why there has been an urge for reuse of the vast amount and still accuulating already available information for the advantage of researchers not specialized in this area.
+Wide interest from the scientific community to make information from phylogenies in general and chronograms  in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
+TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] platforms store and make available published phylogenies and chronograms for easy scientific reuse.
+All of them can be queried using automatised web procedures, which permit personalised, large scale queries that are also very fast.
+OToL stores trees with branch length information from a wide range of living organisms, implementing a very good metadata structure.
+Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data, and their metadata does not allow automatic distinction between types of trees, impairing the automatised access to time of lineage divergence information from the latter two databases
 
-Wide interest from the scientific community to make chronogram information available for consultation and reuse has spurred the creation of public platforms with various goals.
-TreeBASE [@morell1996roots; @Piel2002], Phylomatic [@webb2005phylomatic], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] platforms are dedicated to storing and making available published trees and chronograms for easy scientific reuse.
-For chronogram reuse for a wide range of living organisms, only OToL is generally suitable right now.
-Treebase does not store branch length information and dryad does not distinguish between data sets, so the time of lineage div information cannot be accessed.
-Phylomatic does contain chronograms but is restricted to plants and mammals, and if it tries to add more data it gives many polytomies back.
+Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesised all trees in their repository to expose to the community the phylogenetic relationships among all known lineages in a single tree of all life.
+All or parts of this synthetic tree can be reused for any purpose; however, it currently only shows tree topology, meaning that it does not expose branch length data of any type, including time of divergence data from their stored chronograms.
+Phylomatic [@webb2005phylomatic]
+<!-- In a way, phylomatic synthesises trees based on a query. -->
+is a platform that synthesizes a tree containing taxa of interest from a user supplied set of trees or from its own database.
+It has the advantage that it allows grafting taxa of interest absent from their database, but it is limited to mostly seed plants and mammals.
+If your query does not match its records, it gives many polytomies back.
+The Timetree of Life project focuses on the synthesis of a single chronogram of life [@Hedges2006].
+However, the thousands of chronograms compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
+The latest version of their synthetic chronogram [@Kumar2017] can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
 
-OToL also has the primary goal of synthesizing stored trees into a single tree of life, to show phylogenetic relationships among known extant species and subspecies.
-All or parts of OToL's synthetic tree can be reused for any purpose; however, it currently does not include information on time of divergence, so the platform does not summarize source chronogram data.
-The Timetree of Life platform main goal is the synthesis of a single chronogram of life [@Hedges2006].
-However, both the latest version of their synthetic chronogram [@Kumar2017] and the thousands of chronograms compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
-
-Other platforms such as SuperSmart [@antonelli2017supersmart] and phylogenerator [@pearse2013phylogenerator] are focused in _de novo_ chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
+Other platforms such as SuperSmart [@antonelli2017supersmart] and phylogenerator [@pearse2013phylogenerator] are focused in automatised _de novo_ chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
 However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Moreover, they do not provide information from available chronograms.
+Also, these tools do not provide information from already existent expert chronograms.
 
@@
  -->
-A platform for efficient reuse of expert, published data on time of lineage divergence should have an open source database in a format suitable for scientific reuse, and straightforward means of accessing, comparing and summarizing the source data as needed by the user.
-A prototype service striving for this characteristics was initially developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
-In here we present the first formal description of the `datelife` service.
-Constituted by and R package and web site, it features new methods to summarize chronograms, an improved system for chronogram database maintenance, new functions to visualize source data in the web interface (<http://www.datelife.org/query/>), as well as tools for comparison of source and summary trees.
-R packages for benchmarking of functionalities and exemplifying services were also developped.
+A tool for efficient reuse of expert, published data on time of lineage divergence should have an open source database storing chronograms in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram data as needed by the user.
+A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
+In here we present the formal description and implementation of the `datelife` service, constituted by an R package and a web site (<http://www.datelife.org/>).
+There is still much room for improvement, and flaws and limitations are addressed below.
+We strived for the current implementation of `datelife` to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+<!-- as well as tools for comparison of source and summary trees.  -->
+R packages for benchmarking of functionalities and demonstrating services were also developped.
 
@@
 
-<!--To perform the former tasks:-->
-`datelife`'s chronogram database is currently built from OToL's
-tree repository. Among currently existing repositories (i.e., TreeBASE,
-Dryad), OToL's metadata rich tree store is the only one meeting the requirements
-for automatized handling of source trees.
-TreeBASE trees are stored without branch lengths.
-Dryad metadata does not allow differentiating between different types of data sets, and trees would need to be manually curated anyways.
-`datelife` currently accepts scientific names, from named clades to binomial specifics.
-<!--common names (implemented yet?)-->
-Taxon searches are performed at the species level, so when input names correspond
+<!--To perform all tasks:-->
+To gather, process, and present information, `datelife` builds up from functions
+available in several R packages including rotl [@Michonneau2016], ape [@Paradis2004],
+geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], phylocomr [@Ooms2018] and rphylotastic [@Omeara2019].
+
+`datelife`'s chronogram database currently contains only chronograms from OToL's
+tree repository. Compared to other existing tree repositories, such as TreeBASE and
+Dryad, OToL's metadata rich tree store is the only one that supports automatised proper search, identification and handling of chronograms.
+The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence.
+The Dryad repository not only stores trees and chronograms, but all sorts of biological data, and their metadata does not allow differentiating between trees with branch lengths proportional to time and any other branch length measure that is not relevant to summarize time of divergence.
+
+`datelife` currently accepts scientific names only. It can be any named clade or binomial specific.
+<!--common names (implemented in phylotastic)-->
+Chronogram searches are performed at the species level, so when input names correspond
 to named clades, `datelife` pulls all accepted species names within the
@@
 
-Source chronogram summary tree can be assembled using the Super Distance Matrix
+Source chronogram summary trees can be assembled using the Super Distance Matrix
 (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch
@@
 is implemented to find shared nodes between trees (congruent nodes).
-
-To gather, process, and present information, `datelife` builds up from functions
-available in several R packages including rotl [@Michonneau2016], ape [@Paradis2004],
-geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012],
-taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], phylocomr [@Ooms2018]
-and rphylotastic [@Omeara2019].
 
@@
 \begin{center}
-\textsc{Deficiencies, Limitations and Prospects}
+\textsc{Flaws, Limitations and Prospects}
 \end{center}
@@
 The main goal of `datelife` is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It depends
+It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
+However, it has many flaws.
+Some of them can be overcome, some of them might represent limitations.
+
+There is only one public tree repository from where `datelife` currently pulls chronograms to construct its database.
+Other existing tree repositories are not suitable for time of divergence data mining because they either do not store that data or because their metadata does not have enough information to allow automatised chronogram searches.
+`datelife`'s database has a bit more than 200 chronograms up to the time the manuscript was written.
+<!-- , which probably corresponds to less than 1% (`r 200*100/3987`%) of the information used for TimeTree synthesis  
+Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
+Incorporation of the many chronograms that are available in data repositories such as Dryad and also from supplementary material of journals to `datelife`'s database is crucial to improve `datelife`'s services.
+Methods to automatically mine Dryad repository for chronogram data could be designed and implemented. 
+<!-- What can we do to mine journal databases? -->
+<!-- At the same time, the prospect is to keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
+At the same time, we would like to increase awereness and emphasize on the importance of data sharing in databases that are maintained by the expert scientific community, requiring expert input and manual curation, such as OToL's tree store.
+
+We introduce different ways of summarizing this data, but are these summary chronograms reliable to study evolutionary patterns?
+The short answer is that we do not know to what extent the summary trees can be used for reanalyses and in which areas they can be used more reliably than others.
+Surely they might be useful for ecological analyses, but further studies should be conducted.
+
+As implemented now, we currently by default summarize all source data avilable.
+Users can subset source data if they have reasons to favor some or one source chronogram over others.
+Strictly speaking, a good chronogram reflects the real time of divergence accurately and precisely.
+To our knowledge there is no objective way to determine if an expert chronogram is better than other.
+Some criteria that have been put forward are the level of sampling and the numer of calibrations used.
+Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
+`datelife`'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.
+It has been shown with simulations that in analyses performed with bayesian inference methods, divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations [using secondary calibrations in the same way as primary calibrations are used @schenk2016sec].
+But there are different ways to apply secondary calibrations.
+This information should be included as metadata to facilitate subset of source chronograms.
+This will need to be manually curated as is done in OToL.
+
+It would be important to evaluate how different source chronograms affect summary chronograms, in order to determine if any type of bias is introduced by certain types of chronograms in particular, so this chronograms are not included in the summary.
+
+Even if all source chronograms have been generated by excellent standards and practices, the evolutionary context they depict might be very different.
+The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis [@antonelli2017supersmart].
+Summarizing trees implies summarizing evolutionary hypothesis.
+
+However, the comparison of source trees in the example above shows that even chronograms obtained with primary data can be very different.
+There is some evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
+In other areas of research, such as ecological and conservation biology, it has been demonstrated that the presence of data on lineage divergence represents an improvement for testing alternative hypothesis, compared to testing in a total absence of such time data.
+However, we do not know how using wrong or different kinds of age data will affect the outcomes of ecological tests.
+More work has to be performed in this area.
+The comparison of available data has made evident how much discrepancy in time of lineage divergence can be found across expert knowledge.
+This has been shown for angiosperms previously [@magallon2015metacalibrated], but our examples show that this is a widespread phenomenon that might require further attention.
+In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and software used was very advanced.
+However, this is just an assumption.
+Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used) But we should also be able to choose on
+
+Improvements:
+* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
+[@Schenk2016].
+* bayesian congruification
+* topological congruification
 
@@
 
-Taxon ages are key to many areas of evolutionary studies: trait evolution, species
+Time of lineage divergence data are key to many areas of evolutionary studies: trait evolution,
 diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
 especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to develop the necessary knowledge and data curation skills
-to produce new chronograms. Knowledge on taxon ages is also important for non-biological
-studies and the non-academic community.
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree
-of Life, for a large quantity and diversity of organisms. However, this information
-remains difficult to synthesize for many biologists and the non-academic community<!-- or just non biologists-->.
+do not have the time to acquire and develop the necessary knowledge and data curation skills
+to produce chronograms _de novo_. Knowledge on taxon ages is also crucial for
+science communication and education.
 
-<!--Potential applications shown here-->Here, we have shown that `datelife` allows
-an easy and fast obtention of all publicly available information on taxon ages,
-which can be used to generate new data. <!--Potential applications not shown here-->
-This information can be used to account for the effect of phylogenetic signal in
-studies of trait evolution; to explore potential speciation and extinction dynamics
-of interest within a clade; to obtain a time frame of biogeographical events; for
-science communication and outreach, amongst others.
-<!--Why you should use it-->Compared to similar platforms such as time tree of life
-and supermart, it offers several advantages.
-It is fast;
-source data is completely open;
-it requires no expert biological knowledge from users for any of its functionalities;
+<!--Mention again uses shown here and outline potential uses not shown here-->
+`datelife` allows an easy and fast obtention, as well as comparison of publicly available information
+on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the timeframe of different regions of the tree of life, allowing identification of regions that need more work or that have conflicting information.
+It can also summarize public data, or use it to generate new chronograms based on trees provided by the user.
+Both summary and newly generated trees have the potential to be useful to evaluate evolutionary hypothesis in different areas of research.
+How well this trees work for this purpose still needs to be tested.
+`datelife` helps with awereness on the variation of expert data, and the actual effect of the underlying data on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+
+`datelife` is also part of the reproducibility of science movement.
+It strives to facilitate usage of public data sets as well as encouraging the making of information publicly available.
+Acknowledging source information and describing workflows and analyses.
+
 it allows exploration of alternative taxonomic and phylogenetic schemes;
@@
 it allows rapid synthesis in a number of different formats; <!-- we should add graphical options too-->
-it facilitates reproducibility of analyses;
-
-Improvements, short and long-term:
-* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
-[@Schenk2016].
-* bayesian congruification
-* topological congruification
-
-Problems and caveats:
-Not many databases, only OToL
-Why TreeBase is not very useful for us? Be precise.
-Are these chronograms reliable to study evolutionary patterns, such as species diversification?
-`datelife` can be seen as an open resource to know the current state of knowledge on lineage divergence times.
-Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
-If all, la facilidad para obtener hipotesis de tiempo de divergencia nos ayudará a evaluar la capacidad de los cronogramas para estudiar otros fenomenos evolutivos.
-Por ahora, no podemos aseverar que estos cronogramas puedan usarse para todo tipo de analisis.
 
@@
 <http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
-`datelife`'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+`datelife`'s website is maintained with RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
 `datelife`'s R package stable version is available
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in GitHub repositories at <https://github.com/LunaSare/datelife_paper1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>, respectively.
+Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities can be found in GitHub repositories at <https://github.com/LunaSare/datelife_paper1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>, respectively.
 <!-- Vignettes

### 2019-06-01 Datelife_paper1.tex
discussion

@@
 
-\textbf{abstract.-} Here goes the abstract.
+\textbf{abstract.-}
+The combination of new analytical techniques, availability of more fossil and molecular
+data, and better practices in data sharing has resulted in a steady accumulation
+of chronograms in public and open databases such as Dryad, TreeBASE and Open Tree
+of Life, for a large quantity and diversity of organisms. However, this information
+remains difficult to synthesize for many biologists and the non-academic community.
 
@@
 The primary information needed for time of lineage divergence estimation comes from the fossil record.
-Coupled to molecular phylogenies, molecular dating methods are used to reconstruct the time of divergence of extant lineages.
+Coupled to molecular phylogenies, the time of divergence of extant lineages is reconstructed with molecular dating methods.
 
@@
 \protect\hyperlink{ref-Kumar2017}{2017}).
-Still, generating a chronogram is not easy unless you have specialized training.
-That's why there has been an urge for reuse of the vast amount and still accuulating already available information for the advantage of researchers not specialized in this area.
+Still, generating a chronogram is not an easy task unless you have specialized training.
+That's why there has been an urge for promoting and facilitating reuse of the vast amount and still accumulating information that is already available, for the advantage of researchers not specialized in this area (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al.
+\protect\hyperlink{ref-Stoltzfus2013}{2013}).
 
-Wide interest from the scientific community to make chronogram information available for consultation and reuse has spurred the creation of public platforms with various goals.
+Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
 TreeBASE (Morell \protect\hyperlink{ref-morell1996roots}{1996}; Piel et al.
-\protect\hyperlink{ref-Piel2002}{2002}), Phylomatic (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al.
-\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms are dedicated to storing and making available published trees and chronograms for easy scientific reuse.
-For chronogram reuse for a wide range of living organisms, only OToL is generally suitable right now.
-Treebase does not store branch length information and dryad does not distinguish between data sets, so the time of lineage div information cannot be accessed.
-Phylomatic does contain chronograms but is restricted to plants and mammals, and if it tries to add more data it gives many polytomies back.
+\protect\hyperlink{ref-Piel2002}{2002}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al.
+\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms are committed to storing and making available published phylogenies and chronograms for easy scientific reuse.
+All of them can be queried using automatised web procedures, which permit personalised, large scale queries that are also very fast.
+OToL's stores trees with branch lengths from a wide range of living organisms, using a very good metadata structure.
+Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data, and their metadata does not allow automatic distinction between types of trees, impairing the automatised access to time of lineage divergence information.
 
-OToL also has the primary goal of synthesizing stored trees into a single tree of life, to show phylogenetic relationships among known extant species and subspecies.
-All or parts of OToL's synthetic tree can be reused for any purpose; however, it currently does not include information on time of divergence, so the platform does not summarize source chronogram data.
-The Timetree of Life platform main goal is the synthesis of a single chronogram of life (Hedges et al.
+Besides storing public phylogenetic knowledge, OToL also has the primary goal of exposing a single tree of all life synthesised from trees in their repository, to show phylogenetic relationships of all known lineages.
+All or parts of OToL's synthetic tree can be reused for any purpose; however, it currently only focus on tree topology, meaning that it does not incorporate information on time of divergence from their stored chronograms.
+Phylomatic (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005})
+is a platform that synthesizes a tree containing taxa of interest from a user supplied set of trees or from its own database.
+It has the advantage that it allows grafting taxa of interest absent from their database, but it is limited to mostly seed plants and mammals.
+If your query does not match its records, it gives many polytomies back.
+The Timetree of Life project focuses on the synthesis of a single chronogram of life (Hedges et al.
 \protect\hyperlink{ref-Hedges2006}{2006}).
-However, both the latest version of their synthetic chronogram (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}) and the thousands of chronograms compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
+However, the thousands of chronograms compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
+The latest version of their synthetic chronogram (Kumar et al.
+\protect\hyperlink{ref-Kumar2017}{2017}) can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
 
 Other platforms such as SuperSmart (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}) and phylogenerator (Pearse and Purvis \protect\hyperlink{ref-pearse2013phylogenerator}{2013}) are focused in \emph{de novo} chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
+\protect\hyperlink{ref-antonelli2017supersmart}{2017}) and phylogenerator (Pearse and Purvis \protect\hyperlink{ref-pearse2013phylogenerator}{2013}) are focused in automatised \emph{de novo} chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
 However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Moreover, they do not provide information from available chronograms.
+Also, these tools do not provide information from already existent expert chronograms.
 
-A platform for efficient reuse of expert, published data on time of lineage divergence should have an open source database in a format suitable for scientific reuse, and straightforward means of accessing, comparing and summarizing the source data as needed by the user.
-A prototype service striving for this characteristics was initially developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
+A tool for efficient reuse of expert, published data on time of lineage divergence should have an open source database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing the source data as needed by the user.
+A prototype service aiming to meet this characteristics was initially developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
 \protect\hyperlink{ref-Stoltzfus2013}{2013}).
-In here we present the first formal description of the \texttt{datelife} service.
-Constituted by and R package and web site, it features new methods to summarize chronograms, an improved system for chronogram database maintenance, new functions to visualize source data in the web interface (\url{http://www.datelife.org/query/}), as well as tools for comparison of source and summary trees.
+In here we present the first formal description and implementation of the \texttt{datelife} service, constituted by and R package and a web site (\url{http://www.datelife.org/}).
+There is still much room for improvement, and flaws and limitations are addressed below.
+We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, new functions to visualize and graphically compare source and summary chronograms, as well as tools for comparison of source and summary trees.
 R packages for benchmarking of functionalities and exemplifying services were also developped.
@@
 Dryad metadata does not allow differentiating between different types of data sets, and trees would need to be manually curated anyways.
-\texttt{datelife} currently accepts scientific names, from named clades to binomial specifics.
-Taxon searches are performed at the species level, so when input names correspond
+\texttt{datelife} currently accepts scientific names only. It can be any named clade or binomial specific.
+Chronogram searches are performed at the species level, so when input names correspond
 to named clades, \texttt{datelife} pulls all accepted species names within the
@@
 \begin{center}
-\textsc{Conclusions}
+\textsc{Flaws, Limitations and Prospects}
 \end{center}
 
-Taxon ages are key to many areas of evolutionary studies: trait evolution, species
-diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to develop the necessary knowledge and data curation skills
-to produce new chronograms. Knowledge on taxon ages is also important for non-biological
-studies and the non-academic community.
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as Dryad, TreeBASE or Open Tree
-of Life, for a large quantity and diversity of organisms. However, this information
-remains difficult to synthesize for many biologists and the non-academic community.
+The main goal of \texttt{datelife} is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
+However, it has many flaws.
+Some of them can be overcome, some of them might represent limitations.
 
-Here, we have shown that \texttt{datelife} allows
-an easy and fast obtention of all publicly available information on taxon ages,
-which can be used to generate new data.
-This information can be used to account for the effect of phylogenetic signal in
-studies of trait evolution; to explore potential speciation and extinction dynamics
-of interest within a clade; to obtain a time frame of biogeographical events; for
-science communication and outreach, amongst others.
-Compared to similar platforms such as time tree of life
-and supermart, it offers several advantages.
-It is fast;
-source data is completely open;
-it requires no expert biological knowledge from users for any of its functionalities;
-it allows exploration of alternative taxonomic and phylogenetic schemes;
-it allows rapid exploration of the effect of alternative divergence time hypothesis;
-it allows rapid synthesis in a number of different formats;
-it facilitates reproducibility of analyses;
+There is only one public repository that has the metadata that datelife needs to perform its tasks and services.
+Other databases are not suitable for data mining because of their metadata structure.
+OToL's tree database is very large for phylogenetic trees, but only stores a little bit more than 200 chronograms.
+This database is maintained by the scientific community, so the prospect is to keep inviting and doing work encouraging scientists to make their data sets available in this database.
 
-Improvements, short and long-term:
+We introduce different ways of summarizing this data, but are these summary chronograms reliable to study evolutionary patterns?
+The short answer is that we do not know to what extent the summary trees can be used for reanalyses and in which areas they can be used more reliably than others.
+Surely they might be useful for ecological analyses, but further studies should be conducted.
+
+As implemented now, we currently by default summarize all source data avilable.
+Users can subset source data if they have reasons to favor some or one source chronogram over others.
+To our knowledge there is no objective way to determine if an expert chronogram is better than other.
+Some criteria that have been put forward are the level of sampling and the numer of calibrations used.
+Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
+It would be important to evaluate how different source chronograms affect summary chronograms, in order to determine if any type of bias is introduced by certain types of chronograms in particular, so this chronograms are not included in the summary.
+A first step in that direction would be to add data on chronograms that come from primary or secondary calibrations, and maybe the number of calibrations used. This will need to be manually curated.
+
+Even if all source chronograms have been generated by excellent standards and practices, the evolutionary context they depict might be very different.
+The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis (Antonelli et al.
+\protect\hyperlink{ref-antonelli2017supersmart}{2017}).
+Summarizing trees implies summarizing evolutionary hypothesis.
+
+However, the comparison of source trees in the example above shows that even chronograms obtained with primary data can be very different.
+There is some evidence that different chronograms will result in very different macroevolutionary histories inferred from them (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
+In other areas of research, such as ecological and conservation biology, it has been demonstrated that the presence of data on lineage divergence represents an improvement for testing alternative hypothesis, compared to testing in a total absence of such time data.
+However, we do not know how using wrong or different kinds of age data will affect the outcomes of ecological tests.
+More work has to be performed in this area.
+The comparison of available data has made evident how much discrepancy in time of lineage divergence can be found across expert knowledge.
+This has been shown for angiosperms previously (Magallón et al.
+\protect\hyperlink{ref-magallon2015metacalibrated}{2015}), but our examples show that this is a widespread phenomenon that might require further attention.
+In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and software used was very advanced.
+However, this is just an assumption.
+Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used) But we should also be able to choose on
+
+Improvements:
 * fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
@@
 
-Problems and caveats:
-Not many databases, only OToL
-Why TreeBase is not very useful for us? Be precise.
-Are these chronograms reliable to study evolutionary patterns, such as species diversification?
-\texttt{datelife} can be seen as an open resource to know the current state of knowledge on lineage divergence times.
-Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
-If all, la facilidad para obtener hipotesis de tiempo de divergencia nos ayudará a evaluar la capacidad de los cronogramas para estudiar otros fenomenos evolutivos.
-Por ahora, no podemos aseverar que estos cronogramas puedan usarse para todo tipo de analisis.
+\begin{center}
+\textsc{Conclusions}
+\end{center}
+
+Time of lineage divergence data are key to many areas of evolutionary studies: trait evolution,
+diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
+especially for those who want to use phylogenies but who are not systematists, or
+do not have the time to acquire and develop the necessary knowledge and data curation skills
+to produce chronograms \emph{de novo}. Knowledge on taxon ages is also crucial for
+science communication and education.
+
+\texttt{datelife} allows an easy and fast obtention, as well as comparison of publicly available information
+on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the timeframe of different regions of the tree of life, allowing identification of regions that need more work or that have conflicting information.
+It can also summarize public data, or use it to generate new chronograms based on trees provided by the user.
+Both summary and newly generated trees have the potential to be useful to evaluate evolutionary hypothesis in different areas of research.
+How well this trees work for this purpose still needs to be tested.
+\texttt{datelife} helps with awereness on the variation of expert data, and the actual effect of the underlying data on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+
+\texttt{datelife} is also part of the reproducibility of science movement.
+It strives to facilitate usage of public data sets as well as encouraging the making of information publicly available.
+Acknowledging source information and describing workflows and analyses.
+
+it allows exploration of alternative taxonomic and phylogenetic schemes;
+it allows rapid exploration of the effect of alternative divergence time hypothesis;
+it allows rapid synthesis in a number of different formats;
 
@@
 \url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained by RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+\texttt{datelife}'s website is maintained with RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
 \texttt{datelife}'s R package stable version is available
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the software benchmark can be found in GitHub repositories at \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
+Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities can be found in GitHub repositories at \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
 
@@
 
+\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
+Magallón S., Gómez-Acevedo S., Sánchez-Reyes L.L., Hernández-Hernández T.
+2015.
+A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
+New Phytologist.
+207:437--453.
+
 \leavevmode\hypertarget{ref-Michonneau2016}{}%
@@
 14.
+
+\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
+Title P.O., Rabosky D.L. 2016. Do Macrophylogenies Yield Stable Macroevolutionary Inferences? An Example from Squamate Reptiles. Systematic Biology.:syw102.
 

### 2019-06-03 Datelife_paper1.Rmd
discussion

@@
 All of them can be queried using automatised web procedures, which permit personalised, large scale queries that are also very fast.
-OToL stores trees with branch length information from a wide range of living organisms, implementing a very good metadata structure.
-Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data, and their metadata does not allow automatic distinction between types of trees, impairing the automatised access to time of lineage divergence information from the latter two databases
+OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
+Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information from the latter two databases.
 
-Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesised all trees in their repository to expose to the community the phylogenetic relationships among all known lineages in a single tree of all life.
+Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the the phylogenetic relationships among known lineages.
 All or parts of this synthetic tree can be reused for any purpose; however, it currently only shows tree topology, meaning that it does not expose branch length data of any type, including time of divergence data from their stored chronograms.
-Phylomatic [@webb2005phylomatic]
-<!-- In a way, phylomatic synthesises trees based on a query. -->
-is a platform that synthesizes a tree containing taxa of interest from a user supplied set of trees or from its own database.
-It has the advantage that it allows grafting taxa of interest absent from their database, but it is limited to mostly seed plants and mammals.
-If your query does not match its records, it gives many polytomies back.
 The Timetree of Life project focuses on the synthesis of a single chronogram of life [@Hedges2006].
@@
 The latest version of their synthetic chronogram [@Kumar2017] can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
-
 Other platforms such as SuperSmart [@antonelli2017supersmart] and phylogenerator [@pearse2013phylogenerator] are focused in automatised _de novo_ chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
@@
 tree repository. Compared to other existing tree repositories, such as TreeBASE and
-Dryad, OToL's metadata rich tree store is the only one that supports automatised proper search, identification and handling of chronograms.
+Dryad, OToL's metadata rich tree store is the only one that supports automatised search, identification and handling of chronograms.
 The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence.
-The Dryad repository not only stores trees and chronograms, but all sorts of biological data, and their metadata does not allow differentiating between trees with branch lengths proportional to time and any other branch length measure that is not relevant to summarize time of divergence.
+The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.
+<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->
 
@@
 
-Source chronogram summary trees can be assembled using the Super Distance Matrix
-(SDM) supertree construction approach [@Criscuolo2006] or using the median of branch
-lengths <!-- and the hierarchical clustering method of ttol-->.
-Tree dating and simulation options are performed with various algorithms:
+Information from source chronograms can be summarised using the Super Distance Matrix (SDM) approach for supertree reconstruction [@Criscuolo2006] or using a summary statistic of branch
+lengths (such as median or mean length)<!-- and the hierarchical clustering method of ttol-->.
+Tree dating and simulations are performed with various algorithms:
 Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
 nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
-PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-[@Smith2012].
+PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees that requires initial branch lengths and a set of calibrations.
+
 MrBayes [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
@@
 from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
+
 To apply calibrations to a tree, the congruification algorithm described by @Eastman2013
@@
 It has been shown with simulations that in analyses performed with bayesian inference methods, divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations [using secondary calibrations in the same way as primary calibrations are used @schenk2016sec].
-But there are different ways to apply secondary calibrations.
+But there are different ways to apply secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
 This information should be included as metadata to facilitate subset of source chronograms.
@@
 
-It would be important to evaluate how different source chronograms affect summary chronograms, in order to determine if any type of bias is introduced by certain types of chronograms in particular, so this chronograms are not included in the summary.
-
-Even if all source chronograms have been generated by excellent standards and practices, the evolutionary context they depict might be very different.
-The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis [@antonelli2017supersmart].
-Summarizing trees implies summarizing evolutionary hypothesis.
-
-However, the comparison of source trees in the example above shows that even chronograms obtained with primary data can be very different.
+As observed from the comparison of source chronograms in the example above, even chronograms obtained with primary data can be very different.
 There is some evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
@@
 More work has to be performed in this area.
-The comparison of available data has made evident how much discrepancy in time of lineage divergence can be found across expert knowledge.
-This has been shown for angiosperms previously [@magallon2015metacalibrated], but our examples show that this is a widespread phenomenon that might require further attention.
-In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and software used was very advanced.
+A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
+Comparison of available chronogram data for a wide range of organisms shown in the examples here suggest that this is a widespread phenomenon that requires further attention.
+In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
 However, this is just an assumption.
-Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used) But we should also be able to choose on
+Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used).
+But we should also be able to choose on
+
+Even if all source chronograms have been generated by excellent standards and using same practices, the evolutionary context they depict might be very different.
+The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis [@antonelli2017supersmart].
+Summarizing trees implies summarizing evolutionary hypothesis.
+It is unsure if this is a good or bad thing.
+
+For all these reasons, the ideal would be to use fossil data as primary sources of calibrations.
+This will be an upscale for datelife, taking fossils already cutrated as calibrations to date other trees.
 

### 2019-06-03 Datelife_paper1.tex
discussion

@@
 The primary information needed for time of lineage divergence estimation comes from the fossil record.
-Coupled to molecular phylogenies, the time of divergence of extant lineages is reconstructed with molecular dating methods.
-
+Coupled to molecular phylogenies (and more recenyly to morphological data), the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
 Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al.
@@
 Still, generating a chronogram is not an easy task unless you have specialized training.
-That's why there has been an urge for promoting and facilitating reuse of the vast amount and still accumulating information that is already available, for the advantage of researchers not specialized in this area (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al.
+That's why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al.
 \protect\hyperlink{ref-Stoltzfus2013}{2013}).
@@
 \protect\hyperlink{ref-Piel2002}{2002}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al.
-\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms are committed to storing and making available published phylogenies and chronograms for easy scientific reuse.
+\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms store and make available published phylogenies and chronograms for easy scientific reuse.
 All of them can be queried using automatised web procedures, which permit personalised, large scale queries that are also very fast.
-OToL's stores trees with branch lengths from a wide range of living organisms, using a very good metadata structure.
-Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data, and their metadata does not allow automatic distinction between types of trees, impairing the automatised access to time of lineage divergence information.
+OToL stores trees with branch length information from a wide range of living organisms, implementing a very good metadata structure.
+Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data, and their metadata does not allow automatic distinction between types of trees, impairing the automatised access to time of lineage divergence information from the latter two databases
 
-Besides storing public phylogenetic knowledge, OToL also has the primary goal of exposing a single tree of all life synthesised from trees in their repository, to show phylogenetic relationships of all known lineages.
-All or parts of OToL's synthetic tree can be reused for any purpose; however, it currently only focus on tree topology, meaning that it does not incorporate information on time of divergence from their stored chronograms.
+Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesised all trees in their repository to expose to the community the phylogenetic relationships among all known lineages in a single tree of all life.
+All or parts of this synthetic tree can be reused for any purpose; however, it currently only shows tree topology, meaning that it does not expose branch length data of any type, including time of divergence data from their stored chronograms.
 Phylomatic (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005})
@@
 
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open source database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing the source data as needed by the user.
-A prototype service aiming to meet this characteristics was initially developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
+A tool for efficient reuse of expert, published data on time of lineage divergence should have an open source database storing chronograms in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram data as needed by the user.
+A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
 \protect\hyperlink{ref-Stoltzfus2013}{2013}).
-In here we present the first formal description and implementation of the \texttt{datelife} service, constituted by and R package and a web site (\url{http://www.datelife.org/}).
+In here we present the formal description and implementation of the \texttt{datelife} service, constituted by an R package and a web site (\url{http://www.datelife.org/}).
 There is still much room for improvement, and flaws and limitations are addressed below.
-We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, new functions to visualize and graphically compare source and summary chronograms, as well as tools for comparison of source and summary trees.
-R packages for benchmarking of functionalities and exemplifying services were also developped.
+We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+R packages for benchmarking of functionalities and demonstrating services were also developped.
 
@@
 
-\texttt{datelife}'s chronogram database is currently built from OToL's
-tree repository. Among currently existing repositories (i.e., TreeBASE,
-Dryad), OToL's metadata rich tree store is the only one meeting the requirements
-for automatized handling of source trees.
-TreeBASE trees are stored without branch lengths.
-Dryad metadata does not allow differentiating between different types of data sets, and trees would need to be manually curated anyways.
+To gather, process, and present information, \texttt{datelife} builds up from functions
+available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
+geiger (Harmon et al.
+\protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}), taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl \protect\hyperlink{ref-Heibl2008}{2008}), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018}) and rphylotastic (O'Meara et al.
+\protect\hyperlink{ref-Omeara2019}{2019}).
+
+\texttt{datelife}'s chronogram database currently contains only chronograms from OToL's
+tree repository. Compared to other existing tree repositories, such as TreeBASE and
+Dryad, OToL's metadata rich tree store is the only one that supports automatised proper search, identification and handling of chronograms.
+The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence.
+The Dryad repository not only stores trees and chronograms, but all sorts of biological data, and their metadata does not allow differentiating between trees with branch lengths proportional to time and any other branch length measure that is not relevant to summarize time of divergence.
+
 \texttt{datelife} currently accepts scientific names only. It can be any named clade or binomial specific.
@@
 
-Source chronogram summary tree can be assembled using the Super Distance Matrix
+Source chronogram summary trees can be assembled using the Super Distance Matrix
 (SDM) supertree construction approach (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}) or using the median of branch
@@
 is implemented to find shared nodes between trees (congruent nodes).
-
-To gather, process, and present information, \texttt{datelife} builds up from functions
-available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
-geiger (Harmon et al.
-\protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}),
-taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl \protect\hyperlink{ref-Heibl2008}{2008}), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018})
-and rphylotastic (O'Meara et al. \protect\hyperlink{ref-Omeara2019}{2019}).
 
@@
 
-There is only one public repository that has the metadata that datelife needs to perform its tasks and services.
-Other databases are not suitable for data mining because of their metadata structure.
-OToL's tree database is very large for phylogenetic trees, but only stores a little bit more than 200 chronograms.
-This database is maintained by the scientific community, so the prospect is to keep inviting and doing work encouraging scientists to make their data sets available in this database.
+There is only one public tree repository from where \texttt{datelife} currently pulls chronograms to construct its database.
+Other existing tree repositories are not suitable for time of divergence data mining because they either do not store that data or because their metadata does not have enough information to allow automatised chronogram searches.
+\texttt{datelife}'s database has a bit more than 200 chronograms up to the time the manuscript was written.
+Incorporation of the many chronograms that are available in data repositories such as Dryad and also from supplementary material of journals to \texttt{datelife}'s database is crucial to improve \texttt{datelife}'s services.
+Methods to automatically mine Dryad repository for chronogram data could be designed and implemented.
+At the same time, we would like to increase awereness and emphasize on the importance of data sharing in databases that are maintained by the expert scientific community, requiring expert input and manual curation, such as OToL's tree store.
 
@@
 Users can subset source data if they have reasons to favor some or one source chronogram over others.
+Strictly speaking, a good chronogram reflects the real time of divergence accurately and precisely.
 To our knowledge there is no objective way to determine if an expert chronogram is better than other.
@@
 Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
-It would be important to evaluate how different source chronograms affect summary chronograms, in order to determine if any type of bias is introduced by certain types of chronograms in particular, so this chronograms are not included in the summary.
-A first step in that direction would be to add data on chronograms that come from primary or secondary calibrations, and maybe the number of calibrations used. This will need to be manually curated.
+\texttt{datelife}'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.
+It has been shown with simulations that in analyses performed with bayesian inference methods, divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations (using secondary calibrations in the same way as primary calibrations are used Schenk \protect\hyperlink{ref-schenk2016sec}{2016}\protect\hyperlink{ref-schenk2016sec}{a}).
+But there are different ways to apply secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
+This information should be included as metadata to facilitate subset of source chronograms.
+This will need to be manually curated as is done in OToL.
 
-Even if all source chronograms have been generated by excellent standards and practices, the evolutionary context they depict might be very different.
-The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}).
-Summarizing trees implies summarizing evolutionary hypothesis.
-
-However, the comparison of source trees in the example above shows that even chronograms obtained with primary data can be very different.
+As observed from the comparison of source chronograms in the example above, even chronograms obtained with primary data can be very different.
 There is some evidence that different chronograms will result in very different macroevolutionary histories inferred from them (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
@@
 More work has to be performed in this area.
-The comparison of available data has made evident how much discrepancy in time of lineage divergence can be found across expert knowledge.
-This has been shown for angiosperms previously (Magallón et al.
-\protect\hyperlink{ref-magallon2015metacalibrated}{2015}), but our examples show that this is a widespread phenomenon that might require further attention.
-In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and software used was very advanced.
+A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms (e.g., angiosperms; Magallón et al.
+\protect\hyperlink{ref-magallon2015metacalibrated}{2015}).
+Comparison of available chronogram data for a wide range of organisms shown in the examples here suggest that this is a widespread phenomenon that requires further attention.
+In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
 However, this is just an assumption.
-Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used) But we should also be able to choose on
+Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used).
+But we should also be able to choose on
+
+Even if all source chronograms have been generated by excellent standards and using same practices, the evolutionary context they depict might be very different.
+The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis (Antonelli et al.
+\protect\hyperlink{ref-antonelli2017supersmart}{2017}).
+Summarizing trees implies summarizing evolutionary hypothesis.
+It is unsure if this is a good or bad thing.
+
+For all these reasons, the ideal would be to use fossil data as primary sources of calibrations.
+This will be an upscale for datelife, taking fossils already cutrated as calibrations to date other trees.
 
@@
 * fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
-(Schenk \protect\hyperlink{ref-Schenk2016}{2016}).
+(Schenk \protect\hyperlink{ref-Schenk2016}{2016}\protect\hyperlink{ref-Schenk2016}{b}).
 * bayesian congruification
@@
 
+\leavevmode\hypertarget{ref-schenk2016sec}{}%
+Schenk J.J. 2016a. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
+
 \leavevmode\hypertarget{ref-Schenk2016}{}%
-Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
+Schenk J.J. 2016b. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
 

### 2019-06-03 Intro.Rmd
discussion

@@
 
+Phylomatic [@webb2005phylomatic]
+<!-- In a way, phylomatic synthesises trees based on a query. -->
+is a platform that synthesizes a tree containing taxa of interest from a user supplied set of trees or from its own database.
+It has the advantage that it allows grafting taxa of interest absent from their database, but it is limited to mostly seed plants and mammals.
+If your query does not match its records, it gives many polytomies back.
+
+treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
+[@Smith2012].
 <!--INTRODUCTION: 1st paragraph: general statement that the field knows to be true and important-->

### 2019-06-03 Datelife_paper1.Rmd
discussion

@@
 The primary information needed for time of lineage divergence estimation comes from the fossil record.
-Coupled to molecular phylogenies (and more recenyly to morphological data), the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
+Coupled to molecular phylogenies (and more recently to morphological data), the time of divergence of extant and extinct lineages is reconstructed with molecular dating methods.
 Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
@@
 OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information from the latter two databases.
+Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
 
-Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the the phylogenetic relationships among known lineages.
-All or parts of this synthetic tree can be reused for any purpose; however, it currently only shows tree topology, meaning that it does not expose branch length data of any type, including time of divergence data from their stored chronograms.
+Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the phylogenetic relationships among known lineages.
+All or parts of this synthetic tree can be reused for any purpose. However, it currently only focus on synthesizing tree topology, meaning that it does not expose branch length data of any type. 
+<!-- let alone time of divergence data from their stored chronograms. -->
 The Timetree of Life project focuses on the synthesis of a single chronogram of life [@Hedges2006].
-However, the thousands of chronograms compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
+However, the thousands of chronograms they have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
 The latest version of their synthetic chronogram [@Kumar2017] can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
@@
 However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Also, these tools do not provide information from already existent expert chronograms.
+Moreover, these tools do not provide information from already existent expert chronograms.
 
@@
  -->
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open source database storing chronograms in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram data as needed by the user.
+A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fuly public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
 A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
@@
 <!-- as well as tools for comparison of source and summary trees.  -->
-R packages for benchmarking of functionalities and demonstrating services were also developped.
+R packages for benchmarking of functionalities and demonstrating services were also developed and made available.
 
@@
 
-1) A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
+\begin{enumerate}
+\item A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
 The tree can be in newick or phylo format, and can be with or without branch lengths.
-2) `datelife` then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarization algorithms.
-3) The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
-4) At this point, users can choose to use all or some source data as calibration points to date a tree of their own making or choosing. <!--, a taxonomic tree-->
-5) Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
-6) Finally, users can easily view results graphically as well as construct their own graphs using inbuilt `datelife` graphic generators.
+\item A name search is then performed across the chronogram database; source trees with at least two matching input names are identified; all other taxa that do not match the original query are then dropped from the positively identified source trees --these pruned chronograms are hereafter referred as source chronograms; finally, each source chronogram is transformed to a patristic matrix named by the citation of the original study.
+This format facilitates and greatly speeds up all further analyses and summarising algorithms.
+\item  The user can obtain different summary information from the source chronograms including: a) all source chronogram ages, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, and f) a report of succesful matches of input taxon names across source chronograms.
+Summary information can be used to make decisions on the next steps of the workflow.
+\item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
+\item  If there is no information available for any queried taxa, users can also simulate both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
+\item  Finally, users can easily view and compare results graphically, as well as construct their own graphs using inbuilt `datelife` graphic generation functions.
+\end{enumerate}
 
-<!--To perform all tasks:-->
+<!--To perform each task:-->
 To gather, process, and present information, `datelife` builds up from functions
@@
 
-`datelife`'s chronogram database currently contains only chronograms from OToL's
-tree repository. Compared to other existing tree repositories, such as TreeBASE and
-Dryad, OToL's metadata rich tree store is the only one that supports automatised search, identification and handling of chronograms.
-The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence.
-The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.
-<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->
-
-`datelife` currently accepts scientific names only. It can be any named clade or binomial specific.
-<!--common names (implemented in phylotastic)-->
-Chronogram searches are performed at the species level, so when input names correspond
+A `datelife` search currently accepts scientific names only. It can be any named clade or binomial specific. <!--common names (implemented in phylotastic)-->
+Chronogram search is performed at the species level, so when input names correspond
 to named clades, `datelife` pulls all accepted species names within the
 clade from OToL's reference taxonomy to perform the search.
-Currently, searches at the infraspecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
-`datelife` also processes input names with the taxon name resolution service [TNRS; @Boyle2013],
+Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
+`datelife` processes input names with the taxon name resolution service [TNRS; @Boyle2013],
 <!--which "corrects misspelled names and authorities, standardizes variant spellings,
@@
 
+The chronogram search is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository.
+Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
+Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
+<!-- The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence.
+The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.
+-->
+<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->
+
 Information from source chronograms can be summarised using the Super Distance Matrix (SDM) approach for supertree reconstruction [@Criscuolo2006] or using a summary statistic of branch
-lengths (such as median or mean length)<!-- and the hierarchical clustering method of ttol-->.
+lengths (such as median or mean). The resulting summary distance matrix could be clustered using  <!-- and the hierarchical clustering method of ttol-->
 Tree dating and simulations are performed with various algorithms:
@@
 
-There is only one public tree repository from where `datelife` currently pulls chronograms to construct its database.
-Other existing tree repositories are not suitable for time of divergence data mining because they either do not store that data or because their metadata does not have enough information to allow automatised chronogram searches.
-`datelife`'s database has a bit more than 200 chronograms up to the time the manuscript was written.
-<!-- , which probably corresponds to less than 1% (`r 200*100/3987`%) of the information used for TimeTree synthesis  
+At the moment, `datelife`'s chronogram database is not very large.
+It stores `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written.
+This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is unfortunately not open for scientific reuse and automatised data mining [@Kumar2017].
+OToL is the only public tree repository from where `datelife` can currently pull chronograms to construct its database.
+Other open repositories are not suitable for time of divergence data mining because they either do not store that data (i.e., TreeBASE), or their metadata does not have enough information to allow automatised chronogram searches (i.e., Dryad).
+
+A previous version of TimeTree's synthetic chronogram [@Hedges2015] was made available in OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
+This ensures that at least some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
+
+<!--  
 Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
-Incorporation of the many chronograms that are available in data repositories such as Dryad and also from supplementary material of journals to `datelife`'s database is crucial to improve `datelife`'s services.
-Methods to automatically mine Dryad repository for chronogram data could be designed and implemented. 
+Thus, incorporation of published chronograms available in the Dryad data repository and in supplementary material of journals to `datelife`'s database is crucial to improve its services. 
+Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would need to be determined by "hand".
 <!-- What can we do to mine journal databases? -->
-<!-- At the same time, the prospect is to keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
-At the same time, we would like to increase awereness and emphasize on the importance of data sharing in databases that are maintained by the expert scientific community, requiring expert input and manual curation, such as OToL's tree store.
+<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
+Consequently, we would like to emphasize on the importance for the scientific community of sharing chronogram data in repositories that require expert input and manual curation,\ such as OToL's tree repository.
 
@@
 However, this is just an assumption.
-Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used).
+Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used; support for nodes and ages; age confidence intervals).
 But we should also be able to choose on
@@
 For all these reasons, the ideal would be to use fossil data as primary sources of calibrations.
-This will be an upscale for datelife, taking fossils already cutrated as calibrations to date other trees.
+This will be an upscale for datelife, taking fossils that have already been curated as calibrations to date other trees.
 
@@
 
-Time of lineage divergence data are key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. Obtaining these ages is difficult,
+Time of lineage divergence information is key to many areas of evolutionary studies: trait evolution,
+diversification, biogeography, macroecology and more. Generating this information is difficult,
 especially for those who want to use phylogenies but who are not systematists, or
@@
 <http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
-`datelife`'s website is maintained with RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+`datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
 `datelife`'s R package stable version is available
@@
 Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities can be found in GitHub repositories at <https://github.com/LunaSare/datelife_paper1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>, respectively.
-<!-- Vignettes
+<!-- Vignettes?
 Dryad also?? since it is all in github already, maybe not necessary-->
@@
 
-Funding was provided by US National Science Foundation (NSF) grant 1458603
+Funding was provided by the US National Science Foundation (NSF) grant 1458603
 
@@
 The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in OpenTree.
+the scientists who curate them for deposition in the Open Tree of Life repository.
 The NSF for funding nearly all the above, in addition

### 2019-06-03 Datelife_paper1.tex
discussion

@@
 The primary information needed for time of lineage divergence estimation comes from the fossil record.
-Coupled to molecular phylogenies (and more recenyly to morphological data), the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
+Coupled to molecular phylogenies (and more recently to morphological data), the time of divergence of extant and extinct lineages is reconstructed with molecular dating methods.
 Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al.
@@
 All of them can be queried using automatised web procedures, which permit personalised, large scale queries that are also very fast.
-OToL stores trees with branch length information from a wide range of living organisms, implementing a very good metadata structure.
-Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data, and their metadata does not allow automatic distinction between types of trees, impairing the automatised access to time of lineage divergence information from the latter two databases
+OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
+Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
 
-Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesised all trees in their repository to expose to the community the phylogenetic relationships among all known lineages in a single tree of all life.
-All or parts of this synthetic tree can be reused for any purpose; however, it currently only shows tree topology, meaning that it does not expose branch length data of any type, including time of divergence data from their stored chronograms.
-Phylomatic (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005})
-is a platform that synthesizes a tree containing taxa of interest from a user supplied set of trees or from its own database.
-It has the advantage that it allows grafting taxa of interest absent from their database, but it is limited to mostly seed plants and mammals.
-If your query does not match its records, it gives many polytomies back.
+Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the phylogenetic relationships among known lineages.
+All or parts of this synthetic tree can be reused for any purpose. However, it currently only focus on synthesizing tree topology, meaning that it does not expose branch length data of any type.
 The Timetree of Life project focuses on the synthesis of a single chronogram of life (Hedges et al.
 \protect\hyperlink{ref-Hedges2006}{2006}).
-However, the thousands of chronograms compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
+However, the thousands of chronograms they have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
 The latest version of their synthetic chronogram (Kumar et al.
 \protect\hyperlink{ref-Kumar2017}{2017}) can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
-
 Other platforms such as SuperSmart (Antonelli et al.
@@
 However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Also, these tools do not provide information from already existent expert chronograms.
+Moreover, these tools do not provide information from already existent expert chronograms.
 
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open source database storing chronograms in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram data as needed by the user.
+A tool for efficient reuse of expert, published data on time of lineage divergence should have an open source chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
 A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
@@
 We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
-R packages for benchmarking of functionalities and demonstrating services were also developped.
+R packages for benchmarking of functionalities and demonstrating services were also developed and made available.
 
@@
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
-\tightlist
-\item
-  A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
+\item A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
 The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item
-  \texttt{datelife} then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarization algorithms.
-\item
-  The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
-\item
-  At this point, users can choose to use all or some source data as calibration points to date a tree of their own making or choosing. 
-\item
-  Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
-\item
-  Finally, users can easily view results graphically as well as construct their own graphs using inbuilt \texttt{datelife} graphic generators.
+\item A name search is then performed across the chronogram database; source trees with at least two matching input names are identified; all other taxa that do not match the original query are then dropped from the positively identified source trees --these pruned chronograms are hereafter referred as source chronograms; finally, each source chronogram is transformed to a patristic matrix named by the citation of the original study.
+This format facilitates and greatly speeds up all further analyses and summarising algorithms.
+\item  The user can obtain different summary information from the source chronograms including: a) all source chronogram ages, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, and f) a report of succesful matches of input taxon names across source chronograms.
+Summary information can be used to make decisions on the next steps of the workflow.
+\item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. <!--, a taxonomic tree-->
+\item  If there is no information available for any queried taxa, users can also simulate both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
+\item  Finally, users can easily view and compare results graphically, as well as construct their own graphs using inbuilt `datelife` graphic generation functions.
 \end{enumerate}
@@
 
-\texttt{datelife}'s chronogram database currently contains only chronograms from OToL's
-tree repository. Compared to other existing tree repositories, such as TreeBASE and
-Dryad, OToL's metadata rich tree store is the only one that supports automatised proper search, identification and handling of chronograms.
-The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence.
-The Dryad repository not only stores trees and chronograms, but all sorts of biological data, and their metadata does not allow differentiating between trees with branch lengths proportional to time and any other branch length measure that is not relevant to summarize time of divergence.
-
-\texttt{datelife} currently accepts scientific names only. It can be any named clade or binomial specific.
-Chronogram searches are performed at the species level, so when input names correspond
+A \texttt{datelife} search currently accepts scientific names only. It can be any named clade or binomial specific.
+Chronogram search is performed at the species level, so when input names correspond
 to named clades, \texttt{datelife} pulls all accepted species names within the
 clade from OToL's reference taxonomy to perform the search.
-Currently, searches at the infraspecies level are not allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
-\texttt{datelife} also processes input names with the taxon name resolution service (TNRS; Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}),
+Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
+\texttt{datelife} processes input names with the taxon name resolution service (TNRS; Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}),
 which corrects potentially misspelled names and typos, and standardizes spelling
@@
 
-Source chronogram summary trees can be assembled using the Super Distance Matrix
-(SDM) supertree construction approach (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}) or using the median of branch
-lengths .
-Tree dating and simulation options are performed with various algorithms:
+The chronogram search is performed across \texttt{datelife}'s chronogram database which is assembled from OToL's tree repository.
+Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
+Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
+
+Information from source chronograms can be summarised using the Super Distance Matrix (SDM) approach for supertree reconstruction (Criscuolo et al.
+\protect\hyperlink{ref-Criscuolo2006}{2006}) or using a summary statistic of branch
+lengths (such as median or mean).
+Tree dating and simulations are performed with various algorithms:
 Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
 nodes evenly, which minimizes age variance in the chronogram (Webb et al. \protect\hyperlink{ref-Webb2008}{2008}).
-PATHd8 is a non-clock, rate-smoothing method (Britton et al. \protect\hyperlink{ref-Britton2007}{2007}) to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Smith and O'Meara \protect\hyperlink{ref-Smith2012}{2012}).
+PATHd8 is a non-clock, rate-smoothing method (Britton et al. \protect\hyperlink{ref-Britton2007}{2007}) to date trees that requires initial branch lengths and a set of calibrations.
+
 MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) can be used when adding taxa at
@@
 from a pure birth model, as implemented by Jetz and collaborators (\protect\hyperlink{ref-Jetz2012}{2012}).
+
 To apply calibrations to a tree, the congruification algorithm described by Eastman et al. (\protect\hyperlink{ref-Eastman2013}{2013})
@@
 
-There is only one public tree repository from where \texttt{datelife} currently pulls chronograms to construct its database.
-Other existing tree repositories are not suitable for time of divergence data mining because they either do not store that data or because their metadata does not have enough information to allow automatised chronogram searches.
-\texttt{datelife}'s database has a bit more than 200 chronograms up to the time the manuscript was written.
-Incorporation of the many chronograms that are available in data repositories such as Dryad and also from supplementary material of journals to \texttt{datelife}'s database is crucial to improve \texttt{datelife}'s services.
-Methods to automatically mine Dryad repository for chronogram data could be designed and implemented.
-At the same time, we would like to increase awereness and emphasize on the importance of data sharing in databases that are maintained by the expert scientific community, requiring expert input and manual curation, such as OToL's tree store.
+At the moment, \texttt{datelife}'s chronogram database is not very large.
+It stores 231 chronograms up to the time the manuscript was written.
+This represents 5.79\% of the largest existing chronogram database, which is unfortunately not open for scientific reuse and automatised data mining (Kumar et al.
+\protect\hyperlink{ref-Kumar2017}{2017}).
+OToL is the only public tree repository from where \texttt{datelife} can currently pull chronograms to construct its database.
+Other open repositories are not suitable for time of divergence data mining because they either do not store that data (i.e., TreeBASE), or their metadata does not have enough information to allow automatised chronogram searches (i.e., Dryad).
+
+A previous version of TimeTree's synthetic chronogram (Hedges et al.
+\protect\hyperlink{ref-Hedges2015}{2015}) was made available in OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
+This ensures that at least some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any gien lineage.
+
+Thus, incorporation of the many chronograms that are available in data repositories such as Dryad and also from supplementary material of journals to \texttt{datelife}'s database is crucial to improve \texttt{datelife}'s services.
+Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would need to be determined by ``hand''.
+Consequently, we would like to emphasize on the importance for the scientific community of sharing chronogram data in repositories that require expert input and manual curation,~such as OToL's tree repository.
 
@@
 
-\leavevmode\hypertarget{ref-Smith2012}{}%
-Smith S.A., O'Meara B.C. 2012. TreePL: Divergence time estimation using penalized likelihood for large phylogenies. Bioinformatics. 28:2689--2690.
-
 \leavevmode\hypertarget{ref-Stoltzfus2013}{}%

### 2019-06-03 Intro.Rmd
discussion

@@
 ---
-
+  TimeTree's chronogram database is currently the largest one, but it is not publicly available. ) is close to gather 
 Phylomatic [@webb2005phylomatic]

### 2019-06-03 Datelife_paper1_global.R

@@ -24,3 +24,5 @@ download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/do
 
 source("frin_plan.R")
 loadd() # loads all the targets listed in the workflow plan of the previous `make()`
+library(datelife)
+data(opentree_chronograms)

### 2019-06-03 Datelife_paper1.Rmd
discussion

@@
 
-Information from source chronograms can be summarised using the Super Distance Matrix (SDM) approach for supertree reconstruction [@Criscuolo2006] or using a summary statistic of branch
-lengths (such as median or mean). The resulting summary distance matrix could be clustered using  <!-- and the hierarchical clustering method of ttol-->
-Tree dating and simulations are performed with various algorithms:
-Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
-nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
-PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees that requires initial branch lengths and a set of calibrations.
+Information from source chronograms can be summarised with a summary statistic of tree branch
+lengths (such as median or mean) or using the Super Distance Matrix (SDM) approach for supertree reconstruction [@Criscuolo2006]. <!-- and the hierarchical clustering method of ttol-->
+The resulting summary patristic distance matrix could be clustered with classic algorithms.
+However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package).
+Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
+The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
 
-MrBayes [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
-random, following a reference taxonomy or a topological constraint. It draws ages
-from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
+For tree dating, the congruification algorithm described by @Eastman2013
+is implemented to find shared nodes between trees (congruent nodes).
+The ages of these nodes are then used as calibrations to date any given tree.
+Currently implemented methods for tree dating are BLADJ, MrBayes [@Huelsenbeck2001; @Ronquist2003] and PATHd8 [@Britton2007], a non-clock, rate-smoothing dating method.
 
-To apply calibrations to a tree, the congruification algorithm described by @Eastman2013
-is implemented to find shared nodes between trees (congruent nodes).
+<!-- and simulations are performed with various algorithms: -->
+<!-- MrBayes can be used when adding taxa at -->
+<!-- random, following a reference taxonomy or a topological constraint. It draws ages -->
+<!-- from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012]. -->
 
@@
 <!-- and hierarchical clustering method which appears to push ages back -->
-It is not certain if these chronograms can be used to perform downstream evolutionary analyses.
-There is currently wide interest in determining this.
-However, we know that these chronograms are useful for...
+It is not certain if these chronograms can be used to perform all types of downstream evolutionary analyses and more research is needed in that direction.
 
@@
 
-At the moment, `datelife`'s chronogram database is not very large.
-It stores `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written.
+At the moment, `datelife`'s chronogram database is not very large, storing `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written.
 This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is unfortunately not open for scientific reuse and automatised data mining [@Kumar2017].

### 2019-06-03 Datelife_paper1.tex
discussion

@@
 
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open source chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
+A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fuly public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
 A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
@@
 Summary information can be used to make decisions on the next steps of the workflow.
-\item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. <!--, a taxonomic tree-->
+\item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
 \item  If there is no information available for any queried taxa, users can also simulate both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
@@
 lengths (such as median or mean).
-Tree dating and simulations are performed with various algorithms:
-Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
-nodes evenly, which minimizes age variance in the chronogram (Webb et al. \protect\hyperlink{ref-Webb2008}{2008}).
-PATHd8 is a non-clock, rate-smoothing method (Britton et al. \protect\hyperlink{ref-Britton2007}{2007}) to date trees that requires initial branch lengths and a set of calibrations.
+The resulting summary patristic distance matrix could be clustered with classic algorithms.
+However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data.
+Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
+The Branch Length Adjuster (BLADJ) algorithm (Webb et al.
+\protect\hyperlink{ref-Webb2008}{2008}) was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
 
-MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) can be used when adding taxa at
-random, following a reference taxonomy or a topological constraint. It draws ages
-from a pure birth model, as implemented by Jetz and collaborators (\protect\hyperlink{ref-Jetz2012}{2012}).
-
-To apply calibrations to a tree, the congruification algorithm described by Eastman et al. (\protect\hyperlink{ref-Eastman2013}{2013})
+For tree dating, the congruification algorithm described by Eastman et al. (\protect\hyperlink{ref-Eastman2013}{2013})
 is implemented to find shared nodes between trees (congruent nodes).
+The ages of these nodes are then used as calibrations to date any given tree.
+Currently implemented methods for tree dating are BLADJ, MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) and PATHd8 (Britton et al.
+\protect\hyperlink{ref-Britton2007}{2007}), a non-clock, rate-smoothing dating method.
 
@@
 \protect\hyperlink{ref-Hedges2015}{2015}) was made available in OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
-This ensures that at least some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any gien lineage.
+This ensures that at least some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
 
-Thus, incorporation of the many chronograms that are available in data repositories such as Dryad and also from supplementary material of journals to \texttt{datelife}'s database is crucial to improve \texttt{datelife}'s services.
+Thus, incorporation of published chronograms available in the Dryad data repository and in supplementary material of journals to \texttt{datelife}'s database is crucial to improve its services.
 Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would need to be determined by ``hand''.
@@
 \url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained with RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
 \texttt{datelife}'s R package stable version is available
@@
 
+Supplementary Material
+
 \end{document}

### 2019-06-03 Datelife_paper1_global.R

@@ -20,6 +20,8 @@ download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/do
               destfile = "fig_crossval_bladj.pdf", mode="wb")
 download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_lttplot_crossval_pathd8_summ1.pdf",
               destfile = "fig_crossval_boldsumm.pdf", mode="wb")
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_clusters_both.pdf",
+              destfile = "fig_lttplot_clusters.pdf", mode="wb")
 
 
 source("frin_plan.R")

### 2019-06-03 Datelife_paper1.Rmd
abstract

@@
 data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as Dryad, TreeBASE and Open Tree
-of Life, for a large quantity and diversity of organisms. However, this information
-remains difficult to synthesize for many biologists and the non-academic community.
+of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
+of Life, for a large quantity and diversity of organisms in the last decades.
+However, time of lineage divergence data remains difficult to be obtained and synthesized for many biologists and the non-academic community, despite its importance in many areas of research, education and science communication.
+`datelife` is a service implemented via an R package and a web site
+(<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of expert, published data on time of lineage divergence.
+The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
+A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
+Source chronogram data can be summarised using branch length summary statitiscs or variance minimising approaches to generate a single summary chronogram.
+Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
+If there is no information available for any queried taxa, data can be simulated.
+All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
+Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
+How well this trees work for this purpose still needs to be tested.
+`datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 \item  If there is no information available for any queried taxa, users can also simulate both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily view and compare results graphically, as well as construct their own graphs using inbuilt `datelife` graphic generation functions.
+\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R `phylo` format), as well as view and compare results graphically, or construct their own graphs using inbuilt `datelife` graphic generation functions.
 \end{enumerate}
@@
 <!-- and hierarchical clustering method which appears to push ages back -->
-It is not certain if these chronograms can be used to perform all types of downstream evolutionary analyses and more research is needed in that direction.
+It is not certain if these chronograms should be used to perform all types of downstream evolutionary analyses and more research is needed in that direction.
 
@@
 At the moment, `datelife`'s chronogram database is not very large, storing `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written.
-This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is unfortunately not open for scientific reuse and automatised data mining [@Kumar2017].
+This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is unfortunately not open for scientific reuse nor automatised data mining [@Kumar2017].
 OToL is the only public tree repository from where `datelife` can currently pull chronograms to construct its database.
 Other open repositories are not suitable for time of divergence data mining because they either do not store that data (i.e., TreeBASE), or their metadata does not have enough information to allow automatised chronogram searches (i.e., Dryad).
-
 A previous version of TimeTree's synthetic chronogram [@Hedges2015] was made available in OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
-This ensures that at least some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
-
+This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
 <!--  
@@
 Thus, incorporation of published chronograms available in the Dryad data repository and in supplementary material of journals to `datelife`'s database is crucial to improve its services. 
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would need to be determined by "hand".
+Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by "hand".
 <!-- What can we do to mine journal databases? -->
 <!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
-Consequently, we would like to emphasize on the importance for the scientific community of sharing chronogram data in repositories that require expert input and manual curation,\ such as OToL's tree repository.
+Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.
 
-We introduce different ways of summarizing this data, but are these summary chronograms reliable to study evolutionary patterns?
-The short answer is that we do not know to what extent the summary trees can be used for reanalyses and in which areas they can be used more reliably than others.
-Surely they might be useful for ecological analyses, but further studies should be conducted.
+Another potential concern comes from summary chronograms.
+We currently summarize all source chronogram data by default.
+Users can subset source data if they have reasons to favor some or one source chronogram over others.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there is no objective way to determine if an expert chronogram is better than other.
+Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
+Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
+It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
+Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
+But this remains to be tested.
 
-As implemented now, we currently by default summarize all source data avilable.
-Users can subset source data if they have reasons to favor some or one source chronogram over others.
-Strictly speaking, a good chronogram reflects the real time of divergence accurately and precisely.
-To our knowledge there is no objective way to determine if an expert chronogram is better than other.
-Some criteria that have been put forward are the level of sampling and the numer of calibrations used.
-Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
-`datelife`'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.
-It has been shown with simulations that in analyses performed with bayesian inference methods, divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations [using secondary calibrations in the same way as primary calibrations are used @schenk2016sec].
-But there are different ways to apply secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-This information should be included as metadata to facilitate subset of source chronograms.
-This will need to be manually curated as is done in OToL.
+Furthermore, even chronograms obtained with primary data can be very different, as observed from the comparison of source chronograms in the Fringillidae example. 
+A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
+Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
+<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
+However, this is just an assumption.
+-->
+Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignement (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
+To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
+Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.
+Hence, summarizing chronograms might imply summarizing evolutionary hypothesis.
+This could be good from certain point of view, since it could help to get a single global evolutionary history for a lineage.
+But it could also be bad, since we could be loosing part of the evolutionary history that is only being reflected in some chronograms and not from the summary chronogram.
+Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history [@antonelli2017supersmart].
+This will be implemented in future `datelife` versions.
 
-As observed from the comparison of source chronograms in the example above, even chronograms obtained with primary data can be very different.
-There is some evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
-In other areas of research, such as ecological and conservation biology, it has been demonstrated that the presence of data on lineage divergence represents an improvement for testing alternative hypothesis, compared to testing in a total absence of such time data.
-However, we do not know how using wrong or different kinds of age data will affect the outcomes of ecological tests.
-More work has to be performed in this area.
-A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
-Comparison of available chronogram data for a wide range of organisms shown in the examples here suggest that this is a widespread phenomenon that requires further attention.
-In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
-However, this is just an assumption.
-Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used; support for nodes and ages; age confidence intervals).
-But we should also be able to choose on
-
-Even if all source chronograms have been generated by excellent standards and using same practices, the evolutionary context they depict might be very different.
-The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis [@antonelli2017supersmart].
-Summarizing trees implies summarizing evolutionary hypothesis.
-It is unsure if this is a good or bad thing.
-
-For all these reasons, the ideal would be to use fossil data as primary sources of calibrations.
-This will be an upscale for datelife, taking fossils that have already been curated as calibrations to date other trees.
-
-Improvements:
-* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
-[@Schenk2016].
-* bayesian congruification
-* topological congruification
+Synthetic or summary chronograms are being used more and more frequently to study evolutionary hypothesis.
+We introduce different ways of summarizing time of lineage divergence data, but, are any of these summary chronograms reliable to study evolutionary patterns?
+The short answer is that we do not know to what extent the summary trees should be used for reanalyses and if there bare areas of research where they can be used more reliably than others.
+There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
+In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents an improvement for testing alternative hypothesis using phylogenetic distance.
+It is uncertain, however, how using very different age data could affect the outcomes of these tests, and further research is necessary in this direction.
 
@@
 `datelife` allows an easy and fast obtention, as well as comparison of publicly available information
-on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the timeframe of different regions of the tree of life, allowing identification of regions that need more work or that have conflicting information.
-It can also summarize public data, or use it to generate new chronograms based on trees provided by the user.
-Both summary and newly generated trees have the potential to be useful to evaluate evolutionary hypothesis in different areas of research.
-How well this trees work for this purpose still needs to be tested.
-`datelife` helps with awereness on the variation of expert data, and the actual effect of the underlying data on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, allowing identification of regions that require more research or that have conflicting information.
+<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
+Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
+How well this trees work for this purpose still needs to be further tested.
+If all, `datelife` helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
-`datelife` is also part of the reproducibility of science movement.
+<!-- `datelife` is also part of the reproducibility of science movement.
 It strives to facilitate usage of public data sets as well as encouraging the making of information publicly available.
 Acknowledging source information and describing workflows and analyses.
+-->
 
-it allows exploration of alternative taxonomic and phylogenetic schemes;
-it allows rapid exploration of the effect of alternative divergence time hypothesis;
-it allows rapid synthesis in a number of different formats; <!-- we should add graphical options too-->
+<!-- it allows exploration of alternative taxonomic and phylogenetic schemes; -->
+<!-- it allows rapid ; -->
 

### 2019-06-03 Datelife_paper1.tex
abstract

@@
 data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as Dryad, TreeBASE and Open Tree
-of Life, for a large quantity and diversity of organisms. However, this information
-remains difficult to synthesize for many biologists and the non-academic community.
+of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
+of Life, for a large quantity and diversity of organisms in the last decades.
+However, time of lineage divergence data remains difficult to be obtained and synthesized for many biologists and the non-academic community, despite its importance in many areas of research, education and science communication.
+\texttt{datelife} is a service implemented via an R package and a web site
+(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of expert, published data on time of lineage divergence.
+The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
+A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
+Source chronogram data can be summarised using branch length summary statitiscs or variance minimising approaches to generate a single summary chronogram.
+Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
+If there is no information available for any queried taxa, data can be simulated.
+All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
+Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
+How well this trees work for this purpose still needs to be tested.
+\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 
-Information from source chronograms can be summarised using the Super Distance Matrix (SDM) approach for supertree reconstruction (Criscuolo et al.
-\protect\hyperlink{ref-Criscuolo2006}{2006}) or using a summary statistic of branch
-lengths (such as median or mean).
+Information from source chronograms can be summarised with a summary statistic of tree branch
+lengths (such as median or mean) or using the Super Distance Matrix (SDM) approach for supertree reconstruction (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}).
 The resulting summary patristic distance matrix could be clustered with classic algorithms.
-However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data.
+However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife\_examples package).
 Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
@@
 In both cases, ages are coherent with source ages.
-It is not certain if these chronograms can be used to perform downstream evolutionary analyses.
-There is currently wide interest in determining this.
-However, we know that these chronograms are useful for\ldots{}
+It is not certain if these chronograms should be used to perform all types of downstream evolutionary analyses and more research is needed in that direction.
 
@@
 
-At the moment, \texttt{datelife}'s chronogram database is not very large.
-It stores 231 chronograms up to the time the manuscript was written.
-This represents 5.79\% of the largest existing chronogram database, which is unfortunately not open for scientific reuse and automatised data mining (Kumar et al.
+At the moment, \texttt{datelife}'s chronogram database is not very large, storing 231 chronograms up to the time the manuscript was written.
+This represents 5.79\% of the largest existing chronogram database, which is unfortunately not open for scientific reuse nor automatised data mining (Kumar et al.
 \protect\hyperlink{ref-Kumar2017}{2017}).
@@
 Other open repositories are not suitable for time of divergence data mining because they either do not store that data (i.e., TreeBASE), or their metadata does not have enough information to allow automatised chronogram searches (i.e., Dryad).
-
 A previous version of TimeTree's synthetic chronogram (Hedges et al.
 \protect\hyperlink{ref-Hedges2015}{2015}) was made available in OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
-This ensures that at least some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
+This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
+Thus, incorporation of published chronograms available in the Dryad data repository and in supplementary material of journals to \texttt{datelife}'s database is crucial to improve its services.
+Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by ``hand''.
+Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.
 
-Thus, incorporation of published chronograms available in the Dryad data repository and in supplementary material of journals to \texttt{datelife}'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would need to be determined by ``hand''.
-Consequently, we would like to emphasize on the importance for the scientific community of sharing chronogram data in repositories that require expert input and manual curation,~such as OToL's tree repository.
+Another potential concern comes from summary chronograms.
+We currently summarize all source chronogram data by default.
+Users can subset source data if they have reasons to favor some or one source chronogram over others.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there is no objective way to determine if an expert chronogram is better than other.
+Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
+Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
+It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk \protect\hyperlink{ref-schenk2016sec}{2016}).
+Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
+But this remains to be tested.
 
-We introduce different ways of summarizing this data, but are these summary chronograms reliable to study evolutionary patterns?
-The short answer is that we do not know to what extent the summary trees can be used for reanalyses and in which areas they can be used more reliably than others.
-Surely they might be useful for ecological analyses, but further studies should be conducted.
-
-As implemented now, we currently by default summarize all source data avilable.
-Users can subset source data if they have reasons to favor some or one source chronogram over others.
-Strictly speaking, a good chronogram reflects the real time of divergence accurately and precisely.
-To our knowledge there is no objective way to determine if an expert chronogram is better than other.
-Some criteria that have been put forward are the level of sampling and the numer of calibrations used.
-Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
-\texttt{datelife}'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.
-It has been shown with simulations that in analyses performed with bayesian inference methods, divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations (using secondary calibrations in the same way as primary calibrations are used Schenk \protect\hyperlink{ref-schenk2016sec}{2016}\protect\hyperlink{ref-schenk2016sec}{a}).
-But there are different ways to apply secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-This information should be included as metadata to facilitate subset of source chronograms.
-This will need to be manually curated as is done in OToL.
-
-As observed from the comparison of source chronograms in the example above, even chronograms obtained with primary data can be very different.
-There is some evidence that different chronograms will result in very different macroevolutionary histories inferred from them (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
-In other areas of research, such as ecological and conservation biology, it has been demonstrated that the presence of data on lineage divergence represents an improvement for testing alternative hypothesis, compared to testing in a total absence of such time data.
-However, we do not know how using wrong or different kinds of age data will affect the outcomes of ecological tests.
-More work has to be performed in this area.
+Furthermore, even chronograms obtained with primary data can be very different, as observed from the comparison of source chronograms in the Fringillidae example.
 A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms (e.g., angiosperms; Magallón et al.
 \protect\hyperlink{ref-magallon2015metacalibrated}{2015}).
-Comparison of available chronogram data for a wide range of organisms shown in the examples here suggest that this is a widespread phenomenon that requires further attention.
-In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
-However, this is just an assumption.
-Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used; support for nodes and ages; age confidence intervals).
-But we should also be able to choose on
+Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
+Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignement (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
+To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
+Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.
+Hence, summarizing chronograms might imply summarizing evolutionary hypothesis.
+This could be good from certain point of view, since it could help to get a single global evolutionary history for a lineage.
+But it could also be bad, since we could be loosing part of the evolutionary history that is only being reflected in some chronograms and not from the summary chronogram.
+Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history (Antonelli et al.
+\protect\hyperlink{ref-antonelli2017supersmart}{2017}).
+This will be implemented in future \texttt{datelife} versions.
 
-Even if all source chronograms have been generated by excellent standards and using same practices, the evolutionary context they depict might be very different.
-The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}).
-Summarizing trees implies summarizing evolutionary hypothesis.
-It is unsure if this is a good or bad thing.
-
-For all these reasons, the ideal would be to use fossil data as primary sources of calibrations.
-This will be an upscale for datelife, taking fossils that have already been curated as calibrations to date other trees.
-
-Improvements:
-* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
-(Schenk \protect\hyperlink{ref-Schenk2016}{2016}\protect\hyperlink{ref-Schenk2016}{b}).
-* bayesian congruification
-* topological congruification
+Synthetic or summary chronograms are being used more and more frequently to study evolutionary hypothesis.
+We introduce different ways of summarizing time of lineage divergence data, but, are any of these summary chronograms reliable to study evolutionary patterns?
+The short answer is that we do not know to what extent the summary trees should be used for reanalyses and if there bare areas of research where they can be used more reliably than others.
+There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
+In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents an improvement for testing alternative hypothesis using phylogenetic distance.
+It is uncertain, however, how using very different age data could affect the outcomes of these tests, and further research is necessary in this direction.
 
@@
 \texttt{datelife} allows an easy and fast obtention, as well as comparison of publicly available information
-on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the timeframe of different regions of the tree of life, allowing identification of regions that need more work or that have conflicting information.
-It can also summarize public data, or use it to generate new chronograms based on trees provided by the user.
-Both summary and newly generated trees have the potential to be useful to evaluate evolutionary hypothesis in different areas of research.
-How well this trees work for this purpose still needs to be tested.
-\texttt{datelife} helps with awereness on the variation of expert data, and the actual effect of the underlying data on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\texttt{datelife} is also part of the reproducibility of science movement.
-It strives to facilitate usage of public data sets as well as encouraging the making of information publicly available.
-Acknowledging source information and describing workflows and analyses.
-
-it allows exploration of alternative taxonomic and phylogenetic schemes;
-it allows rapid exploration of the effect of alternative divergence time hypothesis;
-it allows rapid synthesis in a number of different formats;
+on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, allowing identification of regions that require more research or that have conflicting information.
+Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
+How well this trees work for this purpose still needs to be further tested.
+If all, \texttt{datelife} helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 \leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk J.J. 2016a. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
-
-\leavevmode\hypertarget{ref-Schenk2016}{}%
-Schenk J.J. 2016b. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
+Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
 
@@
 
-Supplementary Material
-
 \end{document}

### 2019-06-03 Intro.Rmd
abstract

--- 2019-06-03 Intro.Rmd
@@
 ---
+`datelife`'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.
+
+Improvements:
+* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
+[@Schenk2016].
+* bayesian congruification
+* topological congruification
 

### 2019-06-20 Datelife_paper1.Rmd
tol plan and poster figs

@@
 Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
-The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
+The Branch Length Adjuster (BLADJ) algorithm [@
+2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
 

### 2019-06-20 Discussion.Rmd
tol plan and poster figs

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---
This information can be used to account for the effect of phylogenetic signal in
studies of trait evolution; to explore potential speciation and extinction dynamics
of interest within a clade; to obtain a time frame of biogeographical events; for
science communication and outreach, amongst others.

Problems and caveats:
`datelife` can be seen as an open resource to know the current state of knowledge on lineage divergence times.
Whether chronograms obtained using this original data can be used reliably to study complicated patterns of evolution is still uncertain.
If all, la facilidad para obtener hipotesis de tiempo de divergencia nos ayudará a evaluar la capacidad de los cronogramas para estudiar otros fenomenos evolutivos.
Por ahora, no podemos aseverar que estos cronogramas puedan usarse para todo tipo de analisis.

### 2019-06-20 Intro.Rmd
tol plan and poster figs

@@
 on the global, extant species numbers, which ranges from 3 to 100 million species
-[Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from
+[@Mayr2010; @Mora2011]). One advantage of TimeTree is that it includes taxa from
 across the tree of life, versus more specialized chronogram repositories focusing on land plants

### 2019-06-20 frin_plan.R

@@ -8,9 +8,26 @@ frin_plan <- drake_plan(
   phyloall = summarize_datelife_result(datelife_query = dq,
     datelife_result = dr, summary_format = "phylo_all",
     taxon_summary = "none"),
-  datedotol = get_dated_otol_induced_subtree(input = dq,
-    ott_ids = dq$ott_ids),
+  cb_cols1 = c("#0072B2", "#D55E00", "#440154FF", "#CC79A7", "#440154FF", "#009E73",
+    "#9ad0f3", "#95D840FF", "#F0E442"),
+  cb_cols2 = c("#440154FF", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00",
+    "#CC79A7", "#F0E442", "#95D840FF"),
+  ltt_plot = plot_ltt_phyloall(taxon = "Fringillidae", phy = phyloall, ltt_colors = cb_cols1, tax_datedotol = NULL,
+      file_name = "ltt_plot_poster.pdf", file_dir = getwd(), height = 3, width = 7.9, add_legend = FALSE, add_title = FALSE),
+  datedotol = get_dated_otol_induced_subtree(input = dq, ott_ids = dq$ott_ids),
   otol = get_otol_synthetic_tree(input = dq),
-  bestgrove = get_best_grove(datelife_result = dr)
+  bestgrove = get_best_grove(datelife_result = dr),
+  sdm_matrix = get_sdm_matrix(datelife_result = bestgrove$best_grove),
+  median_matrix = datelife_result_median_matrix(datelife_result = bestgrove$best_grove),
+  phy_sdm = summary_matrix_to_phylo_all(summ_matrix = sdm_matrix, target_tree = otol),
+  phy_median = summary_matrix_to_phylo_all(summ_matrix = median_matrix, target_tree = otol),
+  ltt_summ = plot_ltt_summary(taxon = "Fringillidae", phy = phyloall, phy_sdm, phy_median,
+        file_name = "ltt_summ_poster.pdf", file_dir = "~//Desktop//datelife_paper1//", height = 3, width = 7.9,
+        add_legend = TRUE),
+  each_calibrations = get_all_calibrations(phyloall, each = TRUE),
+  sim_otol = use_each_calibration(phy = otol, calibrations = each_calibrations)$phy,
+  ltt_simotol = plot_ltt_phyloall(taxon = "Fringillidae", sim_otol, summ,
[…]

### 2019-06-20 tol_plan.R

@@ -1,0 +1,18 @@
+library(datelife)
+library(drake)
+tol_plan <- drake_plan(
+  # class(opentree_chronograms$trees) <- "multiPhylo"
+  strings_in_dots = "literals",
+  biggest = which.max(unname(ape::Ntip(opentree_chronograms$trees))),
+  # names(opentree_chronograms$trees[[biggest]])
+  # names(opentree_chronograms$trees[biggest])  
+  tol_dq = make_datelife_query(input = opentree_chronograms$trees[[biggest]]$tip.label, get_spp_from_taxon = FALSE),
+  # tol_dr = get_datelife_result(input = tol_dq), # erroring, so tryCatch:
+  all_dr = lapply(opentree_chronograms$trees, function(x) tryCatch(datelife:::phylo_to_patristic_matrix(x),
+      error = function(e) NULL)), # biggest is the only one that error when going from phylo to patristic matrix.
+  tol_summ = get_taxon_summary(datelife_query = tol_dq, datelife_result = all_dr),
+  tol_otol = get_otol_synthetic_tree(ott_ids = opentree_chronograms$trees[[biggest]]$ott_ids),
+  all_calibs = get_all_calibrations(opentree_chronograms$trees[-93], each = FALSE),
+  tol_otol_sim = use_calibrations(phy = tol_otol, calibrations = all_calibs)
+)
+make(tol_plan)

### 2019-06-23 tol_plan.R
birds plan

@@ -1,18 +1,71 @@
 library(datelife)
 library(drake)
-tol_plan <- drake_plan(
-  # class(opentree_chronograms$trees) <- "multiPhylo"
+class(opentree_chronograms$trees) <- "multiPhylo"
+make_report <- function(rmdname, mdname, placeholder){
+  knitr::knit(knitr_in(rmdname), file_out(mdname), quiet = TRUE)
+}
+render_pdf <- function(reportname, dir, placeholder) {
+    original.dir <- getwd()
+    setwd(dir)
+    system(paste0('pandoc ', paste0(reportname, '.md'), ' -o ', paste0(reportname, '.pdf --pdf-engine=xelatex -V mainfonts="DejaVu Sans"')))
+    setwd(original.dir)
+    # pandoc -o emoji.pdf --pdf-engine=lualatex -V mainfonts="DejaVu Sans"
+    # pandoc -o emoji.pdf --pdf-engine=xelatex  -V mainfonts="DejaVu Sans"
+}
+# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
+# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
+
+tol_query <- drake_plan(
   strings_in_dots = "literals",
   biggest = which.max(unname(ape::Ntip(opentree_chronograms$trees))),
   # names(opentree_chronograms$trees[[biggest]])
-  # names(opentree_chronograms$trees[biggest])  
-  tol_dq = make_datelife_query(input = opentree_chronograms$trees[[biggest]]$tip.label, get_spp_from_taxon = FALSE),
-  # tol_dr = get_datelife_result(input = tol_dq), # erroring, so tryCatch:
-  all_dr = lapply(opentree_chronograms$trees, function(x) tryCatch(datelife:::phylo_to_patristic_matrix(x),
-      error = function(e) NULL)), # biggest is the only one that error when going from phylo to patristic matrix.
+  # names(opentree_chronograms$trees[biggest])
+  # tol_dq = make_datelife_query(input = opentree_chronograms$trees[[biggest]], get_spp_from_taxon = FALSE),
[…]

### 2019-06-23 table0_pdf_template.Rmd
birds plan

---
title: "Datelife Reports"
author: "Luna L. Sanchez Reyes"
date: "`r Sys.Date()`"
output: bookdown::pdf_document2
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
header-includes:
- \usepackage{booktabs}
- \usepackage{makecell}
- \usepackage{multirow}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{array}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{xcolor}
geometry: "left=2.5cm,right=2.5cm,top=2cm,bottom=2cm"
vignette: >
  %\VignetteIndexEntry{DateLife Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---
```{r setup, include = FALSE}
# print(getwd())
# source("report_template_setup.R")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
my_table_format <- function(tt){
  res <- format.data.frame(tt, digits = 4, nsmall = 4, justify = "none")
  return(res)
}
make_table0 <- function(tax_dr, tax_summ, tax_dq, table_caption = "", label = "source_chr"){
    # to determine taxon number, take only the first chronogram from the same publication
    # the most common case is that all chronograms in the same publication have the same species sampling
    # but there are cases in which this is not the case (e.g., De-Nova et al. 2018)
    # and there are several chronograms in the same publication that have different taxon sampling
    # a way to incorporate them would be putting the taxon presence of all source chronograms from the same publication in the same vector, separated by "\n"
    # for now, we will just show the maximum number of sampled species in study
    nn <- sort(unique(names(tax_dr)))
    # taxon_number_all <- sapply(seq(nrow(tax_summ$matrix)), function(x) sum(tax_summ$matrix[x,]))
    # the line above is replaced by tax_summ$summary2$taxon_number
    taxon_number <- unname(sapply(nn, function(x) max(tax_summ$summary2$taxon_number[names(tax_dr) %in% x])))
    xx <- table(names(tax_dr))
    table0 <- data.frame(var1 = paste0(seq(length(xx)), "."),
        var2 = names(xx), var3 = as.numeric(unname(xx)),
        var4 = paste0(taxon_number, "/", length(tax_dq$cleaned_names)))
    table0 <- dplyr::mutate(table0, var1 = kableExtra::cell_spec(var1, "latex", align = "r",
        bold = TRUE, italic = TRUE),
        var2 = kableExtra::cell_spec(var2, "latex", font_size = 8),
        var3 = kableExtra::cell_spec(var3, "latex", align = "c"),
        var4 = kableExtra::cell_spec(var4, "latex", align = "c"))
    t0 <- knitr::kable(table0, escape = FALSE, row.names = FALSE, format = "latex",
        booktabs = TRUE, caption = table_caption, label = label, linesep = "", longtable = TRUE)
    t0 <- gsub("var1", "   ", t0)
    t0 <- gsub("var2", "Citation", t0)
    t0 <- gsub("var3", "Source N", t0)
    t0 <- gsub("var4", "Taxon N", t0)
    kableExtra::kable_styling(kable_input = t0, latex_options = "hold_position") %>%
    column_spec(1, width = "0.4cm") %>%
    column_spec(2, width = "11cm") %>%
    column_spec(3, width = "1.5cm") %>%
    column_spec(4, width = "1.8cm") %>%
    # collapse_rows(columns = 3, latex_hline = "none", valign = "middle") %>%
    row_spec(0, bold = TRUE, align = "c", italic = TRUE, font_size = 9) %>%
    footnote(general = c("Source_N: Number of source chronograms reported in study.",
            "Taxon_N: Number of queried taxa found in source chronograms."),
            general_title = "") -> t0
    t0 <- gsub("Source\\\\_N", "{\\\\textbf{\\\\textit{Source N}}}", t0)
    t0 <- gsub("Taxon\\\\_N", "{\\\\textbf{\\\\textit{Taxon N}}}", t0)
    return(t0)
}
library(magrittr)
loadd(tol_dq)
loadd(tol_dr)
loadd(tol_summ)
loadd(table0_caption)
loadd(taxon)
```

# Datelife result: `r taxon`

```{r echo = FALSE, warning=FALSE, message = FALSE, results = 'asis'}
make_table0(tol_dr, tol_summ, tol_dq, table0_caption)
```

### 2019-06-23 Rebe_Net_diversification.R

@@ -1,0 +1,342 @@
+##Estimación de tasas de diversificación con el método de momentos usando las ecuaciones de 
+##Magallón y Sanderson 2001 aplicado al árbol fechado FBD80 de Malvales.
+install.packages("geiger")
+require ("geiger")
+getwd()
+##################################################
+#2 de febrero de 2018
+#corregí las edades de cada grupo
+#voy a obtener la diversificación neta con el método de 
+#momentos.
+#las edades son del árbol FBD80_v4_comb.treeanot.tree
+data<-read.csv("CrownAges_species_richness.csv")
+data
+class(data)
+##diversificación neta con el comando bd.ms
+#nombre del taxón_r(diversificación neta)_epsilon_c(crown)/s(stem)
+malvoi_r_001_c<-bd.ms(time=data$time[1], n=data$sp[1], epsilon=0.01)
+malvoi_r_001_c
+#voy a tratar de hacer este cálculo para cada familia y subfamilia con un
+#loop:
+#Primero voy a crear un objeto en el que se deben imprimir los resultados
+#del loop
+net_div<-0
+net_div
+#loop
+for(i in 1:19){
+  net_div[i] <- bd.ms(time=data$time[i], n=data$sp[i],epsilon=0.01)
+  print(net_div[i])
+}
+
+print(i)
+net_div
+data_net_div<-data.frame(data$taxon, net_div)
+data_net_div
+#Hecho
+###############################################################################
+#Ahora voy a calcular r (diversificación neta) con el valor de turnoverFBD
+#que estimó el modelo: 0.9195
+net_div_0.91<-0
+for(i in 1:19){
+	net_div_0.91[i]<-bd.ms(time=data$time[i], n=data$sp[i],epsilon=0.9195)
+	print (net_div_0.91[i])
+}
+net_div_e0.91<-data.frame(data$taxon, net_div_0.91)
+net_div_e0.91
+colnames(net_div_e0.91)<- c("taxon", "crown_e0.91")
+##############################################################################
[…]

### 2019-06-23 tol_dating.R

@@ -1,0 +1,105 @@
+# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
+# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
+loadd(tol_jetz1)
+tol_jetz_query <- drake_plan(
+  tol_jetz_dq = make_datelife_query(input = tol_jetz1, get_spp_from_taxon = FALSE),
+  tol_jetz_dr = get_datelife_result(input = tol_dq),
+  tol_summ = get_taxon_summary(datelife_query = tol_dq, datelife_result = tol_dr),
+  # for report table only:
+  tol_dq = tol_jetz_dq,
+  tol_dr = tol_jetz_dr,
+  tol_summ = tol_jetz_summ,
+  reportname = "Jetz_dr_table0",
+  report = make_report("table0_pdf_template.Rmd", paste0("inst/docs/", reportname, ".md"), "try5"),
+  table0_caption = "Jetz et al. 2012 source chronogram original studies information.",
+  taxon = "Birds",
+  table0_pdf = render_pdf(reportname, "inst/docs", "try5")
+)
+make(tol_jetz_query)
+
+load("data/tol_jetz1.RData")
+loadd(tol_dr)
+loadd(tol_jetz2)
+jetz_phylo_all <- vector(mode = "list", length = length(tol_dr))
+is_jetz <- grepl("Jetz", names(tol_dr))
+is_biggest <- which(grepl("Hedges", names(tol_dr)))
+jetz_phylo_all[[which(is_jetz)[1]]] <- tol_jetz1[[1]]
+jetz_phylo_all[[which(is_jetz)[2]]] <- tol_jetz2[[1]]
+
+for(i in seq(jetz_phylo_all)[-is_biggest]){
+  if(!is_jetz[i]){
+    print(i)
+    jetz_phylo_all[[i]] <- datelife::patristic_matrix_to_phylo(tol_dr[[i]])
+  }
+}
+save(jetz_phylo_all, file = "data/jetz_phylo_all.RData")
+
+# the step that takes a lot of time is going from patristic matrix to phylo
+# then get all calibrations for each tree, in a loop, so we save the information
[…]

### 2019-06-23 tol_ott_ids.R

@@ -1,0 +1,39 @@
+# GET OTT IDS FOR JETZ ET AL TREE
+load("data/new.tree_tnrs.RData")
+names(new.tree_tnrs)
+nrow(new.tree_tnrs)
+loadd(tol_jetz1)
+head(new.tree_tnrs)
+head(tol_jetz1[[1]]$tip.label)
+mm_unique <- match(tol_jetz1[[1]]$tip.label, new.tree_tnrs$unique) # position of tip labels in unique
+mm_search <- match(tolower(gsub(" ", "_", tol_jetz1[[1]]$tip.label)), new.tree_tnrs$search)
+sum(is.na(mm_unique)) # 387
+sum(is.na(mm_search))# 436, mm_unique and mm_search match different bird tip labels
+# make a vector of indices of matching tip labels with tnrs table:
+jetz_ott_ids_ii <- rep(0, length(tol_jetz1[[1]]$tip.label))
+jetz_ott_ids_ii[!is.na(mm_unique)] <- mm_unique[!is.na(mm_unique)]
+sum(jetz_ott_ids_ii == 0) # 387, ok
+# those that are not in unique but are on search
+# are the ones that have a different name in ott taxonomy
+ii <- jetz_ott_ids_ii ==0 & !is.na(mm_search)
+sum(ii) # 114
+jetz_ott_ids_ii[ii] <- mm_search[ii]
+sum(jetz_ott_ids_ii == 0) # 273, these are the ones that are not in search nor unique, so search!
+ss <- data.frame(jetz = tol_jetz1[[1]]$tip.label[ii],
+  search = new.tree_tnrs$search[mm_search[ii]])
+  new.tree_tnrs[mm_search[ii],]
+dd <- data.frame(jetz = tol_jetz1[[1]]$tip.label[jetz_ott_ids_ii != 0], search = new.tree_tnrs$search[jetz_ott_ids_ii], unique = new.tree_tnrs$unique[jetz_ott_ids_ii])
+
+jetz_ott_ids <- rep(NA, length(tol_jetz1[[1]]$tip.label))
+jetz_ott_ids[jetz_ott_ids_ii != 0] <- new.tree_tnrs$ott_id[jetz_ott_ids_ii]
+tol_jetz1[[1]]$ott_ids <- as.numeric(jetz_ott_ids)
+new_search <- tol_jetz1[[1]]$tip.label[is.na(tol_jetz1[[1]]$ott_ids)]
[…]

### 2019-06-23 tol_rates.R

@@ -1,0 +1,50 @@
+# hypothetical example
+# phy <- tol_jetz1[[1]]
+# age_and_richness <- jetz_fam_ages_and_richness
+get_all_ms <- function(phy, age_and_richness, epsilon = 0){
+  globalr = geiger::bd.ms(phy= NULL, time = max(ape::branching.times(phy)), n = sum(age_and_richness$richness), missing = 0, crown=TRUE, epsilon = epsilon)
+# c is for crown, global is for global rate, clade is for clade rate, ll is for limits
+    cp <-  geiger::crown.p(phy=NULL, time = age_and_richness$crown_age, n = age_and_richness$richness,
+      r = globalr, epsilon = epsilon)
+    cbd <- geiger::bd.ms(phy=NULL, time = age_and_richness$crown_age, n = age_and_richness$richness,
+      missing = 0, crown = TRUE, epsilon = epsilon)
+
+    cglobalr_ll <- lapply(age_and_richness$crown_age, function(x) tryCatch(
+      as.data.frame(geiger::crown.limits(time = x, r = globalr, epsilon, CI=0.95)),
+      error = function(e) data.frame(lb = NA, ub = NA)))
+    cglobalr_ll <- data.table::rbindlist(cglobalr_ll)
+
+    cclader_ll <- lapply(seq(age_and_richness$crown_age), function(i) tryCatch(
+      as.data.frame(geiger::crown.limits(time = age_and_richness$crown_age[i], r = cbd[i], epsilon, CI=0.95)),
+      error = function(e) data.frame(lb = NA, ub = NA)))
+    cclader_ll <- data.table::rbindlist(cclader_ll)
+
+    all_crown <- data.frame(crown.p = cp, bd.ms = cbd, cglobalr_ll, cclader_ll, age_and_richness$crown_age, age_and_richness$richness)
+    names(all_crown) <- c("crown.p", "bd.ms", "lb_global", "ub_global", "lb_clade", "ub_clade", "crown_age", "richness")
+    row.names(all_crown) <- row.names(age_and_richness)
+
+    sp <-  geiger::stem.p(phy=NULL, time = age_and_richness$stem_age, n = age_and_richness$richness,
+      r = globalr, epsilon = epsilon)
+    sbd <- geiger::bd.ms(phy=NULL, time = age_and_richness$stem_age, n = age_and_richness$richness,
[…]

### 2019-06-23 tol_rates_plots.R

@@ -1,0 +1,16 @@
+ms <- jetz_ms0
+ms <- jetz2012_ms0
+ms <- jetz2012_min_ms0
+ms <- jetz2012_mean_ms0
+ms <- jetz2012_max_ms0
+plot_ms <- function(ms, file_name = NULL, height = 3.5, width = 5){
+  ii <- !is.na(ms$crown$crown.p)
+  cc <- ms$crown$richness[ii] < ms$crown$lb_global[ii] | ms$crown$richness[ii] > ms$crown$ub_global[ii]
+
+  plot(x = ms$crown$crown_age[ii], y = ms$crown$richness[ii], xlab = "Crown age (MYA)",
+  ylab = "log(species number)", col = ifelse(cc, "orchid", "gray") )
+  lines(x = sort(ms$crown$crown_age[ii]), y = ms$crown$lb_global[ii][order(ms$crown$crown_age[ii])], type = "l", col = "grey")
+  lines(x = sort(ms$crown$crown_age[ii]), y = ms$crown$ub_global[ii][order(ms$crown$crown_age[ii])], type = "l", col = "grey")
+
+  hist(ms$crown$crown.p)
+}

### 2019-06-23 tol_taxonomies.R

@@ -1,0 +1,101 @@
+load("data/tol_jetz1.RData")
+
+tol_jetz1[[1]]$ott_id
+which(is.na(tol_jetz1[[1]]$ott_id)) # 17 spp.
+# get ott Families
+jetz_ff <- get_ott_clade(ott_ids = tol_jetz1[[1]]$ott_id, ott_rank = "family")
+save(jetz_ff, file = "~/Desktop/datelife_paper1/data/jetz_ff.RData")
+length(ff$input) == length(tol_jetz1[[1]]$tip.label)
+ff$input[is.na(ff$family)] # tip labels that did not match a family (there are 78)
+# plus the 17 spp that are not in ott gives a total of 95 tips with no family data
+# these will appear as NA in the family vector:
+tol_jetz1[[1]]$ott_family <- ff$family[match(tol_jetz1[[1]]$ott_id, ff$input)]
+save(tol_jetz1, file = "~/Desktop/datelife_paper1/data/tol_jetz1.RData")
+
+# get jetz study Families
+tt <- read.csv("data-raw/nature11631-s2/2012-03-04206D-master_taxonomy.csv")
+nrow(tt)
+new.tree <- rotl::get_study_tree(study_id= "ot_809",tree_id="tree1",
+								tip_label="original_label", deduplicate = TRUE)
+tol_jetz1[[1]]$original.tip.label <- new.tree$tip.label # they are in the same order, I checked it with match
+save(tol_jetz1, file = "~/Desktop/datelife_paper1/data/tol_jetz1.RData")
+
+head(tt)
+mm <- match(tol_jetz1[[1]]$original.tip.label, as.character(tt$TipLabel))
+tol_jetz1[[1]]$jetz2012_family <- as.character(tt$BLFamilyLatin[mm])
+richness <- sapply(unique(as.character(tol_jetz1[[1]]$jetz2012_family)),
+  function(x) sum(as.character(tt$BLFamilyLatin) %in% x))
+mm <- match(tol_jetz1[[1]]$jetz2012_family, names(richness))
+tol_jetz1[[1]]$jetz2012_richness <- richness[mm]
+tail(data.frame(fam = tol_jetz1[[1]]$jetz2012_family, rich = tol_jetz1[[1]]$jetz2012_richness))
+save(tol_jetz1, file = "~/Desktop/datelife_paper1/data/tol_jetz1.RData")
+
+# once with the families, get the crown and stem ages of each Family:
[…]

### 2019-06-23 tol_topologies.R

@@ -1,0 +1,23 @@
+# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
+# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
+tol_topologies <- drake_plan(
+  biggest = which.max(unname(ape::Ntip(opentree_chronograms$trees))),
+  tol_biggest = opentree_chronograms$trees[[biggest]], #tol_biggest$node.label
+  plot_biggest = plot_ltt_phyloall(taxon = "Biggest", phy = tol_biggest, ltt_colors = "#ff0000", tax_datedotol = NULL,
+    file_name = "_lttplot.pdf", file_dir = getwd(), height = 3.5, width = 7, add_legend = FALSE,
+    add_title = TRUE, title_text = " chronogram in datelife database"),
+  # tol_otol_biggest = get_otol_synthetic_tree(input = tol_dq)
+  # tol_otol_biggest = get_otol_synthetic_tree(ott_ids = opentree_chronograms$trees[[biggest]]$ott_ids),
+  # tol_otol_fam =
+  tol_jetz1 = opentree_chronograms$trees[grepl("jetz", tolower(names(opentree_chronograms$trees)))][1],
+  tol_jetz2 = opentree_chronograms$trees[grepl("jetz", tolower(names(opentree_chronograms$trees)))][2],
+  plot_jetz1 = plot_ltt_phyloall(taxon = "Jetz chronogram 1", phy = tol_jetz1,
+    ltt_colors = "#ff0000", tax_datedotol = NULL, file_name = "_lttplot.pdf",
+    file_dir = getwd(), height = 3.5, width = 7, add_legend = FALSE,
+    add_title = TRUE, title_text = " chronogram in datelife database"),
+  plot_jetz2 = plot_ltt_phyloall(taxon = "Jetz chronogram 2", phy = tol_jetz2,
+    ltt_colors = "#ff0000", tax_datedotol = NULL, file_name = "_lttplot.pdf",
+    file_dir = getwd(), height = 3.5, width = 7, add_legend = FALSE,
+    add_title = TRUE, title_text = " chronogram in datelife database")
+)
+make(tol_topologies)

### 2019-08-06 Datelife_paper1.Rmd
Brian comments / modifications

@@
 of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
-of Life, for a large quantity and diversity of organisms in the last decades.
-However, time of lineage divergence data remains difficult to be obtained and synthesized for many biologists and the non-academic community, despite its importance in many areas of research, education and science communication.
+of Life for a large quantity and diversity of organisms in the last few decades.
+However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication.
 `datelife` is a service implemented via an R package and a web site
-(<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of expert, published data on time of lineage divergence.
+(<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of published data on lineage divergence times.
 The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
 A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
-Source chronogram data can be summarised using branch length summary statitiscs or variance minimising approaches to generate a single summary chronogram.
+Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram.
 Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
@@
 \newpage
-Time of lineage divergence constitutes a fundamental piece of information for evolutionary
+Clade ages represent a fundamental piece of information for evolutionary
 understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014].
-The primary information needed for time of lineage divergence estimation comes from the fossil record.
-Coupled to molecular phylogenies (and more recently to morphological data), the time of divergence of extant and extinct lineages is reconstructed with molecular dating methods.
-Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
-Still, generating a chronogram is not an easy task unless you have specialized training.
-That's why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
+The primary information needed these time estimates comes from the fossil record.
+Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
+The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
+Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree.
+That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
 
@@
 TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] platforms store and make available published phylogenies and chronograms for easy scientific reuse.
-All of them can be queried using automatised web procedures, which permit personalised, large scale queries that are also very fast.
+All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
 OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
+Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
 
@@
 <!-- let alone time of divergence data from their stored chronograms. -->
-The Timetree of Life project focuses on the synthesis of a single chronogram of life [@Hedges2006].
-However, the thousands of chronograms they have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
+The Timetree of Life project focuses on the synthesis of a single chronogram of life [@Hedges2006] and presents a very accessible, attractive interface.
+However, the thousands of chronograms this NSF-funded project have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but large scale download remains prohibited by their site.
 The latest version of their synthetic chronogram [@Kumar2017] can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
@@
 However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Moreover, these tools do not provide information from already existent expert chronograms.
+Moreover, these tools do not provide information from already created expert chronograms.
 
@@
 <!-- as well as tools for comparison of source and summary trees.  -->
-R packages for benchmarking of functionalities and demonstrating services were also developed and made available.
+<!-- R packages for benchmarking of functionalities and demonstrating services were also developed and made available.>
 
@@
 The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database; source trees with at least two matching input names are identified; all other taxa that do not match the original query are then dropped from the positively identified source trees --these pruned chronograms are hereafter referred as source chronograms; finally, each source chronogram is transformed to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarising algorithms.
-\item  The user can obtain different summary information from the source chronograms including: a) all source chronogram ages, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, and f) a report of succesful matches of input taxon names across source chronograms.
+\item A name search is then performed across the chronogram database; source trees with at least two matching input names are identified; all other taxa that do not match the original query are then dropped from the positively identified source trees.
+These pruned chronograms are hereafter referred as source chronograms.
+Finally, each source chronogram is transformed to a patristic matrix named by the citation of the original study.
+This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
+\item  The user can obtain different summary information from the source chronograms including: a) all source chronograms, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, f) a report of succesful matches of input taxon names across source chronograms, and g) the single source chronogram with the greatest number of taxa.
 Summary information can be used to make decisions on the next steps of the workflow.
 \item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also simulate both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
+\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
 \item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R `phylo` format), as well as view and compare results graphically, or construct their own graphs using inbuilt `datelife` graphic generation functions.
@@
 variations and synonyms <!--to a taxonomic reference of users choosing, default
-to otol taxonomy -->, increasing the probability to correctly find the
+to otol taxonomy -->, increasing the probability of correctly finding the
 queried taxa in `datelife`'s chronogram database.
@@
 Information from source chronograms can be summarised with a summary statistic of tree branch
-lengths (such as median or mean) or using the Super Distance Matrix (SDM) approach for supertree reconstruction [@Criscuolo2006]. <!-- and the hierarchical clustering method of ttol-->
-The resulting summary patristic distance matrix could be clustered with classic algorithms.
+lengths (such as median or mean). <!-- and the hierarchical clustering method of ttol-->
+The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
 However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package).
@@
 2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
+
+A  much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006], implemented in the ape package [@Paradis2004].
 
@@
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are coherent with source ages.
+In both cases, ages are generally consistent with source ages.
 <!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
 <!-- and hierarchical clustering method which appears to push ages back -->
-It is not certain if these chronograms should be used to perform all types of downstream evolutionary analyses and more research is needed in that direction.
+Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms (_____MANY CITATIONS____)).
+For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results (Title and Rabosky [https://academic.oup.com/sysbio/article/66/5/843/2682290] ).
+Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.
 
@@
 At the moment, `datelife`'s chronogram database is not very large, storing `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written.
-This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is unfortunately not open for scientific reuse nor automatised data mining [@Kumar2017].
+This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining [@Kumar2017].
 OToL is the only public tree repository from where `datelife` can currently pull chronograms to construct its database.
-Other open repositories are not suitable for time of divergence data mining because they either do not store that data (i.e., TreeBASE), or their metadata does not have enough information to allow automatised chronogram searches (i.e., Dryad).
-A previous version of TimeTree's synthetic chronogram [@Hedges2015] was made available in OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
+A previous version of TimeTree's synthetic chronogram [@Hedges2015] was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
 This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
@@
 Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
-Thus, incorporation of published chronograms available in the Dryad data repository and in supplementary material of journals to `datelife`'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by "hand".
+Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to `datelife`'s database is crucial to improve its services.
+Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
 <!-- What can we do to mine journal databases? -->
@@
 Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-But this remains to be tested.
+This remains to be tested.
 
@@
 -->
-Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignement (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
+Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignment (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
 To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
@@
 
-Synthetic or summary chronograms are being used more and more frequently to study evolutionary hypothesis.
-We introduce different ways of summarizing time of lineage divergence data, but, are any of these summary chronograms reliable to study evolutionary patterns?
-The short answer is that we do not know to what extent the summary trees should be used for reanalyses and if there bare areas of research where they can be used more reliably than others.
+We allow accepted ways of creating branch lengths in the absence of starting branch length information for some taxa, such as BLADJ [@ 2008].
+Making up branch lengths in this or other ways is accepted in scientific publications: Rabosky et al.
+[https://www.nature.com/articles/s41586-018-0273-1] created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; Jetz et al.
+[https://www.nature.com/articles/nature11631], created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data; Smith and Brown [https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1019] constructed a tree of 353,185 seed plants where only 23% had molecular data.
+Taken to the extreme, one could make a fully resolved,  calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
+There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.
+
 There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
@@
 
-Time of lineage divergence information is key to many areas of evolutionary studies: trait evolution,
+Divergence time information is key to many areas of evolutionary studies: trait evolution,
 diversification, biogeography, macroecology and more. Generating this information is difficult,
@@
 do not have the time to acquire and develop the necessary knowledge and data curation skills
-to produce chronograms _de novo_. Knowledge on taxon ages is also crucial for
+to produce chronograms _de novo_. Knowledge on clade ages is also crucial for
 science communication and education.
@@
 Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
-How well this trees work for this purpose still needs to be further tested.
-If all, `datelife` helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+`datelife` helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 

### 2019-09-10 tol_dating.R
evolution 2019 poster stuff

@@ -32,8 +32,16 @@ for(i in seq(jetz_phylo_all)[-is_biggest]){
     jetz_phylo_all[[i]] <- datelife::patristic_matrix_to_phylo(tol_dr[[i]])
   }
 }
+names(jetz_phylo_all) <- names(tol_dr)
+class(jetz_phylo_all) <- "multiPhylo"
 save(jetz_phylo_all, file = "data/jetz_phylo_all.RData")
-
+jetz_phylo_all_forplot <- jetz_phylo_all[-c(is_biggest, which(is_jetz))]
+library(RColorBrewer)
+brewer.pal(9, "RdPu")
+colfunc <- colorRampPalette(brewer.pal(9, "RdPu"))
+ltt_colors_here <- sample(colfunc(length(jetz_phylo_all_forplot)), size = length(jetz_phylo_all_forplot))
+datelife::plot_ltt_phyloall(taxon = "Aves", phy = jetz_phylo_all_forplot, study_number_cex = 0.25,
+  lwd_arrows = 1, ltt_colors = ltt_colors_here)
 # the step that takes a lot of time is going from patristic matrix to phylo
 # then get all calibrations for each tree, in a loop, so we save the information
 jetz_all_calibrations <- vector(mode = "list", length = length(tol_dr))
@@ -90,6 +98,12 @@ tol_jetz_chronograms <- drake_plan(
 make(tol_jetz_chronograms)
 
 # run rates analysis and plot
+
+jetz_ms0 <- get_all_ms(phy = tol_jetz1[[1]], age_and_richness = jetz_fam_ages_and_richness, epsilon = 0)
+load("data/jetz_jetz2012fam_age_and_rich.RData")
+jetz2012_ms0 <- get_all_ms(phy = tol_jetz1[[1]], age_and_richness = jetz_jetz2012fam_age_and_rich, epsilon = 0)
+
+
 tip_fams <- tol_jetz1[[1]]$jetz2012_family
 tip_rich <- tol_jetz1[[1]]$jetz2012_richness
 jetz_jetz2012_min_fam_age_and_rich <- get_clade_age_and_rich(phy = tol_jetz1_bladj_min,
@@ -103,3 +117,41 @@ jetz2012_mean_ms0 <- get_all_ms(phy = tol_jetz1_bladj_mean, age_and_richness = j
[…]

### 2019-09-10 tol_rates.R

@@ -44,7 +44,3 @@ get_all_ms <- function(phy, age_and_richness, epsilon = 0){
 
     return(list(crown = all_crown, stem = all_stem))
 }
-
-
-jetz_ms0 <- get_all_ms(phy = tol_jetz1[[1]], age_and_richness = jetz_fam_ages_and_richness, epsilon = 0)
-jetz2012_ms0 <- get_all_ms(phy = tol_jetz1[[1]], age_and_richness = jetz_jetz2012fam_age_and_rich, epsilon = 0)

### 2019-09-10 tol_rates_plots.R

@@ -3,14 +3,79 @@ ms <- jetz2012_ms0
 ms <- jetz2012_min_ms0
 ms <- jetz2012_mean_ms0
 ms <- jetz2012_max_ms0
-plot_ms <- function(ms, file_name = NULL, height = 3.5, width = 5){
+rm(ms)
+plot_ms <- function(ms, file_dir = NULL, file_name = NULL, height = 3.5, width = 5,
+  write_pdf = TRUE, ylab = "Species N", pch = 19, col_significant = NULL, ...){
+  if(!inherits(file_dir, "character")){
+    file_dir <- getwd()
+  }
+  if(!inherits(file_name, "character")){
+    file_name <- "_ms_plot"
+  }
+  file_out <- paste0(file_dir, "//", file_name)
+  message(file_out)
+  if(write_pdf){
+    grDevices::pdf(file = paste0(file_out, "_crown.pdf"), height = height, width = width)
+    graphics::par(mai = c(1.02, 0.82, 0.2, 0.2))
+  }
   ii <- !is.na(ms$crown$crown.p)
   cc <- ms$crown$richness[ii] < ms$crown$lb_global[ii] | ms$crown$richness[ii] > ms$crown$ub_global[ii]
-
   plot(x = ms$crown$crown_age[ii], y = ms$crown$richness[ii], xlab = "Crown age (MYA)",
-  ylab = "log(species number)", col = ifelse(cc, "orchid", "gray") )
-  lines(x = sort(ms$crown$crown_age[ii]), y = ms$crown$lb_global[ii][order(ms$crown$crown_age[ii])], type = "l", col = "grey")
-  lines(x = sort(ms$crown$crown_age[ii]), y = ms$crown$ub_global[ii][order(ms$crown$crown_age[ii])], type = "l", col = "grey")
-
-  hist(ms$crown$crown.p)
+    ylab = ylab, col = ifelse(cc, paste0(col_significant, "80"), "#80808080"), pch = pch)
+  lines(x = sort(ms$crown$crown_age[ii]), y = ms$crown$lb_global[ii][order(ms$crown$crown_age[ii])],
+    type = "l", col = "grey", ...)
+  lines(x = sort(ms$crown$crown_age[ii]), y = ms$crown$ub_global[ii][order(ms$crown$crown_age[ii])],
+    type = "l", col = "grey", ...)
[…]

### 2019-09-29 biorxiv_2019.09.28.Rmd
biorxiv versions

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
    includes:
      # in_header: preamble.tex
      after_body: [biorxiv_2019.09.28_figures.tex]
    # fig_caption: no
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'm gonna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage{caption}
    - \usepackage{blindtext}
    - \usepackage[left]{lineno}
    - \linenumbers
  # - \usepackage{xr}
    # - \usepackage[utf8]{inputenc}
    # - \usepackage[english]{babel}
    # - \usepackage[document]{ragged2e}
nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---
<!-- \begin{\nolinenumbers} -->
<!-- Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE <!--MAX 50 CHARACTERS, ALL CAPITALS-->

**Title**: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life

**Authors**: Luna L. Sánchez-Reyes^1^, Brian C. O’Meara^1^

**Correspondence address**:

1. *Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA*

**Corresponding authors**: sanchez.reyes.luna@gmail.com, bomeara@utk.edu
<!-- \end{nolinenumbers} -->

``` {r child = 'paper.Rmd'}
```
\newpage
\begin{center}
\textsc{References}
\end{center}

### 2019-09-29 biorxiv_2019.09.28.tex
biorxiv versions

--- 2019-06-03 Datelife_paper1.tex
@@
 
+\usepackage{caption}
+\usepackage{blindtext}
 \usepackage[left]{lineno}
@@
 
-Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE
+\textbf{Title}: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
 
-Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
+\textbf{Authors}: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}
 
-Authors: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}
-
-Correspondence address:
+\textbf{Correspondence address}:
 
@@
 
-Corresponding authors: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
+\textbf{Corresponding authors}: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
 
@@
 of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
-of Life, for a large quantity and diversity of organisms in the last decades.
-However, time of lineage divergence data remains difficult to be obtained and synthesized for many biologists and the non-academic community, despite its importance in many areas of research, education and science communication.
+of Life for a large quantity and diversity of organisms in the last few decades.
+However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication.
 \texttt{datelife} is a service implemented via an R package and a web site
-(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of expert, published data on time of lineage divergence.
+(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of published data on lineage divergence times.
 The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
 A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
-Source chronogram data can be summarised using branch length summary statitiscs or variance minimising approaches to generate a single summary chronogram.
+Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram.
 Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
@@
 All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
+Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
 How well this trees work for this purpose still needs to be tested.
@@
 \textbf{Keywords:} Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
-
 \newpage
 
-Time of lineage divergence constitutes a fundamental piece of information for evolutionary
+\begin{center}
+\textsc{Introduction}
+\end{center}
+
+Clade ages represent a fundamental piece of information for evolutionary
 understanding in many areas of research, from developmental to conservation biology (Felsenstein \protect\hyperlink{ref-Felsenstein1985a}{1985}; Webb \protect\hyperlink{ref-Webb2000}{2000}), from historical biogeography to species diversification studies (Posadas et al.
 \protect\hyperlink{ref-posadas2006historical}{2006}; Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-The primary information needed for time of lineage divergence estimation comes from the fossil record.
-Coupled to molecular phylogenies (and more recently to morphological data), the time of divergence of extant and extinct lineages is reconstructed with molecular dating methods.
-Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al.
+The primary information needed for these time estimates comes from the fossil record.
+Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
+The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al.
 \protect\hyperlink{ref-Kumar2017}{2017}).
-Still, generating a chronogram is not an easy task unless you have specialized training.
-That's why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al.
+Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree.
+That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al.
 \protect\hyperlink{ref-Stoltzfus2013}{2013}).
@@
 \protect\hyperlink{ref-Hinchliff2015}{2015}) platforms store and make available published phylogenies and chronograms for easy scientific reuse.
-All of them can be queried using automatised web procedures, which permit personalised, large scale queries that are also very fast.
+All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
 OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
+Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
 
@@
 The Timetree of Life project focuses on the synthesis of a single chronogram of life (Hedges et al.
-\protect\hyperlink{ref-Hedges2006}{2006}).
-However, the thousands of chronograms they have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
+\protect\hyperlink{ref-Hedges2006}{2006}) and presents a very accessible, attractive interface.
+However, the thousands of chronograms this NSF-funded project have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but large scale download remains prohibited by their site.
 The latest version of their synthetic chronogram (Kumar et al.
@@
 We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
-R packages for benchmarking of functionalities and demonstrating services were also developed and made available.
 
@@
 which corrects potentially misspelled names and typos, and standardizes spelling
-variations and synonyms , increasing the probability to correctly find the
+variations and synonyms , increasing the probability of correctly finding the
 queried taxa in \texttt{datelife}'s chronogram database.
@@
 
-Information from source chronograms can be summarised with a summary statistic of tree branch
-lengths (such as median or mean) or using the Super Distance Matrix (SDM) approach for supertree reconstruction (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}).
-The resulting summary patristic distance matrix could be clustered with classic algorithms.
+Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean.
+A much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths (Criscuolo et al.
+\protect\hyperlink{ref-Criscuolo2006}{2006}) is also implemented via the ape package (Paradis et al.
+\protect\hyperlink{ref-Paradis2004}{2004}).
+The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
 However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife\_examples package).
@@
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are coherent with source ages.
-It is not certain if these chronograms should be used to perform all types of downstream evolutionary analyses and more research is needed in that direction.
+In both cases, ages are generally consistent with source ages.
+Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms (Ramshaw et al.
+\protect\hyperlink{ref-ramshaw1972time}{1972}; Sanderson and Doyle \protect\hyperlink{ref-sanderson2001sources}{2001}; Magallón et al.
+\protect\hyperlink{ref-magallon2015metacalibrated}{2015}; Barba-Montoya et al.
+\protect\hyperlink{ref-barba2018constraining}{2018}).
+For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
+Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.
 
@@
 At the moment, \texttt{datelife}'s chronogram database is not very large, storing 231 chronograms up to the time the manuscript was written.
-This represents 5.79\% of the largest existing chronogram database, which is unfortunately not open for scientific reuse nor automatised data mining (Kumar et al.
+This represents 5.79\% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining (Kumar et al.
 \protect\hyperlink{ref-Kumar2017}{2017}).
 OToL is the only public tree repository from where \texttt{datelife} can currently pull chronograms to construct its database.
-Other open repositories are not suitable for time of divergence data mining because they either do not store that data (i.e., TreeBASE), or their metadata does not have enough information to allow automatised chronogram searches (i.e., Dryad).
 A previous version of TimeTree's synthetic chronogram (Hedges et al.
-\protect\hyperlink{ref-Hedges2015}{2015}) was made available in OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
+\protect\hyperlink{ref-Hedges2015}{2015}) was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
 This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
-Thus, incorporation of published chronograms available in the Dryad data repository and in supplementary material of journals to \texttt{datelife}'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by ``hand''.
+Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to \texttt{datelife}'s database is crucial to improve its services.
+Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
 Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.
@@
 Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
-Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignement (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
+Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignment (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
 To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
@@
 
-Synthetic or summary chronograms are being used more and more frequently to study evolutionary hypothesis.
-We introduce different ways of summarizing time of lineage divergence data, but, are any of these summary chronograms reliable to study evolutionary patterns?
-The short answer is that we do not know to what extent the summary trees should be used for reanalyses and if there bare areas of research where they can be used more reliably than others.
-There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
-In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents an improvement for testing alternative hypothesis using phylogenetic distance.
-It is uncertain, however, how using very different age data could affect the outcomes of these tests, and further research is necessary in this direction.
+In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
+Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ (Webb et al.
+\protect\hyperlink{ref-Webb2008}{2008})) for several taxa lacking this information.
+Making up branch lengths in this or other ways is accepted in scientific publications: Rabosky et al.
+(\protect\hyperlink{ref-rabosky2018inverse}{2018}) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data; Smith and Brown (\protect\hyperlink{ref-smith2018constructing}{2018}) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
+There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.
 
@@
 do not have the time to acquire and develop the necessary knowledge and data curation skills
-to produce chronograms \emph{de novo}. Knowledge on taxon ages is also crucial for
+to produce chronograms \emph{de novo}. Knowledge on clade ages is also crucial for
 science communication and education.
@@
 Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
-How well this trees work for this purpose still needs to be further tested.
-If all, \texttt{datelife} helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+\texttt{datelife} helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 \texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version is available
+\texttt{datelife}'s R package stable version is \textbf{not} available yet
 for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities can be found in GitHub repositories at \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
+Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are respectively in the \href{https://github.com/LunaSare/datelife_paper1}{datelife\_paper1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR GitHub account.
 
@@
 
-Funding was provided by the US National Science Foundation (NSF) grant 1458603
-
-NESCent
-
-Open Tree of Life
-
-University of Tennessee, Knoxville
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
 
@@
 that led to initial work on this project.
+The team that assembled datelife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
 The Open Tree of Life project that provides the open, metadata rich repository of
@@
 the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition
-to the ABI grant that funded this project itself.
+The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
 
 \newpage
-
 \begin{center}
@@
 Bapst D.W. 2012. Paleotree: An R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution. 3:803--807.
+
+\leavevmode\hypertarget{ref-barba2018constraining}{}%
+Barba-Montoya J., Reis M.
+dos, Schneider H., Donoghue P.C., Yang Z.
+2018.
+Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
+New Phytologist.
+218:819--834.
 
@@
 
+\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
+Rabosky D.L., Chang J., Title P.O., Cowman P.F., Sallan L., Friedman M., Kaschner K., Garilao C., Near T.J., Coll M., others.
+2018.
+An inverse latitudinal gradient in speciation rate for marine fishes.
+Nature.
+559:392.
+
+\leavevmode\hypertarget{ref-ramshaw1972time}{}%
+Ramshaw J., Richardson D., Meatyard B., Brown R., Richardson M., Thompson E., Boulter D.
+1972.
+The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
+New Phytologist.
+71:773--779.
+
 \leavevmode\hypertarget{ref-RCoreTeam2018}{}%
@@
 
+\leavevmode\hypertarget{ref-sanderson2001sources}{}%
+Sanderson M.J., Doyle J.A. 2001. Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data. American Journal of Botany. 88:1499--1516.
+
 \leavevmode\hypertarget{ref-schenk2016sec}{}%
 Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
+
+\leavevmode\hypertarget{ref-smith2018constructing}{}%
+Smith S.A., Brown J.W. 2018. Constructing a broadly inclusive seed plant phylogeny. American Journal of Botany. 105:302--314.
 
@@
 
+% in this file I leave the figure captions outside\ caption{} because I want them to be formatted in the same way as the general text (double spaced and linenumbered)
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+
 \newpage
-
-\begin{center}
-\textsc{Figure \ref{fig:workflow}}
-\end{center}
+\begin{figure}[!h]
+\includegraphics{Fig1.pdf}
+\caption{}
+\label{fig:workflow}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:workflow}}
+% \end{center}
 Stylized DateLife workflow.
@@
 Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-
-\begin{center}
-\textsc{Figure \ref{fig:runtime1}}
-\end{center}
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{fig_runtime_main.pdf}
+\caption{}
+\label{fig:runtime1}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:runtime1}}
+% \end{center}
 Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
@@
 We also performed a search using the already processed query for comparison (light blue).
-
-\begin{center}
-\textsc{Figure \ref{fig:schronograms1}}
-\end{center}
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
+\caption{}
+\label{fig:schronograms1}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:schronograms1}}
+% \end{center}
 Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
@@
 (\protect\hyperlink{ref-price2014niche}{2014}).
-
-\begin{center}
-\textsc{Figure \ref{fig:summaries}}
-\end{center}
-LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
-
-\begin{center}
-\textsc{Figure \ref{fig:cvbladj}}
-\end{center}
-LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
-The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
-
-\begin{center}
-\textsc{Figure \ref{fig:cvbold}}
-\end{center}
-LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.
-
 \newpage
-
-\begin{figure}[!h]
-\includegraphics{Fig1.pdf}
-\caption{}
-\label{fig:workflow}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_runtime1.pdf}
-\caption{}
-\label{fig:runtime1}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
-\caption{}
-\label{fig:schronograms1}
-\end{figure}
-
-\newpage
-
 \begin{figure}[!h]
@@
 \end{figure}
-
+% \begin{center}
+% \textsc{Figure \ref{fig:summaries}}
+% \end{center}
+LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
 \newpage
-
 \begin{figure}[!h]
@@
 \end{figure}
-
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbladj}}
+% \end{center}
+LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
 \newpage
-
 \begin{figure}[!h]
@@
 \end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbold}}
+% \end{center}
+LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successful in three source chronograms shown here.
+
+% \end{linenumbers}
 

### 2019-09-29 biorxiv_2019.09.28_figures.tex
biorxiv versions

--- 2019-06-03 Datelife_paper1.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{setspace}
-\setstretch{2}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin = 1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{longtable,booktabs}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\providecommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{}
-    \pretitle{\vspace{\droptitle}}
-  \posttitle{}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
-\usepackage[left]{lineno}
-\linenumbers
-
-\begin{document}
-
-Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE
-
-Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
-
-Authors: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}
-
-Correspondence address:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA}
-\end{enumerate}
-
-Corresponding authors: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
+% in this file I leave the figure captions outside\ caption{} because I want them to be formatted in the same way as the general text (double spaced and linenumbered)
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
 
 \newpage
-
-\textbf{abstract.-}
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as Dryad, TreeBASE and Open Tree
-of Life, for a large quantity and diversity of organisms. However, this information
-remains difficult to synthesize for many biologists and the non-academic community.
-
-\textbf{Keywords:} Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
-
-\newpage
-
-Time of lineage divergence constitutes a fundamental piece of information for evolutionary
-understanding in many areas of research, from developmental to conservation biology (Felsenstein \protect\hyperlink{ref-Felsenstein1985a}{1985}; Webb \protect\hyperlink{ref-Webb2000}{2000}), from historical biogeography to species diversification studies (Posadas et al.
-\protect\hyperlink{ref-posadas2006historical}{2006}; Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-The primary information needed for time of lineage divergence estimation comes from the fossil record.
-Coupled to molecular phylogenies (and more recently to morphological data), the time of divergence of extant and extinct lineages is reconstructed with molecular dating methods.
-Probably encouraged by the great developments in DNA sequencing techniques, phylogenetic inference and molecular dating methods, the number of studies publishing phylogenies with branch lengths proportional to geologic time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}).
-Still, generating a chronogram is not an easy task unless you have specialized training.
-That's why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al.
-\protect\hyperlink{ref-Stoltzfus2013}{2013}).
-
-Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
-TreeBASE (Morell \protect\hyperlink{ref-morell1996roots}{1996}; Piel et al.
-\protect\hyperlink{ref-Piel2002}{2002}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al.
-\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms store and make available published phylogenies and chronograms for easy scientific reuse.
-All of them can be queried using automatised web procedures, which permit personalised, large scale queries that are also very fast.
-OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former does not store branch length information and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
-
-Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the phylogenetic relationships among known lineages.
-All or parts of this synthetic tree can be reused for any purpose. However, it currently only focus on synthesizing tree topology, meaning that it does not expose branch length data of any type.
-The Timetree of Life project focuses on the synthesis of a single chronogram of life (Hedges et al.
-\protect\hyperlink{ref-Hedges2006}{2006}).
-However, the thousands of chronograms they have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but not for scientific reuse or reanalysis.
-The latest version of their synthetic chronogram (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}) can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
-Other platforms such as SuperSmart (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}) and phylogenerator (Pearse and Purvis \protect\hyperlink{ref-pearse2013phylogenerator}{2013}) are focused in automatised \emph{de novo} chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
-However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Moreover, these tools do not provide information from already existent expert chronograms.
-
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fuly public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
-A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
-\protect\hyperlink{ref-Stoltzfus2013}{2013}).
-In here we present the formal description and implementation of the \texttt{datelife} service, constituted by an R package and a web site (\url{http://www.datelife.org/}).
-There is still much room for improvement, and flaws and limitations are addressed below.
-We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
-R packages for benchmarking of functionalities and demonstrating services were also developed and made available.
-
-\begin{center}
-\textsc{Description}
-\end{center}
-
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\item A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database; source trees with at least two matching input names are identified; all other taxa that do not match the original query are then dropped from the positively identified source trees --these pruned chronograms are hereafter referred as source chronograms; finally, each source chronogram is transformed to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarising algorithms.
-\item  The user can obtain different summary information from the source chronograms including: a) all source chronogram ages, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, and f) a report of succesful matches of input taxon names across source chronograms.
-Summary information can be used to make decisions on the next steps of the workflow.
-\item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also simulate both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily view and compare results graphically, as well as construct their own graphs using inbuilt `datelife` graphic generation functions.
-\end{enumerate}
-
-To gather, process, and present information, \texttt{datelife} builds up from functions
-available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
-geiger (Harmon et al.
-\protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}), taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl \protect\hyperlink{ref-Heibl2008}{2008}), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018}) and rphylotastic (O'Meara et al.
-\protect\hyperlink{ref-Omeara2019}{2019}).
-
-A \texttt{datelife} search currently accepts scientific names only. It can be any named clade or binomial specific.
-Chronogram search is performed at the species level, so when input names correspond
-to named clades, \texttt{datelife} pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
-Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
-\texttt{datelife} processes input names with the taxon name resolution service (TNRS; Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}),
-which corrects potentially misspelled names and typos, and standardizes spelling
-variations and synonyms , increasing the probability to correctly find the
-queried taxa in \texttt{datelife}'s chronogram database.
-
-The chronogram search is performed across \texttt{datelife}'s chronogram database which is assembled from OToL's tree repository.
-Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
-Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
-
-Information from source chronograms can be summarised using the Super Distance Matrix (SDM) approach for supertree reconstruction (Criscuolo et al.
-\protect\hyperlink{ref-Criscuolo2006}{2006}) or using a summary statistic of branch
-lengths (such as median or mean).
-The resulting summary patristic distance matrix could be clustered with classic algorithms.
-However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data.
-Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
-The Branch Length Adjuster (BLADJ) algorithm (Webb et al.
-\protect\hyperlink{ref-Webb2008}{2008}) was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
-
-For tree dating, the congruification algorithm described by Eastman et al. (\protect\hyperlink{ref-Eastman2013}{2013})
-is implemented to find shared nodes between trees (congruent nodes).
-The ages of these nodes are then used as calibrations to date any given tree.
-Currently implemented methods for tree dating are BLADJ, MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) and PATHd8 (Britton et al.
-\protect\hyperlink{ref-Britton2007}{2007}), a non-clock, rate-smoothing dating method.
-
-\begin{center}
-\textsc{Benchmark}
-\end{center}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS service is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes.
-A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}) that were run locally
---using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 50\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\begin{center}
-\textsc{Example}
-\end{center}
-
-In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain all data on lineage divergence available in \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
-according to the Open Tree of Life taxonomy).
-There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig.
-\ref{fig:schronograms1}).
-Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
-As explained in the \texttt{Description}, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are coherent with source ages.
-It is not certain if these chronograms can be used to perform downstream evolutionary analyses.
-There is currently wide interest in determining this.
-However, we know that these chronograms are useful for\ldots{}
-
-Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs.
-\ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms.
-In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
-Maximum tree ages were only approximately recovered in one case (study 2; Fig.
-\ref{fig:cvbladj}).
-Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
-three cases (studies 3, 5, and 9; Fig.
-\ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and details can be consulted in \url{https://github.com/LunaSare/datelife_examples}.
-
-\begin{center}
-\textsc{Flaws, Limitations and Prospects}
-\end{center}
-
-The main goal of \texttt{datelife} is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
-However, it has many flaws.
-Some of them can be overcome, some of them might represent limitations.
-
-At the moment, \texttt{datelife}'s chronogram database is not very large.
-It stores 231 chronograms up to the time the manuscript was written.
-This represents 5.79\% of the largest existing chronogram database, which is unfortunately not open for scientific reuse and automatised data mining (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}).
-OToL is the only public tree repository from where \texttt{datelife} can currently pull chronograms to construct its database.
-Other open repositories are not suitable for time of divergence data mining because they either do not store that data (i.e., TreeBASE), or their metadata does not have enough information to allow automatised chronogram searches (i.e., Dryad).
-
-A previous version of TimeTree's synthetic chronogram (Hedges et al.
-\protect\hyperlink{ref-Hedges2015}{2015}) was made available in OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
-This ensures that at least some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
-
-Thus, incorporation of published chronograms available in the Dryad data repository and in supplementary material of journals to \texttt{datelife}'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would need to be determined by ``hand''.
-Consequently, we would like to emphasize on the importance for the scientific community of sharing chronogram data in repositories that require expert input and manual curation,~such as OToL's tree repository.
-
-We introduce different ways of summarizing this data, but are these summary chronograms reliable to study evolutionary patterns?
-The short answer is that we do not know to what extent the summary trees can be used for reanalyses and in which areas they can be used more reliably than others.
-Surely they might be useful for ecological analyses, but further studies should be conducted.
-
-As implemented now, we currently by default summarize all source data avilable.
-Users can subset source data if they have reasons to favor some or one source chronogram over others.
-Strictly speaking, a good chronogram reflects the real time of divergence accurately and precisely.
-To our knowledge there is no objective way to determine if an expert chronogram is better than other.
-Some criteria that have been put forward are the level of sampling and the numer of calibrations used.
-Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
-\texttt{datelife}'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.
-It has been shown with simulations that in analyses performed with bayesian inference methods, divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations (using secondary calibrations in the same way as primary calibrations are used Schenk \protect\hyperlink{ref-schenk2016sec}{2016}\protect\hyperlink{ref-schenk2016sec}{a}).
-But there are different ways to apply secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-This information should be included as metadata to facilitate subset of source chronograms.
-This will need to be manually curated as is done in OToL.
-
-As observed from the comparison of source chronograms in the example above, even chronograms obtained with primary data can be very different.
-There is some evidence that different chronograms will result in very different macroevolutionary histories inferred from them (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
-In other areas of research, such as ecological and conservation biology, it has been demonstrated that the presence of data on lineage divergence represents an improvement for testing alternative hypothesis, compared to testing in a total absence of such time data.
-However, we do not know how using wrong or different kinds of age data will affect the outcomes of ecological tests.
-More work has to be performed in this area.
-A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms (e.g., angiosperms; Magallón et al.
-\protect\hyperlink{ref-magallon2015metacalibrated}{2015}).
-Comparison of available chronogram data for a wide range of organisms shown in the examples here suggest that this is a widespread phenomenon that requires further attention.
-In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
-However, this is just an assumption.
-Scores on source data used (quality of alignement: missing data, GC content; tip sampling strategy; phylogenetic inference method; dating inference method; numer of fossils used; support for nodes and ages; age confidence intervals).
-But we should also be able to choose on
-
-Even if all source chronograms have been generated by excellent standards and using same practices, the evolutionary context they depict might be very different.
-The ideal would be to use dates that were obtained from the same DNA context, which should reflect a more homogeneous evolutionary history or hypothesis (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}).
-Summarizing trees implies summarizing evolutionary hypothesis.
-It is unsure if this is a good or bad thing.
-
-For all these reasons, the ideal would be to use fossil data as primary sources of calibrations.
-This will be an upscale for datelife, taking fossils that have already been curated as calibrations to date other trees.
-
-Improvements:
-* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
-(Schenk \protect\hyperlink{ref-Schenk2016}{2016}\protect\hyperlink{ref-Schenk2016}{b}).
-* bayesian congruification
-* topological congruification
-
-\begin{center}
-\textsc{Conclusions}
-\end{center}
-
-Time of lineage divergence information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. Generating this information is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills
-to produce chronograms \emph{de novo}. Knowledge on taxon ages is also crucial for
-science communication and education.
-
-\texttt{datelife} allows an easy and fast obtention, as well as comparison of publicly available information
-on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the timeframe of different regions of the tree of life, allowing identification of regions that need more work or that have conflicting information.
-It can also summarize public data, or use it to generate new chronograms based on trees provided by the user.
-Both summary and newly generated trees have the potential to be useful to evaluate evolutionary hypothesis in different areas of research.
-How well this trees work for this purpose still needs to be tested.
-\texttt{datelife} helps with awereness on the variation of expert data, and the actual effect of the underlying data on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\texttt{datelife} is also part of the reproducibility of science movement.
-It strives to facilitate usage of public data sets as well as encouraging the making of information publicly available.
-Acknowledging source information and describing workflows and analyses.
-
-it allows exploration of alternative taxonomic and phylogenetic schemes;
-it allows rapid exploration of the effect of alternative divergence time hypothesis;
-it allows rapid synthesis in a number of different formats;
-
-\begin{center}
-\textsc{Availability}
-\end{center}
-
-\texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\begin{center}
-\textsc{Supplementary Material}
-\end{center}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities can be found in GitHub repositories at \url{https://github.com/LunaSare/datelife_paper1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}, respectively.
-
-\begin{center}
-\textsc{Funding}
-\end{center}
-
-Funding was provided by the US National Science Foundation (NSF) grant 1458603
-
-NESCent
-
-Open Tree of Life
-
-University of Tennessee, Knoxville
-
-\begin{center}
-\textsc{Acknowledgements}
-\end{center}
-
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition
-to the ABI grant that funded this project itself.
-
-\newpage
-
-\begin{center}
-\textsc{References}
-\end{center}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
-Antonelli A., Hettling H., Condamine F.L., Vos K., Nilsson R.H., Sanderson M.J., Sauquet H., Scharn R., Silvestro D., Töpel M., Bacon C.D., Oxelman B., Vos R.A.
-2017.
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
-Systematic Biology.
-66:153--166.
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst D.W. 2012. Paleotree: An R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution. 3:803--807.
-
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2012.
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-Systematic biology.
-62:298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2015.
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk: Ornithological Advances.
-132:333--348.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle B., Hopkins N., Lu Z., Raygoza Garay J.A., Mozzherin D., Rees T., Matasci N., Narro M.L., Piel W.H., Mckay S.J., Lowry S., Freeland C., Peet R.K., Enquist B.J.
-2013.
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton T., Anderson C.L., Jacquet D., Lundqvist S., Bremer K. 2007. Estimating Divergence Times in Large Phylogenetic Trees. Systematic Biology. 56:741--752.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns K.J., Shultz A.J., Title P.O., Mason N.A., Barker F.K., Klicka J., Lanyon S.M., Lovette I.J.
-2014.
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-Molecular Phylogenetics and Evolution.
-75:41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain S. 2018. bold: Interface to Bold Systems API..
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain S.A., Szöcs E. 2013. taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}. F1000Research. 2:1--29.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo A., Berry V., Douzery E.J., Gascuel O. 2006. SDM: A fast distance-based approach for (super)tree building in phylogenomics. Systematic Biology. 55:740--755.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman J.M., Harmon L.J., Tank D.C. 2013. Congruification: Support for time scaling large phylogenetic trees. Methods in Ecology and Evolution. 4:688--691.
-
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein J. 1985. Phylogenies and the Comparative Method. The American Naturalist. 125:1--15.
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb G.C., England R., Hartig G., McLenachan P.A., Taylor Smith B.L., McComish B.J., Cooper A., Penny D.
-2015.
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-Genome biology and evolution.
-7:2983--2995.
-
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon L., Weir J., Brock C., Glor R., Challenger W. 2008. GEIGER: investigating evolutionary radiations. Bioinformatics. 24:129--131.
-
-\leavevmode\hypertarget{ref-Hedges2006}{}%
-Hedges S.B., Dudley J., Kumar S. 2006. TimeTree: A public knowledge-base of divergence times among organisms. Bioinformatics. 22:2971--2972.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl C. 2008. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..
-
-\leavevmode\hypertarget{ref-Hinchliff2015}{}%
-Hinchliff C.E., Smith S.A., Allman J.F., Burleigh J.G., Chaudhary R., Coghill L.M., Crandall K.A., Deng J., Drew B.T., Gazis R., Gude K., Hibbett D.S., Katz L.A., Laughinghouse H.D., McTavish E.J., Midford P.E., Owen C.L., Ree R.H., Rees J.A., Soltis D.E., Williams T., Cranston K.A.
-2015.
-Synthesis of phylogeny and taxonomy into a comprehensive tree of life.
-Proceedings of the National Academy of Sciences.
-112:12764--12769.
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck J.P., Ronquist F. 2001. MRBAYES: Bayesian inference of phylogenetic trees. Bioinformatics. 17:754--755.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.
-
-\leavevmode\hypertarget{ref-Kumar2017}{}%
-Kumar S., Stecher G., Suleski M., Hedges S.B. 2017. TimeTree: A Resource for Timelines, Timetrees, and Divergence Times. Molecular biology and evolution. 34:1812--1819.
-
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
-Magallón S., Gómez-Acevedo S., Sánchez-Reyes L.L., Hernández-Hernández T.
-2015.
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-New Phytologist.
-207:437--453.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau F., Brown J.W., Winter D.J. 2016. rotl: an R package to interact with the Open Tree of Life data. Methods in Ecology and Evolution. 7:1476--1481.
-
-\leavevmode\hypertarget{ref-morell1996roots}{}%
-Morell V. 1996. The roots of phylogeny. Science. 273:569.
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon H. 2014. Phylogenetic approaches for studying diversification. Ecology Letters. 17:508--525.
-
-\leavevmode\hypertarget{ref-Omeara2019}{}%
-O'Meara B., Md Tayeen A.S., Sanchez Reyes L.L. 2019. Rphylotastic: An r interface to 'phylotastic' web services..
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms J., Chamberlain S. 2018. Phylocomr: Interface to 'phylocom'..
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.
-
-\leavevmode\hypertarget{ref-pearse2013phylogenerator}{}%
-Pearse W.D., Purvis A. 2013. PhyloGenerator: An automated phylogeny generation tool for ecologists. Methods in Ecology and Evolution. 4:692--698.
-
-\leavevmode\hypertarget{ref-Piel2002}{}%
-Piel W.H., Donoghue M., Sanderson M.
-2002.
-TreeBASE : A database of phylogenetic information.
-In: Shimura J., Wilson K., Gordon D., editors.
-To the interoperable ``catalog of life'' with partners.
-Tsukuba, Japan: National Institute for Environmental Studies.
-p.
-41--47.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas P., Crisci J.V., Katinas L. 2006. Historical biogeography: A review of its basic concepts and critical issues. Journal of Arid Environments. 66:389--403.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price T.D., Hooper D.M., Buchanan C.D., Johansson U.S., Tietze D.T., Alström P., Olsson U., Ghosh-Harihar M., Ishtiaq F., Gupta S.K., others.
-2014.
-Niche filling slows the diversification of himalayan songbirds.
-Nature.
-509:222.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. 2018. R: a language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell L.J. 2012. Phytools: An r package for phylogenetic comparative biology (and other things). Methods in Ecology and Evolution. 3:217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist F., Huelsenbeck J.P. 2003. MrBayes 3: Bayesian phylogenetic inference under mixed models. Bioinformatics. 19:1572--1574.
-
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk J.J. 2016a. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
-
-\leavevmode\hypertarget{ref-Schenk2016}{}%
-Schenk J.J. 2016b. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus A., Lapp H., Matasci N., Deus H., Sidlauskas B., Zmasek C.M., Vaidya G., Pontelli E., Cranston K., Vos R., Webb C.O., Harmon L.J., Pirrung M., O'Meara B., Pennell M.W., Mirarab S., Rosenberg M.S., Balhoff J.P., Bik H.M., Heath T.A., Midford P.E., Brown J.W., McTavish E.J., Sukumaran J., Westneat M., Alfaro M.E., Steele A., Jordan G.
-2013.
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
-Title P.O., Rabosky D.L. 2016. Do Macrophylogenies Yield Stable Macroevolutionary Inferences? An Example from Squamate Reptiles. Systematic Biology.:syw102.
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb C.O. 2000. Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. The American Naturalist. 156:145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb C.O., Ackerly D.D., Kembel S.W. 2008. Phylocom: Software for the analysis of phylogenetic community structure and trait evolution. Bioinformatics. 24:2098--2100.
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb C.O., Donoghue M.J. 2005. Phylomatic: Tree assembly for applied phylogenetics. Molecular Ecology Notes. 5:181--183.
-
-\newpage
-
-\begin{center}
-\textsc{Figure \ref{fig:workflow}}
-\end{center}
+\begin{figure}[!h]
+\includegraphics{Fig1.pdf}
+\caption{}
+\label{fig:workflow}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:workflow}}
+% \end{center}
 Stylized DateLife workflow.
@@
 \end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbold}}
+% \end{center}
+LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successful in three source chronograms shown here.
 
-Supplementary Material
-
-\end{document}
+% \end{linenumbers}

### 2019-09-29 paper.Rmd
biorxiv versions

--- 2019-08-06 Datelife_paper1.Rmd
@@
----
-output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    toc: false
-    includes:
-      # in_header: preamble.tex
-      after_body: [Datelife_paper1_figures_noyaml.tex]
-    # fig_caption: no
-textalign: left  # this does not work...
-geometry: margin = 1in
-bibliography: library_red.bib
-csl: systematic-biology.csl
-link-citations: yes
-# indent: true # this works
-# header-includes:
-#    - \setlength\parindent{24pt}  # this works too, but I'm gonna indent manually with 4 spaces
-# nope, 4 spaces indents are not recognized, they do something else
-linestretch: 2
-header-includes:
-    # - \usepackage{caption}
-    - \usepackage[left]{lineno}
-    - \linenumbers
-  # - \documentclass[oupdraft]{sysbio} # this is duplicated in tex generated by knitr
-  # - \usepackage{xr}
-  # - \externaldocument{Datelife_paper1_figures}
-    # - \usepackage[utf8]{inputenc}
-    # - \usepackage[english]{babel}
-    # - \usepackage[document]{ragged2e}
-nocite: |
-  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
----
-
 ```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
-source("Datelife_paper1_global.R")
+# code to generate the paper data and figures:
+# source("data.R")
+# inactivated now just for testing purposes
+# we are using drake to avoid running the examples everytime we change something in the Rmd and knit
+# using this for testing instead:
+source("./frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
+loadd() # loads all the targets listed in the workflow plan of the previous `make()`
+library(datelife)
+data(opentree_chronograms)
+rpackage_available <- FALSE
+availability <- ifelse(rpackage_available, "is available", "is **not** available yet")
+thank_origteam <- TRUE
+origteam <- ifelse(thank_origteam, "The team that assembled datelife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.", "")
 ```
 
-Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE <!--MAX 50 CHARACTERS, ALL CAPITALS-->
-
-Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
-
-Authors: Luna L. Sánchez-Reyes^1^, Brian C. O’Meara^1^
-
-Correspondence address:
-
-1. *Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA*
-
-Corresponding authors: sanchez.reyes.luna@gmail.com, bomeara@utk.edu
-
 \newpage
-
 **abstract.-**
@@
 **Keywords:** Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
-
 \newpage
+\begin{center}
+\textsc{Introduction}
+\end{center}
 Clade ages represent a fundamental piece of information for evolutionary
 understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014].
-The primary information needed these time estimates comes from the fossil record.
+The primary information needed for these time estimates comes from the fossil record.
 Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
@@
 
+<!--(BOLD and OToL species names, are they homogeneous/standardized?-->
+
 \begin{center}
@@
 \end{center}
-<!--(BOLD and OToL species names, are they homogeneous/standardized?-->
+
 The basic `datelife` workflow is shown in figure \ref{fig:workflow} and consists of:
@@
 
-Information from source chronograms can be summarised with a summary statistic of tree branch
-lengths (such as median or mean). <!-- and the hierarchical clustering method of ttol-->
+Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean.
+A  much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006] is also implemented via the ape package [@Paradis2004].<!-- and the hierarchical clustering method of ttol??-->
 The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
@@
 Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
-The Branch Length Adjuster (BLADJ) algorithm [@
-2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
-
-A  much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006], implemented in the ape package [@Paradis2004].
+The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
 
@@
 <!-- and hierarchical clustering method which appears to push ages back -->
-Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms (_____MANY CITATIONS____)).
-For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results (Title and Rabosky [https://academic.oup.com/sysbio/article/66/5/843/2682290] ).
+Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
+For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results [@title2016macrophylogenies].
 Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.
@@
 
-We allow accepted ways of creating branch lengths in the absence of starting branch length information for some taxa, such as BLADJ [@ 2008].
-Making up branch lengths in this or other ways is accepted in scientific publications: Rabosky et al.
-[https://www.nature.com/articles/s41586-018-0273-1] created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; Jetz et al.
-[https://www.nature.com/articles/nature11631], created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data; Smith and Brown [https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1019] constructed a tree of 353,185 seed plants where only 23% had molecular data.
+In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
+Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ [@Webb2008]) for several taxa lacking this information.
+Making up branch lengths in this or other ways is accepted in scientific publications: @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
 Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.
+There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.
 
-There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
-In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents an improvement for testing alternative hypothesis using phylogenetic distance.
-It is uncertain, however, how using very different age data could affect the outcomes of these tests, and further research is necessary in this direction.
+<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
+However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that directions -->
 
@@
 `datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-`datelife`'s R package stable version is available
+`datelife`'s R package stable version `r availability`
 for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities can be found in GitHub repositories at <https://github.com/LunaSare/datelife_paper1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>, respectively.
+Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are respectively in the [datelife_paper1](https://github.com/LunaSare/datelife_paper1), [datelife_examples](https://github.com/LunaSare/datelife_examples), and [datelife_benchmark](https://github.com/LunaSare/datelife_benchmark) repositories in LLSR GitHub account.
 <!-- Vignettes?
@@
 
-Funding was provided by the US National Science Foundation (NSF) grant 1458603
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
 
-NESCent
-
-Open Tree of Life
-
-University of Tennessee, Knoxville
+<!-- Any separate funding from the University of Tennessee, Knoxville?? -->
 
@@
 The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
+that led to initial work on this project. `r origteam`
 The Open Tree of Life project that provides the open, metadata rich repository of
@@
 the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition
-to the ABI grant that funded this project itself.
-
-\newpage
-
-\begin{center}
-\textsc{References}
-\end{center}
+The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

### 2019-09-29 bioarxiv_2019.09/data.R

@@ -1,0 +1,29 @@
+library(datelife)
+data(opentree_chronograms)
+  dq = make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE)
+  dr = get_datelife_result(input = dq)
+  summ = get_taxon_summary(datelife_query = dq, datelife_result = dr)
+  phyloall = summarize_datelife_result(datelife_query = dq,
+    datelife_result = dr, summary_format = "phylo_all",
+    taxon_summary = "none")
+  cb_cols1 = c("#0072B2", "#D55E00", "#440154FF", "#CC79A7", "#440154FF", "#009E73",
+    "#9ad0f3", "#95D840FF", "#F0E442")
+  cb_cols2 = c("#440154FF", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00",
+    "#CC79A7", "#F0E442", "#95D840FF")
+  ltt_plot = plot_ltt_phyloall(taxon = "Fringillidae", phy = phyloall, ltt_colors = cb_cols1, tax_datedotol = NULL,
+      file_name = "ltt_plot_poster.pdf", file_dir = getwd(), height = 3, width = 7.9, add_legend = FALSE, add_title = FALSE)
+  datedotol = get_dated_otol_induced_subtree(input = dq, ott_ids = dq$ott_ids)
+  otol = get_otol_synthetic_tree(input = dq)
+  bestgrove = get_best_grove(datelife_result = dr)
+  sdm_matrix = get_sdm_matrix(datelife_result = bestgrove$best_grove)
+  median_matrix = datelife_result_median_matrix(datelife_result = bestgrove$best_grove)
+  phy_sdm = summary_matrix_to_phylo_all(summ_matrix = sdm_matrix, target_tree = otol)
+  phy_median = summary_matrix_to_phylo_all(summ_matrix = median_matrix, target_tree = otol)
+  ltt_summ = plot_ltt_summary(taxon = "Fringillidae", phy = phyloall, phy_sdm, phy_median,
+        file_name = "ltt_summ_poster.pdf", file_dir = "~//Desktop//datelife_paper1//", height = 3, width = 7.9,
+        add_legend = TRUE)
+  each_calibrations = get_all_calibrations(phyloall, each = TRUE)
+  sim_otol = use_each_calibration(phy = otol, calibrations = each_calibrations)$phy
[…]

### 2019-09-29 bioarxiv_2019.09/figures_download.R

@@ -24,8 +24,3 @@ download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/do
               destfile = "fig_lttplot_clusters.pdf", mode="wb")
 
 
-source("frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
-loadd() # loads all the targets listed in the workflow plan of the previous `make()`
-library(datelife)
-data(opentree_chronograms)
-

### 2019-09-29 frin_plan.R

@@ -15,19 +15,21 @@ frin_plan <- drake_plan(
   ltt_plot = plot_ltt_phyloall(taxon = "Fringillidae", phy = phyloall, ltt_colors = cb_cols1, tax_datedotol = NULL,
       file_name = "ltt_plot_poster.pdf", file_dir = getwd(), height = 3, width = 7.9, add_legend = FALSE, add_title = FALSE),
   datedotol = get_dated_otol_induced_subtree(input = dq, ott_ids = dq$ott_ids),
-  otol = get_otol_synthetic_tree(input = dq),
   bestgrove = get_best_grove(datelife_result = dr),
   sdm_matrix = get_sdm_matrix(datelife_result = bestgrove$best_grove),
   median_matrix = datelife_result_median_matrix(datelife_result = bestgrove$best_grove),
+  otol = get_otol_synthetic_tree(input = dq),
   phy_sdm = summary_matrix_to_phylo_all(summ_matrix = sdm_matrix, target_tree = otol),
   phy_median = summary_matrix_to_phylo_all(summ_matrix = median_matrix, target_tree = otol),
   ltt_summ = plot_ltt_summary(taxon = "Fringillidae", phy = phyloall, phy_sdm, phy_median,
         file_name = "ltt_summ_poster.pdf", file_dir = "~//Desktop//datelife_paper1//", height = 3, width = 7.9,
         add_legend = TRUE),
-  each_calibrations = get_all_calibrations(phyloall, each = TRUE),
-  sim_otol = use_each_calibration(phy = otol, calibrations = each_calibrations)$phy,
-  ltt_simotol = plot_ltt_phyloall(taxon = "Fringillidae", sim_otol, summ,
-        add_title = FALSE, file_dir = getwd(), file_name = "_lttplot_simotol_poster.pdf",
-        ltt_colors = cb_cols1, height = 3, width = 7.9)
+  each_calibrations = get_all_calibrations(phyloall, each = TRUE)
+  # , # inactivating the following because having trouble with getting otol for some reason
+  # it was only used for the poster figures anyways
+  # sim_otol = use_each_calibration(phy = otol, calibrations = each_calibrations)$phy,
+  # ltt_simotol = plot_ltt_phyloall(taxon = "Fringillidae", sim_otol, summ,
+  #       add_title = FALSE, file_dir = getwd(), file_name = "_lttplot_simotol_poster.pdf",
+  #       ltt_colors = cb_cols1, height = 3, width = 7.9)
 )
 make(frin_plan)

### 2019-09-30 paper.Rmd
improved readme

@@
  -->
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fuly public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
+A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
 A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].

### 2019-09-30 biorxiv_2019.09.28.Rmd
for reproducibility's sake

@@
       # in_header: preamble.tex
-      after_body: [biorxiv_2019.09.28_figures.tex]
+      after_body: [biorxiv_2019.09.28figcaps.tex]
     # fig_caption: no
@@
 geometry: margin = 1in
-bibliography: library_red.bib
+bibliography: paper_references.bib
 csl: systematic-biology.csl

### 2019-09-30 biorxiv_2019.09.28.tex
for reproducibility's sake

@@
 
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fuly public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
+A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
 A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
@@
 \texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version is \textbf{not} available yet
+\texttt{datelife}'s R package stable version will be available
 for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
@@
 that led to initial work on this project.
-The team that assembled datelife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
+The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
 The Open Tree of Life project that provides the open, metadata rich repository of
@@
 The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
+\newpage
 
-\newpage
 \begin{center}
@@
 \begin{figure}[!h]
-\includegraphics{Fig1.pdf}
+\includegraphics{./figures/Fig1.pdf}
 \caption{}
@@
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_runtime_main.pdf}
+\includegraphics[width=1\linewidth]{./figures/fig_runtime_main.pdf}
 \caption{}
@@
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
+\includegraphics[width=1\linewidth]{./figures/fig_schronograms1.pdf}
 \caption{}
@@
 \begin{figure}[!h]
-\includegraphics{fig_summaries.pdf}
+\includegraphics{./figures/fig_summaries.pdf}
 %\includegraphics{fig_summaries.pdf}
@@
 % \end{center}
-LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
+LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
 \newpage
 \begin{figure}[!h]
-\includegraphics{fig_crossval_bladj.pdf}
+\includegraphics{./figures/fig_crossval_bladj.pdf}
 \caption{}
@@
 \begin{figure}[!h]
-\includegraphics{fig_crossval_boldsumm.pdf}
+\includegraphics{./figures/fig_crossval_boldsumm.pdf}
 \caption{}

### 2019-09-30 biorxiv_2019.09.28figcaps.tex
for reproducibility's sake

--- 2019-09-30 biorxiv_2019.09.28.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{setspace}
-\setstretch{2}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin = 1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{longtable,booktabs}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\providecommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{}
-    \pretitle{\vspace{\droptitle}}
-  \posttitle{}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
-\usepackage{caption}
-\usepackage{blindtext}
-\usepackage[left]{lineno}
-\linenumbers
-
-\begin{document}
-
-\textbf{Title}: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
-
-\textbf{Authors}: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}
-
-\textbf{Correspondence address}:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA}
-\end{enumerate}
-
-\textbf{Corresponding authors}: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
-
-\newpage
-
-\textbf{abstract.-}
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
-of Life for a large quantity and diversity of organisms in the last few decades.
-However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication.
-\texttt{datelife} is a service implemented via an R package and a web site
-(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of published data on lineage divergence times.
-The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
-Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram.
-Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
-If there is no information available for any queried taxa, data can be simulated.
-All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this trees work for this purpose still needs to be tested.
-\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\textbf{Keywords:} Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
-\newpage
-
-\begin{center}
-\textsc{Introduction}
-\end{center}
-
-Clade ages represent a fundamental piece of information for evolutionary
-understanding in many areas of research, from developmental to conservation biology (Felsenstein \protect\hyperlink{ref-Felsenstein1985a}{1985}; Webb \protect\hyperlink{ref-Webb2000}{2000}), from historical biogeography to species diversification studies (Posadas et al.
-\protect\hyperlink{ref-posadas2006historical}{2006}; Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-The primary information needed for these time estimates comes from the fossil record.
-Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
-The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}).
-Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree.
-That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al.
-\protect\hyperlink{ref-Stoltzfus2013}{2013}).
-
-Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
-TreeBASE (Morell \protect\hyperlink{ref-morell1996roots}{1996}; Piel et al.
-\protect\hyperlink{ref-Piel2002}{2002}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al.
-\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms store and make available published phylogenies and chronograms for easy scientific reuse.
-All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
-OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
-
-Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the phylogenetic relationships among known lineages.
-All or parts of this synthetic tree can be reused for any purpose. However, it currently only focus on synthesizing tree topology, meaning that it does not expose branch length data of any type.
-The Timetree of Life project focuses on the synthesis of a single chronogram of life (Hedges et al.
-\protect\hyperlink{ref-Hedges2006}{2006}) and presents a very accessible, attractive interface.
-However, the thousands of chronograms this NSF-funded project have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but large scale download remains prohibited by their site.
-The latest version of their synthetic chronogram (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}) can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
-Other platforms such as SuperSmart (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}) and phylogenerator (Pearse and Purvis \protect\hyperlink{ref-pearse2013phylogenerator}{2013}) are focused in automatised \emph{de novo} chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
-However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Moreover, these tools do not provide information from already created expert chronograms.
-
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
-A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
-\protect\hyperlink{ref-Stoltzfus2013}{2013}).
-In here we present the formal description and implementation of the \texttt{datelife} service, constituted by an R package and a web site (\url{http://www.datelife.org/}).
-There is still much room for improvement, and flaws and limitations are addressed below.
-We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
-
-\begin{center}
-\textsc{Description}
-\end{center}
-
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\item A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database; source trees with at least two matching input names are identified; all other taxa that do not match the original query are then dropped from the positively identified source trees.
-These pruned chronograms are hereafter referred as source chronograms.
-Finally, each source chronogram is transformed to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-\item  The user can obtain different summary information from the source chronograms including: a) all source chronograms, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, f) a report of succesful matches of input taxon names across source chronograms, and g) the single source chronogram with the greatest number of taxa.
-Summary information can be used to make decisions on the next steps of the workflow.
-\item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R `phylo` format), as well as view and compare results graphically, or construct their own graphs using inbuilt `datelife` graphic generation functions.
-\end{enumerate}
-
-To gather, process, and present information, \texttt{datelife} builds up from functions
-available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
-geiger (Harmon et al.
-\protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}), taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl \protect\hyperlink{ref-Heibl2008}{2008}), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018}) and rphylotastic (O'Meara et al.
-\protect\hyperlink{ref-Omeara2019}{2019}).
-
-A \texttt{datelife} search currently accepts scientific names only. It can be any named clade or binomial specific.
-Chronogram search is performed at the species level, so when input names correspond
-to named clades, \texttt{datelife} pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
-Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
-\texttt{datelife} processes input names with the taxon name resolution service (TNRS; Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}),
-which corrects potentially misspelled names and typos, and standardizes spelling
-variations and synonyms , increasing the probability of correctly finding the
-queried taxa in \texttt{datelife}'s chronogram database.
-
-The chronogram search is performed across \texttt{datelife}'s chronogram database which is assembled from OToL's tree repository.
-Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
-Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
-
-Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean.
-A much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths (Criscuolo et al.
-\protect\hyperlink{ref-Criscuolo2006}{2006}) is also implemented via the ape package (Paradis et al.
-\protect\hyperlink{ref-Paradis2004}{2004}).
-The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
-However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife\_examples package).
-Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
-The Branch Length Adjuster (BLADJ) algorithm (Webb et al.
-\protect\hyperlink{ref-Webb2008}{2008}) was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
-
-For tree dating, the congruification algorithm described by Eastman et al. (\protect\hyperlink{ref-Eastman2013}{2013})
-is implemented to find shared nodes between trees (congruent nodes).
-The ages of these nodes are then used as calibrations to date any given tree.
-Currently implemented methods for tree dating are BLADJ, MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) and PATHd8 (Britton et al.
-\protect\hyperlink{ref-Britton2007}{2007}), a non-clock, rate-smoothing dating method.
-
-\begin{center}
-\textsc{Benchmark}
-\end{center}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS service is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes.
-A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}) that were run locally
---using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 50\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\begin{center}
-\textsc{Example}
-\end{center}
-
-In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain all data on lineage divergence available in \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
-according to the Open Tree of Life taxonomy).
-There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig.
-\ref{fig:schronograms1}).
-Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
-As explained in the \texttt{Description}, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are generally consistent with source ages.
-Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms (Ramshaw et al.
-\protect\hyperlink{ref-ramshaw1972time}{1972}; Sanderson and Doyle \protect\hyperlink{ref-sanderson2001sources}{2001}; Magallón et al.
-\protect\hyperlink{ref-magallon2015metacalibrated}{2015}; Barba-Montoya et al.
-\protect\hyperlink{ref-barba2018constraining}{2018}).
-For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
-Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.
-
-Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs.
-\ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms.
-In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
-Maximum tree ages were only approximately recovered in one case (study 2; Fig.
-\ref{fig:cvbladj}).
-Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
-three cases (studies 3, 5, and 9; Fig.
-\ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and details can be consulted in \url{https://github.com/LunaSare/datelife_examples}.
-
-\begin{center}
-\textsc{Flaws, Limitations and Prospects}
-\end{center}
-
-The main goal of \texttt{datelife} is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
-However, it has many flaws.
-Some of them can be overcome, some of them might represent limitations.
-
-At the moment, \texttt{datelife}'s chronogram database is not very large, storing 231 chronograms up to the time the manuscript was written.
-This represents 5.79\% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}).
-OToL is the only public tree repository from where \texttt{datelife} can currently pull chronograms to construct its database.
-A previous version of TimeTree's synthetic chronogram (Hedges et al.
-\protect\hyperlink{ref-Hedges2015}{2015}) was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
-This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
-Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to \texttt{datelife}'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.
-
-Another potential concern comes from summary chronograms.
-We currently summarize all source chronogram data by default.
-Users can subset source data if they have reasons to favor some or one source chronogram over others.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there is no objective way to determine if an expert chronogram is better than other.
-Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
-Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk \protect\hyperlink{ref-schenk2016sec}{2016}).
-Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-This remains to be tested.
-
-Furthermore, even chronograms obtained with primary data can be very different, as observed from the comparison of source chronograms in the Fringillidae example.
-A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms (e.g., angiosperms; Magallón et al.
-\protect\hyperlink{ref-magallon2015metacalibrated}{2015}).
-Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
-Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignment (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
-To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
-Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.
-Hence, summarizing chronograms might imply summarizing evolutionary hypothesis.
-This could be good from certain point of view, since it could help to get a single global evolutionary history for a lineage.
-But it could also be bad, since we could be loosing part of the evolutionary history that is only being reflected in some chronograms and not from the summary chronogram.
-Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}).
-This will be implemented in future \texttt{datelife} versions.
-
-In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
-Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ (Webb et al.
-\protect\hyperlink{ref-Webb2008}{2008})) for several taxa lacking this information.
-Making up branch lengths in this or other ways is accepted in scientific publications: Rabosky et al.
-(\protect\hyperlink{ref-rabosky2018inverse}{2018}) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data; Smith and Brown (\protect\hyperlink{ref-smith2018constructing}{2018}) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.
-
-\begin{center}
-\textsc{Conclusions}
-\end{center}
-
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. Generating this information is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills
-to produce chronograms \emph{de novo}. Knowledge on clade ages is also crucial for
-science communication and education.
-
-\texttt{datelife} allows an easy and fast obtention, as well as comparison of publicly available information
-on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, allowing identification of regions that require more research or that have conflicting information.
-Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
-\texttt{datelife} helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\begin{center}
-\textsc{Availability}
-\end{center}
-
-\texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version will be available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\begin{center}
-\textsc{Supplementary Material}
-\end{center}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are respectively in the \href{https://github.com/LunaSare/datelife_paper1}{datelife\_paper1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR GitHub account.
-
-\begin{center}
-\textsc{Funding}
-\end{center}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
-
-\begin{center}
-\textsc{Acknowledgements}
-\end{center}
-
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-\newpage
-
-\begin{center}
-\textsc{References}
-\end{center}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
-Antonelli A., Hettling H., Condamine F.L., Vos K., Nilsson R.H., Sanderson M.J., Sauquet H., Scharn R., Silvestro D., Töpel M., Bacon C.D., Oxelman B., Vos R.A.
-2017.
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
-Systematic Biology.
-66:153--166.
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst D.W. 2012. Paleotree: An R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution. 3:803--807.
-
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
-Barba-Montoya J., Reis M.
-dos, Schneider H., Donoghue P.C., Yang Z.
-2018.
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-New Phytologist.
-218:819--834.
-
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2012.
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-Systematic biology.
-62:298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2015.
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk: Ornithological Advances.
-132:333--348.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle B., Hopkins N., Lu Z., Raygoza Garay J.A., Mozzherin D., Rees T., Matasci N., Narro M.L., Piel W.H., Mckay S.J., Lowry S., Freeland C., Peet R.K., Enquist B.J.
-2013.
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton T., Anderson C.L., Jacquet D., Lundqvist S., Bremer K. 2007. Estimating Divergence Times in Large Phylogenetic Trees. Systematic Biology. 56:741--752.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns K.J., Shultz A.J., Title P.O., Mason N.A., Barker F.K., Klicka J., Lanyon S.M., Lovette I.J.
-2014.
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-Molecular Phylogenetics and Evolution.
-75:41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain S. 2018. bold: Interface to Bold Systems API..
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain S.A., Szöcs E. 2013. taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}. F1000Research. 2:1--29.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo A., Berry V., Douzery E.J., Gascuel O. 2006. SDM: A fast distance-based approach for (super)tree building in phylogenomics. Systematic Biology. 55:740--755.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman J.M., Harmon L.J., Tank D.C. 2013. Congruification: Support for time scaling large phylogenetic trees. Methods in Ecology and Evolution. 4:688--691.
-
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein J. 1985. Phylogenies and the Comparative Method. The American Naturalist. 125:1--15.
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb G.C., England R., Hartig G., McLenachan P.A., Taylor Smith B.L., McComish B.J., Cooper A., Penny D.
-2015.
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-Genome biology and evolution.
-7:2983--2995.
-
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon L., Weir J., Brock C., Glor R., Challenger W. 2008. GEIGER: investigating evolutionary radiations. Bioinformatics. 24:129--131.
-
-\leavevmode\hypertarget{ref-Hedges2006}{}%
-Hedges S.B., Dudley J., Kumar S. 2006. TimeTree: A public knowledge-base of divergence times among organisms. Bioinformatics. 22:2971--2972.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl C. 2008. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..
-
-\leavevmode\hypertarget{ref-Hinchliff2015}{}%
-Hinchliff C.E., Smith S.A., Allman J.F., Burleigh J.G., Chaudhary R., Coghill L.M., Crandall K.A., Deng J., Drew B.T., Gazis R., Gude K., Hibbett D.S., Katz L.A., Laughinghouse H.D., McTavish E.J., Midford P.E., Owen C.L., Ree R.H., Rees J.A., Soltis D.E., Williams T., Cranston K.A.
-2015.
-Synthesis of phylogeny and taxonomy into a comprehensive tree of life.
-Proceedings of the National Academy of Sciences.
-112:12764--12769.
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck J.P., Ronquist F. 2001. MRBAYES: Bayesian inference of phylogenetic trees. Bioinformatics. 17:754--755.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.
-
-\leavevmode\hypertarget{ref-Kumar2017}{}%
-Kumar S., Stecher G., Suleski M., Hedges S.B. 2017. TimeTree: A Resource for Timelines, Timetrees, and Divergence Times. Molecular biology and evolution. 34:1812--1819.
-
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
-Magallón S., Gómez-Acevedo S., Sánchez-Reyes L.L., Hernández-Hernández T.
-2015.
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-New Phytologist.
-207:437--453.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau F., Brown J.W., Winter D.J. 2016. rotl: an R package to interact with the Open Tree of Life data. Methods in Ecology and Evolution. 7:1476--1481.
-
-\leavevmode\hypertarget{ref-morell1996roots}{}%
-Morell V. 1996. The roots of phylogeny. Science. 273:569.
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon H. 2014. Phylogenetic approaches for studying diversification. Ecology Letters. 17:508--525.
-
-\leavevmode\hypertarget{ref-Omeara2019}{}%
-O'Meara B., Md Tayeen A.S., Sanchez Reyes L.L. 2019. Rphylotastic: An r interface to 'phylotastic' web services..
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms J., Chamberlain S. 2018. Phylocomr: Interface to 'phylocom'..
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.
-
-\leavevmode\hypertarget{ref-pearse2013phylogenerator}{}%
-Pearse W.D., Purvis A. 2013. PhyloGenerator: An automated phylogeny generation tool for ecologists. Methods in Ecology and Evolution. 4:692--698.
-
-\leavevmode\hypertarget{ref-Piel2002}{}%
-Piel W.H., Donoghue M., Sanderson M.
-2002.
-TreeBASE : A database of phylogenetic information.
-In: Shimura J., Wilson K., Gordon D., editors.
-To the interoperable ``catalog of life'' with partners.
-Tsukuba, Japan: National Institute for Environmental Studies.
-p.
-41--47.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas P., Crisci J.V., Katinas L. 2006. Historical biogeography: A review of its basic concepts and critical issues. Journal of Arid Environments. 66:389--403.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price T.D., Hooper D.M., Buchanan C.D., Johansson U.S., Tietze D.T., Alström P., Olsson U., Ghosh-Harihar M., Ishtiaq F., Gupta S.K., others.
-2014.
-Niche filling slows the diversification of himalayan songbirds.
-Nature.
-509:222.
-
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
-Rabosky D.L., Chang J., Title P.O., Cowman P.F., Sallan L., Friedman M., Kaschner K., Garilao C., Near T.J., Coll M., others.
-2018.
-An inverse latitudinal gradient in speciation rate for marine fishes.
-Nature.
-559:392.
-
-\leavevmode\hypertarget{ref-ramshaw1972time}{}%
-Ramshaw J., Richardson D., Meatyard B., Brown R., Richardson M., Thompson E., Boulter D.
-1972.
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-New Phytologist.
-71:773--779.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. 2018. R: a language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell L.J. 2012. Phytools: An r package for phylogenetic comparative biology (and other things). Methods in Ecology and Evolution. 3:217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist F., Huelsenbeck J.P. 2003. MrBayes 3: Bayesian phylogenetic inference under mixed models. Bioinformatics. 19:1572--1574.
-
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
-Sanderson M.J., Doyle J.A. 2001. Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data. American Journal of Botany. 88:1499--1516.
-
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
-
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
-Smith S.A., Brown J.W. 2018. Constructing a broadly inclusive seed plant phylogeny. American Journal of Botany. 105:302--314.
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus A., Lapp H., Matasci N., Deus H., Sidlauskas B., Zmasek C.M., Vaidya G., Pontelli E., Cranston K., Vos R., Webb C.O., Harmon L.J., Pirrung M., O'Meara B., Pennell M.W., Mirarab S., Rosenberg M.S., Balhoff J.P., Bik H.M., Heath T.A., Midford P.E., Brown J.W., McTavish E.J., Sukumaran J., Westneat M., Alfaro M.E., Steele A., Jordan G.
-2013.
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
-Title P.O., Rabosky D.L. 2016. Do Macrophylogenies Yield Stable Macroevolutionary Inferences? An Example from Squamate Reptiles. Systematic Biology.:syw102.
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb C.O. 2000. Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. The American Naturalist. 156:145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb C.O., Ackerly D.D., Kembel S.W. 2008. Phylocom: Software for the analysis of phylogenetic community structure and trait evolution. Bioinformatics. 24:2098--2100.
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb C.O., Donoghue M.J. 2005. Phylomatic: Tree assembly for applied phylogenetics. Molecular Ecology Notes. 5:181--183.
-
 % in this file I leave the figure captions outside\ caption{} because I want them to be formatted in the same way as the general text (double spaced and linenumbered)
@@
 % \end{linenumbers}
-
-\end{document}

### 2019-09-30 paper.Rmd
for reproducibility's sake

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 ```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
@@
 # using this for testing instead:
-source("./frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
+source("./data/frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
 loadd() # loads all the targets listed in the workflow plan of the previous `make()`
@@
 rpackage_available <- FALSE
-availability <- ifelse(rpackage_available, "is available", "is **not** available yet")
-thank_origteam <- TRUE
-origteam <- ifelse(thank_origteam, "The team that assembled datelife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.", "")
+availability <- ifelse(rpackage_available, "is available", "will be available")
 ```
@@
 The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project. `r origteam`
+that led to initial work on this project.
+The team that assembled `datelife`'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
 The Open Tree of Life project that provides the open, metadata rich repository of

### 2019-09-30 paper_drafts/Datelife_paper1.Rmd
for reproducibility's sake

--- 2019-08-06 Datelife_paper1.Rmd
@@
 ```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
+# code to generate the paper data and figures:
 source("Datelife_paper1_global.R")
@@
 understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014].
-The primary information needed these time estimates comes from the fossil record.
+The primary information needed for these time estimates comes from the fossil record.
 Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.

### 2019-09-30 paper_drafts/Datelife_paper1_figures.Rmd
for reproducibility's sake

--- 2019-05-20 Datelife_paper1_figures.Rmd
@@
   - \linenumbers
-  - \usepackage{caption}
+  # - \usepackage{caption} # no need to use this one here I think, use it on the yaml of the general file
   # - \usepackage{xr}
@@
 \begin{figure}[!h]
-\includegraphics{fig_runtime1.pdf}
+\includegraphics{fig_runtime_main.pdf}
 \caption{}

### 2019-09-30 paper_drafts/Datelife_paper1_figures_noyaml.tex
for reproducibility's sake

--- 2019-05-29 Intro.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{setspace}
-\setstretch{2}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin = 1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{longtable,booktabs}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\providecommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{}
-    \pretitle{\vspace{\droptitle}}
-  \posttitle{}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
-\usepackage[left]{lineno}
-\linenumbers
-
-\begin{document}
-
-Time of lineage divergence
-constitutes a fundamental piece of information for evolutionary
-understanding in many areas of research, from developmental to conservation biology ({\textbf{???}}; Webb \protect\hyperlink{ref-Webb2000}{2000}).
-This information can be obtained from fossils and from molecular dating studies.
-The developments in DNA sequencing and phylogenetic dating inferences have encouraged/spurred the application of molecular dating methods to a very large diversity of organisms, greatly increasing the quantity
-of data on taxon ages across the tree of life.
-Along with phylogenetic relationships, it allows testing alternative
-evolutionary hypothesis in many areas of research, from developmental to conservation biology ({\textbf{???}}; Webb \protect\hyperlink{ref-Webb2000}{2000}).
-It has been shown that only a bit of information on time of divergence can improve inferences of phylogenetic distance Webb et al. (\protect\hyperlink{ref-Webb2008}{2008}) bladj
-Together with geographical distribution patterns, it is a crucial component of historical biogeography studies (Posadas et al. \protect\hyperlink{ref-posadas2006historical}{2006}).
-Coupled to species number, it allows studying the tempo and mode of speciation and extinction,
-central for understanding how biodiversity
-patterns are shaped across space, time, and clades (Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-In the past two decades, the possibility to obtain good quality DNA sequences
-coupled to methodological developments in phylogenetic and dating inference, allowed
-the application of molecular dating methods to a very large diversity
-of organisms, greatly increasing the quantity and quality of data on taxon ages across the tree of life.
-To date, there is a large amount of
-data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE
-and Open Tree of Life (OToL).
-, --\textgreater{}
-
-The TimeTree project (Hedges et al. \protect\hyperlink{ref-Hedges2006}{2006}, \protect\hyperlink{ref-Hedges2015}{2015}; Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) has aggregated chronograms
-from 3,163 studies, encompassing 97,085 species (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}), and continues to grow.
-However, even in this gold standard resource, the included taxa only encompass between
-0.097 and 3.236\% of total species diversity (following taxonomic expert opinion
-on the global, extant species numbers, which ranges from 3 to 100 million species
-{[}Mayr2010; Moran2011{]}). One advantage of TimeTree is that it includes taxa from
-across the tree of life, versus more specialized chronogram repositories focusing on land plants
-(Stevens and Davis \protect\hyperlink{ref-stevens2001apw}{2001}) and mammals (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}), birds (Jetz et al.
-\protect\hyperlink{ref-Jetz2012}{2012}), and fish (Chang et al.
-\protect\hyperlink{ref-chang2019r}{2019}).
-Users can choose
-between a web interface or a mobile app to receive information on divergence times
-for the evolutionary history of a lineage, pairs of taxa, all lineages within a
-taxon, or a list of taxa. As a science communication tool, TimeTree project is
-very powerful: it has a friendly graphical interface, with informative and colorful
-outputs, that allows the general public to satisfy curiosity regarding a particular
-organism of interest or group of them. It is of limited utility for scientific
-studies, however. The thousands of trees that have been entered are unavailable
-for examination or reuse; according to the creators (see TimeTree web FAQ), methods
-for allowing data downloading have been under discussion for the past several years
-yet the primary data remain closed. Moreover, there is no Application Programming
-Interface (API) allowing programmatic access to any data, greatly impairing the
-possibility of large-scale, automated data-mining, which is not allowed under TimeTree
-website's terms of use. The nearly hundred thousand taxon summary chronogram generated
-from TimeTree resources is not available with its publication (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) or the
-TimeTree website, though the still substantial chronogram from a previous publication
-(Hedges et al. \protect\hyperlink{ref-Hedges2015}{2015}) was made available at OToL.
-
-Despite its great importance, analytical tools to summarize available information
-on taxon ages for the scientific community are still lacking.
-We identified several aspects that might have so far delayed the exploitation of
-existing data. First, original chronograms available publicly are scattered across repositories
-and journals supplementary data, usually with different formats too.
-Second, lineage names due to taxonomic idiosincracy can be different among studies
-and manual curation of that is usually necessary.
-Third, data curation
-
-Platforms that contribute to knowledge in this areas:
-
-Supersmart is an open tool but not easy to use, still requires a lot of curation and knowledge.
-
-Phylogenerator ({\textbf{???}})
-
-Generate new dates using all available DNA sequence information;
-Perform one global analysis using all available information;
-
-Problems or downsides: This might be time consuming for large groups and a lot of
-data curation and knowledge on the group of interest is still necessary. For example,
-choosing correct fossils for calibration
-requires a lot of expertise and knowledge on the group. An incorrect use of fossils
-can generate severe bias in dating results (Sauquet et al. \protect\hyperlink{ref-Sauquet2012c}{2012}).
-Hence, data curation is still an important part of any biological study. The research
-community considers it as an important or even crucial step before data analysis.
-Hence, automated processes for large data analysis are frequently received with skepticism.
-
-\texttt{datelife} palliates this by only using information available from already published
-studies, which are ideally constructed using robust information, such as sequence
-data and thoughtfully curated fossil calibrations.
-
-Rapidly increasing data on time of lineage divergence both from molecular and paleontological studies; the increasing importance of use of these data in distant areas of research, often not specialized enough to rapidly obain data on their own; and the lack of an open (both the data sources and the code underlying the analyses) easy to use tool
-inspired the development of a prototype \texttt{datelife} service over a series
-of phylotastic hackathons (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}) at the National Evolutionary Synthesis Center.
-In this paper we present the first formal description of \texttt{datelife}, featuring an improved database of chronograms, more methods to summarize trees, and new functions to visualize data, as well as comparisons of summary trees.
-\texttt{datelife} is the main service for scaling phylogenetic trees in Phylotastic! system
-(Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013})
-It can be used through its R package , and a web interface
-(\url{http://www.datelife.org/query/}).
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2012.
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-Systematic biology.
-62:298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2015.
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk: Ornithological Advances.
-132:333--348.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns K.J., Shultz A.J., Title P.O., Mason N.A., Barker F.K., Klicka J., Lanyon S.M., Lovette I.J.
-2014.
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-Molecular Phylogenetics and Evolution.
-75:41--77.
-
-\leavevmode\hypertarget{ref-chang2019r}{}%
-Chang J., Rabosky D.L., Smith S.A., Alfaro M.E. 2019. An r package and online resource for macroevolutionary studies using the ray-finned fish tree of life. Methods in Ecology and Evolution. 0:1--7.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb G.C., England R., Hartig G., McLenachan P.A., Taylor Smith B.L., McComish B.J., Cooper A., Penny D.
-2015.
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-Genome biology and evolution.
-7:2983--2995.
-
-\leavevmode\hypertarget{ref-Hedges2006}{}%
-Hedges S.B., Dudley J., Kumar S. 2006. TimeTree: A public knowledge-base of divergence times among organisms. Bioinformatics. 22:2971--2972.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.
-
-\leavevmode\hypertarget{ref-Kumar2017}{}%
-Kumar S., Stecher G., Suleski M., Hedges S.B. 2017. TimeTree: A Resource for Timelines, Timetrees, and Divergence Times. Molecular biology and evolution. 34:1812--1819.
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon H. 2014. Phylogenetic approaches for studying diversification. Ecology Letters. 17:508--525.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas P., Crisci J.V., Katinas L. 2006. Historical biogeography: A review of its basic concepts and critical issues. Journal of Arid Environments. 66:389--403.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price T.D., Hooper D.M., Buchanan C.D., Johansson U.S., Tietze D.T., Alström P., Olsson U., Ghosh-Harihar M., Ishtiaq F., Gupta S.K., others.
-2014.
-Niche filling slows the diversification of himalayan songbirds.
-Nature.
-509:222.
-
-\leavevmode\hypertarget{ref-Sauquet2012c}{}%
-Sauquet H., Ho S.Y.W., Gandolfo M.
-a, Jordan G.J., Wilf P., Cantrill D.J., Bayly M.J., Bromham L., Brown G.K., Carpenter R.J., Lee D.M., Murphy D.J., Sniderman J.M.K., Udovicic F.
-2012.
-Testing the impact of calibration on molecular divergence times using a fossil-rich group: the case of Nothofagus (Fagales).
-Systematic Biology.
-61:289--313.
-
-\leavevmode\hypertarget{ref-stevens2001apw}{}%
-Stevens P.F., Davis H. 2001. Angiosperm phylogeny website..
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus A., Lapp H., Matasci N., Deus H., Sidlauskas B., Zmasek C.M., Vaidya G., Pontelli E., Cranston K., Vos R., Webb C.O., Harmon L.J., Pirrung M., O'Meara B., Pennell M.W., Mirarab S., Rosenberg M.S., Balhoff J.P., Bik H.M., Heath T.A., Midford P.E., Brown J.W., McTavish E.J., Sukumaran J., Westneat M., Alfaro M.E., Steele A., Jordan G.
-2013.
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb C.O. 2000. Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. The American Naturalist. 156:145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb C.O., Ackerly D.D., Kembel S.W. 2008. Phylocom: Software for the analysis of phylogenetic community structure and trait evolution. Bioinformatics. 24:2098--2100.
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb C.O., Donoghue M.J. 2005. Phylomatic: Tree assembly for applied phylogenetics. Molecular Ecology Notes. 5:181--183.
-
 \newpage
@@
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_runtime1.pdf}
+\includegraphics[width=1\linewidth]{fig_runtime_main.pdf}
 \caption{}

### 2019-09-30 paper_drafts/Intro.tex
for reproducibility's sake

--- 2019-05-29 Intro.tex
@@
 constitutes a fundamental piece of information for evolutionary
-understanding in many areas of research, from developmental to conservation biology ({\textbf{???}}; Webb \protect\hyperlink{ref-Webb2000}{2000}).
+understanding in many areas of research, from developmental to conservation biology (Felsenstein \protect\hyperlink{ref-Felsenstein1985a}{1985}; Webb \protect\hyperlink{ref-Webb2000}{2000}).
 This information can be obtained from fossils and from molecular dating studies.
@@
 Along with phylogenetic relationships, it allows testing alternative
-evolutionary hypothesis in many areas of research, from developmental to conservation biology ({\textbf{???}}; Webb \protect\hyperlink{ref-Webb2000}{2000}).
+evolutionary hypothesis in many areas of research, from developmental to conservation biology (Felsenstein \protect\hyperlink{ref-Felsenstein1985a}{1985}; Webb \protect\hyperlink{ref-Webb2000}{2000}).
 It has been shown that only a bit of information on time of divergence can improve inferences of phylogenetic distance Webb et al. (\protect\hyperlink{ref-Webb2008}{2008}) bladj
@@
 
-\newpage
-
-\begin{center}
-\textsc{Figure \ref{fig:workflow}}
-\end{center}
-Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-
-\begin{center}
-\textsc{Figure \ref{fig:runtime1}}
-\end{center}
-Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).
-
-\begin{center}
-\textsc{Figure \ref{fig:schronograms1}}
-\end{center}
-Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
-
-\begin{center}
-\textsc{Figure \ref{fig:summaries}}
-\end{center}
-LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
-
-\begin{center}
-\textsc{Figure \ref{fig:cvbladj}}
-\end{center}
-LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
-The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
-
-\begin{center}
-\textsc{Figure \ref{fig:cvbold}}
-\end{center}
-LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics{Fig1.pdf}
-\caption{}
-\label{fig:workflow}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_runtime1.pdf}
-\caption{}
-\label{fig:runtime1}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
-\caption{}
-\label{fig:schronograms1}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics{fig_summaries.pdf}
-%\includegraphics{fig_summaries.pdf}
-\caption{}
-\label{fig:summaries}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics{fig_crossval_bladj.pdf}
-\caption{}
-\label{fig:cvbladj}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\includegraphics{fig_crossval_boldsumm.pdf}
-\caption{}
-\label{fig:cvbold}
-\end{figure}
-
 \end{document}

### 2019-09-30 paper_drafts/bioarxiv_2019.09/paper_figures.Rmd
for reproducibility's sake

--- 2019-05-20 Datelife_paper1_figures.Rmd
@@
----
-output: 
-  bookdown::pdf_document2:
-    keep_tex: yes
-textalign: left
-geometry: margin = 1in
-bibliography: library_red.bib
-csl: systematic-biology.csl
-link-citations: yes
-linestretch: 2
-header-includes:
-  - \usepackage[left]{lineno}
-  - \linenumbers
-  - \usepackage{caption}
-  # - \usepackage{xr}
-  #- \externaldocument{name of external document to reference fugures from there in here}
----
-
-```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
-source("Datelife_paper1_global.R")
-```
-
-<!-- ```{r myfigure, fig.cap = "My figure"} -->
-<!-- # fig.show = "hide" AND include = FALSE # does not generate the label for the figure -->
-<!-- # fig.cap = " " inactivates the label reference -->
-<!-- plot(1:10) -->
-<!-- ``` -->
-
-<!-- My plot is called \@ref(fig:myfigure). -->
-
-<!-- \begin{figure}[!ht] -->
-<!-- \includegraphics{blank.png} -->
-<!-- \caption{} -->
-<!-- \label{fig:workflow} -->
-<!-- \end{figure} -->
-<!-- Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with DateLife, via the R package or through the website.
-Details on the functions involved on each workflow are shown in `datelife`'s R package vignette.
--->
-
-<!-- \begin{figure}[!ht] -->
-<!-- \includegraphics{blank.png} -->
-<!-- \caption{} -->
-<!-- \label{fig:runtime1} -->
-<!-- \end{figure} -->
-<!-- Computation time of input processing and search across `datelife`s  -->
-<!--   chronogram database. -->
-
-<!-- \begin{figure}[!ht] -->
-<!-- \includegraphics{blank.png} -->
-<!-- \caption{} -->
-<!-- \label{fig:schronograms} -->
-<!-- \end{figure} -->
-<!-- Lineage through time (LTT) plots of Fringillidae (bird family of true finches) source chronograms.
-The latter are obtained by searching a chronogram database and subsetting chronograms conaning at least two species.
-Arrows indicate maximum age of each source chronogram and numbers indicate the reference of original studies 1: @barker2012going, 2: @barker2015new, 3: @burns2014phylogenetics, 4: @claramunt2015new, 5: @gibb2015new, 6: @Hedges2015, 7: @hooper2017chromosomal, 8: @Jetz2012, 9: @price2014niche.
--->
-
-<!-- \begin{figure}[!ht] -->
-<!-- \includegraphics{blank.png} -->
-<!-- \caption{} -->
-<!-- \label{fig:cvbladj} -->
-<!-- \end{figure} -->
-<!-- LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarizing information from source chronograms found for the Fringillidae. Arrows indicate maximum age. -->
-
-<!-- \newpage -->
-
 Figure 1. 

### 2019-09-30 paper_drafts/bioarxiv_2019.09/report.Rmd
for reproducibility's sake

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: ../svm-latex-ms.tex
    # includes:
      # in_header: preamble.tex
      # after_body: [Datelife_paper1_figures_noyaml.tex]
link-citations: yes
bibliography: ../library_red.bib
header-includes:
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{caption}
biblio-style: apsr
title: "DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life"
thanks: "Replication files are available on the author's Github account (http://github.com/LunaSare). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding authors**: lsanche7@utk.edu; bomeara@utk.edu; 1. Corresponding adrdress: *Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA*"
author:
- name: Luna L. Sanchez Reyes
  affiliation: University of Tennessee, Knoxville
- name: Brian O'Meara
  affiliation: University of Tennessee, Knoxville
abstract: "The combination of new analytical techniques, availability of more fossil and molecular data, and better practices in data sharing has resulted in a steady accumulation of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree of Life for a large quantity and diversity of organisms in the last few decades. However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication. `datelife` is a service implemented via an R package and a web site (<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of published data on lineage divergence times. The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string. A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices. Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram. Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query. If there is no information available for any queried taxa, data can be simulated. All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis. Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology. How well this trees work for this purpose still needs to be tested. `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results."
keywords: "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: double
endnote: no
nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
```

``` {r child = '../paper.Rmd'}
```

\newpage

\begin{center}
\textsc{References}
\end{center}

``` {r child = 'paper_figures.Rmd'}
```

### 2019-09-30 paper_drafts/bioarxiv_2019.09/report.tex
for reproducibility's sake

--- 2019-09-29 biorxiv_2019.09.28.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{setspace}
-\setstretch{2}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\documentclass[11pt,]{article}
+\usepackage[left=1in,top=1in,right=1in,bottom=1in]{geometry}
+\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
+\usepackage[]{mathpazo}
+
   \usepackage[T1]{fontenc}
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
+
+\usepackage{abstract}
+\renewcommand{\abstractname}{}    % clear the title
+\renewcommand{\absnamepos}{empty} % originally center
+
+\renewenvironment{abstract}
+ {{%
+    \setlength{\leftmargin}{0mm}
+    \setlength{\rightmargin}{\leftmargin}%
+  }%
+  \relax}
+ {\endlist}
+
+\makeatletter
+\def\@maketitle{%
+  \newpage
+%  \null
+%  \vskip 2em%
+%  \begin{center}%
+  \let \footnote \thanks
+    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
+}
+%\fi
+\makeatother
+
+\setcounter{secnumdepth}{0}
+
+\title{DateLife: Leveraging databases and analytical tools to reveal the dated
+Tree of Life \thanks{Replication files are available on the author's Github account
+(\url{http://github.com/LunaSare}). \textbf{Current version}: September
+29, 2019; \textbf{Corresponding authors}:
+\href{mailto:lsanche7@utk.edu}{\nolinkurl{lsanche7@utk.edu}};
+\href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}; 1.
+Corresponding adrdress: \emph{Department of Ecology and Evolutionary
+Biology, University of Tennessee, Knoxville, 425 Hesler Biology
+Building, Knoxville, TN 37996, USA}}  }
+
+\author{\Large Luna L.
+Sanchez Reyes\vspace{0.05in} \newline\normalsize\emph{University of Tennessee, Knoxville}   \and \Large Brian O'Meara\vspace{0.05in} \newline\normalsize\emph{University of Tennessee, Knoxville}  }
+
+\date{}
+
+\usepackage{titlesec}
+
+\titleformat*{\section}{\normalsize\bfseries}
+\titleformat*{\subsection}{\normalsize\itshape}
+\titleformat*{\subsubsection}{\normalsize\itshape}
+\titleformat*{\paragraph}{\normalsize\itshape}
+\titleformat*{\subparagraph}{\normalsize\itshape}
+
+\usepackage{natbib}
+\bibliographystyle{apsr}
+\usepackage[strings]{underscore} % protect underscores in most circumstances
+
+\newtheorem{hypothesis}{Hypothesis}
+\usepackage{setspace}
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\usepackage{hyperref}
+\usepackage[left]{lineno}
+\linenumbers
+\usepackage{caption}
+
+% move the hyperref stuff down here, after header-includes, to allow for - \usepackage{hyperref}
+
+\makeatletter
+\@ifpackageloaded{hyperref}{}{%
 \ifxetex
-    \usepackage{mathspec}
+  \PassOptionsToPackage{hyphens}{url}\usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
 \else
-    \usepackage{fontspec}
+  \PassOptionsToPackage{hyphens}{url}\usepackage[draft,unicode=true]{hyperref}
 \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin = 1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
+}
+
+\@ifpackageloaded{color}{
+    \PassOptionsToPackage{usenames,dvipsnames}{color}
+}{%
+    \usepackage[usenames,dvipsnames]{color}
+}
+\makeatother
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={Luna L. Sanchez Reyes (University of Tennessee, Knoxville) and Brian O'Meara (University of Tennessee, Knoxville)},
+             pdfkeywords = {Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science;
+Congruification; Supertree; Calibrations},  
+            pdftitle={DateLife: Leveraging databases and analytical tools to reveal the dated
+Tree of Life},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage{longtable,booktabs}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
+
+% Add an option for endnotes. -----
+
+% add tightlist ----------
 \providecommand{\tightlist}{%
 \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
+% add some other packages ----------
 
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\providecommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{}
-    \pretitle{\vspace{\droptitle}}
-  \posttitle{}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
-\usepackage{caption}
-\usepackage{blindtext}
-\usepackage[left]{lineno}
-\linenumbers
+% \usepackage{multicol}
+% This should regulate where figures float
+% See: https://tex.stackexchange.com/questions/2275/keeping-tables-figures-close-to-where-they-are-mentioned
+\usepackage[section]{placeins}
 
@@
 
-\textbf{Title}: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
+% \pagenumbering{arabic}% resets `page` counter to 1 
+%
+% \maketitle
 
-\textbf{Authors}: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}
+{% \usefont{T1}{pnc}{m}{n}
+\setlength{\parindent}{0pt}
+\thispagestyle{plain}
+{\fontsize{18}{20}\selectfont\raggedright 
+\maketitle  % title \par
 
-\textbf{Correspondence address}:
+}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA}
-\end{enumerate}
+{
+   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
+\textbf{\authorfont Luna L.
+Sanchez Reyes} \hskip 15pt \emph{\small University of Tennessee, Knoxville}   \par \textbf{\authorfont Brian O'Meara} \hskip 15pt \emph{\small University of Tennessee, Knoxville}
 
-\textbf{Corresponding authors}: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
+}
+
+}
+
+\begin{abstract}
+
+    \hbox{\vrule height .2pt width 39.14pc}
+
+    \vskip 8.5pt % \small
+
+\noindent The combination of new analytical techniques, availability of more
+fossil and molecular data, and better practices in data sharing has
+resulted in a steady accumulation of chronograms in public and open
+databases such as TreeBASE, Dryad, and Open Tree of Life for a large
+quantity and diversity of organisms in the last few decades. However,
+getting a tree with branch lengths proportional to time remains
+difficult for many biologists and the non-academic community, despite
+its importance in many areas of research, education, and science
+communication. \texttt{datelife} is a service implemented via an R
+package and a web site (\url{http://www.datelife.org/}) for efficient
+reuse, summary and reanalysis of published data on lineage divergence
+times. The main workflow starts with at least two taxon names as input,
+either as tip labels on a tree, or as a simple comma separated character
+string. A name search is then performed across the chronogram database
+and positively identified source trees are pruned to maintain queried
+taxa only and stored as a named list of patristic distance matrices.
+Source chronogram data can be summarised using branch length summary
+statistics or variance minimizing approaches to generate a single
+summary chronogram. Source chronogram data can also be used as
+calibration points to date a tree containing some or all names from the
+initial query. If there is no information available for any queried
+taxa, data can be simulated. All source and summary chronograms can be
+saved in formats that permit easy reuse and reanalysis. Summary and
+newly generated trees are potentially useful to evaluate evolutionary
+hypothesis in different areas of research in biology. How well this
+trees work for this purpose still needs to be tested. \texttt{datelife}
+will be useful to increase awereness on the existing variation in expert
+time of divergence data, and might foster exploration of the effect of
+alternative divergence time hypothesis on the results of analyses,
+nurturing a culture of more cautious interpretation of evolutionary
+results.
+
+\vskip 8.5pt \noindent \emph{Keywords}: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science;
+Congruification; Supertree; Calibrations \par
+
+    \hbox{\vrule height .2pt width 39.14pc}
+
+\end{abstract}
+
+\vskip -8.5pt
+
+ % removetitleabstract
+
+\noindent \doublespacing
 
@@
 
-\textbf{abstract.-}
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
-of Life for a large quantity and diversity of organisms in the last few decades.
-However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication.
-\texttt{datelife} is a service implemented via an R package and a web site
-(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of published data on lineage divergence times.
-The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
-Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram.
-Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
-If there is no information available for any queried taxa, data can be simulated.
-All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this trees work for this purpose still needs to be tested.
-\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+\textbf{abstract.-} The combination of new analytical techniques,
+availability of more fossil and molecular data, and better practices in
+data sharing has resulted in a steady accumulation of chronograms in
+public and open databases such as TreeBASE, Dryad, and Open Tree of Life
+for a large quantity and diversity of organisms in the last few decades.
+However, getting a tree with branch lengths proportional to time remains
+difficult for many biologists and the non-academic community, despite
+its importance in many areas of research, education, and science
+communication. \texttt{datelife} is a service implemented via an R
+package and a web site (\url{http://www.datelife.org/}) for efficient
+reuse, summary and reanalysis of published data on lineage divergence
+times. The main workflow starts with at least two taxon names as input,
+either as tip labels on a tree, or as a simple comma separated character
+string. A name search is then performed across the chronogram database
+and positively identified source trees are pruned to maintain queried
+taxa only and stored as a named list of patristic distance matrices.
+Source chronogram data can be summarised using branch length summary
+statistics or variance minimizing approaches to generate a single
+summary chronogram. Source chronogram data can also be used as
+calibration points to date a tree containing some or all names from the
+initial query. If there is no information available for any queried
+taxa, data can be simulated. All source and summary chronograms can be
+saved in formats that permit easy reuse and reanalysis. Summary and
+newly generated trees are potentially useful to evaluate evolutionary
+hypothesis in different areas of research in biology. How well this
+trees work for this purpose still needs to be tested. \texttt{datelife}
+will be useful to increase awereness on the existing variation in expert
+time of divergence data, and might foster exploration of the effect of
+alternative divergence time hypothesis on the results of analyses,
+nurturing a culture of more cautious interpretation of evolutionary
+results.
 
-\textbf{Keywords:} Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
-\newpage
+\textbf{Keywords:} Tree; Phylogeny; Scaling; Dating; Ages; Divergence
+times; Open Science; Congruification; Supertree; Calibrations \newpage
 
@@
 Clade ages represent a fundamental piece of information for evolutionary
-understanding in many areas of research, from developmental to conservation biology (Felsenstein \protect\hyperlink{ref-Felsenstein1985a}{1985}; Webb \protect\hyperlink{ref-Webb2000}{2000}), from historical biogeography to species diversification studies (Posadas et al.
-\protect\hyperlink{ref-posadas2006historical}{2006}; Morlon \protect\hyperlink{ref-Morlon2014}{2014}).
-The primary information needed for these time estimates comes from the fossil record.
-Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
-The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}).
-Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree.
-That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al.
-\protect\hyperlink{ref-Stoltzfus2013}{2013}).
+understanding in many areas of research, from developmental to
+conservation biology \citep{Felsenstein1985a, Webb2000}, from historical
+biogeography to species diversification studies
+\citep{posadas2006historical, Morlon2014}. The primary information
+needed for these time estimates comes from the fossil record. Coupled
+with phylogenies with branch lengths based on molecular and/or
+morphological data, the time of divergence of extant and extinct
+lineages can be reconstructed with molecular dating methods. The number
+of studies publishing phylogenies with branch lengths proportional to
+geological time (hereafter chronograms) have constantly increased in
+number for the last two decades \citep{Kumar2017}. Still, generating a
+chronogram is not an easy task unless you have specialized training: it
+requires inferring a tree, understanding what fossil data are available
+and their limits, and where fossils go on the tree. That is why there
+has been an urge for promoting and facilitating reuse of the vast amount
+of phylogenetic and time of lineage divergence data that has been
+generated and made available in publications, for the advantage of
+research relying on this information
+\citep{webb2005phylomatic, Stoltzfus2013}.
 
-Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
-TreeBASE (Morell \protect\hyperlink{ref-morell1996roots}{1996}; Piel et al.
-\protect\hyperlink{ref-Piel2002}{2002}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al.
-\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms store and make available published phylogenies and chronograms for easy scientific reuse.
-All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
-OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
+Wide interest from the scientific community to make information from
+phylogenies in general and chronograms in particular available for
+consultation and reuse has spurred the creation of public platforms with
+various goals and characteristics. TreeBASE
+\citep{morell1996roots, Piel2002}, the Dryad repository
+(\url{http://datadryad.org/}), and the Open Tree of Life
+\citep[OToL;][]{Hinchliff2015} platforms store and make available
+published phylogenies and chronograms for easy scientific reuse. All of
+them can be queried using automatised web procedures, which permit
+personalized, large scale queries that are also very fast. OToL stores
+trees with branch length information from a wide range of living
+organisms, implementing a metadata structure that stores the branch
+length units (i.e., time or relative susbtitution rates). Treebase and
+Dryad repositories also contain trees from all groups of life, but the
+former did not store branch length information until recently (and lacks
+consistent metadata on what any branchlengths stored mean) and Dryad
+stores many other types of biological data using metadata that does not
+allow automatic distinction of types of trees and branch length units,
+impairing the automatised access to time of lineage divergence
+information.
 
-Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the phylogenetic relationships among known lineages.
-All or parts of this synthetic tree can be reused for any purpose. However, it currently only focus on synthesizing tree topology, meaning that it does not expose branch length data of any type.
-The Timetree of Life project focuses on the synthesis of a single chronogram of life (Hedges et al.
-\protect\hyperlink{ref-Hedges2006}{2006}) and presents a very accessible, attractive interface.
-However, the thousands of chronograms this NSF-funded project have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but large scale download remains prohibited by their site.
-The latest version of their synthetic chronogram (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}) can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
-Other platforms such as SuperSmart (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}) and phylogenerator (Pearse and Purvis \protect\hyperlink{ref-pearse2013phylogenerator}{2013}) are focused in automatised \emph{de novo} chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
-However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Moreover, these tools do not provide information from already created expert chronograms.
+Besides keeping a repository to easily store and share expert
+phylogenetic and chronogram knowledge, OToL also has the primary goal of
+synthesising all trees in their repository to expose to the community a
+single tree of all life depicting the phylogenetic relationships among
+known lineages. All or parts of this synthetic tree can be reused for
+any purpose. However, it currently only focus on synthesizing tree
+topology, meaning that it does not expose branch length data of any
+type. The Timetree of Life project focuses on the synthesis of a single
+chronogram of life \citep{Hedges2006} and presents a very accessible,
+attractive interface. However, the thousands of chronograms this
+NSF-funded project have compiled for synthesis are only publicly
+available for visual examination in their website or for download as
+images, but large scale download remains prohibited by their site. The
+latest version of their synthetic chronogram \citep{Kumar2017} can be
+queried only through their website in a non-automatised fashion, and
+only subsets of it can be reused for analyses with the permission of the
+authors. Other platforms such as SuperSmart
+\citep{antonelli2017supersmart} and phylogenerator
+\citep{pearse2013phylogenerator} are focused in automatised \emph{de
+novo} chronogram inference, by reusing DNA sequence data to reconstruct
+phylogenetic trees. However, expert fossil information necessary for
+subsequent molecular dating analyses still needs to be compiled and
+curated by the user, rendering them a challenging tool to obtain data on
+time of lineage divergence for the non-specialist. Moreover, these tools
+do not provide information from already created expert chronograms.
 
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fuly public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
-A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al.
-\protect\hyperlink{ref-Stoltzfus2013}{2013}).
-In here we present the formal description and implementation of the \texttt{datelife} service, constituted by an R package and a web site (\url{http://www.datelife.org/}).
-There is still much room for improvement, and flaws and limitations are addressed below.
-We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+A tool for efficient reuse of expert, published data on time of lineage
+divergence should have an open and fuly public chronogram database
+storing data in a format suitable for scientific reuse, an automatised
+way of accessing the information, and straightforward means of comparing
+and summarizing chronogram information as needed by the user. A
+prototype service aiming to meet this characteristics was developed over
+a series of hackathons at the National Evolutionary Synthesis Center
+\citep{Stoltzfus2013}. In here we present the formal description and
+implementation of the \texttt{datelife} service, constituted by an R
+package and a web site (\url{http://www.datelife.org/}). There is still
+much room for improvement, and flaws and limitations are addressed
+below. We strived for the current implementation of \texttt{datelife} to
+perform the basic tasks described above, featuring a system for
+maintenance of an open database of chronograms pulled from public
+repositories, methods to summarize and compare source chronograms, and
+new functions to visualize and graphically compare source and summary
+chronograms.
 
@@
 
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
+The basic \texttt{datelife} workflow is shown in figure
+\ref{fig:workflow} and consists of:
 
@@
 
-To gather, process, and present information, \texttt{datelife} builds up from functions
-available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
-geiger (Harmon et al.
-\protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}), taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl \protect\hyperlink{ref-Heibl2008}{2008}), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018}) and rphylotastic (O'Meara et al.
-\protect\hyperlink{ref-Omeara2019}{2019}).
+To gather, process, and present information, \texttt{datelife} builds up
+from functions available in several R packages including rotl
+\citep{Michonneau2016}, ape \citep{Paradis2004}, geiger
+\citep{Harmon2008}, paleotree \citep{Bapst2012a}, bold
+\citep{Chamberlain2018}, phytools \citep{Revell2012}, taxize
+\citep{Chamberlain2013, Chamberlain2018}, phyloch \citep{Heibl2008},
+phylocomr \citep{Ooms2018} and rphylotastic \citep{Omeara2019}.
 
-A \texttt{datelife} search currently accepts scientific names only. It can be any named clade or binomial specific.
-Chronogram search is performed at the species level, so when input names correspond
-to named clades, \texttt{datelife} pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
-Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
-\texttt{datelife} processes input names with the taxon name resolution service (TNRS; Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}),
-which corrects potentially misspelled names and typos, and standardizes spelling
-variations and synonyms , increasing the probability of correctly finding the
-queried taxa in \texttt{datelife}'s chronogram database.
+A \texttt{datelife} search currently accepts scientific names only. It
+can be any named clade or binomial specific. Chronogram search is
+performed at the species level, so when input names correspond to named
+clades, \texttt{datelife} pulls all accepted species names within the
+clade from OToL's reference taxonomy to perform the search. Searches at
+the infraspecies level are not currently allowed, so input names
+belonging to subspecies or any other infraspecific category are
+collapsed to the species level. \texttt{datelife} processes input names
+with the taxon name resolution service \citep[TNRS;][]{Boyle2013}, which
+corrects potentially misspelled names and typos, and standardizes
+spelling variations and synonyms , increasing the probability of
+correctly finding the queried taxa in \texttt{datelife}'s chronogram
+database.
 
-The chronogram search is performed across \texttt{datelife}'s chronogram database which is assembled from OToL's tree repository.
-Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
-Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
+The chronogram search is performed across \texttt{datelife}'s chronogram
+database which is assembled from OToL's tree repository. Compared to
+other existing open tree repositories OToL's metadata rich tree store is
+the only one that supports search, identification, and handling of
+chronograms in an automatised fashion. Also, all their chronograms come
+from peer-reviewed published studies generated by specialists in the
+targeted lineages, arguably representing expert knowledge on time of
+lineage divergence.
 
-Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean.
-A much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths (Criscuolo et al.
-\protect\hyperlink{ref-Criscuolo2006}{2006}) is also implemented via the ape package (Paradis et al.
-\protect\hyperlink{ref-Paradis2004}{2004}).
-The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
-However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife\_examples package).
-Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
-The Branch Length Adjuster (BLADJ) algorithm (Webb et al.
-\protect\hyperlink{ref-Webb2008}{2008}) was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
+Information from source chronograms can be summarised with a summary
+statistic of tree branch lengths, such as median or mean. A much slower,
+but possibly more accurate Super Distance Matrix (SDM) approach for
+supertree reconstruction with branch lengths \citep{Criscuolo2006} is
+also implemented via the ape package \citep{Paradis2004}. The resulting
+summary patristic distance matrix could be clustered with classic
+algorithms to return a tree. However, we noticed that the resulting
+trees are often non-ultrameric and do not reflect the source chronogram
+data (see datelife\_examples package). Instead, we obtained a
+distribution of age data from the summary matrix available for nodes on
+a consensus tree. The Branch Length Adjuster (BLADJ) algorithm
+\citep{Webb2008} was then used to distribute node ages evenly over the
+consensus tree, minimizing age variance in the resulting chronogram.
 
-For tree dating, the congruification algorithm described by Eastman et al. (\protect\hyperlink{ref-Eastman2013}{2013})
-is implemented to find shared nodes between trees (congruent nodes).
-The ages of these nodes are then used as calibrations to date any given tree.
-Currently implemented methods for tree dating are BLADJ, MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) and PATHd8 (Britton et al.
-\protect\hyperlink{ref-Britton2007}{2007}), a non-clock, rate-smoothing dating method.
+For tree dating, the congruification algorithm described by
+\citet{Eastman2013} is implemented to find shared nodes between trees
+(congruent nodes). The ages of these nodes are then used as calibrations
+to date any given tree. Currently implemented methods for tree dating
+are BLADJ, MrBayes \citep{Huelsenbeck2001, Ronquist2003} and PATHd8
+\citep{Britton2007}, a non-clock, rate-smoothing dating method.
 
@@
 
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS service is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes.
-A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}) that were run locally
---using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 50\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
+\texttt{datelife}'s code speed was tested on an Apple iMac with one 3.4
+GHz Intel Core i5 processor. We registered variation in computing time
+of query processing and search through the database relative to number
+of queried taxon names. Query processing increases roughly linearly with
+number of input taxon names, and increases considerably if TNRS service
+is activated. Up to ten thousand names can be processed and searched in
+less than 30 minutes. A name search through the chronogram database with
+an already processed query can be performed in less than a minute, even
+with a very large number of taxon names (Fig. \ref{fig:runtime1}).
+\texttt{datelife}'s code performance was evaluated with a set of unit
+tests designed and implemented with the R package testthat
+\citep{RCoreTeam2018} that were run locally --using the devtools package
+\citep{RCoreTeam2018}, and on a public server --via GitHub, using the
+continuous integration tool Travis CI (\url{https://travis-ci.org}). At
+present, unit tests cover more than 50\% of \texttt{datelife}'s code
+(\url{https://codecov.io/gh/phylotastic/datelife}).
 
@@
 
-In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain all data on lineage divergence available in \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp.
-according to the Open Tree of Life taxonomy).
-There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig.
-\ref{fig:schronograms1}).
-Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
-As explained in the \texttt{Description}, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are generally consistent with source ages.
-Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms (Ramshaw et al.
-\protect\hyperlink{ref-ramshaw1972time}{1972}; Sanderson and Doyle \protect\hyperlink{ref-sanderson2001sources}{2001}; Magallón et al.
-\protect\hyperlink{ref-magallon2015metacalibrated}{2015}; Barba-Montoya et al.
-\protect\hyperlink{ref-barba2018constraining}{2018}).
-For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
-Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.
+In this section we demonstrate the types of outputs that can be obtained
+with \texttt{datelife}, using as an example the bird family Fringillidae
+of true finches. We performed a higher-taxon search to obtain all data
+on lineage divergence available in \texttt{datelife}'s database for all
+recognised species within the Fringillidae (475 spp. according to the
+Open Tree of Life taxonomy). There are 13 chronograms containing at
+least two Fringillidae species, published in 9 different studies (Fig.
+\ref{fig:schronograms1}). Data from these source chronograms was used to
+generate two types of summary chronograms, median and SDM. As explained
+in the \texttt{Description}, data from source chronograms was first
+summarised into a single distance matrix (using either the median or the
+SDM method) and then the available node ages were used as calibrations
+points over a consensus tree topology, to obtain a dated tree with the
+program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are
+older and have wider variation in maximum ages than chronograms obtained
+with SDM. In both cases, ages are generally consistent with source ages.
+Different source chronograms often show substantial variation in ages
+for clades (see, for example, the ongoing debate about crown group age
+of angiosperms
+\citep{barba2018constraining, magallon2015metacalibrated, sanderson2001sources, ramshaw1972time}.
+For some studies, especially ones based on branch lengths (studies of
+diversification, timing of events, trait evolution, and more), using a
+different chronogram may return different results
+\citep{title2016macrophylogenies}. Stitching together these chronograms
+can create a larger tree and uses information from multiple studies, but
+the effect of uncertainties and errors here on downstream analyses still
+requires more research.
 
-Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs.
-\ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms.
-In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
-Maximum tree ages were only approximately recovered in one case (study 2; Fig.
-\ref{fig:cvbladj}).
-Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
-three cases (studies 3, 5, and 9; Fig.
-\ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and details can be consulted in \url{https://github.com/LunaSare/datelife_examples}.
+Data from source chronograms was also used to date tree topologies with
+no branch length information and trees with branch lengths in relative
+substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a
+form of cross validation, we used tree topologies from each study and
+calibrated them using information from all other source chronograms. In
+the absence of branch length data, the ages of internal nodes were
+approximately recovered in almost all cases (except for studies 3, and
+5; Fig. \ref{fig:cvbladj}). Maximum tree ages were only approximately
+recovered in one case (study 2; Fig. \ref{fig:cvbladj}). Branch lengths
+were successfully generated using the BOLD database for all source
+chronograms. However, dating with PATHd8 (using congruified
+calibrations) was only successful in three cases (studies 3, 5, and 9;
+Fig. \ref{fig:cvbold}). From these, two trees have a different sampling
+than the original source chronogram, mainly because DNA data for some
+species is absent from the BOLD. Maximum ages are quite different from
+source chronograms, but this might be explained also by the differences
+in sampling between source chronograms and BOLD trees. More examples and
+details can be consulted in
+\url{https://github.com/LunaSare/datelife_examples}.
 
@@
 
-The main goal of \texttt{datelife} is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
-However, it has many flaws.
-Some of them can be overcome, some of them might represent limitations.
+The main goal of \texttt{datelife} is to make expert information on time
+of lineage divergence easily accesible for comparison, reuse, and
+reanalysis, to researchers in all areas of science and with all levels
+of expertise in the matter. It is a very fast tool that fulfills the
+quality of openness and does not require any expert biological knowledge
+from users --besides the names of the organisms they want to work with--
+for any of its functionalities. However, it has many flaws. Some of them
+can be overcome, some of them might represent limitations.
 
-At the moment, \texttt{datelife}'s chronogram database is not very large, storing 231 chronograms up to the time the manuscript was written.
-This represents 5.79\% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining (Kumar et al.
-\protect\hyperlink{ref-Kumar2017}{2017}).
-OToL is the only public tree repository from where \texttt{datelife} can currently pull chronograms to construct its database.
-A previous version of TimeTree's synthetic chronogram (Hedges et al.
-\protect\hyperlink{ref-Hedges2015}{2015}) was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
-This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
-Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to \texttt{datelife}'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.
+At the moment, \texttt{datelife}'s chronogram database is not very
+large, storing 231 chronograms up to the time the manuscript was
+written. This represents 5.79\% of the largest existing chronogram
+database, which is not open for scientific reuse nor automatised data
+mining \citep{Kumar2017}. OToL is the only public tree repository from
+where \texttt{datelife} can currently pull chronograms to construct its
+database. A previous version of TimeTree's synthetic chronogram
+\citep{Hedges2015} was made available in the OToL repository, hence the
+amount of lineages represented in datelife's database is at least as
+substantial as TimeTree's. This ensures that some information will be
+available for any given query, but it does not ensure that the full
+state of knowledge of time of divergence data will be available for any
+given lineage. Thus, incorporation of more published chronograms
+deposited in OpenTree or perhaps pulled from Dryad directly, to
+\texttt{datelife}'s database is crucial to improve its services. Methods
+to automatically mine chronogram data from the Dryad repository could be
+designed and implemented. However, the unit of branch lengths would
+still need to be determined by hand. Consequently, we would like to
+emphasize on the importance of sharing chronogram data for the
+scientific community, in repositories that require expert input and
+manual curation, such as OToL's tree repository.
 
-Another potential concern comes from summary chronograms.
-We currently summarize all source chronogram data by default.
-Users can subset source data if they have reasons to favor some or one source chronogram over others.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there is no objective way to determine if an expert chronogram is better than other.
-Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
-Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk \protect\hyperlink{ref-schenk2016sec}{2016}).
-Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-This remains to be tested.
+Another potential concern comes from summary chronograms. We currently
+summarize all source chronogram data by default. Users can subset source
+data if they have reasons to favor some or one source chronogram over
+others. Strictly speaking, a good chronogram should reflect the real
+time of lineage divergence accurately and precisely. To our knowledge,
+there is no objective way to determine if an expert chronogram is better
+than other. Some criteria that have been put forward are the level of
+lineage sampling and the number of calibrations used. Scientists usually
+also favor chronograms coming from studies with primary calibrations to
+ones from secondary calibrations. It has been observed with simulations
+that divergence times inferred with secondary calibrations are
+significantly younger than those inferred with primary calibrations in
+analyses performed with bayesian inference methods when priors are
+implemented in similar ways in both analyses \citep{schenk2016sec}. Yet,
+there are different ways to use secondary calibrations and the bias
+might not be encountered with other dating methods that do not require
+prior assumptions (such as ML methods). This remains to be tested.
 
-Furthermore, even chronograms obtained with primary data can be very different, as observed from the comparison of source chronograms in the Fringillidae example.
-A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms (e.g., angiosperms; Magallón et al.
-\protect\hyperlink{ref-magallon2015metacalibrated}{2015}).
-Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
-Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignment (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
-To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
-Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.
-Hence, summarizing chronograms might imply summarizing evolutionary hypothesis.
-This could be good from certain point of view, since it could help to get a single global evolutionary history for a lineage.
-But it could also be bad, since we could be loosing part of the evolutionary history that is only being reflected in some chronograms and not from the summary chronogram.
-Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history (Antonelli et al.
-\protect\hyperlink{ref-antonelli2017supersmart}{2017}).
+Furthermore, even chronograms obtained with primary data can be very
+different, as observed from the comparison of source chronograms in the
+Fringillidae example. A large discrepancy in time of lineage divergence
+across expert knowledge is well known for different groups of organisms
+\citep[e.g., angiosperms;][]{magallon2015metacalibrated}. Comparison of
+available chronogram data for a wide range of organisms shown here
+suggest that this is a widespread phenomenon that requires further
+attention. Characteristics of the data used for dating analyses as well
+as from the output chronogram itself, such as quality of alignment
+(missing data, GC content), lineage sampling strategy and proportion,
+phylogenetic and dating inference method, number of fossils used as
+calibrations, support for nodes and ages, and confidence intervals could
+be used to score quality of source chronograms. To facilitate subsetting
+of source chronograms following different criteria, this information
+should be included as metadata manually entered by curators (as is done
+in OToL) in the future. Still, even if all source chronograms have been
+generated by excellent standards and using similar methods, the
+evolutionary history they depict might be very different. Hence,
+summarizing chronograms might imply summarizing evolutionary hypothesis.
+This could be good from certain point of view, since it could help to
+get a single global evolutionary history for a lineage. But it could
+also be bad, since we could be loosing part of the evolutionary history
+that is only being reflected in some chronograms and not from the
+summary chronogram. Ideally, we should still rely on time of lineage
+divergence data obtained from a single analysis using fossil data as
+primary sources of calibrations, and using fossils that have already
+been curated as calibrations to date other trees, which should reflect a
+more homogeneous evolutionary history \citep{antonelli2017supersmart}.
 This will be implemented in future \texttt{datelife} versions.
 
-In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
-Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ (Webb et al.
-\protect\hyperlink{ref-Webb2008}{2008})) for several taxa lacking this information.
-Making up branch lengths in this or other ways is accepted in scientific publications: Rabosky et al.
-(\protect\hyperlink{ref-rabosky2018inverse}{2018}) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data; Smith and Brown (\protect\hyperlink{ref-smith2018constructing}{2018}) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.
+In other areas of biological research, such as ecology and conservation
+biology, it has been indicated that at least some data on lineage
+divergence represents a relevant improvement for testing alternative
+hypothesis using phylogenetic distance. Hence, we allow accepted ways of
+creating branch lengths in the absence of starting branch length
+information (such as BLADJ \citep{Webb2008}) for several taxa lacking
+this information. Making up branch lengths in this or other ways is
+accepted in scientific publications: \citet{rabosky2018inverse} created
+a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\%
+had molecular data; \citet{Jetz2012}, created a time-calibrated tree of
+all 9,993 bird species, where 67\% had molecular data;
+\citet{smith2018constructing} constructed a tree of 353,185 seed plants
+where only 23\% had molecular data. Taken to the extreme, one could make
+a fully resolved, calibrated tree of all modern and extinct taxa using a
+single taxonomy and a single calibration with the polytomy resolution
+and branch imputation methods. There has yet to be a thorough analysis
+of what can go wrong when one goes beyond the data in this way, so we
+urge caution; we also urge readers to follow the example of many of the
+large tree papers cited above and make sure results are substantially
+similar between trees containing only taxa with molecular or other data
+and trees that combine those taxa with taxa from taxonomy alone.
 
@@
 
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. Generating this information is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills
-to produce chronograms \emph{de novo}. Knowledge on clade ages is also crucial for
-science communication and education.
+Divergence time information is key to many areas of evolutionary
+studies: trait evolution, diversification, biogeography, macroecology
+and more. Generating this information is difficult, especially for those
+who want to use phylogenies but who are not systematists, or do not have
+the time to acquire and develop the necessary knowledge and data
+curation skills to produce chronograms \emph{de novo}. Knowledge on
+clade ages is also crucial for science communication and education.
 
-\texttt{datelife} allows an easy and fast obtention, as well as comparison of publicly available information
-on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, allowing identification of regions that require more research or that have conflicting information.
-Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
-\texttt{datelife} helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+\texttt{datelife} allows an easy and fast obtention, as well as
+comparison of publicly available information on time of lineage
+divergence, providing a straightforward way to get an informed idea on
+the state of knowledge of the time frame of evolution of different
+regions of the tree of life, allowing identification of regions that
+require more research or that have conflicting information. Both summary
+and newly generated trees are potentially useful to evaluate
+evolutionary hypothesis in different areas of research.
+\texttt{datelife} helps with awereness on the existing variation in
+expert time of divergence data, and might foster exploration of the
+effect of alternative divergence time hypothesis on the results of
+analyses, nurturing a culture of more cautious interpretation of
+evolutionary results.
 
@@
 
-\texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version is \textbf{not} available yet
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
+\texttt{datelife} is free and open source and it can be used through its
+current website \url{http://www.datelife.org/query/}, through its R
+package, and through Phylotastic's project web portal
+\url{http://phylo.cs.nmsu.edu:3000/}. \texttt{datelife}'s website is
+maintained using RStudio's shiny server and the shiny package open
+infrastructure, as well as Docker. \texttt{datelife}'s R package stable
+version is \textbf{not} available yet for installation from the CRAN
+repository (\url{https://cran.r-project.org/package=datelife}) using the
+command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
+Development versions are available from the GitHub repository
+(\url{https://github.com/phylotastic/datelife}) and can be installed
+using the command
+\texttt{devtools::install\_github("phylotastic/datelife")}.
 
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are respectively in the \href{https://github.com/LunaSare/datelife_paper1}{datelife\_paper1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR GitHub account.
+Code used to generate all versions of this manuscript, the biological
+examples, as well as the benchmark of functionalities are respectively
+in the
+\href{https://github.com/LunaSare/datelife_paper1}{datelife\_paper1},
+\href{https://github.com/LunaSare/datelife_examples}{datelife\_examples},
+and
+\href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark}
+repositories in LLSR GitHub account.
 
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
+Funding was provided by the US National Science Foundation (NSF) grants
+ABI-1458603 to Datelife project and DBI-0905606 to the National
+Evolutionary Synthesis Center (NESCent), and the Phylotastic project
+Grant ABI-1458572.
 
@@
 
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled datelife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
+We thank colleagues from the O'Meara Lab at the University of Tennesse
+Knoxville for suggestions, discussions and software testing. The late
+National Evolutionary Synthesis Center (NESCent), which sponsored
+hackathons that led to initial work on this project. The team that
+assembled datelife's first proof of concept: Tracy Heath, Jonathan
+Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and
+Luke Harmon. The Open Tree of Life project that provides the open,
+metadata rich repository of trees used for \texttt{datelife}. The many
+scientists who publish their chronograms in an open, reusable form, and
+the scientists who curate them for deposition in the Open Tree of Life
+repository. The NSF for funding nearly all the above, in addition to the
+ABI grant that funded this project itself.
 
 \newpage
+
 \begin{center}
@@
 
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
-Antonelli A., Hettling H., Condamine F.L., Vos K., Nilsson R.H., Sanderson M.J., Sauquet H., Scharn R., Silvestro D., Töpel M., Bacon C.D., Oxelman B., Vos R.A.
-2017.
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
-Systematic Biology.
-66:153--166.
+Figure 1. Stylized DateLife workflow. This shows the general worflows
+and analyses that can be performed with DateLife, via the R package or
+through the website. Details on the functions involved on each workflow
+are shown in \texttt{datelife}'s R package vignette.
 
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst D.W. 2012. Paleotree: An R package for paleontological and phylogenetic analyses of evolution. Methods in Ecology and Evolution. 3:803--807.
+Figure 2. Computation time of input processing and search across
+\texttt{datelife}s chronogram database.
 
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
-Barba-Montoya J., Reis M.
-dos, Schneider H., Donoghue P.C., Yang Z.
-2018.
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-New Phytologist.
-218:819--834.
+Figure 3. Lineage through time (LTT) plots of source chronograms
+containing all or a subset of species from the bird family Fringillidae
+of true finches. Arrows indicate maximum age of each chronogram. Numbers
+reference to chronograms' original publications 1:
+\citet{barker2012going}, 2: \citet{barker2015new}, 3:
+\citet{burns2014phylogenetics}, 4: \citet{claramunt2015new}, 5:
+\citet{gibb2015new}, 6: \citet{Hedges2015}, 7:
+\citet{hooper2017chromosomal}, 8: \citet{Jetz2012}, 9:
+\citet{price2014niche}.
 
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2012.
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-Systematic biology.
-62:298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker F.K., Burns K.J., Klicka J., Lanyon S.M., Lovette I.J.
-2015.
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk: Ornithological Advances.
-132:333--348.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle B., Hopkins N., Lu Z., Raygoza Garay J.A., Mozzherin D., Rees T., Matasci N., Narro M.L., Piel W.H., Mckay S.J., Lowry S., Freeland C., Peet R.K., Enquist B.J.
-2013.
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton T., Anderson C.L., Jacquet D., Lundqvist S., Bremer K. 2007. Estimating Divergence Times in Large Phylogenetic Trees. Systematic Biology. 56:741--752.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns K.J., Shultz A.J., Title P.O., Mason N.A., Barker F.K., Klicka J., Lanyon S.M., Lovette I.J.
-2014.
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-Molecular Phylogenetics and Evolution.
-75:41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain S. 2018. bold: Interface to Bold Systems API..
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain S.A., Szöcs E. 2013. taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}. F1000Research. 2:1--29.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt S., Cracraft J. 2015. A new time tree reveals earth history's imprint on the evolution of modern birds. Science advances. 1:e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo A., Berry V., Douzery E.J., Gascuel O. 2006. SDM: A fast distance-based approach for (super)tree building in phylogenomics. Systematic Biology. 55:740--755.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman J.M., Harmon L.J., Tank D.C. 2013. Congruification: Support for time scaling large phylogenetic trees. Methods in Ecology and Evolution. 4:688--691.
-
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein J. 1985. Phylogenies and the Comparative Method. The American Naturalist. 125:1--15.
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb G.C., England R., Hartig G., McLenachan P.A., Taylor Smith B.L., McComish B.J., Cooper A., Penny D.
-2015.
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-Genome biology and evolution.
-7:2983--2995.
-
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon L., Weir J., Brock C., Glor R., Challenger W. 2008. GEIGER: investigating evolutionary radiations. Bioinformatics. 24:129--131.
-
-\leavevmode\hypertarget{ref-Hedges2006}{}%
-Hedges S.B., Dudley J., Kumar S. 2006. TimeTree: A public knowledge-base of divergence times among organisms. Bioinformatics. 22:2971--2972.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges S.B., Marin J., Suleski M., Paymer M., Kumar S. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution. 32:835--845.
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl C. 2008. PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages..
-
-\leavevmode\hypertarget{ref-Hinchliff2015}{}%
-Hinchliff C.E., Smith S.A., Allman J.F., Burleigh J.G., Chaudhary R., Coghill L.M., Crandall K.A., Deng J., Drew B.T., Gazis R., Gude K., Hibbett D.S., Katz L.A., Laughinghouse H.D., McTavish E.J., Midford P.E., Owen C.L., Ree R.H., Rees J.A., Soltis D.E., Williams T., Cranston K.A.
-2015.
-Synthesis of phylogeny and taxonomy into a comprehensive tree of life.
-Proceedings of the National Academy of Sciences.
-112:12764--12769.
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper D.M., Price T.D. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature ecology \& evolution. 1:1526.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck J.P., Ronquist F. 2001. MRBAYES: Bayesian inference of phylogenetic trees. Bioinformatics. 17:754--755.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz W., Thomas G., Joy J.J., Hartmann K., Mooers A. 2012. The global diversity of birds in space and time. Nature. 491:444--448.
-
-\leavevmode\hypertarget{ref-Kumar2017}{}%
-Kumar S., Stecher G., Suleski M., Hedges S.B. 2017. TimeTree: A Resource for Timelines, Timetrees, and Divergence Times. Molecular biology and evolution. 34:1812--1819.
-
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
-Magallón S., Gómez-Acevedo S., Sánchez-Reyes L.L., Hernández-Hernández T.
-2015.
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-New Phytologist.
-207:437--453.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau F., Brown J.W., Winter D.J. 2016. rotl: an R package to interact with the Open Tree of Life data. Methods in Ecology and Evolution. 7:1476--1481.
-
-\leavevmode\hypertarget{ref-morell1996roots}{}%
-Morell V. 1996. The roots of phylogeny. Science. 273:569.
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon H. 2014. Phylogenetic approaches for studying diversification. Ecology Letters. 17:508--525.
-
-\leavevmode\hypertarget{ref-Omeara2019}{}%
-O'Meara B., Md Tayeen A.S., Sanchez Reyes L.L. 2019. Rphylotastic: An r interface to 'phylotastic' web services..
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms J., Chamberlain S. 2018. Phylocomr: Interface to 'phylocom'..
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis E., Claude J., Strimmer K. 2004. APE: analyses of phylogenetics and evolution in R language. Bioinformatics. 20:289--290.
-
-\leavevmode\hypertarget{ref-pearse2013phylogenerator}{}%
-Pearse W.D., Purvis A. 2013. PhyloGenerator: An automated phylogeny generation tool for ecologists. Methods in Ecology and Evolution. 4:692--698.
-
-\leavevmode\hypertarget{ref-Piel2002}{}%
-Piel W.H., Donoghue M., Sanderson M.
-2002.
-TreeBASE : A database of phylogenetic information.
-In: Shimura J., Wilson K., Gordon D., editors.
-To the interoperable ``catalog of life'' with partners.
-Tsukuba, Japan: National Institute for Environmental Studies.
-p.
-41--47.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas P., Crisci J.V., Katinas L. 2006. Historical biogeography: A review of its basic concepts and critical issues. Journal of Arid Environments. 66:389--403.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price T.D., Hooper D.M., Buchanan C.D., Johansson U.S., Tietze D.T., Alström P., Olsson U., Ghosh-Harihar M., Ishtiaq F., Gupta S.K., others.
-2014.
-Niche filling slows the diversification of himalayan songbirds.
-Nature.
-509:222.
-
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
-Rabosky D.L., Chang J., Title P.O., Cowman P.F., Sallan L., Friedman M., Kaschner K., Garilao C., Near T.J., Coll M., others.
-2018.
-An inverse latitudinal gradient in speciation rate for marine fishes.
-Nature.
-559:392.
-
-\leavevmode\hypertarget{ref-ramshaw1972time}{}%
-Ramshaw J., Richardson D., Meatyard B., Brown R., Richardson M., Thompson E., Boulter D.
-1972.
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-New Phytologist.
-71:773--779.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. 2018. R: a language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell L.J. 2012. Phytools: An r package for phylogenetic comparative biology (and other things). Methods in Ecology and Evolution. 3:217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist F., Huelsenbeck J.P. 2003. MrBayes 3: Bayesian phylogenetic inference under mixed models. Bioinformatics. 19:1572--1574.
-
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
-Sanderson M.J., Doyle J.A. 2001. Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data. American Journal of Botany. 88:1499--1516.
-
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
-
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
-Smith S.A., Brown J.W. 2018. Constructing a broadly inclusive seed plant phylogeny. American Journal of Botany. 105:302--314.
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus A., Lapp H., Matasci N., Deus H., Sidlauskas B., Zmasek C.M., Vaidya G., Pontelli E., Cranston K., Vos R., Webb C.O., Harmon L.J., Pirrung M., O'Meara B., Pennell M.W., Mirarab S., Rosenberg M.S., Balhoff J.P., Bik H.M., Heath T.A., Midford P.E., Brown J.W., McTavish E.J., Sukumaran J., Westneat M., Alfaro M.E., Steele A., Jordan G.
-2013.
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-BMC Bioinformatics.
-14.
-
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
-Title P.O., Rabosky D.L. 2016. Do Macrophylogenies Yield Stable Macroevolutionary Inferences? An Example from Squamate Reptiles. Systematic Biology.:syw102.
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb C.O. 2000. Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. The American Naturalist. 156:145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb C.O., Ackerly D.D., Kembel S.W. 2008. Phylocom: Software for the analysis of phylogenetic community structure and trait evolution. Bioinformatics. 24:2098--2100.
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb C.O., Donoghue M.J. 2005. Phylomatic: Tree assembly for applied phylogenetics. Molecular Ecology Notes. 5:181--183.
-
-% in this file I leave the figure captions outside\ caption{} because I want them to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+Figure 4. LTT plots of median and Supermatrix Distance Method (SDM)
+chronograms summarizing information from source chronograms found for
+the Fringillidae. Arrows indicate maximum age.
 
 \newpage
+
 \begin{figure}[!h]
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:workflow}}
-% \end{center}
-Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
+
 \newpage
+
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_runtime_main.pdf}
+\includegraphics{fig_runtime1.pdf}
 \caption{}
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:runtime1}}
-% \end{center}
-Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).
+
 \newpage
+
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{fig_schronograms1.pdf}
+\includegraphics{fig_schronograms1.pdf}
 \caption{}
-\label{fig:schronograms1}
+\label{fig:schronograms}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:schronograms1}}
-% \end{center}
-Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
+
 \newpage
-\begin{figure}[!h]
-\includegraphics{fig_summaries.pdf}
-%\includegraphics{fig_summaries.pdf}
-\caption{}
-\label{fig:summaries}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:summaries}}
-% \end{center}
-LTT plots of median and Supermatrix Distance Method (SDM) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate maximum age.
-\newpage
-\begin{figure}[!h]
+
+\begin{figure}[!ht]
 \includegraphics{fig_crossval_bladj.pdf}
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbladj}}
-% \end{center}
-LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
-The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
+
 \newpage
-\begin{figure}[!h]
-\includegraphics{fig_crossval_boldsumm.pdf}
-\caption{}
-\label{fig:cvbold}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbold}}
-% \end{center}
-LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.
-
-% \end{linenumbers}
+\singlespacing 
+\bibliography{../library\_red.bib}
 

### 2019-09-30 Datelife_paper1.R

@@ -1,11 +1,0 @@
-# Domingo 14 de enero 2018
-library(citer)
-library(rmarkdown)
-setwd("~/Library/Mobile Documents/com~apple~CloudDocs/Desktop/datelife_paper1")
-pandoc_convert("Datelife_paper1.md", to="pdf", citeproc = TRUE, output = "Datelife_paper1.pdf")
-# pandoc: Unknown writer: pdf
-# To create a pdf with pandoc, use the latex or beamer writer and specify
-# an output file with .pdf extension (pandoc -t latex -o filename.pdf).
-# Error: pandoc document conversion failed with error 9
-
-rmarkdown::render(input = "~/Desktop/datelife_paper1/Datelife_paper1.Rmd", output_format = "pdf_document")

### 2019-09-30 Datelife_paper1_global.R

@@ -1,31 +1,0 @@
-# fig_workflow_cap <- paste("Stylized DateLife workflow. This shows the general worflows 
-#                           and analyses that can be performed with DateLife, via the R 
-#                           package or through the website. Details on the functions involved 
-#                           on each workflow are shown in datelife's R package vignette.")
-download.file(url = "https://github.com/phylotastic/datelife/raw/b3532e70b679272093658731d249534e84e15e28/data-raw/benchmark/runtime_tests/2_tests/2_random_spp_names/mdq_plus_gdr_2018.05.29_fill.pdf", 
-              destfile = "fig_runtime1.pdf", mode="wb")
-# fig_runtime1_cap <- "Computation time of input processing and search across `DateLife`s 
-#   chronogram database"
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_phyloall.pdf",
-              destfile = "fig_schronograms1.pdf", mode="wb")
-# fig_schronograms_cap <- "Lineage through time (LTT) plots of source chronograms containing 
-#   all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers 
-#   indicate chrinograms' original publication: 1) [@barker2012going], 2) [@barker2015new], 3) [@burns2014phylogenetics],
-#   4) [@claramunt2015new], 5) [@gibb2015new], 6) [@Hedges2015], 7) [@hooper2017chromosomal],
-#   8) [@Jetz2012], 9) [@price2014niche]"
-  
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_summary_chronograms2.pdf",
-              destfile = "fig_summaries.pdf", mode="wb")
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_crossval_bladj.pdf",
-              destfile = "fig_crossval_bladj.pdf", mode="wb")
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_lttplot_crossval_pathd8_summ1.pdf",
-              destfile = "fig_crossval_boldsumm.pdf", mode="wb")
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_clusters_both.pdf",
[…]

### 2019-09-30 Rebe_Net_diversification.R

@@ -1,342 +1,0 @@
-##Estimación de tasas de diversificación con el método de momentos usando las ecuaciones de 
-##Magallón y Sanderson 2001 aplicado al árbol fechado FBD80 de Malvales.
-install.packages("geiger")
-require ("geiger")
-getwd()
-##################################################
-#2 de febrero de 2018
-#corregí las edades de cada grupo
-#voy a obtener la diversificación neta con el método de 
-#momentos.
-#las edades son del árbol FBD80_v4_comb.treeanot.tree
-data<-read.csv("CrownAges_species_richness.csv")
-data
-class(data)
-##diversificación neta con el comando bd.ms
-#nombre del taxón_r(diversificación neta)_epsilon_c(crown)/s(stem)
-malvoi_r_001_c<-bd.ms(time=data$time[1], n=data$sp[1], epsilon=0.01)
-malvoi_r_001_c
-#voy a tratar de hacer este cálculo para cada familia y subfamilia con un
-#loop:
-#Primero voy a crear un objeto en el que se deben imprimir los resultados
-#del loop
-net_div<-0
-net_div
-#loop
-for(i in 1:19){
-  net_div[i] <- bd.ms(time=data$time[i], n=data$sp[i],epsilon=0.01)
-  print(net_div[i])
-}
-
-print(i)
-net_div
-data_net_div<-data.frame(data$taxon, net_div)
-data_net_div
-#Hecho
-###############################################################################
-#Ahora voy a calcular r (diversificación neta) con el valor de turnoverFBD
-#que estimó el modelo: 0.9195
-net_div_0.91<-0
-for(i in 1:19){
-	net_div_0.91[i]<-bd.ms(time=data$time[i], n=data$sp[i],epsilon=0.9195)
-	print (net_div_0.91[i])
-}
-net_div_e0.91<-data.frame(data$taxon, net_div_0.91)
-net_div_e0.91
-colnames(net_div_e0.91)<- c("taxon", "crown_e0.91")
-##############################################################################
[…]

### 2019-09-30 bioarxiv_2019.09/data.R

@@ -1,29 +1,0 @@
-library(datelife)
-data(opentree_chronograms)
-  dq = make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE)
-  dr = get_datelife_result(input = dq)
-  summ = get_taxon_summary(datelife_query = dq, datelife_result = dr)
-  phyloall = summarize_datelife_result(datelife_query = dq,
-    datelife_result = dr, summary_format = "phylo_all",
-    taxon_summary = "none")
-  cb_cols1 = c("#0072B2", "#D55E00", "#440154FF", "#CC79A7", "#440154FF", "#009E73",
-    "#9ad0f3", "#95D840FF", "#F0E442")
-  cb_cols2 = c("#440154FF", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00",
-    "#CC79A7", "#F0E442", "#95D840FF")
-  ltt_plot = plot_ltt_phyloall(taxon = "Fringillidae", phy = phyloall, ltt_colors = cb_cols1, tax_datedotol = NULL,
-      file_name = "ltt_plot_poster.pdf", file_dir = getwd(), height = 3, width = 7.9, add_legend = FALSE, add_title = FALSE)
-  datedotol = get_dated_otol_induced_subtree(input = dq, ott_ids = dq$ott_ids)
-  otol = get_otol_synthetic_tree(input = dq)
-  bestgrove = get_best_grove(datelife_result = dr)
-  sdm_matrix = get_sdm_matrix(datelife_result = bestgrove$best_grove)
-  median_matrix = datelife_result_median_matrix(datelife_result = bestgrove$best_grove)
-  phy_sdm = summary_matrix_to_phylo_all(summ_matrix = sdm_matrix, target_tree = otol)
-  phy_median = summary_matrix_to_phylo_all(summ_matrix = median_matrix, target_tree = otol)
-  ltt_summ = plot_ltt_summary(taxon = "Fringillidae", phy = phyloall, phy_sdm, phy_median,
-        file_name = "ltt_summ_poster.pdf", file_dir = "~//Desktop//datelife_paper1//", height = 3, width = 7.9,
-        add_legend = TRUE)
-  each_calibrations = get_all_calibrations(phyloall, each = TRUE)
-  sim_otol = use_each_calibration(phy = otol, calibrations = each_calibrations)$phy
[…]

### 2019-09-30 bioarxiv_2019.09/figures_download.R

@@ -1,26 +1,0 @@
-# fig_workflow_cap <- paste("Stylized DateLife workflow. This shows the general worflows 
-#                           and analyses that can be performed with DateLife, via the R 
-#                           package or through the website. Details on the functions involved 
-#                           on each workflow are shown in datelife's R package vignette.")
-download.file(url = "https://github.com/phylotastic/datelife/raw/b3532e70b679272093658731d249534e84e15e28/data-raw/benchmark/runtime_tests/2_tests/2_random_spp_names/mdq_plus_gdr_2018.05.29_fill.pdf", 
-              destfile = "fig_runtime1.pdf", mode="wb")
-# fig_runtime1_cap <- "Computation time of input processing and search across `DateLife`s 
-#   chronogram database"
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_phyloall.pdf",
-              destfile = "fig_schronograms1.pdf", mode="wb")
-# fig_schronograms_cap <- "Lineage through time (LTT) plots of source chronograms containing 
-#   all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers 
-#   indicate chrinograms' original publication: 1) [@barker2012going], 2) [@barker2015new], 3) [@burns2014phylogenetics],
-#   4) [@claramunt2015new], 5) [@gibb2015new], 6) [@Hedges2015], 7) [@hooper2017chromosomal],
-#   8) [@Jetz2012], 9) [@price2014niche]"
-  
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_summary_chronograms2.pdf",
-              destfile = "fig_summaries.pdf", mode="wb")
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_crossval_bladj.pdf",
-              destfile = "fig_crossval_bladj.pdf", mode="wb")
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_lttplot_crossval_pathd8_summ1.pdf",
-              destfile = "fig_crossval_boldsumm.pdf", mode="wb")
-download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_clusters_both.pdf",
-              destfile = "fig_lttplot_clusters.pdf", mode="wb")
-
-

### 2019-09-30 data/frin_plan.R

@@ -1,0 +1,35 @@
+library(datelife)
+library(drake)
+frin_plan <- drake_plan(
+  strings_in_dots = "literals",
+  dq = make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE),
+  dr = get_datelife_result(input = dq),
+  summ = get_taxon_summary(datelife_query = dq, datelife_result = dr),
+  phyloall = summarize_datelife_result(datelife_query = dq,
+    datelife_result = dr, summary_format = "phylo_all",
+    taxon_summary = "none"),
+  cb_cols1 = c("#0072B2", "#D55E00", "#440154FF", "#CC79A7", "#440154FF", "#009E73",
+    "#9ad0f3", "#95D840FF", "#F0E442"),
+  cb_cols2 = c("#440154FF", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00",
+    "#CC79A7", "#F0E442", "#95D840FF"),
+  ltt_plot = plot_ltt_phyloall(taxon = "Fringillidae", phy = phyloall, ltt_colors = cb_cols1, tax_datedotol = NULL,
+      file_name = "ltt_plot_poster.pdf", file_dir = getwd(), height = 3, width = 7.9, add_legend = FALSE, add_title = FALSE),
+  datedotol = get_dated_otol_induced_subtree(input = dq, ott_ids = dq$ott_ids),
+  bestgrove = get_best_grove(datelife_result = dr),
+  sdm_matrix = get_sdm_matrix(datelife_result = bestgrove$best_grove),
+  median_matrix = datelife_result_median_matrix(datelife_result = bestgrove$best_grove),
+  otol = get_otol_synthetic_tree(input = dq),
+  phy_sdm = summary_matrix_to_phylo_all(summ_matrix = sdm_matrix, target_tree = otol),
+  phy_median = summary_matrix_to_phylo_all(summ_matrix = median_matrix, target_tree = otol),
+  ltt_summ = plot_ltt_summary(taxon = "Fringillidae", phy = phyloall, phy_sdm, phy_median,
+        file_name = "ltt_summ_poster.pdf", file_dir = "~//Desktop//datelife_paper1//", height = 3, width = 7.9,
+        add_legend = TRUE),
+  each_calibrations = get_all_calibrations(phyloall, each = TRUE)
+  # , # inactivating the following because having trouble with getting otol for some reason
[…]

### 2019-09-30 figures_download.R

@@ -1,0 +1,24 @@
+# fig_workflow_cap <- paste("Stylized DateLife workflow. This shows the general worflows
+#                           and analyses that can be performed with DateLife, via the R
+#                           package or through the website. Details on the functions involved
+#                           on each workflow are shown in datelife's R package vignette.")
+download.file(url = "https://github.com/phylotastic/datelife/raw/b3532e70b679272093658731d249534e84e15e28/data-raw/benchmark/runtime_tests/2_tests/2_random_spp_names/mdq_plus_gdr_2018.05.29_fill.pdf",
+              destfile = "./figures/fig_runtime1.pdf", mode="wb")
+# fig_runtime1_cap <- "Computation time of input processing and search across `DateLife`s
+#   chronogram database"
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_phyloall.pdf",
+              destfile = "./figures/fig_schronograms1.pdf", mode="wb")
+# fig_schronograms_cap <- "Lineage through time (LTT) plots of source chronograms containing
+#   all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers
+#   indicate chrinograms' original publication: 1) [@barker2012going], 2) [@barker2015new], 3) [@burns2014phylogenetics],
+#   4) [@claramunt2015new], 5) [@gibb2015new], 6) [@Hedges2015], 7) [@hooper2017chromosomal],
+#   8) [@Jetz2012], 9) [@price2014niche]"
+
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_summary_chronograms2.pdf",
+              destfile = "./figures/fig_summaries.pdf", mode="wb")
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_crossval_bladj.pdf",
+              destfile = "./figures/fig_crossval_bladj.pdf", mode="wb")
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_lttplot_crossval_pathd8_summ1.pdf",
+              destfile = "./figures/fig_crossval_boldsumm.pdf", mode="wb")
[…]

### 2019-09-30 frin_plan.R

@@ -1,35 +1,0 @@
-library(datelife)
-library(drake)
-frin_plan <- drake_plan(
-  strings_in_dots = "literals",
-  dq = make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE),
-  dr = get_datelife_result(input = dq),
-  summ = get_taxon_summary(datelife_query = dq, datelife_result = dr),
-  phyloall = summarize_datelife_result(datelife_query = dq,
-    datelife_result = dr, summary_format = "phylo_all",
-    taxon_summary = "none"),
-  cb_cols1 = c("#0072B2", "#D55E00", "#440154FF", "#CC79A7", "#440154FF", "#009E73",
-    "#9ad0f3", "#95D840FF", "#F0E442"),
-  cb_cols2 = c("#440154FF", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00",
-    "#CC79A7", "#F0E442", "#95D840FF"),
-  ltt_plot = plot_ltt_phyloall(taxon = "Fringillidae", phy = phyloall, ltt_colors = cb_cols1, tax_datedotol = NULL,
-      file_name = "ltt_plot_poster.pdf", file_dir = getwd(), height = 3, width = 7.9, add_legend = FALSE, add_title = FALSE),
-  datedotol = get_dated_otol_induced_subtree(input = dq, ott_ids = dq$ott_ids),
-  bestgrove = get_best_grove(datelife_result = dr),
-  sdm_matrix = get_sdm_matrix(datelife_result = bestgrove$best_grove),
-  median_matrix = datelife_result_median_matrix(datelife_result = bestgrove$best_grove),
-  otol = get_otol_synthetic_tree(input = dq),
-  phy_sdm = summary_matrix_to_phylo_all(summ_matrix = sdm_matrix, target_tree = otol),
-  phy_median = summary_matrix_to_phylo_all(summ_matrix = median_matrix, target_tree = otol),
-  ltt_summ = plot_ltt_summary(taxon = "Fringillidae", phy = phyloall, phy_sdm, phy_median,
-        file_name = "ltt_summ_poster.pdf", file_dir = "~//Desktop//datelife_paper1//", height = 3, width = 7.9,
-        add_legend = TRUE),
-  each_calibrations = get_all_calibrations(phyloall, each = TRUE)
-  # , # inactivating the following because having trouble with getting otol for some reason
[…]

### 2019-09-30 paper_drafts/Datelife_paper1.R

@@ -1,0 +1,11 @@
+# Domingo 14 de enero 2018
+library(citer)
+library(rmarkdown)
+setwd("~/Library/Mobile Documents/com~apple~CloudDocs/Desktop/datelife_paper1")
+pandoc_convert("Datelife_paper1.md", to="pdf", citeproc = TRUE, output = "Datelife_paper1.pdf")
+# pandoc: Unknown writer: pdf
+# To create a pdf with pandoc, use the latex or beamer writer and specify
+# an output file with .pdf extension (pandoc -t latex -o filename.pdf).
+# Error: pandoc document conversion failed with error 9
+
+rmarkdown::render(input = "~/Desktop/datelife_paper1/Datelife_paper1.Rmd", output_format = "pdf_document")

### 2019-09-30 paper_drafts/Datelife_paper1_global.R

@@ -1,0 +1,31 @@
+# fig_workflow_cap <- paste("Stylized DateLife workflow. This shows the general worflows 
+#                           and analyses that can be performed with DateLife, via the R 
+#                           package or through the website. Details on the functions involved 
+#                           on each workflow are shown in datelife's R package vignette.")
+download.file(url = "https://github.com/phylotastic/datelife/raw/b3532e70b679272093658731d249534e84e15e28/data-raw/benchmark/runtime_tests/2_tests/2_random_spp_names/mdq_plus_gdr_2018.05.29_fill.pdf", 
+              destfile = "fig_runtime1.pdf", mode="wb")
+# fig_runtime1_cap <- "Computation time of input processing and search across `DateLife`s 
+#   chronogram database"
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_phyloall.pdf",
+              destfile = "fig_schronograms1.pdf", mode="wb")
+# fig_schronograms_cap <- "Lineage through time (LTT) plots of source chronograms containing 
+#   all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers 
+#   indicate chrinograms' original publication: 1) [@barker2012going], 2) [@barker2015new], 3) [@burns2014phylogenetics],
+#   4) [@claramunt2015new], 5) [@gibb2015new], 6) [@Hedges2015], 7) [@hooper2017chromosomal],
+#   8) [@Jetz2012], 9) [@price2014niche]"
+  
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_summary_chronograms2.pdf",
+              destfile = "fig_summaries.pdf", mode="wb")
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_crossval_bladj.pdf",
+              destfile = "fig_crossval_bladj.pdf", mode="wb")
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_lttplot_crossval_pathd8_summ1.pdf",
+              destfile = "fig_crossval_boldsumm.pdf", mode="wb")
+download.file(url = "https://github.com/LunaSare/datelife_examples/raw/master/docs/plots/Fringillidae_LTTplot_clusters_both.pdf",
[…]

### 2019-09-30 tol_dating.R

@@ -1,157 +1,0 @@
-# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
-# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
-loadd(tol_jetz1)
-tol_jetz_query <- drake_plan(
-  tol_jetz_dq = make_datelife_query(input = tol_jetz1, get_spp_from_taxon = FALSE),
-  tol_jetz_dr = get_datelife_result(input = tol_dq),
-  tol_summ = get_taxon_summary(datelife_query = tol_dq, datelife_result = tol_dr),
-  # for report table only:
-  tol_dq = tol_jetz_dq,
-  tol_dr = tol_jetz_dr,
-  tol_summ = tol_jetz_summ,
-  reportname = "Jetz_dr_table0",
-  report = make_report("table0_pdf_template.Rmd", paste0("inst/docs/", reportname, ".md"), "try5"),
-  table0_caption = "Jetz et al. 2012 source chronogram original studies information.",
-  taxon = "Birds",
-  table0_pdf = render_pdf(reportname, "inst/docs", "try5")
-)
-make(tol_jetz_query)
-
-load("data/tol_jetz1.RData")
-loadd(tol_dr)
-loadd(tol_jetz2)
-jetz_phylo_all <- vector(mode = "list", length = length(tol_dr))
-is_jetz <- grepl("Jetz", names(tol_dr))
-is_biggest <- which(grepl("Hedges", names(tol_dr)))
-jetz_phylo_all[[which(is_jetz)[1]]] <- tol_jetz1[[1]]
-jetz_phylo_all[[which(is_jetz)[2]]] <- tol_jetz2[[1]]
-
-for(i in seq(jetz_phylo_all)[-is_biggest]){
-  if(!is_jetz[i]){
-    print(i)
-    jetz_phylo_all[[i]] <- datelife::patristic_matrix_to_phylo(tol_dr[[i]])
-  }
-}
-names(jetz_phylo_all) <- names(tol_dr)
-class(jetz_phylo_all) <- "multiPhylo"
-save(jetz_phylo_all, file = "data/jetz_phylo_all.RData")
[…]

### 2019-09-30 tol_ott_ids.R

@@ -1,39 +1,0 @@
-# GET OTT IDS FOR JETZ ET AL TREE
-load("data/new.tree_tnrs.RData")
-names(new.tree_tnrs)
-nrow(new.tree_tnrs)
-loadd(tol_jetz1)
-head(new.tree_tnrs)
-head(tol_jetz1[[1]]$tip.label)
-mm_unique <- match(tol_jetz1[[1]]$tip.label, new.tree_tnrs$unique) # position of tip labels in unique
-mm_search <- match(tolower(gsub(" ", "_", tol_jetz1[[1]]$tip.label)), new.tree_tnrs$search)
-sum(is.na(mm_unique)) # 387
-sum(is.na(mm_search))# 436, mm_unique and mm_search match different bird tip labels
-# make a vector of indices of matching tip labels with tnrs table:
-jetz_ott_ids_ii <- rep(0, length(tol_jetz1[[1]]$tip.label))
-jetz_ott_ids_ii[!is.na(mm_unique)] <- mm_unique[!is.na(mm_unique)]
-sum(jetz_ott_ids_ii == 0) # 387, ok
-# those that are not in unique but are on search
-# are the ones that have a different name in ott taxonomy
-ii <- jetz_ott_ids_ii ==0 & !is.na(mm_search)
-sum(ii) # 114
-jetz_ott_ids_ii[ii] <- mm_search[ii]
-sum(jetz_ott_ids_ii == 0) # 273, these are the ones that are not in search nor unique, so search!
-ss <- data.frame(jetz = tol_jetz1[[1]]$tip.label[ii],
-  search = new.tree_tnrs$search[mm_search[ii]])
-  new.tree_tnrs[mm_search[ii],]
-dd <- data.frame(jetz = tol_jetz1[[1]]$tip.label[jetz_ott_ids_ii != 0], search = new.tree_tnrs$search[jetz_ott_ids_ii], unique = new.tree_tnrs$unique[jetz_ott_ids_ii])
-
-jetz_ott_ids <- rep(NA, length(tol_jetz1[[1]]$tip.label))
-jetz_ott_ids[jetz_ott_ids_ii != 0] <- new.tree_tnrs$ott_id[jetz_ott_ids_ii]
-tol_jetz1[[1]]$ott_ids <- as.numeric(jetz_ott_ids)
-new_search <- tol_jetz1[[1]]$tip.label[is.na(tol_jetz1[[1]]$ott_ids)]
[…]

### 2019-09-30 tol_plan.R

@@ -1,71 +1,0 @@
-library(datelife)
-library(drake)
-class(opentree_chronograms$trees) <- "multiPhylo"
-make_report <- function(rmdname, mdname, placeholder){
-  knitr::knit(knitr_in(rmdname), file_out(mdname), quiet = TRUE)
-}
-render_pdf <- function(reportname, dir, placeholder) {
-    original.dir <- getwd()
-    setwd(dir)
-    system(paste0('pandoc ', paste0(reportname, '.md'), ' -o ', paste0(reportname, '.pdf --pdf-engine=xelatex -V mainfonts="DejaVu Sans"')))
-    setwd(original.dir)
-    # pandoc -o emoji.pdf --pdf-engine=lualatex -V mainfonts="DejaVu Sans"
-    # pandoc -o emoji.pdf --pdf-engine=xelatex  -V mainfonts="DejaVu Sans"
-}
-# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
-# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
-
-tol_query <- drake_plan(
-  strings_in_dots = "literals",
-  biggest = which.max(unname(ape::Ntip(opentree_chronograms$trees))),
-  # names(opentree_chronograms$trees[[biggest]])
-  # names(opentree_chronograms$trees[biggest])
-  # tol_dq = make_datelife_query(input = opentree_chronograms$trees[[biggest]], get_spp_from_taxon = FALSE),
-  # tol_dr = get_datelife_result(input = tol_dq), # erroring, so tryCatch and divide it because it is too large:
-  # all_dr_1 = lapply(opentree_chronograms$trees[1:100], function(x) tryCatch(datelife:::phylo_to_patristic_matrix(x),
-  #     error = function(e) NULL)) # biggest is the only one that error when going from phylo to patristic matrix.
-  # all_dr_2 = lapply(opentree_chronograms$trees[101:200], function(x) tryCatch(datelife:::phylo_to_patristic_matrix(x),
-  #     error = function(e) NULL)), # biggest is the only one that error when going from phylo to patristic matrix.
[…]

### 2019-09-30 tol_rates.R

@@ -1,46 +1,0 @@
-# hypothetical example
-# phy <- tol_jetz1[[1]]
-# age_and_richness <- jetz_fam_ages_and_richness
-get_all_ms <- function(phy, age_and_richness, epsilon = 0){
-  globalr = geiger::bd.ms(phy= NULL, time = max(ape::branching.times(phy)), n = sum(age_and_richness$richness), missing = 0, crown=TRUE, epsilon = epsilon)
-# c is for crown, global is for global rate, clade is for clade rate, ll is for limits
-    cp <-  geiger::crown.p(phy=NULL, time = age_and_richness$crown_age, n = age_and_richness$richness,
-      r = globalr, epsilon = epsilon)
-    cbd <- geiger::bd.ms(phy=NULL, time = age_and_richness$crown_age, n = age_and_richness$richness,
-      missing = 0, crown = TRUE, epsilon = epsilon)
-
-    cglobalr_ll <- lapply(age_and_richness$crown_age, function(x) tryCatch(
-      as.data.frame(geiger::crown.limits(time = x, r = globalr, epsilon, CI=0.95)),
-      error = function(e) data.frame(lb = NA, ub = NA)))
-    cglobalr_ll <- data.table::rbindlist(cglobalr_ll)
-
-    cclader_ll <- lapply(seq(age_and_richness$crown_age), function(i) tryCatch(
-      as.data.frame(geiger::crown.limits(time = age_and_richness$crown_age[i], r = cbd[i], epsilon, CI=0.95)),
-      error = function(e) data.frame(lb = NA, ub = NA)))
-    cclader_ll <- data.table::rbindlist(cclader_ll)
-
-    all_crown <- data.frame(crown.p = cp, bd.ms = cbd, cglobalr_ll, cclader_ll, age_and_richness$crown_age, age_and_richness$richness)
-    names(all_crown) <- c("crown.p", "bd.ms", "lb_global", "ub_global", "lb_clade", "ub_clade", "crown_age", "richness")
-    row.names(all_crown) <- row.names(age_and_richness)
-
-    sp <-  geiger::stem.p(phy=NULL, time = age_and_richness$stem_age, n = age_and_richness$richness,
-      r = globalr, epsilon = epsilon)
-    sbd <- geiger::bd.ms(phy=NULL, time = age_and_richness$stem_age, n = age_and_richness$richness,
-      missing = 0, crown = TRUE, epsilon = epsilon)
-
[…]

### 2019-09-30 tol_rates_plots.R

@@ -1,81 +1,0 @@
-ms <- jetz_ms0
-ms <- jetz2012_ms0
-ms <- jetz2012_min_ms0
-ms <- jetz2012_mean_ms0
-ms <- jetz2012_max_ms0
-rm(ms)
-plot_ms <- function(ms, file_dir = NULL, file_name = NULL, height = 3.5, width = 5,
-  write_pdf = TRUE, ylab = "Species N", pch = 19, col_significant = NULL, ...){
-  if(!inherits(file_dir, "character")){
-    file_dir <- getwd()
-  }
-  if(!inherits(file_name, "character")){
-    file_name <- "_ms_plot"
-  }
-  file_out <- paste0(file_dir, "//", file_name)
-  message(file_out)
-  if(write_pdf){
-    grDevices::pdf(file = paste0(file_out, "_crown.pdf"), height = height, width = width)
-    graphics::par(mai = c(1.02, 0.82, 0.2, 0.2))
-  }
-  ii <- !is.na(ms$crown$crown.p)
-  cc <- ms$crown$richness[ii] < ms$crown$lb_global[ii] | ms$crown$richness[ii] > ms$crown$ub_global[ii]
-  plot(x = ms$crown$crown_age[ii], y = ms$crown$richness[ii], xlab = "Crown age (MYA)",
-    ylab = ylab, col = ifelse(cc, paste0(col_significant, "80"), "#80808080"), pch = pch)
-  lines(x = sort(ms$crown$crown_age[ii]), y = ms$crown$lb_global[ii][order(ms$crown$crown_age[ii])],
-    type = "l", col = "grey", ...)
-  lines(x = sort(ms$crown$crown_age[ii]), y = ms$crown$ub_global[ii][order(ms$crown$crown_age[ii])],
-    type = "l", col = "grey", ...)
-  text(labels = paste("n =", sum(ms$crown$richness[ii] > ms$crown$ub_global[ii])), cex = 0.8,
-    x = 0.07*max(ms$crown$crown_age, na.rm = TRUE), y = 0.9*max(ms$stem$richness, na.rm = TRUE), adj = 0)
-  text(labels = paste("n =", sum(ms$crown$richness[ii] < ms$crown$lb_global[ii])), cex = 0.8,
-    x = 0.99*max(ms$crown$crown_age, na.rm = TRUE), y = 0.1*max(ms$stem$richness, na.rm = TRUE), adj = 1)
[…]

### 2019-09-30 tol_taxonomies.R

@@ -1,101 +1,0 @@
-load("data/tol_jetz1.RData")
-
-tol_jetz1[[1]]$ott_id
-which(is.na(tol_jetz1[[1]]$ott_id)) # 17 spp.
-# get ott Families
-jetz_ff <- get_ott_clade(ott_ids = tol_jetz1[[1]]$ott_id, ott_rank = "family")
-save(jetz_ff, file = "~/Desktop/datelife_paper1/data/jetz_ff.RData")
-length(ff$input) == length(tol_jetz1[[1]]$tip.label)
-ff$input[is.na(ff$family)] # tip labels that did not match a family (there are 78)
-# plus the 17 spp that are not in ott gives a total of 95 tips with no family data
-# these will appear as NA in the family vector:
-tol_jetz1[[1]]$ott_family <- ff$family[match(tol_jetz1[[1]]$ott_id, ff$input)]
-save(tol_jetz1, file = "~/Desktop/datelife_paper1/data/tol_jetz1.RData")
-
-# get jetz study Families
-tt <- read.csv("data-raw/nature11631-s2/2012-03-04206D-master_taxonomy.csv")
-nrow(tt)
-new.tree <- rotl::get_study_tree(study_id= "ot_809",tree_id="tree1",
-								tip_label="original_label", deduplicate = TRUE)
-tol_jetz1[[1]]$original.tip.label <- new.tree$tip.label # they are in the same order, I checked it with match
-save(tol_jetz1, file = "~/Desktop/datelife_paper1/data/tol_jetz1.RData")
-
-head(tt)
-mm <- match(tol_jetz1[[1]]$original.tip.label, as.character(tt$TipLabel))
-tol_jetz1[[1]]$jetz2012_family <- as.character(tt$BLFamilyLatin[mm])
-richness <- sapply(unique(as.character(tol_jetz1[[1]]$jetz2012_family)),
-  function(x) sum(as.character(tt$BLFamilyLatin) %in% x))
-mm <- match(tol_jetz1[[1]]$jetz2012_family, names(richness))
-tol_jetz1[[1]]$jetz2012_richness <- richness[mm]
-tail(data.frame(fam = tol_jetz1[[1]]$jetz2012_family, rich = tol_jetz1[[1]]$jetz2012_richness))
-save(tol_jetz1, file = "~/Desktop/datelife_paper1/data/tol_jetz1.RData")
-
-# once with the families, get the crown and stem ages of each Family:
[…]

### 2019-09-30 tol_topologies.R

@@ -1,23 +1,0 @@
-# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
-# @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @ @
-tol_topologies <- drake_plan(
-  biggest = which.max(unname(ape::Ntip(opentree_chronograms$trees))),
-  tol_biggest = opentree_chronograms$trees[[biggest]], #tol_biggest$node.label
-  plot_biggest = plot_ltt_phyloall(taxon = "Biggest", phy = tol_biggest, ltt_colors = "#ff0000", tax_datedotol = NULL,
-    file_name = "_lttplot.pdf", file_dir = getwd(), height = 3.5, width = 7, add_legend = FALSE,
-    add_title = TRUE, title_text = " chronogram in datelife database"),
-  # tol_otol_biggest = get_otol_synthetic_tree(input = tol_dq)
-  # tol_otol_biggest = get_otol_synthetic_tree(ott_ids = opentree_chronograms$trees[[biggest]]$ott_ids),
-  # tol_otol_fam =
-  tol_jetz1 = opentree_chronograms$trees[grepl("jetz", tolower(names(opentree_chronograms$trees)))][1],
-  tol_jetz2 = opentree_chronograms$trees[grepl("jetz", tolower(names(opentree_chronograms$trees)))][2],
-  plot_jetz1 = plot_ltt_phyloall(taxon = "Jetz chronogram 1", phy = tol_jetz1,
-    ltt_colors = "#ff0000", tax_datedotol = NULL, file_name = "_lttplot.pdf",
-    file_dir = getwd(), height = 3.5, width = 7, add_legend = FALSE,
-    add_title = TRUE, title_text = " chronogram in datelife database"),
-  plot_jetz2 = plot_ltt_phyloall(taxon = "Jetz chronogram 2", phy = tol_jetz2,
-    ltt_colors = "#ff0000", tax_datedotol = NULL, file_name = "_lttplot.pdf",
-    file_dir = getwd(), height = 3.5, width = 7, add_legend = FALSE,
-    add_title = TRUE, title_text = " chronogram in datelife database")
-)
-make(tol_topologies)

### 2020-04-25 biorxiv_2019.09.28.Rmd
coming back

@@
 ---
-<!-- \begin{\nolinenumbers} -->
-<!-- Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE <!--MAX 50 CHARACTERS, ALL CAPITALS-->
+```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
+# code to generate the paper data and figures:
+# source("data.R")
+# inactivated now just for testing purposes
+# we are using drake to avoid running the examples everytime we change something in the Rmd and knit
+# using this for testing instead:
+source("./data/frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
+loadd() # loads all the targets listed in the workflow plan of the previous `make()`
+library(datelife)
+data(opentree_chronograms)
+rpackage_available <- FALSE
+availability <- ifelse(rpackage_available, "is available", "will be available")
+```
 
-**Title**: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
-
-**Authors**: Luna L. Sánchez-Reyes^1^, Brian C. O’Meara^1^
-
-**Correspondence address**:
-
-1. *Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA*
-
-**Corresponding authors**: sanchez.reyes.luna@gmail.com, bomeara@utk.edu
-<!-- \end{nolinenumbers} -->
-
-``` {r child = 'paper.Rmd'}
+``` {r child = 'paper_1-header.Rmd'}
+```
+\newpage
+``` {r child = 'paper_2-abstract.Rmd'}
+```
+\newpage
+\begin{center}
+\textsc{Introduction}
+\end{center}
+``` {r child = 'paper_3-main.Rmd'}
 ```

### 2020-04-25 biorxiv_2019.09.28.tex
coming back

@@
 \textbf{Corresponding authors}: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}
-
 \newpage
@@
 \protect\hyperlink{ref-Piel2002}{2002}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al.
-\protect\hyperlink{ref-Hinchliff2015}{2015}) platforms store and make available published phylogenies and chronograms for easy scientific reuse.
-All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
+\protect\hyperlink{ref-Hinchliff2015}{2015}) are platforms that store and make available published phylogenies and chronograms for easy scientific reuse.
+Tree in all of them can be queried using automatised web procedures, which permit personalized, large scale searches that are also very fast.
 OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
+Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branch lengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
 
@@
 This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-\item  The user can obtain different summary information from the source chronograms including: a) all source chronograms, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, f) a report of succesful matches of input taxon names across source chronograms, and g) the single source chronogram with the greatest number of taxa.
+\item  The user can obtain different summary information including: a) all source chronograms, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, f) a report of succesful matches of input taxon names across source chronograms, and g) the single source chronogram with the greatest number of taxa.
 Summary information can be used to make decisions on the next steps of the workflow.
-\item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
+\item  Then, source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
 \item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R `phylo` format), as well as view and compare results graphically, or construct their own graphs using inbuilt `datelife` graphic generation functions.
+\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using inbuilt \texttt{datelife} graphic generation functions.
 \end{enumerate}
@@
 A \texttt{datelife} search currently accepts scientific names only. It can be any named clade or binomial specific.
-Chronogram search is performed at the species level, so when input names correspond
+The search is performed at the species level, so when input names correspond
 to named clades, \texttt{datelife} pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
+clade from OToL's reference taxonomy and uses all of those.
 Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
@@
 
-The chronogram search is performed across \texttt{datelife}'s chronogram database which is assembled from OToL's tree repository.
+Then, a name matching procedure is performed across \texttt{datelife}'s chronogram database which is assembled from OToL's tree repository.
 Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
-Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
+Also, the tip names in their stored trees correspond to scientific names at or below the species level.
+Finally, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
 
@@
 The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
-However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife\_examples package).
+However, the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife\_examples package).
 Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
 The Branch Length Adjuster (BLADJ) algorithm (Webb et al.
-\protect\hyperlink{ref-Webb2008}{2008}) was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
+\protect\hyperlink{ref-Webb2008}{2008}) was then used to fix known node ages and then distribute nodes with unknown ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
 
@@
 Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
-As explained in the \texttt{Description}, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
+As explained in the ``Description'' section, data from source chronograms was first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages were used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
 \ref{fig:summaries}).
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are generally consistent with source ages.
-Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms (Ramshaw et al.
-\protect\hyperlink{ref-ramshaw1972time}{1972}; Sanderson and Doyle \protect\hyperlink{ref-sanderson2001sources}{2001}; Magallón et al.
-\protect\hyperlink{ref-magallon2015metacalibrated}{2015}; Barba-Montoya et al.
-\protect\hyperlink{ref-barba2018constraining}{2018}).
-For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
-Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.
+In both cases, ages are generally consistent with source ages, but see next section.
 
@@
 \ref{fig:cvbladj}).
-Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
-three cases (studies 3, 5, and 9; Fig.
+To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data from the Barcode of Life Database (BOLD) and branch lengths were successfully generated for all source chronograms.
+However, dating with PATHd8 (using congruified calibrations) was only successful in
+three cases (studies 3, 5, and 9, shown in Fig.
 \ref{fig:cvbold}).
@@
 Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and details can be consulted in \url{https://github.com/LunaSare/datelife_examples}.
+More examples and code used to generate these trees can be consulted in \url{https://github.com/LunaSare/datelife_examples}.
 
@@
 This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
-Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to \texttt{datelife}'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
+Thus, incorporation of more published chronograms deposited in OpenTree, or perhaps directly pulled from the Dryad repository, to \texttt{datelife}'s database is crucial to improve its services.
+Methods to automatically mine chronogram data from Dryad could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
 Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.
@@
 Another potential concern comes from summary chronograms.
-We currently summarize all source chronogram data by default.
-Users can subset source data if they have reasons to favor some or one source chronogram over others.
+We currently summarize by default all source chronograms that overlap with at least two taxa.
+Users can subset source data if they have reasons to favor some source chronograms over others.
 Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
@@
 It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk \protect\hyperlink{ref-schenk2016sec}{2016}).
-Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-This remains to be tested.
+Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require setting priors (such as ML methods).
+In sum, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
 
-Furthermore, even chronograms obtained with primary data can be very different, as observed from the comparison of source chronograms in the Fringillidae example.
-A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms (e.g., angiosperms; Magallón et al.
-\protect\hyperlink{ref-magallon2015metacalibrated}{2015}).
-Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
-Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignment (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
-To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
-Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.
-Hence, summarizing chronograms might imply summarizing evolutionary hypothesis.
-This could be good from certain point of view, since it could help to get a single global evolutionary history for a lineage.
-But it could also be bad, since we could be loosing part of the evolutionary history that is only being reflected in some chronograms and not from the summary chronogram.
+Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
+This observation is often encountered in the literature
+(see, for example, the ongoing debate about crown group age of angiosperms (Ramshaw et al.
+\protect\hyperlink{ref-ramshaw1972time}{1972}; Sanderson and Doyle \protect\hyperlink{ref-sanderson2001sources}{2001}; Magallón et al.
+\protect\hyperlink{ref-magallon2015metacalibrated}{2015}; Barba-Montoya et al.
+\protect\hyperlink{ref-barba2018constraining}{2018}).
+For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}).
+Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.
+
+Summarizing high-quality chronograms might also imply summarizing evolutionary hypothesis.
+This could be enlightening from certain point of view, since it could help getting a single global evolutionary history for a lineage.
+It could also be really misleading, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in some chronograms and not from the summary chronogram.
 Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history (Antonelli et al.
@@
 
-In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
+Alternatively, one could try to choose the ``best'' chronogram.
+Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
+Some characteristics that are often cited as a measure of improved quality in dating studies addressing long-studied organisms are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the near future.
+
+In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
 Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ (Webb et al.
@@
 Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.
+There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data and trees that are reconstructed using taxonomy.
 
@@
 Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website.
+This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
 Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.

### 2020-04-25 paper_1-header.Rmd
coming back

---
output:
  pdf_document: default
  html_document: default
---

**Title**: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life

**Authors**: Luna L. Sánchez-Reyes^1^, Brian C. O’Meara^1^

**Correspondence address**:

1. *Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA*

**Corresponding authors**: sanchez.reyes.luna@gmail.com, bomeara@utk.edu
<!-- \end{nolinenumbers} -->

### 2020-04-25 paper_2-abstract.Rmd
coming back

--- 2019-08-06 Datelife_paper1.Rmd
@@
 output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    toc: false
-    includes:
-      # in_header: preamble.tex
-      after_body: [Datelife_paper1_figures_noyaml.tex]
-    # fig_caption: no
-textalign: left  # this does not work...
-geometry: margin = 1in
-bibliography: library_red.bib
-csl: systematic-biology.csl
-link-citations: yes
-# indent: true # this works
-# header-includes:
-#    - \setlength\parindent{24pt}  # this works too, but I'm gonna indent manually with 4 spaces
-# nope, 4 spaces indents are not recognized, they do something else
-linestretch: 2
-header-includes:
-    # - \usepackage{caption}
-    - \usepackage[left]{lineno}
-    - \linenumbers
-  # - \documentclass[oupdraft]{sysbio} # this is duplicated in tex generated by knitr
-  # - \usepackage{xr}
-  # - \externaldocument{Datelife_paper1_figures}
-    # - \usepackage[utf8]{inputenc}
-    # - \usepackage[english]{babel}
-    # - \usepackage[document]{ragged2e}
-nocite: |
-  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
+  pdf_document: default
+  html_document: default
 ---
-
-```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
-source("Datelife_paper1_global.R")
-```
-
-Running head: DATELIFE: REVEALING THE DATED TREE OF LIFE <!--MAX 50 CHARACTERS, ALL CAPITALS-->
-
-Title: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life
-
-Authors: Luna L. Sánchez-Reyes^1^, Brian C. O’Meara^1^
-
-Correspondence address:
-
-1. *Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA*
-
-Corresponding authors: sanchez.reyes.luna@gmail.com, bomeara@utk.edu
-
-\newpage
 
@@
 **Keywords:** Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
-
-\newpage
-Clade ages represent a fundamental piece of information for evolutionary
-understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014].
-The primary information needed these time estimates comes from the fossil record.
-Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
-The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
-Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree.
-That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
-
-Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
-TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] platforms store and make available published phylogenies and chronograms for easy scientific reuse.
-All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
-OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
-
-Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the phylogenetic relationships among known lineages.
-All or parts of this synthetic tree can be reused for any purpose. However, it currently only focus on synthesizing tree topology, meaning that it does not expose branch length data of any type.
-<!-- let alone time of divergence data from their stored chronograms. -->
-The Timetree of Life project focuses on the synthesis of a single chronogram of life [@Hedges2006] and presents a very accessible, attractive interface.
-However, the thousands of chronograms this NSF-funded project have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but large scale download remains prohibited by their site.
-The latest version of their synthetic chronogram [@Kumar2017] can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
-Other platforms such as SuperSmart [@antonelli2017supersmart] and phylogenerator [@pearse2013phylogenerator] are focused in automatised _de novo_ chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees.
-However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Moreover, these tools do not provide information from already created expert chronograms.
-
-<!-- discussion:
-The Angiosperm Phylogeny Website [APW; @stevens2001apw] also specializes in constructing a single synthetic tree, but it focuses in showing phylogenetic relationships between seed plant families only, and it does not store or make available the source trees.
-
-Platforms dedicated to store data on fossils and fossil calibrations: PBDB and The fossil calibration database
-can provide data on time of lineage divergence, but it is often not in a phylogenetic context.
-
-To our knowledge, there is no system allowing the comparison of available source information (apw does this a little bit, but it is a verbal comparison more than a graphical or visual comparison).
-Other types of technical comparisons: which trees are better (which trees represent reality better), but this is of course a very difficult issue.
- -->
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fuly public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
-A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
-In here we present the formal description and implementation of the `datelife` service, constituted by an R package and a web site (<http://www.datelife.org/>).
-There is still much room for improvement, and flaws and limitations are addressed below.
-We strived for the current implementation of `datelife` to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
-<!-- as well as tools for comparison of source and summary trees.  -->
-<!-- R packages for benchmarking of functionalities and demonstrating services were also developed and made available.>
-
-\begin{center}
-\textsc{Description}
-\end{center}
-<!--(BOLD and OToL species names, are they homogeneous/standardized?-->
-The basic `datelife` workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\item A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database; source trees with at least two matching input names are identified; all other taxa that do not match the original query are then dropped from the positively identified source trees.
-These pruned chronograms are hereafter referred as source chronograms.
-Finally, each source chronogram is transformed to a patristic matrix named by the citation of the original study.
-This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-\item  The user can obtain different summary information from the source chronograms including: a) all source chronograms, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, f) a report of succesful matches of input taxon names across source chronograms, and g) the single source chronogram with the greatest number of taxa.
-Summary information can be used to make decisions on the next steps of the workflow.
-\item  Source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R `phylo` format), as well as view and compare results graphically, or construct their own graphs using inbuilt `datelife` graphic generation functions.
-\end{enumerate}
-
-<!--To perform each task:-->
-To gather, process, and present information, `datelife` builds up from functions
-available in several R packages including rotl [@Michonneau2016], ape [@Paradis2004],
-geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], phylocomr [@Ooms2018] and rphylotastic [@Omeara2019].
-
-A `datelife` search currently accepts scientific names only. It can be any named clade or binomial specific. <!--common names (implemented in phylotastic)-->
-Chronogram search is performed at the species level, so when input names correspond
-to named clades, `datelife` pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
-Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
-`datelife` processes input names with the taxon name resolution service [TNRS; @Boyle2013],
-<!--which "corrects misspelled names and authorities, standardizes variant spellings,
-and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
-which corrects potentially misspelled names and typos, and standardizes spelling
-variations and synonyms <!--to a taxonomic reference of users choosing, default
-to otol taxonomy -->, increasing the probability of correctly finding the
-queried taxa in `datelife`'s chronogram database.
-
-The chronogram search is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository.
-Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
-Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
-<!-- The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence.
-The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.
--->
-<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->
-
-Information from source chronograms can be summarised with a summary statistic of tree branch
-lengths (such as median or mean). <!-- and the hierarchical clustering method of ttol-->
-The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
-However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package).
-Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
-The Branch Length Adjuster (BLADJ) algorithm [@
-2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
-
-A  much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006], implemented in the ape package [@Paradis2004].
-
-For tree dating, the congruification algorithm described by @Eastman2013
-is implemented to find shared nodes between trees (congruent nodes).
-The ages of these nodes are then used as calibrations to date any given tree.
-Currently implemented methods for tree dating are BLADJ, MrBayes [@Huelsenbeck2001; @Ronquist2003] and PATHd8 [@Britton2007], a non-clock, rate-smoothing dating method.
-
-<!-- and simulations are performed with various algorithms: -->
-<!-- MrBayes can be used when adding taxa at -->
-<!-- random, following a reference taxonomy or a topological constraint. It draws ages -->
-<!-- from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012]. -->
-
-<!-- Details on each step are further developed in `datelife`'s R package documentation -->
-<!-- `datelife workflow` vignette at (<https://LINK>). -->
-
-\begin{center}
-\textsc{Benchmark}
-\end{center}
-<!--**Benchmark: Testing DateLife computing performance**-->
-<!-- Good to test 3 features on a package: performance, speed and scalability-->
-
-`datelife`'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS service is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes.
-A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-<!--Speed of other datelife workflows? NO, bc it refkects speed from other software mainly. But I will put a sentence about it...
-Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
-generating new trees: shows bladj speed;
-Adding dates processing time: reflects more mrbayes speed than anything;
-get_bold_otol_tree running time: reflects bold downloading data speed.-->
-<!--Speed with different number of chronograms in database? NO-->
-<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
-`datelife`'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat [@RCoreTeam2018] that were run locally
---using the devtools package [@RCoreTeam2018], and on a public server --via
-GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
-present, unit tests cover more than 50% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
-
-\begin{center}
-\textsc{Example}
-\end{center}
-<!--**Biological example: Testing DateLife accuracy**-->
-<!--Bird (or reptile) chronograms, too long time...
-finches is good
-nothofagus
-Look for all chronograms containing any birds
-Or, look for chronograms containing small and old lineages
-Determine which clade of birds has the more chronograms (have been dated
-more times) and use that as biological example. Fringillidae is a good candidate
-Find a clade with at least one chronogram containing all clade's species.
-(Penguins look good, but they are giving weird results in SDM)
-
-Remove this chronogram from datelife Results.
-
-Make sdm and median trees and Compare
-
-add taxa with different methods and Compare
-
-Use ltts to compare for now.
-
-think of a test to compare trees, topology- and date-wise
-
-test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
-Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
--->
-In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain all data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp.
-according to the Open Tree of Life taxonomy).
-There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig.
-\ref{fig:schronograms1}).
-<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig.
-\ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
--->
-Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
-As explained in the `Description`, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are generally consistent with source ages.
-<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
-<!-- and hierarchical clustering method which appears to push ages back -->
-Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms (_____MANY CITATIONS____)).
-For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results (Title and Rabosky [https://academic.oup.com/sysbio/article/66/5/843/2682290] ).
-Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.
-
-Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs.
-\ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms.
-In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
-Maximum tree ages were only approximately recovered in one case (study 2; Fig.
-\ref{fig:cvbladj}).
-Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
-<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
-three cases (studies 3, 5, and 9; Fig.
-\ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and details can be consulted in <https://github.com/LunaSare/datelife_examples>.
-
-<!--Show code here, in the paper? Or show it only in the examples? Or only in the vignette? Write down what about datelife will be shown in each of these.
-Paper should show outputs from web site and package.
-Vignette will show the R code used to generate the outputs displayed here and in the website.
-datelife examples will explain more on what is done in every step.
--->
-
-\begin{center}
-\textsc{Flaws, Limitations and Prospects}
-\end{center}
-
-The main goal of `datelife` is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
-However, it has many flaws.
-Some of them can be overcome, some of them might represent limitations.
-
-At the moment, `datelife`'s chronogram database is not very large, storing `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written.
-This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining [@Kumar2017].
-OToL is the only public tree repository from where `datelife` can currently pull chronograms to construct its database.
-A previous version of TimeTree's synthetic chronogram [@Hedges2015] was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
-This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
-<!--
-Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
-Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to `datelife`'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
-<!-- What can we do to mine journal databases? -->
-<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.
-
-Another potential concern comes from summary chronograms.
-We currently summarize all source chronogram data by default.
-Users can subset source data if they have reasons to favor some or one source chronogram over others.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there is no objective way to determine if an expert chronogram is better than other.
-Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
-Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
-Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-This remains to be tested.
-
-Furthermore, even chronograms obtained with primary data can be very different, as observed from the comparison of source chronograms in the Fringillidae example.
-A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
-Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
-<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
-However, this is just an assumption.
--->
-Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignment (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
-To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
-Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.
-Hence, summarizing chronograms might imply summarizing evolutionary hypothesis.
-This could be good from certain point of view, since it could help to get a single global evolutionary history for a lineage.
-But it could also be bad, since we could be loosing part of the evolutionary history that is only being reflected in some chronograms and not from the summary chronogram.
-Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history [@antonelli2017supersmart].
-This will be implemented in future `datelife` versions.
-
-We allow accepted ways of creating branch lengths in the absence of starting branch length information for some taxa, such as BLADJ [@ 2008].
-Making up branch lengths in this or other ways is accepted in scientific publications: Rabosky et al.
-[https://www.nature.com/articles/s41586-018-0273-1] created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; Jetz et al.
-[https://www.nature.com/articles/nature11631], created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data; Smith and Brown [https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1019] constructed a tree of 353,185 seed plants where only 23% had molecular data.
-Taken to the extreme, one could make a fully resolved,  calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.
-
-There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
-In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents an improvement for testing alternative hypothesis using phylogenetic distance.
-It is uncertain, however, how using very different age data could affect the outcomes of these tests, and further research is necessary in this direction.
-
-\begin{center}
-\textsc{Conclusions}
-\end{center}
-
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. Generating this information is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills
-to produce chronograms _de novo_. Knowledge on clade ages is also crucial for
-science communication and education.
-
-<!--Mention again uses shown here and outline potential uses not shown here-->
-`datelife` allows an easy and fast obtention, as well as comparison of publicly available information
-on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, allowing identification of regions that require more research or that have conflicting information.
-<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
-Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
-`datelife` helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-<!-- `datelife` is also part of the reproducibility of science movement.
-It strives to facilitate usage of public data sets as well as encouraging the making of information publicly available.
-Acknowledging source information and describing workflows and analyses.
--->
-
-<!-- it allows exploration of alternative taxonomic and phylogenetic schemes; -->
-<!-- it allows rapid ; -->
-
-\begin{center}
-\textsc{Availability}
-\end{center}
-
-`datelife` is free and open source and it can be used through its current website
-<http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
-`datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-`datelife`'s R package stable version is available
-for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
-using the command `install.packages(pkgs = "datelife")` from within R. Development versions
-are available from the GitHub repository (<https://github.com/phylotastic/datelife>)
-and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
-
-\begin{center}
-\textsc{Supplementary Material}
-\end{center}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities can be found in GitHub repositories at <https://github.com/LunaSare/datelife_paper1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>, respectively.
-<!-- Vignettes?
-Dryad also?? since it is all in github already, maybe not necessary-->
-
-\begin{center}
-\textsc{Funding}
-\end{center}
-
-Funding was provided by the US National Science Foundation (NSF) grant 1458603
-
-NESCent
-
-Open Tree of Life
-
-University of Tennessee, Knoxville
-
-\begin{center}
-\textsc{Acknowledgements}
-\end{center}
-
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for `datelife`.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition
-to the ABI grant that funded this project itself.
-
-\newpage
-
-\begin{center}
-\textsc{References}
-\end{center}

### 2020-04-25 paper_3-main.Rmd
coming back

--- 2019-09-30 paper.Rmd
@@
 ---
-```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
-# code to generate the paper data and figures:
-# source("data.R")
-# inactivated now just for testing purposes
-# we are using drake to avoid running the examples everytime we change something in the Rmd and knit
-# using this for testing instead:
-source("./data/frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
-loadd() # loads all the targets listed in the workflow plan of the previous `make()`
-library(datelife)
-data(opentree_chronograms)
-rpackage_available <- FALSE
-availability <- ifelse(rpackage_available, "is available", "will be available")
-```
 
-\newpage
-**abstract.-**
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
-of Life for a large quantity and diversity of organisms in the last few decades.
-However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication.
-`datelife` is a service implemented via an R package and a web site
-(<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of published data on lineage divergence times.
-The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
-Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram.
-Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
-If there is no information available for any queried taxa, data can be simulated.
-All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this trees work for this purpose still needs to be tested.
-`datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-**Keywords:** Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
-\newpage
-\begin{center}
-\textsc{Introduction}
-\end{center}
 Clade ages represent a fundamental piece of information for evolutionary
@@
 Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
-TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] platforms store and make available published phylogenies and chronograms for easy scientific reuse.
-All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
+TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] are platforms that store and make available published phylogenies and chronograms for easy scientific reuse.
+Tree in all of them can be queried using automatised web procedures, which permit personalized, large scale searches that are also very fast.
 OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
-Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
+Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branch lengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.
 
@@
 We strived for the current implementation of `datelife` to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
-<!-- as well as tools for comparison of source and summary trees.  -->
+<!-- , for the benefit of the broader community. -->
+<!-- as well as tools for comparison of source and summary trees.  NOT YET u_u-->
 <!-- R packages for benchmarking of functionalities and demonstrating services were also developed and made available.>
 
-<!--(BOLD and OToL species names, are they homogeneous/standardized?-->
+<!--(BOLD and OToL species names, are they standardized with each other? I think yes-->
 
@@
 A `datelife` search currently accepts scientific names only. It can be any named clade or binomial specific. <!--common names (implemented in phylotastic)-->
-Chronogram search is performed at the species level, so when input names correspond
+The search is performed at the species level, so when input names correspond
 to named clades, `datelife` pulls all accepted species names within the
-clade from OToL's reference taxonomy to perform the search.
+clade from OToL's reference taxonomy and uses all of those.
 Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
@@
 
-The chronogram search is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository.
+Then, a name matching procedure is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository.
 Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
-Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
+Also, the tip names in their stored trees correspond to scientific names at or below the species level.
+Finally, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
 <!-- The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence.
@@
 The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
-However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package).
+However, the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package).
 Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree.
-The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
+The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to fix known node ages and then distribute nodes with unknown ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.
 
@@
 Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
-As explained in the `Description`, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig.
+As explained in the "Description" section, data from source chronograms was first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages were used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
 \ref{fig:summaries}).
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are generally consistent with source ages.
+In both cases, ages are generally consistent with source ages, but see next section.
 <!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
 <!-- and hierarchical clustering method which appears to push ages back -->
-Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
-For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results [@title2016macrophylogenies].
-Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.
 
@@
 \ref{fig:cvbladj}).
-Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
+To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data from the Barcode of Life Database (BOLD) and branch lengths were successfully generated for all source chronograms.
+However, dating with PATHd8 (using congruified calibrations) was only successful in
 <!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
-three cases (studies 3, 5, and 9; Fig.
+three cases (studies 3, 5, and 9, shown in Fig.
 \ref{fig:cvbold}).
@@
 Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and details can be consulted in <https://github.com/LunaSare/datelife_examples>.
+More examples and code used to generate these trees can be consulted in <https://github.com/LunaSare/datelife_examples>.
 
@@
 Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
-Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to `datelife`'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
+Thus, incorporation of more published chronograms deposited in OpenTree, or perhaps directly pulled from the Dryad repository, to `datelife`'s database is crucial to improve its services.
+Methods to automatically mine chronogram data from Dryad could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
 <!-- What can we do to mine journal databases? -->
@@
 It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
-Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods).
-This remains to be tested.
+Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require setting priors (such as ML methods).
+In sum, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
 
-Furthermore, even chronograms obtained with primary data can be very different, as observed from the comparison of source chronograms in the Fringillidae example.
-A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
-Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
+Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
+This observation is often encountered in the literature 
+<!-- Different source chronograms often show substantial variation in clade ages  -->
+(see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
+For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies].
+Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.
+
+Summarizing high-quality chronograms might also imply summarizing evolutionary hypothesis.
+This could be enlightening from certain point of view, since it could help getting a single global evolutionary history for a lineage.
+It could also be really misleading, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in some chronograms and not from the summary chronogram.
+Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history [@antonelli2017supersmart].
+This will be implemented in future `datelife` versions.
+
+<!-- still requires more research. -->
+<!-- A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].  -->
+<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
 <!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
@@
 -->
-Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignment (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms.
-To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
-Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.
-Hence, summarizing chronograms might imply summarizing evolutionary hypothesis.
-This could be good from certain point of view, since it could help to get a single global evolutionary history for a lineage.
-But it could also be bad, since we could be loosing part of the evolutionary history that is only being reflected in some chronograms and not from the summary chronogram.
-Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history [@antonelli2017supersmart].
-This will be implemented in future `datelife` versions.
 
-In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
+Alternatively, one could try to choose the "best" chronogram.
+Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
+Some characteristics that are often cited as a measure of improved quality in dating studies addressing long-studied organisms are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the near future.
+
+<!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->
+
+In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
 Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ [@Webb2008]) for several taxa lacking this information.

### 2020-04-25 paper_drafts/Intro.Rmd
coming back

--- 2019-06-20 Intro.Rmd
@@
 geometry: margin = 1in
-bibliography: library_red.bib
-csl: systematic-biology.csl
+bibliography: ../paper_references.bib
+csl: ../systematic-biology.csl
 link-citations: yes

### 2020-04-25 paper_drafts/Intro.tex
coming back

@@
 
+\texttt{datelife}'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.
+
+Improvements:
+* fossils as calibrations: Using secondary calibrations can generate biased ages when using bayesian methods, mainly because we don't know what prior to give to secondary calibrations
+(Schenk \protect\hyperlink{ref-Schenk2016}{2016}).
+* bayesian congruification
+* topological congruification
+
+Phylomatic (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005})
+is a platform that synthesizes a tree containing taxa of interest from a user supplied set of trees or from its own database.
+It has the advantage that it allows grafting taxa of interest absent from their database, but it is limited to mostly seed plants and mammals.
+If your query does not match its records, it gives many polytomies back.
+
+treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
+(Smith and O'Meara \protect\hyperlink{ref-Smith2012}{2012}).
 Time of lineage divergence
@@
 on the global, extant species numbers, which ranges from 3 to 100 million species
-{[}Mayr2010; Moran2011{]}). One advantage of TimeTree is that it includes taxa from
+(Mayr \protect\hyperlink{ref-Mayr2010}{2010}; Mora et al. \protect\hyperlink{ref-Mora2011}{2011})). One advantage of TimeTree is that it includes taxa from
 across the tree of life, versus more specialized chronogram repositories focusing on land plants
@@
 
+\leavevmode\hypertarget{ref-Mayr2010}{}%
+Mayr R.M. 2010. Tropical Arthropod Species, More or Less? Science. 329:41--42.
+
+\leavevmode\hypertarget{ref-Mora2011}{}%
+Mora C., Tittensor D.P., Adl S., Simpson A.G., Worm B. 2011. How many species are there on earth and in the ocean? PLoS Biology. 9:1--8.
+
 \leavevmode\hypertarget{ref-Morlon2014}{}%
@@
 61:289--313.
+
+\leavevmode\hypertarget{ref-Schenk2016}{}%
+Schenk J.J. 2016. Consequences of secondary calibrations on divergence time estimates. PLoS ONE. 11.
+
+\leavevmode\hypertarget{ref-Smith2012}{}%
+Smith S.A., O'Meara B.C. 2012. TreePL: Divergence time estimation using penalized likelihood for large phylogenies. Bioinformatics. 28:2689--2690.
 

### 2020-04-25 paper_drafts/paper.Rmd
coming back

--- 2019-09-30 paper.Rmd
@@
 
-\newpage
-**abstract.-**
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing has resulted in a steady accumulation
-of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
-of Life for a large quantity and diversity of organisms in the last few decades.
-However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication.
-`datelife` is a service implemented via an R package and a web site
-(<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of published data on lineage divergence times.
-The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
-Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram.
-Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
-If there is no information available for any queried taxa, data can be simulated.
-All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this trees work for this purpose still needs to be tested.
-`datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-**Keywords:** Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
-\newpage
-\begin{center}
-\textsc{Introduction}
-\end{center}
 Clade ages represent a fundamental piece of information for evolutionary

### 2020-04-25 paper_3-main.Rmd
ejm comments

@@
 Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
-TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] are platforms that store and make available published phylogenies and chronograms for easy scientific reuse.
-Tree in all of them can be queried using automatised web procedures, which permit personalized, large scale searches that are also very fast.
+TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life (OToL) repository [@Hinchliff2015; @mctavish2015phylesystem] are platforms that store and make available published phylogenies and chronograms for easy scientific reuse.
+Trees in all of them can be queried using automatised web procedures, which permit personalized, large scale searches that are also very fast.
 OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates).
@@
 However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist.
-Moreover, these tools do not provide information from already created expert chronograms.
+Additionally, neither Supersmart nor phylogenerator are designed to provide information from already created expert chronograms, since it might not be their objective.
 
@@
 A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
-In here we present the formal description and implementation of the `datelife` service, constituted by an R package and a web site (<http://www.datelife.org/>).
-There is still much room for improvement, and flaws and limitations are addressed below.
-We strived for the current implementation of `datelife` to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+Here we present the formal description and implementation of the `datelife` service, constituted by an R package and a web site (<http://www.datelife.org/>).
+<!-- There is still much room for improvement, and flaws and limitations are addressed below.  -->
+The current implementation of `datelife` performs the basic tasks described above.
+It features a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
 <!-- , for the benefit of the broader community. -->
@@
 \begin{enumerate}
-\item A user providing at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-The tree can be in newick or phylo format, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database; source trees with at least two matching input names are identified; all other taxa that do not match the original query are then dropped from the positively identified source trees.
-These pruned chronograms are hereafter referred as source chronograms.
-Finally, each source chronogram is transformed to a patristic matrix named by the citation of the original study.
+\item A user providing at least two taxon names as input.
+Taxon names can be given at the species level or above.
+The user can provide them as a simple comma separated character string, or as tip labels on a tree.
+The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
+\item A name search is then performed across the chronogram database.
+Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
+<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
+Finally, each pruned source chronogram is transformed to a patristic distance matrix.
 This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-\item  The user can obtain different summary information including: a) all source chronograms, b) maximum ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all or a subset of source chronograms, f) a report of succesful matches of input taxon names across source chronograms, and g) the single source chronogram with the greatest number of taxa.
-Summary information can be used to make decisions on the next steps of the workflow.
+For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
+\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow.
+Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
 \item  Then, source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
 \item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using inbuilt \texttt{datelife} graphic generation functions.
+\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
 \end{enumerate}
@@
 to named clades, `datelife` pulls all accepted species names within the
-clade from OToL's reference taxonomy and uses all of those.
+clade from OToL's reference taxonomy [@rees2017automated] and uses all of those.
 Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
@@
 We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing increases roughly linearly with number of input taxon names, and
+Query processing time increases roughly linearly with number of input taxon names, and
 increases considerably if TNRS service is activated.

### 2020-04-27 paper_3-main.Rmd
ejm comments

@@
 
-Then, a name matching procedure is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository.
+Then, a name matching procedure is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository [known as the phylesystem; @mctavish2015phylesystem].
 Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion.
@@
 
-Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean.
-A  much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006] is also implemented via the ape package [@Paradis2004].<!-- and the hierarchical clustering method of ttol??-->
+Information from pruned source chronograms can then be summarised with simple summary statistics of tree branch length, such as the median or the mean.
+A  much slower, Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006] is also implemented via the ape package [@Paradis2004].<!-- and the hierarchical clustering method of ttol??-->
+<!-- describe what SDM does and why clustering does not work here -->
 The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree.
@@
 Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS service is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes.
-A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig.
+increases considerably if taxonomic name resolution with TNRS [@Boyle2013] is activated.
+<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
+Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
+Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
 \ref{fig:runtime1}).
@@
 `datelife`'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat [@RCoreTeam2018] that were run locally
---using the devtools package [@RCoreTeam2018], and on a public server --via
+implemented with the R package testthat [@RCoreTeam2018] that were run both locally
+with the devtools package [@RCoreTeam2018], and on a public server --via
 GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
@@
 In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain all data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp.
-according to the Open Tree of Life taxonomy).
+We performed a higher-taxon search to obtain data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` species according to the Open Tree of Life taxonomy).
 There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig.
@@
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-In both cases, ages are generally consistent with source ages, but see next section.
+With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Flaws, Limitations and Prospects section).
 <!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
@@
 \ref{fig:cvbladj}).
-Maximum tree ages were only approximately recovered in one case (study 2; Fig.
-\ref{fig:cvbladj}).
-To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data from the Barcode of Life Database (BOLD) and branch lengths were successfully generated for all source chronograms.
-However, dating with PATHd8 (using congruified calibrations) was only successful in
+Maximum tree ages were only approximately recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
+<!-- We need to think abput the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
+To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data using the Barcode of Life Data System [BOLD; ratnasingham2007bold] and branch lengths were successfully generated for all source chronograms.
+However, dating with PATHd8 [@Britton2007] using congruified calibrations, was only successful in
 <!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
@@
 Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and code used to generate these trees can be consulted in <https://github.com/LunaSare/datelife_examples>.
+More examples and code used to generate these trees were developes on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.
 

### 2020-04-29 paper_3-main.Rmd
ejm comments

@@
 
-At the moment, `datelife`'s chronogram database is not very large, storing `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written.
-This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining [@Kumar2017].
-OToL is the only public tree repository from where `datelife` can currently pull chronograms to construct its database.
-A previous version of TimeTree's synthetic chronogram [@Hedges2015] was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
-This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
+Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OToL's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database.
+This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(sys.Date(), format="%B %d %Y")`.
+Unforutnately, TimeTree's database is not open for scientific reuse nor automatised data mining [@Kumar2017].
+In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015].
+This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OToL repository, and is part of datelife's database now.
+Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage.
+Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
 <!--
 Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
-Thus, incorporation of more published chronograms deposited in OpenTree, or perhaps directly pulled from the Dryad repository, to `datelife`'s database is crucial to improve its services.
-Methods to automatically mine chronogram data from Dryad could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
-<!-- What can we do to mine journal databases? -->
+Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services.
+One option to increase our database is the Dryad data repository.
+Methods to automatically mine chronograms from Dryad could be designed and implemented.
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
+<!-- Can we mine journal databases? -->
 <!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.
+Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OToL's tree repository [@mctavish2015phylesystem].
 
@@
 We currently summarize by default all source chronograms that overlap with at least two taxa.
-Users can subset source data if they have reasons to favor some source chronograms over others.
+Users can subset source data if they have reasons to choose some source chronograms over others.
 Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there is no objective way to determine if an expert chronogram is better than other.
+To our knowledge, there is no objective way to determine if an expert chronogram is better than another.
 Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
-Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations.
+Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms).
 It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
-Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require setting priors (such as ML methods).
-In sum, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
+Yet, there are different ways to use secondary calibrations
+<!-- give some examples or alternative ways of using calibrations-->
+and that same bias might not be encountered with dating methods that do not require setting priors, i.e., Maximum Likelihood methods such as r8s [@sanderson2003r8s].
+Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
 

### 2020-04-30 paper_3-main.Rmd
ejm comments

@@
 
-Summarizing high-quality chronograms might also imply summarizing evolutionary hypothesis.
-This could be enlightening from certain point of view, since it could help getting a single global evolutionary history for a lineage.
-It could also be really misleading, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in some chronograms and not from the summary chronogram.
-Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history [@antonelli2017supersmart].
-This will be implemented in future `datelife` versions.
+Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
+For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
+Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence).
+Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
+Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart].
+<!-- Will this be implemented in future `datelife` versions?
+Probably not-->
+However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
+Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
+<!-- Example -->
 
 <!-- still requires more research. -->
-<!-- A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].  -->
+<!-- Often, there is a large discrepancy in time of lineage divergence estimates across studies from expert knowledge, that is well known and documented for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
+-->
 <!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
@@
 
-Alternatively, one could try to choose the "best" chronogram.
+Alternatively, one could try to choose the "best" chronogram from a set of possible evolutionary hypotheses.
 Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited as a measure of improved quality in dating studies addressing long-studied organisms are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the near future.
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.
 
@@
 
-In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance.
-Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ [@Webb2008]) for several taxa lacking this information.
-Making up branch lengths in this or other ways is accepted in scientific publications: @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
+In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
+Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option).
+Making up branch lengths in this or other ways is accepted in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data adn the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
 Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data and trees that are reconstructed using taxonomy.
+There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.
 
@@
 Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. Generating this information is difficult,
+diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms _de novo_ is difficult,
 especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills
-to produce chronograms _de novo_. Knowledge on clade ages is also crucial for
-science communication and education.
+do not have the time to acquire and develop the necessary knowledge and data curation skills.
+Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
 
 <!--Mention again uses shown here and outline potential uses not shown here-->
-`datelife` allows an easy and fast obtention, as well as comparison of publicly available information
-on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, allowing identification of regions that require more research or that have conflicting information.
+`datelife` allows easy and fast summarization of publicly available information
+on time of lineage divergence.
+This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
 <!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
-Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.
-`datelife` helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
+`datelife` helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 

### 2020-06-11 paper_3-main.Rmd
missing references

@@
 Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics.
-TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life (OToL) repository [@Hinchliff2015; @mctavish2015phylesystem] are platforms that store and make available published phylogenies and chronograms for easy scientific reuse.
+TreeBASE [@piel2009treebase; @vos2012nexml], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life (OToL) repository [@Hinchliff2015; @mctavish2015phylesystem] are platforms that store and make available published phylogenies and chronograms for easy scientific reuse.
 Trees in all of them can be queried using automatised web procedures, which permit personalized, large scale searches that are also very fast.
